### 2015-08-11 research_class/acmtog/acmtog-sample.tex
Add upstream acmtog upstream package/class.
Downloaded from: https://www.acm.org/publications/article-templates

% v2-acmtog-sample.tex, dated March 7 2012
% This is a sample file for ACM Transactions on Graphics
%
% Compilation using 'acmtog.cls' - version 1.2 (March 2012), Aptara Inc.
% (c) 2010 Association for Computing Machinery (ACM)
%
% Questions/Suggestions/Feedback should be addressed to => "acmtexsupport@aptaracorp.com".
% Users can also go through the FAQs available on the journal's submission webpage.
%
% Steps to compile: latex, bibtex, latex latex
%
% For tracking purposes => this is v1.2 - March 2012
\documentclass{acmtog} % V1.2

%\acmVolume{VV}
%\acmNumber{N}
%\acmYear{YYYY}
%\acmMonth{Month}
%\acmArticleNum{XXX}
%\acmdoi{10.1145/XXXXXXX.YYYYYYY}

\acmVolume{28}
\acmNumber{4}
\acmYear{2009}
\acmMonth{September}
\acmArticleNum{106}
\doi{10.1145/1559755.1559763}

%ISSN
\issn{1234-56789}

% Copyright
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

\begin{document}

\markboth{V. F. Pamplona et al.}{Photorealistic Models for Pupil Light Reflex and Iridal Pattern Deformation}

\title{Photorealistic Models for Pupil Light Reflex and\\ Iridal Pattern Deformation} % title

\author{VITOR F. PAMPLONA {\upshape and} MANUEL M. OLIVEIRA
\affil{Universidade Federal do Rio Grande do Sul}
\and
GLADIMIR V. G. BARANOSKI
\affil{University of Waterloo}
\and
SEAN FOGARTY
\affil{University of Illinois at Urbana-Champaign}
% NOTE! Affiliations placed here should be for the institution where the
%       BULK of the research was done. If the author has gone to a new
%       institution, before publication, the (above) affiliation should NOT be changed.
%       The authors 'current' address may be given in the "Author's addresses:" block (below).
%       So for example, Mr. Fogarty, the bulk of the research was done at UIUC, and he is
%       currently affiliated with NASA.
}

%
% The code below should be generated by the tool at
% http://dl.acm.org/ccs.cfm
% Please copy and paste the code instead of the example below. 
%
\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>10010520.10010553.10010562</concept_id>
  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010575.10010755</concept_id>
  <concept_desc>Computer systems organization~Redundancy</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010553.10010554</concept_id>
  <concept_desc>Computer systems organization~Robotics</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
 <concept>
  <concept_id>10003033.10003083.10003095</concept_id>
  <concept_desc>Networks~Network reliability</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
</ccs2012>  
\end{CCSXML}

\ccsdesc[500]{Computer systems organization~Embedded systems}
\ccsdesc[300]{Computer systems organization~Redundancy}
\ccsdesc{Computer systems organization~Robotics}
\ccsdesc[100]{Networks~Network reliability}

%
% End generated code
%

% We no longer use \terms command
%\terms{Experimentation, Human Factors}

\keywords{Face animation, image-based modelling, iris animation, photorealism, physiologically-based modelling}

\acmformat{Vitor F. Pamplona, Manuel M. Oliveira, Gladimir V. G. Baranoski,
and Sean Fogarty. 2009. Photorealistic models for pupil light reflex and iridal pattern deformation.
{\em ACM Trans. Graph.} 28, 4, Article 106 (September 2009), 10 pages.\\
\doiline}

\maketitle

\begin{bottomstuff}
Manuel M. Oliveira acknowledges a CNPq-Brazil fellowship (305613/2007-3). Gladimir V. G. Baranoski acknowledges a
NSERC-Canada grant (238337). Microsoft Brazil provided additional support.
Authors' addresses: Sean Fogarty, (Current address) NASA Ames Research Center, Moffett Field, California 94035.
\end{bottomstuff}

\begin{abstract}
We introduce a physiologically-based model for pupil light reflex (PLR)
and an image-based model for iridal pattern deformation. Our PLR model
expresses the pupil diameter as a function of the lighting of the
environment, and is described by a delay-differential equation,
naturally adapting the pupil diameter even to abrupt changes in light
conditions. Since the parameters of our PLR model were derived from
measured data, it correctly simulates the actual behavior of the human
pupil. Another contribution of our work is a model for realistic
deformation of the iris pattern as a function of pupil dilation and
constriction. Our models produce high-fidelity appearance effects and
can be used to produce real-time predictive animations of the pupil and
iris under variable lighting conditions. We assess the predictability
and quality of our simulations through comparisons of modeled results
against measured data derived from experiments also described in this
work. Combined, our models can bring facial animation to new
photorealistic standards. Another contribution of our work is a model
for realistic deformation of the iris pattern as a function of pupil
dilation and constriction. Another contribution of our work is a model
for realistic deformation of the iris pattern as a function of pupil
dilation and constriction. Another contribution of our work is a model
for realistic deformation of the iris pattern as a function of pupil
dilation and constriction.
\end{abstract}

\section{Introduction}
Arguably, the most important features in facial animation are the eyes,
which are essential not only in directing the gaze of the
audience~\cite{BLR-2007}, but also in conveying the appropriate degree
of expression through pupil dilation and constriction movements. Hence,
for animations depicting close-up views of the face, natural-looking
eyes and pupil movements are highly desirable.

\begin{quote}
``Walt Disney once said to his animation team that the audience watches
the eyes and this is where the time and money must be spent if the
character is to act convincingly".
\end{quote}

Differently from most of the body, the human eye is subject to some
involuntary movements of the pupil, which are determined by ambient
illumination, drug action, and emotional conditions, among
others~\cite{BDDG-03}. Pupillary light reflex (PLR) is responsible for
the constriction of the pupil area in highly lit environments and for
its dilation in dimmed ones. PLR is an integral part of our daily
experience, and except for drug-induced action, is the single most
noticeable of such involuntary movements of the pupil.

\begin{table*}[t]
\tabcolsep22pt
\tbl{Summary of the  Main Mathematical and Physical Quantities Considered in the Development of the Proposed
Models}{%
\begin{tabular}{@{}ccc@{}}\hline
Symbol &{Description} &{Physical Unit} \\
\hline
$L_{b}$ & luminance & blondels (B) \\
$L_{fL}$ & luminance & foot-Lambert (fL) \\
$I_{l}$ & illuminance & lumens/$mm^2$  (lm/$mm^2$)\\
$R$ & light frequency & {Hertz (Hz)} \\
%% $\phi$ & retinal light flux & lumens  ({\it lm}) \\
%% $\bar{\phi}$ & retinal light flux threshold & lumens  ({\it lm}) \\
$D$ & pupil diameter & millimeters  ({\it mm}) \\
$A$ & pupil area & square millimeters  ($mm^2$) \\
$r_I$ & individual variability index & $r_I \in$ [0,1] \\
$t$ & current simulation time & milliseconds  ({\it ms}) \\
$\tau$ & pupil latency & milliseconds  ({\it ms}) \\
$x$ & muscular activity & none\\
$\rho_{i}$ & ratio describing the relative position & none \\
$\beta, \alpha, \gamma, k$ & constants of proportionality & none \\\hline
\end{tabular}}
\label{tab:symbols}
\begin{tabnote}
This is an example of table footnote. This is an example of table footnote. This is an example of table footnote.
This is an example of table footnote. This is an example of table footnote.
\end{tabnote}
\end{table*}

\begin{figure*}[t]
\centerline{\includegraphics[width=11cm]{tog-sample-mouse}}
\caption{Comparison of the results predicted by our models against video
of a human iris. (left) One frame of an animation simulating the changes
in pupil diameter and iridal pattern deformation. (center) One frame
from a video of a human iris. (right) Graph comparing the measured pupil
diameters from each individual frame of a nine-second-long video
sequence (green line) against the behavior predicted by our model (red
line). The gray bars indicate the periods in which the light was kept on
and off. The complete video sequence and corresponding animation are
shown in the accompanying video.}
  \label{fig:videocomparison}
\end{figure*}

The human iris is a muscular tissue containing several easily
identifiable structures. Together, they define patterns that are
deformed as a result of changes in the pupil diameter. Although pupil
light reflex and iridal deformations could be animated using standard
computer graphics techniques, which in turn, may result in more
realistic and reproducible of these movements.

In this article, we present a physiologically-based model for realistic
animation of PLR. Our model combines and extends some theoretical
results from the field of mathematical biology~\cite{CDFKMS-06} with
experimental data collected by several {researchers} relating pupil
diameter to the intensity of environmental
light~\cite{BCHH:Collewijn:1988}. The resulting model produces
high-fidelity appearance effects and can be used to produce real-time
predictive animations of the pupil and iris under variable lighting
conditions (Section~\ref{sec:PLRValidation}). We model the iridal
pattern deformation process by acquiring a set of high-resolution
photographs of real irises at different levels of pupillary dilation and
by tracking their features across the set of images. By analyzing the
tracked positions, we obtained a simple analytical expression for the
iridal deformation pattern as a function of the pupil diameter
(Section~\ref{sec:patterndeformations}). To the best our knowledge, ours
is the first physiologically-based model for simulating pupil light
reflex presented in the graphics literature (the first model ever to
simulate individual variability in terms of PLR
sensitivity---Section~\ref{subsec:Variations}), as well as the first
model for iridal pattern deformation. Moreover, ours are the first
practical models (providing actual coefficient values) in the literature
for simulating the dynamics of the pupil and iris under variable
lighting conditions. We demonstrate the effectiveness of our approach by
comparing the results predicted by our models against photographs and
videos captured from real human irises
(Figures~\ref{fig:videocomparison} and 9). Table~\ref{tab:symbols}
summarizes the main mathematical and physical quantities used in the
derivation of the proposed models and which are considered throughout
this work.

\section{Related Work in Computer Graphics}
\label{sec:relatedwork}
%
\looseness-1A few researchers have addressed the issue of realistic
human iris synthesis. Lefohn et al. blend several textures created by an
artist, each containing some eye feature. Other image-based approaches
have been proposed by Cui et al., Wecker et~al., and Makthal and Ross.
Essentially, they decompose a set of iris images using techniques such
as principal component analysis, multiresolution and wavelets, and
Markov random fields, and recombine the obtained data to generate new
images of irises. Zuo and Schmid created a fiber-based 3D model of the
iris. Lam and Baranoski introduced a predictive light transport model
for the human iris, which computes the spectral responses of iridal
tissues described by biophysical parameters. Fran\c{c}ois et~al.
estimate iris height maps from gray-scale images. All these approaches
use stationary pupil sizes.

Sagar et al. developed an anatomically detailed model of the eye to be
used in a surgical simulator. In their model, Gaussian perturbations
were used to simulate the waviness of ciliary fibers and the retraction
of pupillary fibers during pupil dilation. Alternatively, depending on
the level of object manipulation, a texture mapping approach was used to
model the iridal appearance. It is worth noting, however, that their
goal was to achieve functional realism \cite{DS-02} as opposed to
physical or photorealism.

\section{Brief Overview of the Human Iris\break and Pupil}
\label{sec:biologicalreview}

The human iris has a diameter of about 12\,mm and forms a disc that
controls how much light reaches the retina. Under high levels of
lighting, the iris dilates, flattening itself and decreasing the pupil
size. Under low levels of illumination, it constricts, folding itself
and increasing the pupil area. The pupil diameter varies from 1.5\,mm to
8\,mm on average, and in general, it is not a perfect circle. Also, its
center may deviate from the center of the iris by an offset of up to
20\%. According to Newsome and Loewenfeld, there are no observable
differences in the iris regarding light-induced or drug-induced pupil
dilation/constriction.

The human iris is divided in two zones by the \emph{collarette}, a
delicate zig-zag line also known as the iris frill. The \emph{pupillary}
zone is bounded by the pupil, while the \emph{ciliary} zone extends to
the outer border of the iris. Each zone is characterized by a muscle.
The \emph{sphincter}, located in the pupillary zone, is a concentric
muscle that constricts to decrease the pupil size. The \emph{dilator},
found in the ciliary zone, is a radial muscle that constricts to
increase the pupil size. These two muscles overlap at the collarette.

The sphincter and dilator muscles are independently connected to the
autonomous nervous system (ANS) and the pupil size results from a
balance of the separately incoming stimuli to the two
muscles~\cite{Dod-1900}. The ANS conducts the pupillary light reflex and
\emph{hippus} neural actions. Hippus are spontaneously irregular
variations in pupil diameter, which can essentially be characterized as
random noise in the 0.05 to 0.3~Hz frequency range. In PLR, when light
reaches the retina, neural signals are sent to the brain, which sends
back a signal for closing or opening the pupil. Thus, PLR can be modeled
in two phases: perception, and after some time delay, adjustment.

\section{Models of Pupil Dynamics}
\label{sub:models_of_pupil_dynamics}

The pupillometry literature describes several models built around
experiments designed to measure the values of some parameters as a
function of incident light intensity. Link and Stark performed a study
where a light source was placed in front of the subjects' irises and, by
varying the intensity and frequency of the light, they measured the
pupillary latency (the time delay between the instant in which the light
pulse reaches the retina and the beginning of iridal reaction):
\begin{eqnarray}
  \tau(R,L_{fL}) & = & 253 - 14\,{{ln}}(L_{fL}) + 70 R - 29\, R\, {{ln}}(L_{fL}),
  \label{eq:Link_n_Stark}
\end{eqnarray}
where $\tau$ is the latency in milliseconds, $L_{fL}$ is the luminance
measured in foot-Lamberts (fL), and $R$ is the light frequency measured
in Hz.

Other similar models predict an average pupil size as a function of the
light intensity using a few experimental measurements~\cite{DC-1901}.
Among those, the most popular one is the Moon and Spencer model, which
is expressed as:
\begin{equation}
\label{eq:moon}
 D = 4.9 - 3 {{tanh}}[0.4 ({{log}}_{10}(L_{b}) - 0.5)],
\end{equation}
where the pupil diameter, $D$, varies from 2 to 8~mm, and $L_{b}$ is the
background luminance level expressed in blondels, varying from $10^5$
blondels in sunny days to $10^{-5}$ blondels in dark nights. $tanh$ is
the hyperbolic tangent.

\subsection{Physiologically-Based Models}
\label{subsub:theoretical_models}
%
In Mathematical Biology and related fields, models based on
physiological and anatomical observations were derived to express the
relationships among the pupillary action variables without relying on
quantitative experimental data. For example, Usui and Stark proposed a
parametric model of the iris to describe the static characteristics of
pupil response to light stimuli, and to explain its random fluctuations
in terms of probability density functions. Recently, Tilmant {et al.}
proposed a model of PLR based on physiological knowledge and guided by
experiments. Although they have obtained plausible results, Tilmant {et
al.}\ have recommended the use of another physiologically-based model to
more accurately monitor pupillary dynamics, namely the time-dependent
model developed by Longtin and Milton.

Longtin and Milton define the efferent neural signal $E(t)$ arriving at the iris per unit of time $t$,
as:
\begin{equation}
\label{eq:et}
    E(t) = \beta \; {{ln}} \left[ \frac{ \phi(t-\tau) }{ \bar{\phi} } \right],
    \label{eq:efferent_neural_signal_1}
\end{equation}
where $\beta$ is a constant of proportionality and $\phi$ is the retinal
light flux measured in lumens and defined by Stark and Sherman as $\phi
= I_{l}A$: illuminance ($I_{l}$, in lumens/$mm^2$) times the pupil area
($A$, in $mm^2$). $\tau$ is the latency, and $\bar{\phi}$ is the retinal
light level threshold (the light level below which there is no change in
the pupil area). The notation $\phi(t-\tau)$ indicates that the current
effect depends on the retinal light flux at a time $\tau$ milliseconds
in the past. As the efferent neural signal reaches the iris, it induces
some muscular activity $x$ that may cause the pupil to dilate or
constrict. According to Partridge and Benton, the relationship between
$E(t)$ and $x$ can be approximated by:
\begin{equation}
E(t) \tilde{=} k \left( \frac{dx}{dt} + \alpha x \right),
    \label{eq:efferent_neural_signal_2}
\end{equation}
where $k$ is a proportionality factor and $\alpha$ is a rate constant
that depends on the definition and units of $x$ used in the model.
Longtin and Milton combine Equations~(\ref{eq:efferent_neural_signal_1})
and (\ref{eq:efferent_neural_signal_2}) as:
\begin{equation}
  \label{eq:ativ}
  \frac{dx}{dt} + \alpha x = \gamma \; {{ln}} \left[ \frac{ \phi(t-\tau) }{ \bar{\phi} } \right].
  \label{eq:delay_differential}
\end{equation}
They express the pupil area as $A = f(x)$ and use the inverse $f^{-1}(A)
= g(A) = x$ to remove $x$ from Equation~(\ref{eq:delay_differential}).
In their paper, Longtin and Milton use a Hill function \cite{Duc-2002}
(Equation~\ref{eq:Hill_function}) as the function $f$, since it can
approximate the elasto-mechanical properties of the iris during the
pupillary activity:
\begin{equation}
\label{eq:area}
  A = f(x) = \frac{\Lambda\theta^n}{\theta^n+x^n} + \Lambda'.
  \label{eq:Hill_function}
\end{equation}
Here, $\Lambda'$ and $\Lambda + \Lambda'$ are, respectively, the minimum
and the maximum pupil areas, and $\theta$ is the value of $x$
corresponding to the average pupil area. Longtin and Milton's model then
becomes:
\begin{equation}
  \frac{dg}{dA}\frac{dA}{dt} + \alpha g(A) = \gamma \; {{ln}} \left[ \frac{ \phi(t-\tau) }{ \bar{\phi} }
\right],
\label{eq:longtin}
\end{equation}
where
\begin{equation}
  g(A) = x = \sqrt[n]{ \frac {\Lambda \theta^n} { A - \Lambda'} - \theta^n }.
\label{eq:inverseHill}
\end{equation}

An S-shaped curve similar to the Hill function has been described in the
physiologically-based model of Usui and Stark to approximate the pupil
diameter of an average individual under static illumination conditions.

\section{The Proposed Physiological-Based Model}
\label{sec:proposed_model}
%
The model of Moon and Spencer (Equation~(\ref{eq:moon})) is based on a
set of discrete measurements and approximates the response on an average
individual under various lighting conditions. Their measurements were
made after the pupil size had stabilized for each illumination level,
and therefore, their model does not describe the pupil behavior outside
the equilibrium state. Moreover, pupil size, latency, constriction, and
redilation velocities tend to vary among individuals exposed to the same
lighting stimulus~\cite{GPMBH-2005}. We remark that such variations are
not captured by the model of Moon and Spencer.

Longtin and Milton's model (Equation~(\ref{eq:longtin})) is time
dependent and adaptive, with the potential to handle abrupt lighting
changes. It is a theoretical model, and unfortunately, Longtin and
Milton did not provide the values for the various parameters in their
model (\emph{i.e.}, $\gamma$, $\alpha$, $\theta$, $n$, $\bar{\phi}$), as
these, in principle, depend on the abstract notion of iridal muscular
activity $x$, as well as on the use of the Hill function. The use of
incorrect parameter values will not produce realistic results and may
cause Equation~(\ref{eq:longtin}) to not converge.

Starting from Longtin and Milton's and from Moon and Spencer's models,
we derive a practical model that predicts the pupil diameter for the
nonequilibrium case based on experimental data
(Section~\ref{subsec:DynamicCase}). In Section \ref{subsec:Variations},
we show how we can extend this basic model to take individual
variability into account.

\subsection{Equilibrium Case}
\label{subsec:EquilibriumCase}
Under constant lighting conditions, the pupil area in Longtin and
Milton's model will converge to an equilibrium state, where:
\begin{eqnarray*}
  \frac{dg}{dA}\frac{dA}{dt} = 0.
  \label{eq:longtinArea}
\end{eqnarray*}
%
Under such a circumstance, and assuming there is no occurrence of
hippus, $\phi$ becomes time invariant. Also, recall that
${{ln}}(m/n)={{ln}}(m) - ln(n)$, and therefore, one can rewrite Longtin
and Milton's model (Equation~(\ref{eq:longtin})) for the equilibrium
case as:
\begin{equation}
 \alpha g(A) = \gamma \; ({{ln}}(\phi) - {{ln}}(\bar{\phi})).
 \label{eq:Milton_for_experiment}
\end{equation}
%
In turn, the Moon and Spencer model can be rewritten as
\begin{eqnarray*}
  \left(\frac{D - 4.9}{3} \right) = \left[0.4 \left( \frac{ln}{(L_{b})}ln(10) - 0.5 \left(\frac{ln}{(10)}ln(10)\right)
\right)\right],
  \label{eq:moonAndSpencerRewrote}
\end{eqnarray*}
and since the hyperbolic tangent is an odd function, we can rewrite this equation as:
\begin{equation}
  -2.3026 \; {{atanh}} \left(\frac{D-4.9}{3}\right) = 0.4 ({{ln}}(L_{b}) -1.1513),
\label{eq:moonChanged}
\end{equation}
where ${{atanh}}$ is the arc-hyperbolic tangent. Comparing Equations~(\ref{eq:Milton_for_experiment}) and
(\ref{eq:moonChanged}), in order for  Longtin and Milton's model to fit the response of Moon and Spencer's average
subject under  equilibrium conditions, one has:
\begin{eqnarray}
\label{eq:moonsEqualsLongtins1}
 -2.3026 \; {{atanh}} \left(\frac{D-4.9}{3}\right) & \approx & \alpha g(A) \\
\label{eq:moonsEqualsLongtins2}
 0.4 ( {{ln}}(L_{b}) -1.1513)  & \approx &  \gamma ({{ln}}(\phi) - {{ln}}(\bar{\phi})).
\end{eqnarray}
%
From Equation~(\ref{eq:moonsEqualsLongtins2}) we can estimate the value
of the parameter $\gamma$. One should note that $L_b$ is expressed in
blondels while $\phi$ is given in lumens. Although, in general one
cannot convert between two photometric quantities, this can be done
under some well-defined situations. Since Moon and Spencer's data were
collected with the subject seated before a large white screen of uniform
intensity which covers most of their field of view, we assume that the
light reaching a person's pupil has been reflected by a perfect
(Lambertian) diffuse surface. Recall that an ideal (lossless) diffuse
reflector returns all of the incident flux so that its reflectance $\rho
= 1$ and its BRDF $f = 1/\pi$. For such a reflector, $1$~blondel $=
10^{-6}$ lumens/mm$^2$.

Since the light flux, $\phi$, depends on the area of the pupil, in order
to estimate $\gamma$, we first evaluate the left-hand side of
Equation~(\ref{eq:moonsEqualsLongtins2}) for the entire range of
illumination covered by Moon and Spencer's model: $L_b \in [10^{-5},
10^{5}]$ blondels. For each value of $L_b$, we then use
Equation~(\ref{eq:moon}) to estimate $D$, from which the pupil area $A =
\pi(D/2)^2$, and then $\phi$, are computed. The retinal light level
threshold $\bar{\phi} = 4.8118\times10^{-10}$ lumens was obtained using
the pupil diameter $D_t=7.8272$~mm, predicted by Equation~\ref{eq:moon}
for $L_b = 10^{-5}$ blondels ($\bar{\phi} = \pi(7.8272 / 2)^2~mm^2
\times 10^{-5} 10^{-6}$ lumens/mm$^2$). Using the tabulated data for the
left-hand side of Equation~(\ref{eq:moonsEqualsLongtins2}) and the
conversion scheme just described, we get the following fitting:
\begin{equation}
  0.4 ( {{ln}}(L_{b}) -1.1513)      \approx  0.45 \; ({{ln}}(\phi) - {{ln}}(\bar{\phi})) -5.2,
 \label{eq:approx_log_terms}
\end{equation}
whose quality of approximation is illustrated in
Figure~\ref{fig:Longtin_v_Moon_right_and_comparison} (left). The
vertical axis of the graph (scaled muscular activity) represents $\alpha
g(A)$, where $g(A) = x$ is the muscular activity. The extra constant
$-5.2$ translates the function on the right-hand side of
Equation~\ref{eq:moonsEqualsLongtins2} vertically, improving the
fitting. Given Equation~(\ref{eq:approx_log_terms}), we can replace
$g(A)$ with $M(D)$ (Equation~(\ref{eq:moonsEqualsLongtins1})), with
$\alpha = -2.3026$, where $M(D)$ is given by:
\begin{equation}
  M(D) = {{atanh}} \left( \frac{D-4.9}{3} \right).
  \label{eq:our_function_for_elasto_mechanical_prop}
\end{equation}
Thus, the equilibrium situation can be expressed by
Equation~(\ref{eq:our_model}). As expected, it approximates Moon and
Spencer's function (Equation~(\ref{eq:moon})) for the pupil diameter of
the average subject quite well. The absolute value of the difference
between Equations~(\ref{eq:moon}) and (\ref{eq:our_model}) is under
$2\%$ over the entire range of $[10^{-5}, 10^{5}]$ blondels
(Figure~\ref{fig:Longtin_v_Moon_right_and_comparison} right).
\begin{equation}
\label{eq:our_model}
  2.3026 \; M(D)  =  5.2 - 0.45 \; {{ln}} \left[\frac{\phi}{\bar{\phi}} \right]
\end{equation}
%
\begin{figure}[t]
\centerline{\includegraphics[width=7cm]{tog-sample-mouse}}
\caption{High-quality fittings: (left) Both sides of
Equation~\ref{eq:approx_log_terms}. (right) Equations~\ref{eq:moon} and
\ref{eq:our_model}, whose difference in absolute values is under $2\%$
over the entire range $[10^{-5}, 10^{5}]$ blondels.}
    \label{fig:Longtin_v_Moon_right_and_comparison}
\end{figure}
%
\subsection{The Dynamic Case}
\label{subsec:DynamicCase}
%
Equation~(\ref{eq:our_model}) cannot be used to describe the evolution
of the pupil diameter in time as a function of instantaneous variations
of the light intensity arriving at the pupil. Nevertheless, the obtained
constants are still valid for the dynamic case, since the equilibrium is
just a special case of the more general pupil behavior, for which the
constants should also hold.

In general, one cannot take an equation obtained for equilibrium and
generalize it to the dynamic case. In our model, however, this is
possible because of the following constraints:
\begin{enumerate}
\item[(1)] $g(A)$ and $M(D)$ have no explicit time dependence;
\item[(2)] the range of values assumed by $A$ (or $D$) is the same for
both the equilibrium and the nonequilibrium cases;
\item[(3)] there is a one-to-one mapping between $A$ and $D$.
\end{enumerate}

By introducing time in Equation~(\ref{eq:our_model}), we obtain a delay differential equation that
corresponds to our solution for the dynamic case:
\begin{equation}
\label{eq:dTEval}
  {{dt}}_{c} = \frac{T_{c} - T_{p}}{S} \quad
  {{dt}}_{d} = \frac{T_{c} - T_{p}}{3S},
\end{equation}
where $T_{c}$ and $T_{p}$ are respectively the current and previous
simulation times (times since the simulation started) measured in
milliseconds, $S$ is a constant that affects the constriction/dilation
velocity and varies among individuals. The higher the $S$ value, the
smaller the time step used in the simulation and, consequently, the
smaller the pupil constriction/dilation velocity.

Figure~4 shows pupil diameter values corresponding to Moon and Spencer's average
subject simulated using Equation~(16) considering some abrupt changes in the environment luminance.
For this example, our results are compared to results provided by the static models of Moon and Spencer
(Equation~(\ref{eq:moon})) and of De Groot and Gebhard.
%

\subsection{Modeling Individual Differences}
\label{subsec:Variations}

While Equation~(16) simulates dynamic pupil behavior, it only does so
for the average individual represented by the Moon and Spencer model.
There are, however, substantial differences in the way pupils from
different individuals react to a given light stimulus. Such variations
include differences in diameter~\cite{google}, latency, and constriction
and redilation velocities. In order to simulate individual differences,
we cannot just arbitrarily change the parameter values of our model, as
Equation~(16) may not converge.

Figure~3 shows the original data used by Moon and Spencer. The curve
$C_m$ (shown in black), was obtained by converting the values of $L_b$
in the range of $[10^{-5}, 10^{5}]$ blondels to lumens (see
Section~\ref{subsec:EquilibriumCase}) and then using
Equation~(\ref{eq:our_model}) to compute the corresponding pupil
diameter values used for plotting. The top and bottom curves, $C_t$ and
$C_b$, respectively, define an envelope containing all pupil diameter
values used by Moon and Spencer. $C_b$ was obtained by fitting a 5
degree polynomial to 11 of the smallest pupil diameter values along the
entire luminance range. Likewise, $C_t$ was obtained by fitting a 5
degree polynomial to 11 of the largest pupil diameter values. We treat
$C_b$, $C_m$, and $C_t$ as isocurves $C(p)$ for some parameter $p \in
[0,1]$, so that $C(0) = C_b$, and $C(1) = C_t$. We then model individual
differences by associating to each individual $I$, an index $r_I \in
[0,1]$, which corresponds to an isocurve, $C(r_I)$. This index can be
randomly generated or, alternatively, it can be recovered from
experimental data as described in Section~\ref{sec:PLRValidation}. To
avoid convergence problems and still achieve the results corresponding
to isocurve $C(r_I)$, we rewrite $C_t$ and $C_b$, respectively, as new
functions $C_{tD}$ and $C_{bD}$ of the pupil diameter:
\begin{eqnarray}
\label{eq:ctD}
  C_{tD}(D) &=& \,{-}\,0.013 D^5  +  0.322 D^4 - 3.096 D^3\nonumber\\
              &&\,{+}\,13.655 D^2 - 25.347 D + 18.179\\
\label{eq:cbD}
  C_{bD}(D) &=& \,{-}\,5.442 D^5 + 1.387 D^4 - 1.343 D^3\nonumber\\
              &&\,{+}\,6.219 D^2 - 1.317 D   + 1.219.
\end{eqnarray}
%

\looseness-1In order to obtain $C_{tD}$, we evaluate the functions $C_m$
and $C_t$ for $L_b$ in the range $[10^{-5}, 10^5]$ blondels, creating
ordered pairs of diameter values $(D_m, D_t) = (C_m(L_b), C_t(L_b))$.
Given enough of these pairs, we fit a curve expressing $D_t$ as a
function of $D_m$ (or $D$ for short). The resulting curve is $C_{tD}$
(Equation~(\ref{eq:ctD})). The case of $C_{bD}$ is similar. The final
pupil diameter at any time is then obtained solving Equation~16 for $D$
and then evaluating:
\begin{equation}
\label{eq:isocurve}
D_{final} = C_{bD}(D) + (C_{tD}(D) - C_{bD}(D)) r_I.
\end{equation}
We have adopted this solution due to its simplicity and generality: we
can easily replace the curves $C_{bD}(D)$ and $C_{tD}(D)$ with new ones,
covering new data as they become available, or representing other models
(e.g., De Groot and Gebhard). Since the relative distances of $C_m$ to
$C_b$ and $C_t$ vary for different values of $D$, no value of $r_I$ will
exactly recover $C_m$. This is not a problem, however, as $C_m$
corresponds to the average subject. Other parameterizations are
possible, including ones that interpolate $C_m$ for a given value of
$r_I$.

Although our model properly simulates the elastic behavior of the iris
muscular activity during changes in lighting conditions, it does not
model hippus (Equation~16 will converge to some pupil diameter value if
the lighting conditions remain constant). As random fluctuations whose
causes are still unknown, it is currently not possible to define a
physiologically-based model for hippus. We visually approximate the
hippus effect by adding small random variations to the light intensity
(between $-10^{0.3}$ and $10^{0.3}$ blondels), to induce small
variations in the pupil diameter (of the order of 0.2 mm), in the
frequency range of 0.05Hz to 0.3Hz. This significantly improves the
realism of the resulting simulations and animations. According to Usui
and Stark, the standard deviation of the noise corresponds to
approximately 10\% of the pupil diameter.

\subsection{The PLR Model Validation}
\label{sec:PLRValidation}
%
In order to validate our PLR model under nonequilibrium conditions and
to show that it is capable of representing individual variability, we
performed some qualitative comparisons between actual pupil behavior and
the results of simulations produced by our model. For this, we captured
videos of normal subjects presenting significantly different light
sensitivities (different PLR responses), while a light was turned {on}
and {off} several times. Since pupil constriction is bigger when both
eyes are stimulated, the subjects kept both eyes opened. To avoid
fatigue and habituation of the iris, in each experiment we recorded less
than one minute of video per subject.
\begin{itemize}
\item The image texel size of surface textures that represent 3D
elements (e.g. forest) should vary with distance, but should not match
true perspective (\emph{texel size} in Section~3.2, Texture Gradients).
\item The image space distribution of texel elements of 3D textures
(e.g. forest) should mimic one that would result from the projection of
homogeneously distributed surface elements (\emph{texel density} in
Section~3.2, Texture Gradients).
\item Image space texel spacing of 3D textures should ensure that texels
overlap, especially in steep areas (\emph{texel occlusion} in
Section~3.2, Texture Gradients).
\item Fall lines follow essential structures of terrain. They act as
surface contours, and are used by panorama artists to paint cliff and
snow textures (\emph{fall lines} in Section~3.2, Surface Contours).
\item Fall lines are used as imaginary lines along which tree strokes
are placed, acting as texture \emph{meta-strokes} (\emph{meta-strokes}
in Section~3.2, Surface Contours).
\item Shading tone should have a good distribution of light, medium, and
dark values (\emph{shading} in Section~3.2, Shading).
\item Light position should be placed so that the rendering of the
terrain exhibits a good balance of light and shade, as seen from the
selected viewpoint (\emph{light direction} in Section~3.2, Shading).
\item For extended terrain areas, indicating silhouettes, especially
between occluding hills, is useful (\emph{silhouettes} in Section~3.2,
Silhouettes).
\item Water surfaces should reflect the environment (\emph{water
textures} in Section~3.3, Brushstroke Colors).
\item Geometry should be emphasized by use of vertical exaggeration
(\emph{vertical exaggeration} in Section~3.4).
\end{itemize}

We computed the pupil diameters of the subjects at each frame of the
video sequences. Lighting measurements made during video capture were
used as input to our PLR model for simulating pupil behavior. The pupil
diameters resulting from these simulations were then compared to the
pupil diameters computed at individual video frames. Note that the
simulated results are not expected to quantitatively match the observed
ones, but rather be in qualitative agreement with observed behavior.

The videos were captured using a Cannon ELURA2 miniDV camcorder (NTSC,
720$\times$576 pixels) with progressive scan, connected to a PC through
a firewire connection. We kept the room's light dimmed so that the
subjects' pupils could dilate naturally to some extent, but not so dark
that we could not see the pupils in the individual video frames. Because
of these constraints, we used two subjects (both males) with light eyes
(a 24-year-old with green eyes, and a 26-year-old with blue eyes). For
each frame, the pupil diameters were estimated from the set of dark
pixels (pupil area $P_{area}$) inside a specified rectangle containing
solely the subject's pupil and part of the iris (Figure~5). Given
$P_{area}$, the pupil diameter was obtained (assuming the pupil is a
circle) as $d = 2(\sqrt{P_{area}/\pi})$ pixels. The conversion from
pixels to millimeters was performed considering a typical iris diameter
of 12mm. According to our experience, computing the pupil diameter as
described produces more accurate results than computing it as the number
of pixels in the largest straight segment in the set of dark pixels (the
pupil).

Since the video frames were captured at approximately 30Hz, in practice
no variation is expected between the pupil diameters in neighbor frames
under constant illumination, even in the presence of hippus. Thus, we
estimated the average error in the computed pupil diameters to be
approximately 0.1mm by computing the average difference between
estimated pupil diameters for neighbor frames. Based on the video
sequences, we set $S = 600$ (Equation~(\ref{eq:dTEval})) for the two
subjects in all experiments, as this value made their simulated
constriction velocities approximate the ones in the video sequences. We
empirically set the frequency of the two light sources used in our
experiments to $R = 0.4$Hz, a value that made the latency estimated by
Equation~(\ref{eq:Link_n_Stark}) approximate the latency observed in the
video frames.

To evaluate the quality of our simulations, we performed experiments
with both subjects using two different kinds of light sources to induce
pupil constriction: a small flashlight and a 100-watt incandescent white
light bulb. For light measurements, we used an LD-200 Instrutemp digital
lux meter (precision $\pm3\%$, frequency 2\,Hz).

\subsubsection{The Flashlight Experiments}
\label{sec:first_flashlightExperiment}

In these experiments, we used a light source to induce significant
changes in the subjects' pupil diameters without introducing
considerable changes in the lighting conditions of the environment. For
this purpose, we used a small flashlight powered by a single AAA battery
(1.5 Volt) kept at about 20cm from the subject's right eye and pointed
at it. Given the small area illuminated by the flashlight as well as its
reduced power, the readings from the lux meter were very sensitive to
even small changes in the relative position and orientation of the
flashlight with respect to lux meter sensor. Thus, we decided to run two
simulations using the recorded data: (1) considering the light intensity
estimated using Equation~(\ref{eq:moon}), and (2) considering the
readings from the lux meter. These two experiments are explained next.

In this experiment, we used the Moon and Spencer equation
(Equation~(\ref{eq:moon})) to solve for the light intensities during the
{on} and {off} states of the flashlight, based on the measured pupil
diameters (from the video). Since the Moon and Spencer function (curve
$C_m$ in Figure~3) represents the pupil behavior of an average
individual, we estimated the {on} ({off}) light intensity as the average
of the computed {on} ({off}) intensities for both subjects. Using this
procedure, we obtained estimates of $10^{1.1}$~blondels when the
flashlight was on, and $10^{-0.5}$~blondels when the flashlight was off.
Given the average luminance value for the {on} ({off}) state and the
{corresponding} pupil diameter for a given subject, we used
Equation~(\ref{eq:isocurve}) to estimate the $r_{I_{on}}$
($r_{I_{off}}$) index for that subject. The subject's final $r_I$ index
was computed as the average between his \smash{$r_{I_{on}}$} and
\smash{$r_{I_{off}}$} indices. Using this procedure, we obtained $r_I =
0.4$ for the green-eye subject and $r_I = 0.03$ for the blue-eye
subject.

Figure~6 shows the actual pupil diameter measurements performed on a
frame-by-frame basis along 9-second-long sequences captured for each
subject. The green ``+" marks on top represent the measurements for the
green-eye subject, while the blue ``x" marks show the measurements of
the blue-eye subject. This example illustrates the intersubject
variability in terms of light sensitivity and shows the ability of our
model to appropriately represent such individual differences. The
vertical dotted lines delimit the intervals in which the flashlight was
kept on and off for each subject.
%Note that the duration of these intervals varied as the video sequences were acquired.
The solid and dashed lines represent the simulated results produced by
our model for the green-eye and blue-eye subjects, respectively, and
closely agree with the actual measured values. These curves were
produced automatically from Equations~(16) and~(\ref{eq:isocurve}), on
top of which we added small random variations (hippus effect) as
described in the previous section. The accompanying video shows
side-by-side comparisons of our simulated results and videos captured
for the two subjects.

\paragraph{The second flashlight experiment}
In this experiment, we used the readings provided by the lux meter for
the {on} and {off} states of the flashlight. These illuminance values
were 350lux\footnote{$1~{{lux}} = 1~{{lumen}} /m^2$.} and 90lux,
respectively. One should recall that in such a setup, small changes in
the position and orientation of the subject's head produce changes in
the illuminance at the pupil. Therefore, these values are only
approximations to the actual illuminance reaching each subject's lit
eye. Given the illuminance values and the subjects' corresponding pupil
diameters estimated from the video frames, we obtained the actual
pupil's luminous flux (in lumens) at the two flashlight states, for each
individual. These values were then converted to blondels according to
the assumption described in Section~\ref{subsec:EquilibriumCase}. We
then used Equations~(16) and~(\ref{eq:isocurve}) to estimate their
corresponding $r_I$ indices (by averaging \smash{$r_{I_{on}}$} and
\smash{$r_{I_{off}}$}), obtaining $r_I = 0.54$ for the blue-eye subject
and $r_I = 0.92$ for the green-eye subject. Figure~7 compares the actual
pupil measurements (same as in Figure~6) with the results simulated by
our model using the lux meter readings as input. The differences between
the simulated curves shown in Figures~6 and 7 are primarily due to the
added random noise (hippus).

\subsubsection{The 100-Watt Lightbulb Experiment}
\label{sec:luxMeterExperiment}
For this experiment we used a more stable light source to induce pupil
constriction: a spot with a 100-watt incandescent white lightbulb, kept
at about one meter in front and one meter to the right of the subject's
head. This setup allowed the subjects to remain comfortable with their
eyes opened while the light was on.

We measured the environment light intensity during the {on} and {off}
states by positioning the digital lux meter at approximately the same
position and orientation as the subject's right eye. During the blue-eye
subject experiment, we found the illuminance to be equal to $140$ lux
when the light was off and $315$ lux when it was on. During the
green-eye subject experiment, the readings were $91$ and $540$ lux,
respectively. These differences resulted from a darker environment and a
slight approximation of the green-eye subject to the light source.
Again, we used the illuminance values and the subjects' corresponding
pupil diameters (measured from the video) as input to Equations~(16)
and~(\ref{eq:isocurve}) to estimate their corresponding $r_I$ indices
(by averaging $r_{I_{on}}$ and $r_{I_{off}}$). We obtained $r_I = 0.9$
for the blue-eye subject and $r_I = 1.0$ for the green-eye subject.

Figure~8 (top) shows the actual pupil diameter measurements performed on
a frame-by-frame basis along 56- and 50-second-long sequences captured
for the blue-eye and for the green-eye subjects, respectively. The
vertical lines delimit the intervals in which the light was kept {on}
and {off} for each subject. The solid and dashed lines represent the
simulated results produced automatically by our model (Equations~16
and~\ref{eq:isocurve}) with and without hippus, respectively, and
closely agree with the actual measurements. Figure~8 (bottom) shows
zoomed versions of portions of the graphs shown on top, exhibiting
{off-on-off} transitions.

One should note that the simulated results produced by our PLR model
closely approximate the actual behaviors of the subjects' pupils in all
three experiments, illustrating the effectiveness of our model. The
differences in the $r_I$ indices for a given subject among the
experiments can be explained as follows.
\begin{itemize}
\item In the two flashlight experiments, the pupil diameters used for
the {on} and {off} states were the same, but the illuminance values
provided by Equation~\ref{eq:moon} and by the lux meter were different.
The different indices simply reflect the different light sensitivities
presented to our model as input.
\item When comparing the 100-watt lightbulb and the flashlight
experiments, both the lighting and the pupil sizes varied for the {on}
and {off} states of the light sources. For instance, for the green-eye
subject, the pupil diameters were approximately 4.3mm and 5.7mm for the
{on} and {off} states of the flashlight, respectively (Figure~7). This
resulted in an $r_I$ index of 0.92. In the case of the 100-watt
lightbulb experiment, these values were approximately 4.3mm and 6.0mm,
respectively (Figure~8), with $r_I = 1.0$. These two indices are
relatively close and reflect the difference in the maximum pupil
diameters between the two experiments. The difference in the $r_I$
indices for the blue-eye subject were considerably larger, from 0.54 to
0.9. Again, this can be explained by comparing the measured pupil
diameters in the two experiments. These values went from approximately
3.2mm and 4.2mm in the {on} and {off} states of the flashlight
(Figure~7) to 4.4mm and 5.2mm in the {on} and {off} states of the
100-watt lightbulb (Figure~8).
\end{itemize}
An important point to note is that by using an average of the estimated
$r_I$ indices for the {on} and {off} states of the light source, our
model is capable of realistically simulating the pupil behavior of
individuals with considerable differences in PLR responses under
different and variable lighting conditions.

\section{Modeling the Iris Deformation}
\label{sec:patterndeformations}

Although the iris is a well-known structure, there is no general
agreement about a model of its behavior. He suggested that the collagen
fibers are arranged in a series of parallel arcs, connecting the iris
root with the pupil border, clockwise and counterclockwise in an angle
of 90 degrees oriented by the center of the pupil. These fibers would be
interwoven with other iris components, such as blood vessels. Based on
Rohen's fiber arrangement, Wyatt proposed a 2D nonlinear model for iris
deformation. Such a model has been validated on canine, porcine, and
monkey irises, but so far not on human irises~[Wyatt private
communication].

Figure~9 (right) shows how the positions of the individually tracked
iridal feature points changed along the dilation process. The
trajectories of the points both on the pupillary and ciliary zones move
on approximately radial paths. Although some imprecision in the exact
location of the points might have resulted from the manual
specification, most of the deviation from the radial paths result from
the existence of blood vessels under the iris, and from crypts, and
folds (the iris folds its tissue as a result of pupil dilation) that
prevent iris points from always moving along radial lines. Such
structures vary considerably among individuals but, according to our
experience, their influence on the paths of the feature points usually
has small magnitude (Figure~9 right). Therefore, as a first
approximation, we can assume that the iris points move along straight
lines in the radial directions. It is worth noting that\break Wyatt's 2D
model does not take the influence of these structures into account
either.

In order to find how fast the feature points moved, we computed the
following measures during the dilation process: (1) the distance from
the tracked feature point to the pupil center; (2) the distance from the
tracked feature point to the pupil border; and (3) the ratio between the
distance from the tracked point to the pupil border and the local width
of the iridal disk (the distance from the pupil border to the external
iris border measured along the radial segment passing through the
feature point). One should recall that the pupil is not necessarily
circular and that its center does not necessarily coincide with the
center of the iris. While measurements (1) and (2) presented a pretty
much linear behavior, the ratio represented by (3) was approximately
constant for all feature points (Figure~10 right). The same behavior was
observed in the irises of all five volunteers. Like the variations in
the trajectories of the points shown in Figure~9 (right), the deviations
from horizontal lines in Figure~10 (right) are caused by the subjects'
iris structures, specially the iridal folds. Again, as a first
approximation, the following ratio can be assumed constant for any
iridal point $p_i$, for all values of pupil diameters:
\begin{equation}
\rho_i = \frac{\|p_i - c_i\|}{\|E_i - c_i\|},
\label{eq:invariance_iris_deformation}
\end{equation}
where $p_i$ is a point on the iris disk, $c_i$ and $E_i$ are the points
on the pupil border and on the iris outer circle, respectively, such
that they are collinear to the radial segment passing through $p_i$.
$\|.\|$ is the $L^2$ (Euclidean) norm. The invariance expressed by
Equation~\ref{eq:invariance_iris_deformation} summarizes the
observations illustrated in Figure~10 (right) and is the basis of our
image-based model for iridal pattern deformation.

%
\subsection{Animating the Deformed Iridal Patterns}
\label{sec:iris_animation}
%
As an approximation to the behaviors depicted in Figures~9 (right) and
10 (right), we use texture mapping to animate the iris deformation
process. Note that this is a natural and efficient way of implementing
the behavior modeled by Equation~(\ref{eq:invariance_iris_deformation}):
as the pupil dilates/constricts, the iris ring is compressed/stretched,
but the parameterization (in the $[0,1] \times [0,1]$ domain) of the
points inside the ring remains the same. Thus, for animation purposes,
we model the iris as a planar triangle-strip mesh on the disk defined by
the two circles %(Figure~\ref{fig:iris_mesh}) and use a photograph of an
iris with a small pupil diameter as a texture. Texture coordinates map
the border of the pupil to the inner circle of the mesh, and outer
border of the iris to the mesh's outer circle. Currently, we tessellate
the mesh creating a pair of triangles at every five degrees. The
animation proceeds by computing the new pupil diameter $D$ as a function
of the incident lighting using Equation~(\ref{eq:isocurve}). We then
reposition each vertex $v_i$, located on the inner circle of the mesh,
at a distance $D/2$ along the radial line connecting the center of the
pupil to $v_i$, while keeping their original texture coordinates
unchanged. One should recall that the center of the pupil does not
necessarily match the center of the iris, and thus, it is important to
keep the coordinates of the center of the pupil. Figure~8 shows the
renderings of an iris created using our models for different lighting
conditions. Note that the patterns deform in a natural way. No light
reflection on a corneal surface has been simulated, to avoid masking
iris details.

\section{Discussion}
\label{sec:results}

We have implemented the proposed models and used them to render
synthetic images of the human iris and pupil. The resulting animations
are very convincing and run in real time. We have compared the images
produced by our models with photographs and videos of real human irises.
The results produced by our models are in qualitative agreement with
observed behavior in humans.

In order to demonstrate the potential use of the proposed models in
computer graphics, we built an application that renders a human head
model in an environment illuminated by HDR cube maps (see accompanying
video). The head model was obtained and its original irises were
replaced by our textured triangle-strip model. The HDR images were
obtained from Paul Debevec's web site and are used to approximate the
environment's radiance. As the head looks at different parts of the
environment, its pupil diameters adapt to the irradiance in the solid
angle defined by its field of view, producing pleasing animation
effects.

Accommodation and age affect the pupil diameter and iris color
influences some PLR parameters, such as {maximum} pupil diameter,
latency, and constriction velocity. These aspects are currently not
taken into account by our model because of the lack of reliable data
over a large range of lighting conditions. For instance, discuss the
effect of age on the size of the pupil. Their study, however, only
considered luminance values from $10^1$ to $10^4$ blondels, which
corresponds to only about 30\% of the luminance range used by our model.
Currently, we model variations in pupil diameters for the same light
stimulus using Equation~\ref{eq:isocurve}, which can be used to simulate
the age-related miosis effect reported by Winn. Also, since our model
covers the entire range of valid pupil diameter values, it safely covers
the pupillary sizes resulting from influence of attentional and other
cognitive factors. Extending our model to handle other phenomena based
on biophysical parameters is an interesting direction for future work.

No relief data representing the iris folds are used in the current
version of the model, as it is done in the technique presented by. Also,
no corneal refraction is used. Thus, at grazing angles, in addition to
the distortion resulting from pupil dilation/constriction, one would
perceive the projective distortion due to texture mapping. Relief
information could be added to our model in a straightforward way,
allowing some interesting shading effects such as projected shadows and
self-occlusions.

We use a linear model for iridal pattern deformation even though the
actual deformation is nonlinear. However, such nonlinearity contributes
approximately only 1\% of the diameter of a typical iris (12.0mm). Most
of the nonlinear behavior seen in Figure~9 (right) and Figure~10 (right)
is due to the interference of folds and blood vessels, which varies
among individuals. To the best of our knowledge, no model in the
literature takes those factors into account.

Many other factors affect pupil size, including particular states of
mind, such as interest and curiosity, spectral sensitivity, respiratory
and heart rate, and spatial patterns in the visual field. Taking all
these aspects into account seems to be impractical due to their inherent
complexity and limited supporting data. We should emphasize that PLR
causes the single most noticeable involuntary movements of the pupil. As
the graphs depicted in Figures~7 and 8 and the accompanying video show,
our PLR model alone can produce predictable animations of the pupil
dynamics.

\section{Conclusion}
\label{sec:conclusion}
%
We have presented new models for realistic renderings of the human iris
and pupil. Our physiologically-based model of the pupil light reflex
combines and extends theoretical results from the Mathematical Biology
field with experimental data collected by several researchers. The
resulting model is expressed in terms of a nonlinear delay-differential
equation that describes the changes in the pupil diameter as function of
the environment lighting. Our model is also original in the sense that
it can simulate individual differences with respect to light
sensitivity. As all parameters of our models were derived from
experimental data, they correctly simulate the actual behavior of the
human iris and pupil. They also produce high-fidelity appearance
effects, which can be used to create\break real-time predictive
animations of the pupil and iris under variable lighting conditions. We
have validated our models through comparisons of our simulated results
against videos and photographs captured from human irises. The quality
of these simulations qualitatively matched the actual behaviors of human
pupils and irises.

To the best of our knowledge, ours is the first physiologically-based
model for simulating pupil light reflex presented in the graphics
literature. It is also the first practical model (providing actual
coefficient values) in the literature for simulating the dynamics of
pupil and iris under variable lighting conditions, and the first
integrated model in all of the literature to consider individual
variability in pupil diameter using general equations for latency and
velocity. Our image-based model for iridal pattern deformation is also
the first model of its kind in the graphics literature.\vskip21pt

% Start of "Sample References" section

\section{Typical references in new ACM Reference Format}
A paginated journal article \cite{Abril07}, an enumerated
journal article \cite{Cohen07}, a reference to an entire issue \cite{JCohen96},
a monograph (whole book) \cite{Kosiur01}, a monograph/whole book in a series (see 2a in spec. document)
\cite{Harel79}, a divisible-book such as an anthology or compilation \cite{Editor00}
followed by the same example, however we only output the series if the volume number is given
\cite{Editor00a} (so Editor00a's series should NOT be present since it has no vol. no.),
a chapter in a divisible book \cite{Spector90}, a chapter in a divisible book
in a series \cite{Douglass98}, a multi-volume work as book \cite{Knuth97},
an article in a proceedings (of a conference, symposium, workshop for example)
(paginated proceedings article) \cite{Andler79}, a proceedings article
with all possible elements \cite{Smith10}, an example of an enumerated
proceedings article \cite{VanGundy07},
an informally published work \cite{Harel78}, a doctoral dissertation \cite{Clarkson85},
a master's thesis: \cite{anisi03}, an online document / world wide web resource \cite{Thornburg01}, \cite{Ablamowicz07},
\cite{Poker06}, a video game (Case 1) \cite{Obama08} and (Case 2) \cite{Novak03}
and \cite{Lee05} and (Case 3) a patent \cite{JoeScientist001},
work accepted for publication \cite{rous08}, 'YYYYb'-test for prolific author
\cite{SaeediMEJ10} and \cite{SaeediJETC10}. Other cites might contain
'duplicate' DOI and URLs (some SIAM articles) \cite{Kirschmer:2010:AEI:1958016.1958018}.
Boris / Barbara Beeton: multi-volume works as books
\cite{MR781536} and \cite{MR781537}.

\appendix

\section{Classical Multidimensional Scaling}
\label{sec:cmds}

Let $\mathrm{D}$ be an $n\times n$ matrix of pairwise distances. The
matrix $D$ is symmetric with a zero diagonal. We are interested in
finding a $d \times n$ matrix $\mathrm{X}$ where each column
$\bm{x}_{i}$ is the representation of the point $i$ in $R^{d}$ and
$\mathrm{D}_{ij} = \|\bm{x}_{i}-\bm{x}_{j}\|_{2}$. Denote the inner
product (or Gram matrix) for this set of points by $\mathrm{K} =
\mathrm{X}^{\top}\mathrm{X}$.

$\mathrm{K}$ is an $n\times n$ symmetric positive semidefinite matrix.
Let us now abuse notation and use $\mathrm{D}^{2}$ to indicate the
matrix of squared pairwise distances $\mathrm{K} =
-\frac{1}{2}(\mathrm{I} -
\mathrm{1}\mathrm{1}^{\top})\mathrm{D}^{2}(\mathrm{I} -
\mathrm{1}\mathrm{1}^{\top})$. Here, $\mathrm{I}$ is the $n \times n$
identity matrix and $\mathrm{1}$ is the $n$-vector of all ones.

\begin{acks}
We are grateful to the following people for resources, discussions and
suggestions: Prof. Jacobo Melamed Cattan (Ophthalmology-UFRGS), Prof.
Roberto da Silva (UFRGS), Prof. Luis A. V. Carvalho (Optics-USP/SC),
Prof. Anatolio Laschuk (UFRGS), Leandro Fernandes, Marcos
Slomp, Leandro Lichtenfelz, Renato Silveira,
Eduardo Gastal, and Denison Tavares. We also thank the volunteers who
allowed us to collect pictures and videos of their irises: Alex Gimenes,
Boris Starov, Christian Pagot, Claudio Menezes, Giovane Kuhn, Jo\~{a}o
Paulo Gois, Leonardo Schmitz, Rodrigo Mendes, and Tiago Etiene.
\end{acks}

% Bibliography
\bibliographystyle{ACM-Reference-Format-Journals}
\bibliography{acmtog-sample-bibfile}
                                % Sample .bib file with references that match those in
                                % the 'Specifications Document (V1.5)' as well containing
                                % 'legacy' bibs and bibs with 'alternate codings'.
                                % Gerry Murray - March 2012

\received{September 2008}{March 2009}

\end{document}
% End of v2-acmtog-sample.tex (March 2012) - Gerry Murray, ACM

### 2015-08-11 research_class/paper.tex
First pass of paper.md converted to paper.tex

% Based on: v2-acmtog-sample.tex, dated March 7 2012
%
% Compilation using 'acmtog.cls' - version 1.2 (March 2012), Aptara Inc.
% (c) 2010 Association for Computing Machinery (ACM)
%
\documentclass{acmtog} % V1.2

\usepackage{hyperref}

\usepackage{graphicx}
\DeclareGraphicsExtensions{.png,.jpg,.pdf,.eps}

%\acmVolume{VV}
%\acmNumber{N}
%\acmYear{YYYY}
%\acmMonth{Month}
%\acmArticleNum{XXX}
%\acmdoi{10.1145/XXXXXXX.YYYYYYY}

\acmVolume{28}
\acmNumber{4}
\acmYear{2009}
\acmMonth{September}
\acmArticleNum{106}
\doi{10.1145/1559755.1559763}

%ISSN
\issn{1234-56789}

% Copyright
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

\begin{document}

\markboth{J. D. Martin et al.}{Raft Distributed Consensus Algorithm over WebRTC (Web Real-Time Communication)}

\title{Raft Distributed Consensus Algorithm over WebRTC (Web Real-Time Communication)} % title

\author{Joel D. Martin {\upshape and} David Levine
\affil{University of Texas at Arglington}
}

%
% The code below should be generated by the tool at
% http://dl.acm.org/ccs.cfm
% Please copy and paste the code instead of the example below. 
%
\begin{CCSXML}
    <ccs2012>
    <concept>
    <concept_id>10003752.10003809.10010172</concept_id>
    <concept_desc>Theory of computation~Distributed
    algorithms</concept_desc>
    <concept_significance>500</concept_significance>
    </concept>
    </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Distributed algorithms}

%
% End generated code
%

\keywords{Distributed systems, Real-time chat, Distributed consensus, WWW, World Wide Web, Browsers}

%\acmformat{Vitor F. Pamplona, Manuel M. Oliveira, Gladimir V. G. Baranoski,
%and Sean Fogarty. 2009. Photorealistic models for pupil light reflex and iridal pattern deformation.
%{\em ACM Trans. Graph.} 28, 4, Article 106 (September 2009), 10 pages.\\
%\doiline}

\maketitle

%\begin{bottomstuff}
%Manuel M. Oliveira acknowledges a CNPq-Brazil fellowship (305613/2007-3). Gladimir V. G. Baranoski acknowledges a
%NSERC-Canada grant (238337). Microsoft Brazil provided additional support.
%Authors' addresses: Sean Fogarty, (Current address) NASA Ames Research Center, Moffett Field, California 94035.
%\end{bottomstuff}

\begin{abstract}
TODO: write an abstract.
\end{abstract}

\section{Introduction}
In the past few years, browsers have quickly evolved from document
rendering engines into powerful platforms for building sophisticated
applications. Modern browsers are integrating new protocols for
communication including WebSockets for full-duplex browser-to-server
communication and WebRTC for direct browser-to-browser communication.
These new capabilities enable software developers to create
distributed network applications with browser applications as first
class citizens.

These new browser capabilities also bring with them the traditional
challenges of building reliable distributed applications such as:
concurrency, security, scalability, naming, discovery, replication,
relocation, etc. The solutions to many of these challenges involve
the proper management of distributed state. Distributed consensus is
one category of distributed state management in which multiple nodes
must come to agreement on the state of single value or set of values.
The grandparent of distributed consensus algorithms is the Paxos
algorithm described by Leslie Lamport. However, Paxos has been
notoriously difficult to understand and to implement correctly. Raft
is a more recent distributed consensus algorithm that aims for
understandability and practicality of implementation.

In this study we will explore an implementation of the Raft protocol
in a browser context using WebRTC as the communication channel. We
will describe the results of the implementation and look at some of
unique challenges (and opportunities) we encountered with the
implementation. We will also explore some ideas for further study.

\section{Use Cases}
The traditional use of distributed consensus is in the data-center
(private or cloud based) especially as the core coordination component
of a replicated data-storage system. In these examples, the set of
nodes that compose the cluster tend to be fairly static and changes to
cluster membership are usually for the purpose of replacing nodes.
Changes to the long term size of the cluster are uncommon.

Most existing distributed browser applications are extensions of this
model in which browsers connect to load-balanced web servers which are
in close network proximity to the distributed consensus cluster. The
backend web system delivers the web application to the browser and
also serves as the conduit for all browser-to-browser communication
and state management.

However, it is not always desirable for application messages and state
to traverse the server back-end for various reasons including the
following:

\begin{enumerate}
\item \emph{security (privacy)}: application state and user messages may need to traverse directly from browser to browser rather than traversing through a server.
\item \emph{locality}: the nature of the application may allow clusters of applications that are nearby on a network (e.g. intranet) to communicate and coordinate directly rather than passing all messages and state through a server back-end which may be considerable more remote.
\item \emph{bandwidth and scale}: a distributed browser based application may be able to avoid expensive bandwidth into and out of the backend data-center by passing messages and state directly between browsers.  Depending on the design of the application this may also enable better scalability then a more centralized model.
\end{enumerate}

Here are some specific use cases in which Raft over WebRTC may be
applicable:

\begin{enumerate}
\item \emph{Consistent order secure chat}: a chat room system where all users see the same message order but messages never traverse through a central server even though the application code itself is delivered from a central web service.
\item \emph{Private multiuser document editing/whiteboard system}: Google Docs among other systems has demonstrated a powerful model for collaborative document creation. However, these systems all user central web services to coordinate shared state and handle browser-to-browser communication. Using a Raft over WebRTC could enable tools with similar functionality but without the central coordination and communication.
\item \emph{Browser based network games}: most network games have certain states that must be agreed upon by all nodes (e.g. avatar dead or alive, etc). This can be achieved with Raft over WebRTC without requiring any data to traverse a central server. In addition to reducing browser to server bandwidth requirements, this may also allow lower latency gaming if games instances are partitioned by latency locality.
\end{enumerate}

\section{Background}

\subsection{Raft}

Raft is a new distributed consensus algorithm designed to be
understandable and practical without sacrificing safety and
correctness.

The Raft algorithm breaks the problem of consensus into three core
subproblems: leader election, log replication, and safety. A full Raft
solution must also address: cluster membership changes, log
compaction, and client interaction. The Raft algorithm implemented and
discussed in this paper is based on "Consensus: Bridging Theory and
Practice" by Diego Ongaro (PhD Dissertation) TODO/CITE.

The Raft protocol is based on the concept of a distributed transaction
log. Entries in the log represent a sequence of changes to an internal
state machine. If all members of the Raft cluster have the same log
entries, then their state machine on each node will have exactly the
same state. Each node of a Raft cluster can be in one of three states:
follower, candidate, or leader. The responsibility of the leader is to
accept new transaction log entries and then replicated these entries
in the same order to all other members of the cluster. The leader is
the only member of the cluster that may make changes to the
transaction log. In order to maintain leadership, the leader sends
heartbeat RPCs to all the followers in the cluster. Followers that
do not receive a heartbeat RPC within a certain time period become
candidates and attempt to be elected by the other nodes as the new
leader of the cluster.

\subsubsection{Leader Election}

Each node of the Raft cluster maintains an ordinally increasing
current term value. When a Raft follower node does not receive
a heartbeat from a leader within a random election timeout period, it
transitions to candidate state, votes for itself, increases its
current term value by one and sends a requestVote RPC (with the
new term) to all the other nodes in the cluster. When a node receives
a requestVote RPC with a term that is greater than its own it will
become a follower (if not already), update its term to the new term,
and send a RPC response back that indicates a vote for that candidate.
However, a node may only vote for one candidate in a given term. If
a candidate receives votes from more than half of the cluster, then it
immediately becomes a leader and sends out a appendEntries RPC to
confirm leadership and reset the election timers on all the other
nodes.

\subsubsection{Log Replication}

Each Raft node maintains a transaction log. The entries in the
transaction log each contain a term value in addition to the action to
apply to the state machine. Each node also keep track of the most
recent log entry which is known to be committed (exists on more than
half the nodes in the cluster).

When a new leader is elected, it begins sending appendEntries RPCs
to other nodes in the Raft cluster. These RPC messages contain
information about the state of its transaction log: index and term of
the latest entry, and the most recently committed entry. If the node
does not have an entry matching the most recent entry on the leader,
then it replies with false to indicate that its log and not up to
date. The next appendEntries RPC that the leader sends to that
node will contain the next oldest index and term. This will continue
until the leader discovers the latest log entry which is agreement
(the node may not have any entries if it is new). Then the leader will
begin to propagate entries to the node in subsequent appendEntries
RPCs until the node is caught up. The leader continues sending
empty appendEntries as heartbeat RPCs to all the nodes until it
receives a new entry in its transaction log from a client.

\subsubsection{Safety}

In order to ensure that each state machine on every node executes
exactly the same commands in exactly the same order, the Raft system
provides safety guarantees. In particular, there are some restrictions
on which Raft nodes are actually eligible to become a leader based on
the state of their transaction log compared to other node transaction
logs. A Raft node will only vote for a candidate if the candidate has
a log that is more up-to-date than the voter. A log entry with a later
term is always more up-to-date. If the log entries have the same term,
then the entry with a higher index is more up-to-date. In addition,
Raft leaders may not consider entries from a previous term to be
committed until it has committed at least one an entry from its own
term.

Raft nodes must also persist certain properties to durable storage
before sending or receiving any RPCs. The properties that must be
persisted are current term, current vote for this term (if any), and
either the full transaction log or the state machine plus any
unapplied transaction log entries.

\subsubsection{Membership Changes}

Raft leverages the replicated transaction log to accomplish live
cluster membership changes. Membership changes are accomplished by
adding or removing one Raft node at a time using special add/remove
log entries. As soon as the new entry is added to the log it becomes
effective without waiting for the entry to be fully committed across
the cluster. However, a new cluster change entry may not be added to
the log until the previous one is committed. Once the change entry is
committed, the initiator of the change is notified, removed servers
can be shut down and another change entry may be added to the log.

\subsubsection{Log Compaction}

In order to keep the replicated transaction log from growing
indefinitely, Raft nodes should periodically compact their logs. There
are many different ways to accomplish log compaction and the best way
will depend on application requirements and on the specific nature of
the distributed state machine. For example, if the state machine is
simply a single shared counter, then any log entries before the most
recently committed entry may be safely discarded. The simplest and
most generic solution is often to provide some way of snapshotting the
entire state machine (including the term and log index it represents)
to disk at which point the all previously applied log entries can be
discarded. Snapshotting the state machine also means that the
implementation must provide a way for the leader to serialize and send
the current state machine to other nodes if their transaction log is
too old (e.g. when a new node is added).

\subsubsection{Client Interaction}

Clients of the Raft cluster are able to interact with the distributed
state machine by sending RPCs to the leader in order to add command
entries to the transaction log. Clients first find the address of any
node in the cluster either via broadcast or via an external directory
service. Once the client discovers a node of the cluster, that node
can either forward messages directly to the leader node (if it is not
the leader) or it can reject client requests with a response
indicating the address of the leader for the client to redirect to.

The Raft protocol should also provide strict linearizable semantics for
client commands/requests (reads and writes). In order to accomplish
this, each client is given a unique ID, and the client assigns
a unique serial number to each command. This prevents a single client
command from being applied twice to the state machine (in case of
a leader crash, or network duplication).

Linearizability is important not just for writes/updates to the state
machine, but also for reads, which means that read requests must also
be entered as transactions in the log and committed to more than half
the Raft nodes before the leader can respond to the client. If reads
bypass the transaction log then those reads are serializable but not
necessarily linearizable. For example, a leader may have been deposed
and not realize it if the cluster is partitioned which could result in
reads of stale data (the leader of the larger partition may have
already committed new entries).

\subsection{Raft.js}

Raft.js is an implementation of the Raft algorithm in JavaScript that
was creator by the author of this paper TODO/CITE. Raft.js is designed
to run in either a browser environment or within node.js (server-side
JavaScript). Raft.js implements the Raft algorithm as described in
"Consensus", Ongaro (dissertation TODO/CITE).

\subsubsection{Modular Design}

The implementation of the Raft algorithm is implemented in the
RaftServerBase class (in base.js). The base class does not directly
implement stateful functions such as RPC communication, durable
storage, scheduling/timeouts, or state machine command/management. In
order to create a working implementation these functions must be
provided either as configuration parameters when the class is
instantiated or by sub-classing the class.

The modular design of the Raft.js implementation allows it to be easily
used in several different contexts. For example, the RaftServerLocal
class (local.js) implements RPC calls using plain JavaScript function
calls between different instances of the class (nodes) in the same
execution context. The online Raft visualization at
http://kanaka.github.io/raft.js/ uses the RaftServerLocal class and
instantiates it using a scheduling functions that fire when the user
clicks on the "Step" button rather than due to the passage of time.

The RaftServerHttp class (`http.js`) embeds a simple web server into
each Raft node and then uses normal HTTP requests to send RPC messages
between Raft nodes.

For this paper, the RaftServerLocal class was extended with the
capability of sending RPCs over the WebRTC Data Channel.

\subsubsection{Differences between Raft and Raft.js}

The current Raft.js implementation is based on "Consensus", Ongaro and
implements most of the Raft algorithm including: leader election, log
replication, safety, membership changes, client interaction, and
read-only operation optimizations. Log compaction and full client
linearizability are not currently implemented (clients and client
transactions are not assigned unique IDs). These author intends to
implement these functions in the future, however these do not have
a significant impact on the ability to test Raft combined with WebRTC.

The Raft protocol description in "Consensus", Ongaro, has undergone
some revisions since the original draft paper (including incorporation
of suggestions by the author of this paper). These changes include
cluster membership simplifications, separating the concepts of
committing entries to the log and applying entries to the state
machine, etc. As part of the work for implementing Raft
over WebRTC, the Raft.js implementation was updated to reflect the
most recent Raft paper.

The are also some other differences between Raft.js and the Ongaro
paper:

\begin{enumerate}
\item The paper describes the Raft RPCs in terms of the traditional request/response model where requests and responses are automatically correlated together by a lower level of the software stack. For example with HTTP 1.0 a separate network session is established for each request/response pair (HTTP 1.0). Other examples including using a unique per session ID to correlate request and response.  WebRTC DataChannel connections are message based rather than request/response based. The original Raft definition of RPC and RPC response messages almost contain enough information to function correctly in an environment without builtin request/response correlation. Raft.js includes a small modification to the response RPCs for requestVote and appendEntries to add a 'sourceId' field.
\item The Raft paper is clear that each membership change must be delayed until all the previous membership change is committed (sequential).  It is implied that the leader node should track all pending membership change requests and apply them one at a time. However, in Raft.js, if a client requests a membership change while another change request is pending (uncommitted), the leader will reject the change request with a status of 'PENDING\_CONFIG\_CHANGE'. This keeps the core Raft implementation simpler with the trade-off being that the client implementation of membership change requests is slightly more complex.
\end{enumerate}

\subsection{WebRTC}

WebRTC is name of a collection of browser APIs and network protocols
that are currently being standardized the W3C and IETF respectively.
Collectively, these APIs and protocols work together to enable
real-time peer-to-peer video, audio and data communication between
browsers.

\subsubsection{WebRTC Discovery and Signaling}

One of the challenges with browser-to-browser communication (and
direct peer-to-peer communication in general) is that the web browser
environment or user agent (UA) is usually running behind a firewall in
a private internet subnet. This complicates direct communication in
several ways: the public Internet address of the user agent is unknown
and dynamic, inbound connections are denied by default, and outbound
connections are network address translated (NATed).

To establish a direct communication channel between browsers, WebRTC
uses mechanism that is similar to the Session Initiation Protocol
(SIP) commonly used for Voice-over-IP (VoIP) communication. This
involves a third party server known as a signaling server. Each user
agent (browser) connects to this server at a well known address in
order to establish a direct connection to other user agents. The
signaling server is also uses to communicate session control messages
(open, close, error) and to negotiate media capabilities. The signaling
server may optionally provide other services such as higher level
session management, HTTP serving of the web application, address book
service, client presence/availability, etc.

\subsubsection{WebRTC APIs}

The WebRTC browser APIs enable web applications written in JavaScript
to establish browser-to-browser network connectivity in order to
stream real-time video, audio and data communication. These APIs are
currently in the process of being standardized by the W3C organization
(TODO/CITE: http://www.w3.org/TR/webrtc/)

There are three primary WebRTC API definitions:

\begin{enumerate}
\item \href{http://www.w3.org/TR/webrtc/\#rtcpeerconnection-interface}{RTCPeerConnection}: The RTCPeerConnection API represents a direct connection between two browsers. Each RTCPeerConnection object has an associated Interactive Connectivity Establishment (ICE) agent that is used to establish network connectivity through firewalls and NAT services.

\item \href{http://www.w3.org/TR/webrtc/\#media-stream-api-extensions-for-network-use}{MediaStream (aka getUserMedia}: The MediaStream API represents a stream of audio or video data.  MediaStreams can easily be sent and received over RTCPeerConnections. The MediaStream API definition provides a mechanism that allows JavaScript to enumerate the native MediaStreams that are provided by the browser such as a local web camera or a live video stream of the desktop (for desktop sharing).  Although support for MediaStreams was the original impetus for the creation of the WebRTC APIs and protocols, the Raft over WebRTC implementation does not make use of the MediaStream API.

\item \href{http://www.w3.org/TR/webrtc/\#rtcdatachannel}{RTCDataChannel}: The RTCDataChannel API provides a mechanism for sending and receiving generic data. A single RTCPeerConnection can contain multiple RTCDataChannels. The RTCDataChannel API is modeled after the WebSocket API, however, unlike the WebSocket API which is reliable and guaranteed order transport (like TCP), each RTCDataChannel can be configured with different reliability and ordering settings. For example, by relaxing the reliability and ordering constraints on a RTCDataChannel, the average latency and jitter per message can be decreased even though some messages may be dropped or arrive out of order. The Raft protocol is specifically designed to support dropped or out-of-order messages so the transport options available with RTCDataChannel can be used to gain extra efficiency.
\end{enumerate}

\subsubsection{WebRTC Protocol}

There is a large suite of protocols that are required for a working
WebRTC system. The Internet Engineering Task Force (IETF) organization
is responsible for specifying and/or standardizing (where necessary)
the protocols that are part of WebRTC. In many cases an existing
protocol is applicable and is referenced or extended as part of the
WebRTC suite.

Describing all the protocols that make up the WebRTC suite is beyond
the scope of the this document. However, here is a (incomplete) list
of WebRTC protocols drafts and related protocols that are
part of WebRTC:

\begin{enumerate}
\item General:
    \begin{enumerate}
    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-overview-13}{WebRTC: Overview}
    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-security-arch-11}{WebRTC: Security Architecture}
    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-security-08}{WebRTC: Security Considerations}
    \end{enumerate}
\item Data transport:
    \begin{enumerate}
    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-transports-08}{WebRTC: Data transport overview}
    \item \href{https://tools.ietf.org/html/draft-ietf-tsvwg-rtcweb-qos-03}{WebRTC: DSCP and other packet markings for RTCWeb QoS}
    \item \href{https://tools.ietf.org/html/draft-ietf-tsvwg-sctp-dtls-encaps-09}{DTLS Encapsulation of SCTP Packets}
    \item \href{https://tools.ietf.org/html/draft-ietf-mmusic-sctp-sdp-14}{Stream Control Transmission Protocol (SCTP)-Based Media Transport in the Session Description Protocol (SDP)}
    \item \href{https://tools.ietf.org/html/draft-ietf-tsvwg-sctp-ndata-02}{Stream Schedulers and a New Data Chunk for the Stream Control Transmission Protocol}
    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-rtp-usage-23}{WebRTC: Media Transport and Use of RTP}
    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-data-channel-13}{WebRTC: Data Channels}
    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-data-protocol-09}{WebRTC: Data Channel Establishment Protocol}
    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-alpn-01}{WebRTC: Application Layer Protocol Negotiation for Web Real-Time Communications (WebRTC)}
    \item \href{https://tools.ietf.org/html/rfc5245}{RFC5245: Interactive Connectivity Establishment (ICE)}
    \item \href{https://tools.ietf.org/html/rfc5128}{RFC5128: Peer-to-Peer (P2P) Communication across Network Address Translators (NATs)}
    \item \href{https://tools.ietf.org/html/rfc5389}{RFC5389: Session Traversal Utilities for NAT (STUN)}
    \item \href{https://tools.ietf.org/html/rfc3489}{RFC3489: STUN - Simple Traversal of User Datagram Protocol (UDP) Through Network Address Translators (NATs)}
    \item \href{https://tools.ietf.org/html/rfc5766}{RFC5766: Traversal Using Relays around NAT (TURN)}
    \item \href{https://tools.ietf.org/html/rfc6156}{RFC6156: Traversal Using Relays around NAT (TURN) Extension for IPv6}
    \item \href{https://tools.ietf.org/html/rfc6544}{RFC6544: TCP Candidates with Interactive Connectivity  Establishment (ICE)}
    \item \href{https://tools.ietf.org/html/RFC4571}{RFC4571: Framing Real-time Transport Protocol (RTP) and RTP Control Protocol (RTCP) Packets over Connection-Oriented Transport}
    \item \href{https://tools.ietf.org/html/rfc5764}{RFC5764: Datagram Transport Layer Security (DTLS) Extension to Establish Keys for the Secure Real-time Transport Protocol (SRTP)}
    \end{enumerate}

\item Data framing and security
    \begin{enumerate}
    \item \href{https://tools.ietf.org/html/rfc3550}{RFC3550: Real-time Transport Protocol (RTP)}
    \item \href{https://tools.ietf.org/html/rfc3711}{RFC3711: Secure Real-time Transport Protocol (SRTP)}
    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-rtp-usage-23}{WebRTC: Media Transport and Use of RTP}
    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-data-channel-13}{WebRTC: Data Channels}
    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-data-protocol-09}{WebRTC: Data Channel Establishment Protocol}
    \end{enumerate}
\item Data formats (audio)
    \begin{enumerate}
    \item \href{https://tools.ietf.org/html/draft-cbran-rtcweb-codec-02}{WebRTC: WebRTC Codec and Media Processing Requirements}
    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-audio-07}{WebRTC: Audio Codec and Processing Requirements}
    \item \href{https://tools.ietf.org/html/rfc6716}{RFC6716: Opus Audio Codec}
    \item \href{https://tools.ietf.org/html/draft-ietf-payload-rtp-opus-11}{RTP Payload Format for Opus Speech and Audio Codec}
    \item \href{https://tools.ietf.org/html/rfc3551}{RFC3551: RTP Profile for Audio and Video Conference}
    \item \href{https://tools.ietf.org/html/rfc4733}{RFC4733: RTP Payload for DTMF Digits, Telephony Tones, and Telephony Signals}
    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-audio-codecs-for-interop-01}{WebRTC: Additional WebRTC audio codecs for interoperability with legacy networks}
    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-video-05}{WebRTC: Video Processing and Codec Requirements}
    \item \href{https://tools.ietf.org/html/rfc6386}{RFC6386: VP8 Data Format and Decoding Guide}
    \item \href{http://www.itu.int/rec/T-REC-H.264}{ITU H.264: Advanced video coding for generic audiovisual services}
    \item \href{https://tools.ietf.org/html/draft-ietf-payload-vp8-15}{RTP Payload Format for VP8 Video}
    \item \href{https://tools.ietf.org/html/rfc6236}{RFC6236: Negotiation of Generic Image Attributes in the Session Description Protocol (SDP)}
    \end{enumerate}

\item Connection management
    \begin{enumerate}
    \item \href{https://tools.ietf.org/html/rfc3264}{RFC3264: An Offer/Answer Model with the Session Description Protocol (SDP)}
    \item \href{https://tools.ietf.org/html/rfc2327}{RFC2327: SDP: Session Description Protocol}
    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-jsep-09}{WebRTC: Javascript Session Establishment Protocol (JSEP)}
    \item \href{https://tools.ietf.org/html/rfc5245}{RFC5245: Interactive Connectivity Establishment (ICE)}
    \item \href{https://tools.ietf.org/html/draft-ietf-mmusic-trickle-ice-02}{RFC5763: Trickle ICE: Incremental Provisioning of Candidates for the Interactive Connectivity Establishment (ICE) Protocol}
    \end{enumerate}
\end{enumerate}

Note: the signaling transport itself is not defined as part of WebRTC.
This is up to the application designer.

\subsection{PeerJS}

PeerJS is a project that provides a simplified abstraction for using
WebRTC. The first component of PeerJS is a node.js server library that
implements a WebRTC signaling server. The second component is
a JavaScript library that interacts with that signaling server and
also provides a simpler interface for using the WebRTC APIs.

The JavaScript interface use for establishing the initial connection
(signaling) in the WebRTC 1.0 standard is somewhat complicated.
A competing standard called Object Real-time Communication (ORTC) is
working to define a simpler and more modular JavaScript API for WebRTC
signaling. However, at the time of this work, the ORTC interface was
not yet widely supported and still in a state of flux.  The use of
PeerJS provides a suitable abstraction in the meantime (and will
likely be modified to support ORTC).

\subsubsection{PeerJS Server}

The PeerJS project provides a simple WebRTC signaling server called
PeerServer. This server uses HTTP and WebSockets transport protocols
to perform WebRTC signaling on behalf of browser clients. The
PeerServer server is also extensible so that more advanced services
can be built with it. The PeerJS project includes an extension to
PeerServer called ExpressPeerServer that combines a PeerJS signaling
server with the Node.js based Express web application framework
(TODO/CITE).

\subsubsection{PeerJS Client Library}

The PeerJS client library provides an simple abstraction over the
native browser WebRTC APIs. The WebRTC API standards are not yet
finalized and so different browsers and browser releases may support
different versions of the WebRTC draft APIs. The PeerJS client library
abstracts over these differences and provides a common interface so
that the application developer does not have to deal directly with
browser differences.

The PeerJS client library also provides a default signaling mechanism
that is designed to operate with a PeerJS signaling server. This is
particularly helpful because much of the WebRTC signaling transport
and protocol is not defined as part of the WebRTC standardization
effort. The PeerJS organization provides a cloud-hosted version of
a PeerJS compatible signaling server, however an application developer
can also run their own PeerJS signaling service using the PeerServer
server described above.

\section {Design and Implementation}

\begin{figure*}[Ht]
\centerline{\includegraphics[width=15cm]{raft_rtc_architecture}}
\caption{Raft over RTC Architecture}
  \label{fig:raft_rtc_architecture}
\end{figure*}

\subsection{Incremental Steps}

During the planning phase, the project was broken down into six
incremental steps:

\begin{enumerate}
\item Simple WebRTC messaging: implement a small JavaScript application that uses the PeerJS client library and an unmodified PeerJS signaling server. Demonstrate communication over WebRTC DataChannel using two browser instances (or browser tabs).

\item Raft.js over WebRTC: combine the existing Raft.js implementation with the small JavaScript application above to demonstrate Raft over WebRTC. This step uses a statically configured three node cluster (three browser tabs). The PeerJS signaling server is polled to retrieve the list and other clients and the Raft cluster is started when the number of PeerJS clients reaches three.

\item Updated Raft.js: update the Raft.js implementation to match the latest description of the Raft protocol (as described in "Consensus", Ongaro dissertation). This includes the implementation of the simplified membership change algorithm (sequential changes) and the separation of the log commit and state machine apply concepts. [Additionally, the new Raft.js implementation always starts a new Raft cluster as a cluster with a single node and additional nodes are dynamically added to the cluster (as suggested in section 4.4 of the paper).]

\item Dynamic Raft.js over WebRTC: enhance the signaling server with support for sessions/channels to enable multiple separate Raft clusters. Start each cluster with a single node and automatically add and remove Raft nodes as they come and go.

\item Visual representation: display Raft node state and RPC statistics in addition to the internal Raft log messages.

\item Server as Raft.js peer (aspirational): enable server nodes (Raft.js running on Node.js) to fully participate as Raft cluster nodes.
\end{enumerate}

Steps 1-5 were successfully implemented for this paper. Step 6 was an
aspirational goal because this functionality is not yet supported in
PeerJS although it is being tracked by issue report:
https://github.com/peers/peerjs/issues/103

\subsection{Server}

The file `rtc\_server.js` implements the signaling server for the Raft
over WebRTC application. The server extends the PeerJS
`ExpressPeerServer` object in order to provide WebRTC signaling. In
addition to the signaling function, the server also provides the
following service endpoints:

\begin{enumerate}
\item Static web service endpoint: this serves the static HTML, CSS and JavaScript that make up the Raft over WebRTC web application.

\item New channel endpoint: when a browser loads this endpoint, a new PeerJS channel is created and the browser is redirected to a URL for loading the main application (via static file service above).  The redirect URL has a query string appended that contains the channel ID. In addition the redirect URL has a fragment identifier (hash/history) that communicates to the web application that this is the first server in the channel.

\item Peer list endpoint: this endpoint takes a channel ID and returns the list of WebRTC peers (browser user agents) that are currently registered in that channel.

Refer to the "4.3 Client" section for a description of how these
endpoints interact.
\end{enumerate}

\subsection{Client}

\begin{figure*}[Ht]
\centerline{\includegraphics[width=15cm]{raft_rtc_sequence}}
\caption{Raft over RTC Sequence Diagram}
  \label{fig:raft_rtc_sequence}
\end{figure*}

The Raft over WebRTC testing application is designed as a single web
page (`rtc.html`). The application has different starting behavior
depending on the value of the URL query string (the URL component
following the "?" and up to the "\#") and the fragment identifier (the
URL component after the "\#").

When a new Raft cluster is created, the browser (user agent) connects
to the server new channel endpoint. This endpoint causes the server to
create a new channel for PeerJS signaling. Once this is complete the
server responds to the browser with a redirect message that contains
the new channel identifier in the query parameter and an indication
that this is the very first Raft cluster node in the fragment
identifier.

The browser follows the redirect and loads the static web application
at the static file endpoint provided by the server. This includes the
main application page (`rtc.html`), the core raft algorithm
(`base.js` and `local.js`), the PeerJS client library (`peer.js`), the
actual Raft over WebRTC implementation (`rtc.js`), and a layout
stylesheet (`web/demo.css`).

If the URL fragment identifier is set to "firstServer", then this Raft
node is initialized as the leader of a single node cluster and all
normal timers are started. The fragment identifier is then unset. The
query parameter contains the ID of this cluster (the PeerJS channel).

Subsequent nodes are added to the same Raft cluster by starting a new
browser context (separate browser, browser window, or browser tab) and
loading the same URL (which no longer has the "firstServer" fragment
identifier). There is a convenience link provided in the application
page that opens a new window with a new cluster node.

When the page first loads, a PeerJS signaling connection is
established with the server. Once this connection is established, the
application will be notified when any PeerJS clients connect to or
disconnect from the server. After the PeerJS signaling connection is
established, the client immediately queries the Peer list endpoint to
discover any clients that were connected prior to when this node
connected to the server.

The web application also begins a periodic async polling function
(`addRemoveServersAsync`) when the page loads. When the node is a Raft
leader, the function compares the current active list of PeerJS clients to
the Raft cluster node list. If there are any differences between the
two lists, then any new clients are added to the Raft cluster and
missing clients are removed from the Raft cluster one at a time using
the `addServer` and `removeServer` RPC calls. Servers additions are
prioritized before removals (as recommended in Ongaro 4.4 TODO/CITE)
in order to maximize availability.

When an instance of the web application starts that is not
a "firstServer", the normal Raft timers are not automatically enabled
until the server receives an RPC. The prevents the server from
possibly becoming a leader before it is incorporated into the cluster
by an appendEntries RPC from the "true" leader.

In addition to the low level Raft message log, the following
statistics are shown in the application:

\begin{enumerate}
\item \emph{Term}: the current Raft leadership term.
\item \emph{State}: the current Raft node state (follower, candidate, or leader)
\item \emph{Cluster Size}: a count of the current Raft cluster membership (number of nodes currently participating in the cluster)
\item \emph{Log Length}: number of entries in the transaction log.
\item \emph{Request Votes Sent/Received}: the number of `requestVote` RPCs that have been sent from and received by this node.
\item \emph{Append Entries Sent/Received}: the number of `appendEntries` RPCs that have been sent from and received by this node.
\end{enumerate}

\section{Outcome}

\subsection{Testing}

The initial way that the application was tested was by loading the
initial landing page to create a Raft over WebRTC cluster with
a single leader. Here is a screenshot of the browser after the first
node is created:

\begin{figure*}[Ht]
\centerline{\includegraphics[width=15cm]{browser_1}}
\caption{Raft over WebRTC Browser with One Node}
  \label{fig:browser_1}
\end{figure*}

Then four more Raft nodes were then added to the cluster by clicking
on the embedded "create a node" link in the application. Here is
a screenshot of the browser once four new nodes are added to the
cluster (5 total):

\begin{figure*}[Ht]
\centerline{\includegraphics[width=15cm]{browser_2}}
\caption{Raft over WebRTC Browser with Five Node}
  \label{fig:browser_2}
\end{figure*}

After every change was made to the cluster all the nodes were checked
to ensure that the following steady state properties were observed:
\begin{enumerate}
\item only a single leader
\item all nodes have the same term and transaction log count
\item all non-leader nodes are receiving `appendEntries` RPCs from the leader
\item all nodes have the same perception of the cluster size
\item the transaction log entries are identical on all nodes
\item the state machine is identical on all nodes
\end{enumerate}

The current leader was then terminated. The other nodes were checked
for the steady state properties. This was repeated until only two
nodes remained in the cluster.

After the five node cluster was brought up, state machine commands
were manually sent to leader to add to the transaction log. This was
done using the JavaScript console as there is not yet a way to do this
via the interface. Again the nodes were checked for the steady state
properties.

Then the application was tested was by loading the initial cluster
creation landing page in a browser tab. Then nodes were randomly added
to the cluster (by clicking on the link embedded in the page) or
removed from the cluster (by closing a tab). After each change the
steady state properties were checked.

Similar testing to the above was done with browser instances on two
separate computers. In addition, cross-browser functionality was
verified by running nodes from the same cluster running simultaneously
in Chrome 42 and Firefox 37.

\subsection{Results}

The results of the manual testing above showed that Raft over WebRTC
was able to perform leader elections correctly, the log and state
machine was replicated correctly to all nodes, and the safety
properties were maintained. During testing the membership of the
cluster was changed numerous times (both node additions and removals)
and the Raft properties were maintained throughout. In addition,
client interaction was tested although the application does not
provide a friendly interface for doing so.

One limitation of the system is that the if the cluster membership
drops from two nodes to one node then the cluster will become
unavailable without manual intervention. This is because when there
are two nodes in the cluster, both nodes are required for for
a majority vote and when one member is forcible shutdown then the
remaining cluster node will be unable to commit transaction entries
and the cluster will become unavailable. There are several options to
get around this problem:

\begin{enumerate}
\item The node can ask the leader to be removed and then wait for the
  transaction to become effective before it disconnects.  This may not
  always be an option especially in a browser context where the user
  may immediately kill the window/tab without giving the application
  a change to remove itself from the cluster.
\item The signaling server can serve as a tie-breaker. This would allow
  the remaining leader successfully commit the removal of the other
  node. In the case of a partition, the signaling server must only
  vote for one of the nodes.
\item The node with the lower ID could have an implicit tie-breaker vote.
  This would allow one node of the partition to continue in the case
  of a network partition. If a node dies or is killed, this would only
  allow the cluster to continue to operate if the node that dies
  happens to have the higher ID.
\item The system could allow manual user intervention. However, this is
  only likely to work correctly in the case where the user of the last
  remaining node can be certain that the other node will never return
  otherwise they is the risk of data-corruption.
\end{enumerate}

Running a Raft cluster in a browser context leads to some level of
timer instability. The JavaScript execution context is a fully event
driven environment with a single thread of execution for a given
JavaScript context. JavaScript code is run when browser (or Node.js)
events trigger JavaScript code that is registered to that event. In
order to implement logic that runs periodically, JavaScript code
is register as a callback timer using the setTimer, setInterval or
requestAnimationFrame APIs. However, none of these mechanisms is
a precise timer. When a timer fires, the callback code they refer to
is added to the JavaScript event queue and events on the queue are
processed in the order they arrive. This means callbacks may be
delayed by other code that is already running or queued first.

However, there is another browser mechanism that is less well known
but that causes even larger delays in timer code execution. Modern
browsers often have dozens or even hundreds of windows and/or tabs
open simultaneously. In order to reduce CPU load, increase user
responsiveness and improve battery life, most modern browsers will
throttle the JavaScript callback timers for all windows and tabs that
are hidden or have not been directly interacted with for some period
of time. This throttling can cause timers to slow down by an order of
magnitude or more.

In the context of Raft over WebRTC, the slowdown of background page
timers means that whenever the leader node is residing on a page that
is in the background, the propagation of transaction log entries slows
down substantially. This also means that other nodes which are in the
foreground are more likely to have an election timeout timer fire and
they are more likely to become a leader because they are more likely
to be able to send and receive votes quickly than nodes that are in
the background. In this sense the problem is self-correcting because
over the long term because higher performing nodes more are likely to
start and win elections. This also indicates dynamic adjustment of
timeout values may be a fruitful area of further study.

\section{Next Steps}

One of the useful outcomes of this project was that it brought to
light many areas where further study of Raft, WebRTC, and Raft over
WebRTC should be pursued.

Here are some interesting areas for further exploration:

\begin{enumerate}
\item Explore the use of alternate WebRTC transport modes for relaxing
  ordering constraints and delivery guarantees. Since the Raft
  protocol is tolerant of out-of-order and dropped messages, it is
  possible that Raft over WebRTC could be more efficient with these
  one or both of these constraints relaxed.
\item Test the survivability of the cluster when the signaling server is
  becomes completely unresponsive or just very slow.
\item Implement a full online chat system with linearizable guarantees for
  the chat message ordering.
\item Explore models for dynamically adjusting Raft timeout values to
  account for changing network and browser conditions.
\item Test the ability of the system to catch up stale nodes by
  disconnecting one or more nodes, run the cluster for some period
  time with transactions created periodically, and then reconnect the stale
  clients.
\item Implement and test forwarding client requests rather than default
  redirecting the client to the leader. In other words, allow client
  requests to originate from any node of the cluster and be redirected
  automatically to the leader (rather than the current manual
  process).
\item Perform quantitative and real-world testing to better characterize
  the performance, availability and scalability characteristics of the
  system.
\item Implement a user interface for sending commands to the transaction
  log and for viewing the current state of the Raft state machine.
\item Explore models for dealing with the case where the cluster goes from
  two nodes down to one node. Can usefulness and availability be
  improved without sacrificing Raft safety constraints.
\item Test the system in more various network environments. Determine the
  reliability of establishing direct browser-to-browser communication.
  Determine if corporate firewalls make this as difficult as is
  implied in some of the documentation and standards texts.
\item Implement and test forwarding client requests rather than default
  redirecting the client to the leader. In other words, allow client
  requests to originate from any node of the cluster and be redirected
  automatically to the leader (rather than the current manual
  process).
\item Perform quantitative and real-world testing to better characterize
  the performance, availability and scalability characteristics of the
  system.
\item Implement a user interface for sending commands to the transaction
  log and for viewing the current state of the Raft state machine.
\item Explore models for dealing with the case where the cluster goes from
  two nodes down to one node. Can usefulness and availability be
  improved without sacrificing Raft safety constraints.
\item Test the system in more various network environments. Determine the
  reliability of establishing direct browser-to-browser communication.
  Determine if corporate firewalls make this as difficult as is
  implied in some of the documentation and standards texts.
\item Explore models for timing out and automatically removing nodes
  when they become unresponsive without needing to be notified by the
  signaling server that the nodes have disconnected.
\item Test larger scale deployments across multiple systems. Determine the
  factors that affect scalability and characterize scaling limits.
\item Perform wider cross-browser testing including Internet Explorer and
  Opera.
\item Implement Raft log compaction.
\item Explore models that allow for a combination of strict shared state
  and high performance bulk data such as media or object blobs. For
  example, the data with full consensus might contain hashes that to
  the refer to bulk data.
\item Implement a Node.js based Raft over WebRTC application. This would
  allow non-browser contexts to participate in the Raft cluster
  together with the browser clients. PeerJS issue
  [103](https://github.com/peers/peerjs/issues/103) must be resolved
  for this to be feasible.
\item Explore a design where some nodes of the cluster are non-voting and
  are not counted as part of transaction commit quorum. However, these
  nodes would still be included in the log replication algorithm. One
  model this would enable would be a configuration where a small group
  of voting nodes serve as the hub for a larger group of non-voting
  nodes. This may enable higher scaling of the system. It also enables
  interesting applications such as a logging or backup nodes.
\item Implement and test forwarding client requests rather than default
  redirecting the client to the leader. In other words, allow client
  requests to originate from any node of the cluster and be redirected
  automatically to the leader (rather than the current manual
  process).
\item Perform quantitative and real-world testing to better characterize
  the performance, availability and scalability characteristics of the
  system.
\item Implement a user interface for sending commands to the transaction
  log and for viewing the current state of the Raft state machine.
\item Explore models for dealing with the case where the cluster goes from
  two nodes down to one node. Can usefulness and availability be
  improved without sacrificing Raft safety constraints.
\item Test the system in more various network environments. Determine the
  reliability of establishing direct browser-to-browser communication.
  Determine if corporate firewalls make this as difficult as is
  implied in some of the documentation and standards texts.
\item Explore models for timing out and automatically removing nodes
  when they become unresponsive without needing to be notified by the
  signaling server that the nodes have disconnected.
\item Test larger scale deployments across multiple systems. Determine the
  factors that affect scalability and characterize scaling limits.
\item Perform wider cross-browser testing including Internet Explorer and
  Opera.
\item Implement Raft log compaction.
\item Explore models that allow for a combination of strict shared state
  and high performance bulk data such as media or object blobs. For
  example, the data with full consensus might contain hashes that to
  the refer to bulk data.
\item Implement a Node.js based Raft over WebRTC application. This would
  allow non-browser contexts to participate in the Raft cluster
  together with the browser clients. PeerJS issue
  [103](https://github.com/peers/peerjs/issues/103) must be resolved
  for this to be feasible.
\item Explore a design where some nodes of the cluster are non-voting and
  are not counted as part of transaction commit quorum. However, these
  nodes would still be included in the log replication algorithm. One
  model this would enable would be a configuration where a small group
  of voting nodes serve as the hub for a larger group of non-voting
  nodes. This may enable higher scaling of the system. It also enables
  interesting applications such as a logging or backup nodes.
\end{enumerate}

\begin{acks}
I would like to thank my academic supervisor, David Levine, for his
excellent input and oversight of this project. I would also like to
thank Diego Ongaro for his amazing work on the Raft protocol and paper
which made implementing a Raft implementation a pleasure. There are
too many people involved in the creation and standardization of WebRTC
but I would like to thank them for their tireless work to make
browser-to-browser communication possible. Last but not least, I would
like to thank my family for putting up with the long nights and
weekends I spent on this project. Thanks!
\end{acks}

% Bibliography
\bibliographystyle{ACM-Reference-Format-Journals}
% TODO: references
\bibliography{acmtog-sample-bibfile}
                                % Sample .bib file with references that match those in
                                % the 'Specifications Document (V1.5)' as well containing
                                % 'legacy' bibs and bibs with 'alternate codings'.
                                % Gerry Murray - March 2012

%\received{September 2008}{March 2009}

\end{document}
% End of v2-acmtog-sample.tex (March 2012) - Gerry Murray, ACM

### 2015-08-11 rtc.js
test: fix channel force and addServer reporting.

@@ -150,7 +150,18 @@ function addRemoveServersAsync() {
         var diffNodes = function(map1, map2, rpc, argKey) {
             var peerIds = Object.keys(map1);
             for (var i=0; i<peerIds.length; i++) {
-                var peerId = peerIds[i];
+                var peerId = peerIds[i],
+                    cb = (function() {
+                        var id=peerId; // capture current val
+                        return function(res) {
+                            if (res.status === 'OK') {
+                                log("finished " + rpc + " of " + id);
+                            } else {
+                                log("could not " + rpc + " of " +
+                                    id + ": " + res.status);
+                            }
+                        };
+                    })();
                 if (!(peerId in map2)) {
                     changes += 1;
                     // Only make one modification each round
@@ -159,14 +170,7 @@ function addRemoveServersAsync() {
                     log(rpc + " of " + peerId);
                     var args = {};
                     args[argKey] = peerId;
-                    node[rpc](args, function(res) {
-                        if (res.status === 'OK') {
-                            log("finished " + rpc + " of " + peerId);
-                        } else {
-                            log("could not " + rpc + " of " +
-                                peerId + ": " + res.status);
-                        }
-                    });
+                    node[rpc](args, cb);
                 }
             }
         }

### 2015-08-11 rtc_server.js

@@ -38,13 +38,20 @@ function newPeerServerChannel(channel) {
 var channelCnt = 0;
 
 app.get('/', function(req, res, next) {
-    var channel = channelCnt++;
-    newPeerServerChannel(channel);
+    var channel = null;
     // Pass along query parameters
     var query = "";
     for (var k in req.query) {
         query += '&'+k+'='+req.query[k];
+        if (k === 'channel') {
+            channel = req.query[k];
+        }
     }
+    if (channel === null) {
+        channel = channelCnt++;
+    }
+    newPeerServerChannel(channel);
+
     // Full redirect URL
     var url = opts.home + '?channel=' + channel + query + '#firstServer';
     res.send('<html><head><meta http-equiv="refresh" content="0; url=' + url + '"/></head></html>');

### 2015-08-11 test/rtc.js

@@ -8,10 +8,11 @@
 
 var system = require('system'),
     webpage = require('webpage'),
+    channel = Math.round(Math.random()*1000000),
     base_address = system.args[1],
     server_count = (system.args.length >= 3) ? parseInt(system.args[2]) : 3;
 
-var query = "?channel=0&console_logging=true";
+var query = '?channel='+ channel + '&console_logging=true';
 
 function new_instance(page_id, firstServer) {
     var page = webpage.create();
@@ -33,7 +34,7 @@ function new_instance(page_id, firstServer) {
     if (!firstServer) {
         full_address += "/rtc.html";
     }
-    full_address += query; //+ (firstServer ? "#firstServer" : "");
+    full_address += query + (firstServer ? "#firstServer" : "");
 
     console.log("Opening " + full_address);
 

### 2015-08-11 test/chat.js
test/chat.js: load chat app.

@@ -1,0 +1,61 @@
+// Usage:
+// - Start rtc_server.js first:
+//   ./rtc_server --port 8000
+//
+// - Then use a docker build of slimerjs to run the test:
+//   IP_ADDR=$(hostname -I | awk '{print $1}')
+//   docker run -it -v `pwd`/test/rtc.js:/rtc.js fentas/slimerjs slimerjs /rtc.js http://${IP_ADDR}:8000/
+
+var system = require('system'),
+    webpage = require('webpage'),
+    channel = Math.round(Math.random()*1000000),
+    base_address = system.args[1],
+    server_count = (system.args.length >= 3) ? parseInt(system.args[2]) : 3;
+
+var query = '?channel='+ channel + '&console_logging=true';
+
+function new_instance(page_id, firstServer) {
+    var page = webpage.create();
+
+    // Register external handlers
+    page.onConsoleMessage = function (msg, line, origin) {
+        console.log("CONSOLE " + page_id + ": " + msg);
+    };
+    page.onCallback = function(msg) {
+        console.log("CALLBACK " + page_id + ": " + msg);
+        if (msg === 'QUIT') {
+            console.log("Normal exit");
+            slimer.exit(0);
+        }
+    };
+
+    var full_address = base_address;
+
+    if (!firstServer) {
+        full_address += "/chat.html";
+    }
+    full_address += query + (firstServer ? "#firstServer" : "");
+
+    console.log("Opening " + full_address);
+
+    page.open(full_address, function(status) {
+        if (status !== 'success') {
+            console.log('Unable to load the address!');
+            phantom.exit(1);
+        }
+        var mainTitle = page.evaluate(function () {
+            return document.title;
+        });
+        setTimeout(function() {
+            console.log('Saving image /tmp/chat' + page_id + '.png');
+            page.render('/tmp/chat' + page_id + '.png');
+        }, 10000);
+    });
+
+    return page;
+}
+
+new_instance(0, true);
+for (var i=1; i < server_count; i++) {
+    new_instance(i, false);
+}

### 2015-08-11 test/chat.js
test/chat.js: poll each node and exit when up.

@@ -1,30 +1,31 @@
 // Usage:
 // - Start rtc_server.js first:
-//   ./rtc_server --port 8000
+//   ./rtc_server --port 8000 --home chat.html
 //
 // - Then use a docker build of slimerjs to run the test:
 //   IP_ADDR=$(hostname -I | awk '{print $1}')
-//   docker run -it -v `pwd`/test/rtc.js:/rtc.js fentas/slimerjs slimerjs /rtc.js http://${IP_ADDR}:8000/
+//   docker run -it -v `pwd`/test/chat.js:/chat.js fentas/slimerjs slimerjs /chat.js http://${IP_ADDR}:8000/
 
 var system = require('system'),
     webpage = require('webpage'),
     channel = Math.round(Math.random()*1000000),
     base_address = system.args[1],
-    server_count = (system.args.length >= 3) ? parseInt(system.args[2]) : 3;
+    server_count = (system.args.length >= 3) ? parseInt(system.args[2]) : 3,
+    timeout = 10*1000;
 
 var query = '?channel='+ channel + '&console_logging=true';
 
-function new_instance(page_id, firstServer) {
+function new_page(page_id, firstServer) {
     var page = webpage.create();
 
     // Register external handlers
     page.onConsoleMessage = function (msg, line, origin) {
-        console.log("CONSOLE " + page_id + ": " + msg);
+        console.log('CONSOLE ' + page_id + ': ' + msg);
     };
     page.onCallback = function(msg) {
-        console.log("CALLBACK " + page_id + ": " + msg);
+        console.log('CALLBACK ' + page_id + ': ' + msg);
         if (msg === 'QUIT') {
-            console.log("Normal exit");
+            console.log('Normal exit');
             slimer.exit(0);
         }
     };
@@ -32,11 +33,11 @@ function new_instance(page_id, firstServer) {
     var full_address = base_address;
 
     if (!firstServer) {
-        full_address += "/chat.html";
+        full_address += '/chat.html';
     }
-    full_address += query + (firstServer ? "#firstServer" : "");
+    full_address += query + (firstServer ? '#firstServer' : '');
 
-    console.log("Opening " + full_address);
+    console.log('Opening ' + full_address);
 
     page.open(full_address, function(status) {
         if (status !== 'success') {
@@ -46,16 +47,90 @@ function new_instance(page_id, firstServer) {
         var mainTitle = page.evaluate(function () {
             return document.title;
         });
-        setTimeout(function() {
-            console.log('Saving image /tmp/chat' + page_id + '.png');
[…]

### 2015-08-18 rtc.js
test/chat.js: Add chat line, show timing information.

@@ -223,7 +223,7 @@ peer.on('open', function(id) {
                 rtcReceive(data);
             });
             conn.on('close', function(data) {
-                log("RTC connection closed:" + conn.peer);
+                //log("RTC connection closed:" + conn.peer);
                 delete nodeMap[peer_id];
             });
         });
@@ -259,7 +259,7 @@ peer.on('connection', function(conn) {
         rtcReceive(data);
     });
     conn.on('close', function(data) {
-        log("RTC connection closed:" + conn.peer);
+        //log("RTC connection closed:" + conn.peer);
         delete nodeMap[conn.peer];
     });
 });

### 2015-08-18 test/chat.js

@@ -13,7 +13,8 @@ var home = '/chat.html',
     base_address = system.args[1],
     server_count = (system.args.length >= 3) ? parseInt(system.args[2]) : 3,
     page_create_delay = 1,
-    timeout = (10 + server_count)*1000;
+    up_timeout = (10 + (server_count*server_count)/4)*1000;
+    pred_timeout = (1 + (server_count*server_count)/6)*1000;
 
 var query = '?channel='+ channel + '&console_logging=true';
 
@@ -42,7 +43,7 @@ function new_page(page_id, home, firstServer, callback) {
     }
     full_address += query + (firstServer ? '#firstServer' : '');
 
-    console.log('Opening ' + full_address);
+    console.log('Opening ' + page_id + ': ' + full_address);
 
     page.open(full_address, function(status) {
         if (status !== 'success') {
@@ -52,7 +53,7 @@ function new_page(page_id, home, firstServer, callback) {
         //var mainTitle = page.evaluate(function () {
         //    return document.title;
         //});
-        callback(page);
+        if (callback) { callback(page); }
     });
 
     return page;
@@ -77,13 +78,6 @@ function get_node_info(full) {
 function wait_cluster_up(timeout, callback) {
     var start_time = Date.now();
     var checkfn = function () {
-        /*
-        // Wait until all pages/nodes created
-        if (pages.length < server_count) {
-            setTimeout(checkfn, 100);
-            return;
-        }
-        */
         // Gather data from the nodes
         var nodes = [];
         for (var i=0; i < pages.length; i++) {
@@ -112,11 +106,35 @@ function wait_cluster_up(timeout, callback) {
             states.candidate.length === 0 && 
             states.follower.length === server_count-1 &&
             totalNodeCnt === server_count*server_count) {
-            callback(true, nodes);
+            callback(true, nodes, Date.now() - start_time);
         } else if (Date.now() - start_time > timeout) {
-            callback(false, nodes);
+            callback(false, nodes, Date.now() - start_time);
         } else {
             setTimeout(checkfn, 500);
+        }
+    }
+    checkfn();
+}
+
+function wait_cluster_predicate(predicate, timeout, callback) {
+    var start_time = Date.now();
+    var checkfn = function () {
+        // Gather data from the nodes
+        var results = [];
+        for (var i=0; i < pages.length; i++) {
+            results.push(pages[i].evaluate(predicate));
+        }
+
[…]

### 2015-08-18 test/rtc.js

@@ -13,7 +13,8 @@ var home = '/rtc.html',
     base_address = system.args[1],
     server_count = (system.args.length >= 3) ? parseInt(system.args[2]) : 3,
     page_create_delay = 1,
-    timeout = (10 + server_count)*1000;
+    up_timeout = (10 + (server_count*server_count)/4)*1000;
+    pred_timeout = (1 + (server_count*server_count)/6)*1000;
 
 var query = '?channel='+ channel + '&console_logging=true';
 
@@ -42,7 +43,7 @@ function new_page(page_id, home, firstServer, callback) {
     }
     full_address += query + (firstServer ? '#firstServer' : '');
 
-    console.log('Opening ' + full_address);
+    console.log('Opening ' + page_id + ': ' + full_address);
 
     page.open(full_address, function(status) {
         if (status !== 'success') {
@@ -52,7 +53,7 @@ function new_page(page_id, home, firstServer, callback) {
         //var mainTitle = page.evaluate(function () {
         //    return document.title;
         //});
-        callback(page);
+        if (callback) { callback(page); }
     });
 
     return page;
@@ -77,13 +78,6 @@ function get_node_info(full) {
 function wait_cluster_up(timeout, callback) {
     var start_time = Date.now();
     var checkfn = function () {
-        /*
-        // Wait until all pages/nodes created
-        if (pages.length < server_count) {
-            setTimeout(checkfn, 100);
-            return;
-        }
-        */
         // Gather data from the nodes
         var nodes = [];
         for (var i=0; i < pages.length; i++) {
@@ -112,11 +106,35 @@ function wait_cluster_up(timeout, callback) {
             states.candidate.length === 0 && 
             states.follower.length === server_count-1 &&
             totalNodeCnt === server_count*server_count) {
-            callback(true, nodes);
+            callback(true, nodes, Date.now() - start_time);
         } else if (Date.now() - start_time > timeout) {
-            callback(false, nodes);
+            callback(false, nodes, Date.now() - start_time);
         } else {
             setTimeout(checkfn, 500);
+        }
+    }
+    checkfn();
+}
+
+function wait_cluster_predicate(predicate, timeout, callback) {
+    var start_time = Date.now();
+    var checkfn = function () {
+        // Gather data from the nodes
+        var results = [];
+        for (var i=0; i < pages.length; i++) {
+            results.push(pages[i].evaluate(predicate));
[…]

### 2015-08-19 research_class/pgfplots/tex/generic/pgfplots/libs/tikzlibrarypgfplots.fillbetween.code.tex
Add pgfplots_1.12.1.tds.zip

%--------------------------------------------
%
% Package pgfplots
%
% Provides a user-friendly interface to create function plots (normal
% plots, semi-logplots and double-logplots).
% 
% It is based on Till Tantau's PGF package.
%
% Copyright 2013 by Christian Feuersaenger
%
% This program is free software: you can redistribute it and/or modify
% it under the terms of the GNU General Public License as published by
% the Free Software Foundation, either version 3 of the License, or
% (at your option) any later version.
% 
% This program is distributed in the hope that it will be useful,
% but WITHOUT ANY WARRANTY; without even the implied warranty of
% MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
% GNU General Public License for more details.
% 
% You should have received a copy of the GNU General Public License
% along with this program.  If not, see <http://www.gnu.org/licenses/>.
%
%--------------------------------------------

% This library adds support for high-level instructions for "fill area
% between two arbitrary plots of functions".
%
% It activates the syntax
%
% 	\addplot fill between[of=A and B];
%
% where A and B are two plots named by 'name path='.
%
% In fact, this here is not much more than a low-level invocation of
% \pgfpathfillbetween
% and a couple of styles. It could become a TikZ library because it
% actually works on any two named paths, but it has its restrictions
% regarding the supported input paths: both need to be plots of
% functions (non-intersecting, should have at most one function value
% for each canvas X coord)

\pgfutil@IfUndefined{pgfplotsset}{%
	\pgferror{Please load pgfplots before pgfplots.fillbetween.}%
	\endinput
}{}%

\usetikzlibrary{fillbetween}
\usepgfplotslibrary{decorations.softclip}

\pgfkeys{
	/pgfplots/every fill between plot/.style={/pgfplots/area legend,/tikz/fill},
	/tikz/soft clip assign/domain/.code args={#1:#2}{%
		\pgfplotsifisvisualizationphase{%
			\edef\pgfplots@loc@TMPa{%
				(axis cs:#1,\pgfkeysvalueof{/pgfplots/ymin}) 
				rectangle (axis cs:#2,\pgfkeysvalueof{/pgfplots/ymax})}%
			\pgfkeysalso{/tikz/soft clip assign/path/.expand once=\pgfplots@loc@TMPa}%
		}{}%
	},%
	/tikz/soft clip assign/y domain y/.style={/tikz/soft clip assign/domain y={#1}},
	/tikz/soft clip assign/domain y/.code args={#1:#2}{%
		\pgfplotsifisvisualizationphase{%
			\edef\pgfplots@loc@TMPa{%
				(axis cs:\pgfkeysvalueof{/pgfplots/xmin},#1) 
				rectangle (axis cs:\pgfkeysvalueof{/pgfplots/xmax},#2)}%
			\pgfkeysalso{/tikz/soft clip assign/path/.expand once=\pgfplots@loc@TMPa}%
		}{}%
	},%
	%
	/pgfplots/execute at begin axis@@/.add={}{%
		\gdef\pgfplotsfillbetween@list@of@layers@with@clippath{}%
		\gdef\b@pgfplotsfillbetween@list@has@set@layers{0}%
	},
}

% #1: the value of /tikz/fill between/on layer
\def\pgfplotsfillbetween@ensure@clipping@on@layer{%
	\ifpgfplots@clip
		\pgfkeysgetvalue{/tikz/fill between/on layer}\pgfplots@loc@TMPa
		\edef\pgfplots@loc@TMPa{\pgfplots@loc@TMPa}%
		%
		% check if 'pre main,' is inside of the already prepared layers:
		\def\pgfplots@loc@TMPb{\expandafter\pgfutil@in@\expandafter{\pgfplots@loc@TMPa,}}%
		\expandafter\pgfplots@loc@TMPb\expandafter{\pgfplotsfillbetween@list@of@layers@with@clippath}%
		%
		\ifpgfutil@in@
			% ah - already prepared. Good, nothing left to do.
		\else
			\pgfplots@log3{fill between: found 'clip=true'. Preparing pgfplotsextra instruction with clip path for layer '\pgfplots@loc@TMPa'...}%
			\xdef\pgfplotsfillbetween@list@of@layers@with@clippath{\pgfplots@loc@TMPa,\pgfplotsfillbetween@list@of@layers@with@clippath}%
			%
			\edef\pgfplots@loc@TMPb{%
				\noexpand\pgfplotsextra{%
					\noexpand\pgfplotsonlayer{\pgfplots@loc@TMPa}%
					\noexpand\pgfplotspathaxisoutline
					\noexpand\pgfusepath{clip}%
					\noexpand\endpgfplotsonlayer
				}%
			}%
			\pgfplots@loc@TMPb
			%
		\fi
	\fi
}%

% The options are from 
% \addplot[#1] fill between[#2] #3;
%
% This method is invoked from the global \addplot dispatching.
\def\pgfplotslibraryfillbetween@addplot#1#2#3{%
	%
	\pgfplots@start@plot@with@behavioroptions{/pgfplots/every fill between plot,#1}%
	%
	\pgfkeysvalueof{/pgfplots/execute at end survey}%
	\pgfplots@remember@survey@option@list
	%
	%
	\pgfplotslibraryfillbetweenpreparecurrentlayer
	%
	\pgfplots@addplot@enqueue@coords{%
		% precmd
	}{%
		% empty - this here is a TikZ instruction, not a "real" \addplot command
	}{%
		% the path instruction:
		\tikzfillbetween[#2]{%
			/pgfplots/.search also=/tikz,%
			/pgfplots/.cd,%
			/tikz/fill between/every last segment/.append style={/tikz/fill between/path after segment={#3}},%
			#1%
		}%
	}{%	
		% post command.
	}%
}

% ONLY TO BE USED AS LATER OPTION!
\pgfkeysdef{/pgfplots/fill between/@ensure layers}{%
	\ifpgfplots@layered@graphics
	\else
		\pgfplots@log3{fill between: activating layered graphics}%
		\pgfplotsset{set layers}%
		\pgfplots@set@options@of@layered@graphics
	\fi
}%

% SIDE-EFFECT: modifies \b@pgfplotsfillbetween@list@has@set@layers and
% \pgfplotsfillbetween@list@of@layers@with@clippath
\def\pgfplotslibraryfillbetweenpreparecurrentlayer{%
	\pgfkeysgetvalue{/tikz/fill between/on layer}\pgfplots@loc@TMPa
	\ifx\pgfplots@loc@TMPa\pgfutil@empty
	\else
		\if0\b@pgfplotsfillbetween@list@has@set@layers
			%
			\pgfplotssetlateoptions{fill between/@ensure layers}%
			\gdef\b@pgfplotsfillbetween@list@has@set@layers{1}%
		\fi
		\pgfplotsfillbetween@ensure@clipping@on@layer%
	\fi
}

\def\pgfplots@tikzfillbetween{\pgfutil@ifnextchar[{\pgfplots@tikzfillbetween@opt}{\pgfplots@tikzfillbetween@opt[]}}
\def\pgfplots@tikzfillbetween@opt[#1]#2{%
	\pgfplots@addplot@enqueue@coords{%
		% precmd
	}{%
		% empty - this here is a TikZ instruction, not a "real" \addplot command
	}{%
		% the path instruction:
		\tikzfillbetween[#1]{#2}%
	}{%	
		% post command.
	}%
}%

\expandafter\def\expandafter\pgfplots@replace@path@commands\expandafter{\pgfplots@replace@path@commands
	\let\pgfplots@orig@tikzfillbetween=\tikzfillbetween
	\let\tikzfillbetween=\pgfplots@tikzfillbetween
}%
\expandafter\def\expandafter\pgfplots@restore@path@commands\expandafter{\pgfplots@restore@path@commands
	\let\tikzfillbetween=\pgfplots@orig@tikzfillbetween
}

\endinput

### 2015-08-19 research_class/pgfplots/tex/generic/pgfplots/libs/tikzlibrarypgfplots.groupplots.code.tex
Add pgfplots_1.12.1.tds.zip

%--------------------------------------------
%
% Package pgfplots
%
% Provides a user-friendly interface to create function plots (normal
% plots, semi-logplots and double-logplots).
% 
% It is based on Till Tantau's PGF package.
%
% Copyright 2009 by Nick Papior Andersen.
%
% This program is free software: you can redistribute it and/or modify
% it under the terms of the GNU General Public License as published by
% the Free Software Foundation, either version 3 of the License, or
% (at your option) any later version.
% 
% This program is distributed in the hope that it will be useful,
% but WITHOUT ANY WARRANTY; without even the implied warranty of
% MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
% GNU General Public License for more details.
% 
% You should have received a copy of the GNU General Public License
% along with this program.  If not, see <http://www.gnu.org/licenses/>.
%
%--------------------------------------------

\newif\ifpgfplots@group@sharedlabels

\def\pgfplots@group@xticklabels{}
\def\pgfplots@group@yticklabels{}
\def\pgfplots@group@xlabels{}
\def\pgfplots@group@ylabels{}
\def\pgfplots@group@totalplots{1}
\def\pgfplots@group@columns{1}
\def\pgfplots@group@rows{1}
\def\pgfplots@group@name{group}
\def\pgfplots@group@style{}
\gdef\pgfplots@group@rmopts{}
\newcount\pgfplots@group@current@plot
\newcount\pgfplots@group@current@row
\newcount\pgfplots@group@current@column

\usetikzlibrary{calc}

\pgfkeys{
    /pgfplots/group/.is family,
    /pgfplots/group style/.store in=\pgfplots@group@style,
    /pgfplots/group/horizontal sep/.initial=1cm,
    /pgfplots/group/vertical sep/.initial=1cm,
    /pgfplots/group/group size/.style args={#1 by #2}{/pgfplots/group/columns=#1,/pgfplots/group/rows=#2},
    /pgfplots/group/group size/.default={1 by 1},
    /pgfplots/group/columns/.store in=\pgfplots@group@columns,
    /pgfplots/group/rows/.store in=\pgfplots@group@rows,
    /pgfplots/group/every plot/.style=,
    /pgfplots/group/shared labels/.is if=pgfplots@group@sharedlabels,
    /pgfplots/group/shared labels=false,
    /pgfplots/group/xticklabels at/.is choice,
    /pgfplots/group/xticklabels at/all/.code=\def\pgfplots@group@xticklabels{all},
    /pgfplots/group/xticklabels at/edge top/.code=\def\pgfplots@group@xticklabels{top},
    /pgfplots/group/xticklabels at/edge bottom/.code=\def\pgfplots@group@xticklabels{lower},
    /pgfplots/group/xticklabels at/edge left/.code=\def\pgfplots@group@xticklabels{left}, % TODO
    /pgfplots/group/xticklabels at/edge right/.code=\def\pgfplots@group@xticklabels{right}, % TODO
    /pgfplots/group/xticklabels at=all,
    /pgfplots/group/yticklabels at/.is choice,
    /pgfplots/group/yticklabels at/all/.code=\def\pgfplots@group@yticklabels{all},
    /pgfplots/group/yticklabels at/edge left/.code=\def\pgfplots@group@yticklabels{left},
    /pgfplots/group/yticklabels at/edge right/.code=\def\pgfplots@group@yticklabels{right},
    /pgfplots/group/yticklabels at/edge top/.code=\def\pgfplots@group@yticklabels{top}, % TODO
    /pgfplots/group/yticklabels at/edge bottom/.code=\def\pgfplots@group@yticklabels{bottom}, % TODO
    /pgfplots/group/yticklabels at=all,
    /pgfplots/group/xlabels at/.is choice,
    /pgfplots/group/xlabels at/all/.code=\def\pgfplots@group@xlabels{all},
    /pgfplots/group/xlabels at/edge top/.code=\def\pgfplots@group@xlabels{top},
    /pgfplots/group/xlabels at/edge bottom/.code=\def\pgfplots@group@xlabels{lower},
    /pgfplots/group/xlabels at/edge right/.code=\def\pgfplots@group@xlabels{right}, % TODO
    /pgfplots/group/xlabels at/edge left/.code=\def\pgfplots@group@xlabels{left},% TODO
    /pgfplots/group/xlabels at=all,
    /pgfplots/group/ylabels at/.is choice,
    /pgfplots/group/ylabels at/all/.code=\def\pgfplots@group@ylabels{all},
    /pgfplots/group/ylabels at/edge left/.code=\def\pgfplots@group@ylabels{left},
    /pgfplots/group/ylabels at/edge right/.code=\def\pgfplots@group@ylabels{right},
    /pgfplots/group/ylabels at/edge top/.code=\def\pgfplots@group@ylabels{top}, % TODO
    /pgfplots/group/ylabels at/edge bottom/.code=\def\pgfplots@group@ylabels{bottom}, % TODO
    /pgfplots/group/ylabels at=all,
    /pgfplots/group/x descriptions at/.is choice,
    /pgfplots/group/x descriptions at/all/.style={
        /pgfplots/group/xlabels at=all,
        /pgfplots/group/xticklabels at=all
    },
    /pgfplots/group/x descriptions at/edge bottom/.style={
        /pgfplots/group/xlabels at=edge bottom,
        /pgfplots/group/xticklabels at=edge bottom
    },
    /pgfplots/group/x descriptions at/edge top/.style={
        /pgfplots/group/xlabels at=edge top,
        /pgfplots/group/xticklabels at=edge top
    },
    /pgfplots/group/y descriptions at/.is choice,
    /pgfplots/group/y descriptions at/all/.style={
        /pgfplots/group/ylabels at=all,
        /pgfplots/group/yticklabels at=all
    },
    /pgfplots/group/y descriptions at/edge left/.style={
        /pgfplots/group/ylabels at=edge left,
        /pgfplots/group/yticklabels at=edge left
    },
    /pgfplots/group/y descriptions at/edge right/.style={
        /pgfplots/group/ylabels at=edge right,
	/pgfplots/group/yticklabels at=edge right
    },
    /pgfplots/group/group name/.store in=\pgfplots@group@name,
    /pgfplots/group/empty plot/.style={/pgfplots/hide axis=true},
	/pgfplots/trim axis group left/.code={%
		\pgfutil@ifundefined{pgfsettrimleft}{%
			\pgfplots@error{Sorry, the 'trim axis group left' requires a more recent PGF version.}%
		}{%
			\pgfkeysalso{/tikz/trim left={(group c1r1.south west)}}%
		}%
	},%
	/tikz/trim axis group left/.style={/pgfplots/trim axis group left},
	/pgfplots/trim axis group right/.code={%
		\pgfutil@ifundefined{pgfsettrimleft}{%
			\pgfplots@error{Sorry, the 'trim axis group right' requires a more recent PGF version.}%
		}{%
			\pgfkeysalso{/pgfplots/trim axis right}%
		}%
	},%
	/tikz/trim axis group right/.style={/pgfplots/trim axis group right},
}
\pgfkeys{
	/pgfplots/warning/groupplots/too many plots/.code 2 args={%
		\pgfplotsthrow@warning{You are adding more plots than you specified, current plots=#1\space vs. total allowed
          plots=#2.}%
	},
}

\def\nextgroupplot{%
    \pgfutil@ifnextchar[{\pgfplots@group@nextplot}{\pgfplots@group@nextplot[]}%}
}

%
% The actual addition of the new plots
%
\def\pgfplots@group@nextplot[#1]{%
    % Check if it legal to produce another plot
    \ifnum\pgfplots@group@current@plot=\pgfplots@group@totalplots\relax
      \pgfplotswarning{groupplots/too many plots}{\the\pgfplots@group@current@plot}{\pgfplots@group@totalplots}\pgfeov
    \else
      % 
      % If there has been other plots before then end the plot
      % 
      \ifnum0<\pgfplots@group@current@plot\relax
        % DEBUG: \message{DEBUG: Ending plot}
        \endpgfplots@environment@opt
      \fi
      % Prepare for next plot, increment plot, row and column
      \pgfplots@group@increment@numbers
      % DEBUG: \message{DEBUG: Cur. Plot: \the\pgfplots@group@current@plot. Current Row: \the\pgfplots@group@current@row. Current Column: \the\pgfplots@group@current@column.}    
      % Creation of the new axis environment
      \pgfplots@group@environment@create{#1}
    \fi
}
%
% This creates an environment according to which plots that have been typeset.
% The code utilizes previous plots anchor points to position themselves, unless it's the 
% first plot.
% Using the figure below to describe the code implementation:
% ____    ____   ____
%|        | |        | |        |
%|   1   | |   2   | |   3   |
%|___| |___| |___|
% ____    ____   ____
%|        | |        | |        |
%|   4   | |   5   | |   6   |
%|___| |___| |___|
% ____    ____   ____
%|        | |        | |        |
%|   7   | |   8   | |   9   |
%|___| |___| |___|
% 
% The plots anchors are as follows: 
%    1: plot is typeset normally
%    2: the "1" plot anchor "east" is anchored to "west"
%    3: the "2" plot anchor "east" is anchored to "west"
%    4: the "1" plot anchor "south" is anchored to "north"
%    5: the "2" plot anchor "south" is anchored to "north"
%    6: the "3" plot anchor "south" is anchored to "north"
%    7: the "4" plot anchor "south" is anchored to "north"
%    8: the "5" plot anchor "south" is anchored to "north"
%    9: the "6" plot anchor "south" is anchored to "north"
%
\def\pgfplots@group@environment@create#1{%
    \ifnum1=\pgfplots@group@current@row\relax
      \ifnum1=\pgfplots@group@current@column\relax
        \pgfplots@environment@opt[name={\pgfplots@group@name\space c\the\pgfplots@group@current@column r\the\pgfplots@group@current@row},%
          /pgfplots/group/every plot,%
          /pgfplots/group/plot c\the\pgfplots@group@current@column r\the\pgfplots@group@current@row,%
          #1]
      \else
        \pgfplotsgroupplotsmath@decrease\pgfplots@group@current@column
        \xdef\pgfplots@glob@TMPb{\pgfplots@group@name\space c\pgfmathresult r\the\pgfplots@group@current@row}
        \pgfplots@environment@opt[name={\pgfplots@group@name\space c\the\pgfplots@group@current@column r\the\pgfplots@group@current@row},%
          /pgfplots/group/every plot,%
          /pgfplots/group/plot c\the\pgfplots@group@current@column r\the\pgfplots@group@current@row,%
          at={($(\pgfplots@glob@TMPb.east)+(\pgfkeysvalueof{/pgfplots/group/horizontal sep},0)$)},anchor=west,%
          #1]
      \fi
    \else
      \pgfplotsgroupplotsmath@decrease\pgfplots@group@current@row
      \xdef\pgfplots@glob@TMPb{\pgfplots@group@name\space c\the\pgfplots@group@current@column r\pgfmathresult}
      \pgfplots@environment@opt[name={\pgfplots@group@name\space c\the\pgfplots@group@current@column r\the\pgfplots@group@current@row},%
        /pgfplots/group/every plot,%
        /pgfplots/group/plot c\the\pgfplots@group@current@column r\the\pgfplots@group@current@row,%
        at={($(\pgfplots@glob@TMPb.south)-(0,\pgfkeysvalueof{/pgfplots/group/vertical sep})$)},anchor=north,%
        #1]%
    \fi%
}

%
% Determine the labels of both x and y.
%
\def\pgfplots@group@determine@labels{%
    \def\pgfplots@loc@TMPa{all}
    \def\pgfplots@loc@TMPb{lower}
    \ifx\pgfplots@loc@TMPa\pgfplots@group@xlabels%
      % xlabels at=all
      % Do nothing as all should have xlabels
    \fi
    \gdef\pgfplots@glob@TMPa{}
    \ifx\pgfplots@loc@TMPb\pgfplots@group@xlabels%
	  % xlabels at=lower
      \ifnum1<\pgfplots@group@rows\relax
        \foreach \pgfplots@column in {1,...,\pgfplots@group@columns} {%
            \foreach \pgfplots@row in {1,...,\pgfplots@group@rows} {%
                \ifnum\pgfplots@row=\pgfplots@group@rows\relax%
                \else
                \expandafter\xdef\expandafter\pgfplots@glob@TMPa\expandafter{%
                    \pgfplots@glob@TMPa,/pgfplots/group/plot c\pgfplots@column r\pgfplots@row/.append style={/pgfplots/xlabel={}}}%
                \fi
            }%
        }%
      \fi%
	\else
		\def\pgfplots@loc@TMPb{top}%
		\ifx\pgfplots@loc@TMPb\pgfplots@group@xlabels%
			% xlabels at=top
			\ifnum1<\pgfplots@group@rows\relax
				\foreach \pgfplots@column in {1,...,\pgfplots@group@columns} {%
					\foreach \pgfplots@row in {1,...,\pgfplots@group@rows} {%
						\ifnum\pgfplots@row=1 %
						\expandafter\xdef\expandafter\pgfplots@glob@TMPa\expandafter{%
							\pgfplots@glob@TMPa,/pgfplots/group/plot c\pgfplots@column r\pgfplots@row/.append style={/pgfplots/xticklabel pos=upper}}%
						\else
						\expandafter\xdef\expandafter\pgfplots@glob@TMPa\expandafter{%
							\pgfplots@glob@TMPa,/pgfplots/group/plot c\pgfplots@column r\pgfplots@row/.append style={/pgfplots/xlabel={}}}%
						\fi
					}%
				}%
			\fi%
		\fi
    \fi
    % 
    % Defining styles for the y's
    % 
    \ifx\pgfplots@loc@TMPa\pgfplots@group@ylabels%
	  % ylabels at=all
      % Do nothing as all should have ylabels
	\else
		\def\pgfplots@loc@TMPb{left}%
		\ifx\pgfplots@loc@TMPb\pgfplots@group@ylabels%
			% ylabels at=left
			\ifnum1<\pgfplots@group@columns\relax
				\foreach \pgfplots@row in {1,...,\pgfplots@group@rows} {%
					\foreach \pgfplots@column in {2,...,\pgfplots@group@columns} {%
						\expandafter\xdef\expandafter\pgfplots@glob@TMPa\expandafter{%
							\pgfplots@glob@TMPa,/pgfplots/group/plot c\pgfplots@column r\pgfplots@row/.append style={/pgfplots/ylabel={}}}%
					}%
				}%
			\fi%
		\else
			\def\pgfplots@loc@TMPb{right}%
			\ifx\pgfplots@loc@TMPb\pgfplots@group@ylabels%
				% ylabels at=right
				\ifnum1<\pgfplots@group@columns\relax
					\foreach \pgfplots@row in {1,...,\pgfplots@group@rows} {%
						\foreach \pgfplots@column in {1,...,\pgfplots@group@columns} {%
							\ifnum\pgfplots@column=\pgfplots@group@columns\relax
							\expandafter\xdef\expandafter\pgfplots@glob@TMPa\expandafter{%
								\pgfplots@glob@TMPa,/pgfplots/group/plot c\pgfplots@column r\pgfplots@row/.append style={/pgfplots/yticklabel pos=upper}}%
							\else
							\expandafter\xdef\expandafter\pgfplots@glob@TMPa\expandafter{%
								\pgfplots@glob@TMPa,/pgfplots/group/plot c\pgfplots@column r\pgfplots@row/.append style={/pgfplots/ylabel={}}}%
							\fi
						}%
					}%
				\fi
			\fi%
		\fi
    \fi
    \expandafter\pgfkeys\expandafter{\pgfplots@glob@TMPa}
}

%
% Determine the ticklabels of both x and y.
%
\def\pgfplots@group@determine@ticklabels{%
    \def\pgfplots@loc@TMPa{all}
    \def\pgfplots@loc@TMPb{top}
    \def\pgfplots@loc@TMPc{lower}
    \ifx\pgfplots@loc@TMPa\pgfplots@group@xticklabels%
      % Do nothing as all should have xticklabels
    \fi
    \gdef\pgfplots@glob@TMPa{}
    \ifx\pgfplots@loc@TMPb\pgfplots@group@xticklabels%
      \foreach \pgfplots@column in {1,...,\pgfplots@group@columns} {%
          \expandafter\xdef\expandafter\pgfplots@glob@TMPa\expandafter{%
              \pgfplots@glob@TMPa,/pgfplots/group/plot c\pgfplots@column r1/.append style={xticklabel pos=upper}}%
          \ifnum1<\pgfplots@group@rows\relax
            \foreach \pgfplots@row in {2,...,\pgfplots@group@rows} {%
                \expandafter\xdef\expandafter\pgfplots@glob@TMPa\expandafter{%
                    \pgfplots@glob@TMPa,/pgfplots/group/plot c\pgfplots@column r\pgfplots@row/.append style={/pgfplots/xticklabels={}}}%
            }%
          \fi%
      }%
    \fi
    \ifx\pgfplots@loc@TMPc\pgfplots@group@xticklabels%
      \foreach \pgfplots@column in {1,...,\pgfplots@group@columns} {%
          \expandafter\xdef\expandafter\pgfplots@glob@TMPa\expandafter{%
              \pgfplots@glob@TMPa,/pgfplots/group/plot c\pgfplots@column r\pgfplots@group@rows/.append style={xticklabel pos=lower}}%
          \ifnum1<\pgfplots@group@rows\relax
            \pgfplotsgroupplotsmath@decrease\pgfplots@group@rows
            \foreach \pgfplots@row in {1,...,\pgfmathresult} {%
                \expandafter\xdef\expandafter\pgfplots@glob@TMPa\expandafter{%
                    \pgfplots@glob@TMPa,/pgfplots/group/plot c\pgfplots@column r\pgfplots@row/.append style={/pgfplots/xticklabels={}}}%
            }%
          \fi%
      }%
    \fi%
    % 
    % Defining styles for the ytick's
    % 
    \def\pgfplots@loc@TMPb{left}
    \def\pgfplots@loc@TMPc{right}
    \ifx\pgfplots@loc@TMPa\pgfplots@group@yticklabels%
      % Do nothing as all should have yticklabels
    \fi
    \ifx\pgfplots@loc@TMPb\pgfplots@group@yticklabels%
      \foreach \pgfplots@row in {1,...,\pgfplots@group@rows} {%
          \expandafter\xdef\expandafter\pgfplots@glob@TMPa\expandafter{%
              \pgfplots@glob@TMPa,/pgfplots/group/plot c1r\pgfplots@row/.append style={yticklabel pos=left}}%
          \ifnum1<\pgfplots@group@columns\relax
            \foreach \pgfplots@column in {2,...,\pgfplots@group@columns} {%
                \expandafter\xdef\expandafter\pgfplots@glob@TMPa\expandafter{%
                    \pgfplots@glob@TMPa,/pgfplots/group/plot c\pgfplots@column r\pgfplots@row/.append style={/pgfplots/yticklabels={}}}%
            }%
          \fi%
      }%
    \fi
    \ifx\pgfplots@loc@TMPc\pgfplots@group@yticklabels%
      \foreach \pgfplots@row in {1,...,\pgfplots@group@rows} {%
          \expandafter\xdef\expandafter\pgfplots@glob@TMPa\expandafter{%
              \pgfplots@glob@TMPa,/pgfplots/group/plot c\pgfplots@group@columns r\pgfplots@row/.append style={yticklabel pos=right}}%
          \ifnum1<\pgfplots@group@columns\relax
            \pgfplotsgroupplotsmath@decrease\pgfplots@group@columns
            \foreach \pgfplots@column in {1,...,\pgfmathresult} {%
                \expandafter\xdef\expandafter\pgfplots@glob@TMPa\expandafter{%
                    \pgfplots@glob@TMPa,/pgfplots/group/plot c\pgfplots@column r\pgfplots@row/.append style={/pgfplots/yticklabels={}}}%
            }%
          \fi%
      }%
    \fi%
    \expandafter\pgfkeys\expandafter{\pgfplots@glob@TMPa}
}

%
% The counters which are used to determine the current plot:
%    \pgfplots@group@current@plot
%    \pgfplots@group@current@column
%    \pgfplots@group@current@row
% gets incremented according to the scheme below:
% ____    ____   ____
%|        | |        | |        |
%|   1   | |   2   | |   3   |
%|___| |___| |___|
% ____    ____   ____
%|        | |        | |        |
%|   4   | |   5   | |   6   |
%|___| |___| |___|
% ____    ____   ____
%|        | |        | |        |
%|   7   | |   8   | |   9   |
%|___| |___| |___|
% where the number matches the plot number.
%
\def\pgfplots@group@increment@numbers{%
    \advance\pgfplots@group@current@plot by1
    \ifnum\pgfplots@group@current@column=\pgfplots@group@columns\relax
      \ifnum1=\pgfplots@group@current@plot\relax
      \else
        \advance\pgfplots@group@current@row by1
      \fi
      \pgfplots@group@current@column=1
    \else
      \advance\pgfplots@group@current@column by1
    \fi
}

\def\pgfplots@group@get@plot@name#1,#2\in#3{%
    \edef#3{c#1r#2}%
}

%
% Setting up all possible styles in current plot
% I.e. Setting:
% /pgfplots/group/plot c1r1/.style=
% /pgfplots/group/plot c2r1/.style=
% ...
% /pgfplots/group/plot c<columns>r<rows>/.style=
%
\def\pgfplots@group@individual@setup{%
    \gdef\pgfplots@glob@TMPa{}
    \foreach \pgfplots@column in {1,...,\pgfplots@group@columns} {%
        \foreach \pgfplots@row in {1,...,\pgfplots@group@rows} {%
            \expandafter\xdef\expandafter\pgfplots@glob@TMPa\expandafter{%
                \pgfplots@glob@TMPa,/pgfplots/group/plot c\pgfplots@column r\pgfplots@row/.style=}%
        }%
    }%
    \expandafter\pgfkeys\expandafter{\pgfplots@glob@TMPa}%
}

%
% Implementation of the environment {groupplots}
\def\pgfplots@environment@groupplot{%
	\pgfutil@ifnextchar[{\pgfplots@@environment@groupplot}{\pgfplots@@environment@groupplot[\pgfutil@empty]}%
}

\def\endpgfplots@environment@groupplot{\endpgfplots@environment@opt\endgroup}

\def\pgfplots@@environment@groupplot[#1]{%
    \begingroup%
    % Only find the keys column and row
    % the following things appear to be complete... ?
    \pgfkeysinstallkeyfilterhandler{/pgf/key filter handlers/append filtered to}{\pgfplots@group@rmopts}%
    \let\pgfplots@group@rmopts=\pgfutil@empty
    \pgfkeysinstallkeyfilter{/pgf/key filters/equals}{/pgfplots/group style}
    \pgfkeysinstallkeyfilter{/pgf/key filters/is descendant of}{/pgfplots/group}
    \pgfqkeysfiltered{/pgfplots}{#1}
    \edef\pgfplots@loc@TMPa{%
        \noexpand\pgfkeys{/pgfplots/group/.cd,\pgfplots@group@style}
    }\pgfplots@loc@TMPa
	%
	\let\pgfplots@set@options@sanitizemode@old=\pgfplots@set@options@sanitizemode
	\let\pgfplots@set@options@sanitize@old=\pgfplots@set@options@sanitize
	\let\pgfplots@set@options@sanitize=\relax
	\let\pgfplots@set@options@sanitizemode=\relax
    \expandafter\pgfplots@set@options\expandafter{\pgfplots@group@rmopts}
	\let\pgfplots@set@options@sanitize=\pgfplots@set@options@sanitize@old
	\let\pgfplots@set@options@sanitizemode=\pgfplots@set@options@sanitizemode@old
	%
    \pgfplots@group@individual@setup
    \pgfplots@group@determine@ticklabels
    \pgfplots@group@determine@labels
    \pgfplots@group@current@plot=0%
    \pgfplots@group@current@row=1%
    \pgfplots@group@current@column=0%
    \begingroup
      \c@pgf@counta=\pgfplots@group@rows\relax
      \multiply \c@pgf@counta by\pgfplots@group@columns\relax
      \edef\pgfmathresult{\the\c@pgf@counta}%
      \pgfmath@smuggleone\pgfmathresult
    \endgroup
    \edef\pgfplots@group@totalplots{\pgfmathresult}%
    %DEBUG: \message{TEST: \pgfplots@group@totalplots, rows=\pgfplots@group@rows, columns=\pgfplots@group@columns}%
}

% Assigns \pgfmathresult as #1-1 (in integer arithmetics)
%
% #1 an integer (a register or something which expands to an integer).
\def\pgfplotsgroupplotsmath@decrease#1{%
    \begingroup
      \c@pgf@counta=#1\relax
      \advance\c@pgf@counta by-1
      \edef\pgfmathresult{\the\c@pgf@counta}%
      \pgfmath@smuggleone\pgfmathresult
    \endgroup
}

\tikzaddtikzonlycommandshortcutlet\groupplot\pgfplots@environment@groupplot
\tikzaddtikzonlycommandshortcutlet\endgroupplot\endpgfplots@environment@groupplot

\endinput

### 2015-08-19 research_class/pgfplots/tex/generic/pgfplots/pgfplotscore.code.tex
Add pgfplots_1.12.1.tds.zip

%--------------------------------------------
%
% Package pgfplots
%
% Provides a user-friendly interface to create function plots (normal
% plots, semi-logplots and double-logplots).
% 
% It is based on Till Tantau's PGF package.
%
% Copyright 2007-2014 by Christian Feuersänger.
%
% This program is free software: you can redistribute it and/or modify
% it under the terms of the GNU General Public License as published by
% the Free Software Foundation, either version 3 of the License, or
% (at your option) any later version.
% 
% This program is distributed in the hope that it will be useful,
% but WITHOUT ANY WARRANTY; without even the implied warranty of
% MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
% GNU General Public License for more details.
% 
% You should have received a copy of the GNU General Public License
% along with this program.  If not, see <http://www.gnu.org/licenses/>.
%
%--------------------------------------------

\def\pgfplotsset#{\pgfqkeys{/pgfplots}}

% This is *identical* to \pgfutil@IfUndefined . I copied it here
% because pgf up to and including version 2.10 does not contain it.
\def\pgfplotsutil@IfUndefined#1{%
  \begingroup\expandafter\expandafter\expandafter\endgroup
  \expandafter\ifx\csname#1\endcsname\relax
    \expandafter\pgfutil@firstoftwo
  \else
    \expandafter\pgfutil@secondoftwo
  \fi
}
\pgfplotsutil@IfUndefined{pgfutil@IfUndefined}{\let\pgfutil@IfUndefined=\pgfplotsutil@IfUndefined}{}

\def\pgfplotsloglevel{4}%

\def\pgfplots@log#1#2{%
	\ifnum#1>\pgfplotsloglevel\relax
	\else
		\ifnum\inputlineno=1
			\pgfutil@IfUndefined{@currenvline}{%
				\def\pgfplots@log@tmp{}%
			}{%
				\let\pgfplots@log@tmp=\@currenvline
			}%
		\else
			\def\pgfplots@log@tmp{ on input line \the\inputlineno}%
		\fi
		\immediate\write-1{Package pgfplots info\pgfplots@log@tmp: #2}%
	\fi
}%

\newif\ifpgfplots@LUA@supported
\pgfutil@IfUndefined{directlua}{%
	\pgfplots@LUA@supportedfalse
}{%
	\pgfplots@LUA@supportedtrue
}%

% this is an emergency key which allows to switch off LUA support in
% pgfplots early at boot time. 
%
% use \def\pgfplotsenablelua{0} to switch it off. '1' switches it on
% (or uncommenting the line).
\pgfutil@IfUndefined{pgfplotsenablelua}{}{%
	\ifpgfplots@LUA@supported
		\def\pgfplots@glob@TMPc{1}%
		\ifx\pgfplots@glob@TMPc\pgfplotsenablelua
			\message{Pgfplots: found emergency control flag \string\pgfplotsenablelua=1. Keeping LUA active.^^J}%
		\else
			\message{Pgfplots: found emergency control flag \string\pgfplotsenablelua!=1. Disabling LUA support.^^J}%
			\pgfplots@LUA@supportedfalse
		\fi
	\fi
}%

\def\pgfplots@alloc@catcodetable{%
	% it is task of the TeX format to load this.
	% context comes with it out-of-the box, and pgfplots.sty loads luatexbase.sty
	\pgfutil@IfUndefined{newcatcodetable}{%
		% Oh. The TeX format does not have support for
		% \newcatcodetable and the associated pgfplots module did not
		% load an addon-package!?
		%
		% We _need_ the catcodes which are active when pgfplots starts.
		% And there is no central mechanism to allocate catcodes. I will
		% take one for me now and hope that it will not be used (soon).
		\def\pgfplotsglobalretval{41}%
	}{%
		% the macro \pgfplots@luabackend@table is never used by TeX
		% code.
		% I only need the value inside of the .lua files: whenever
		% they print something, they need to use it.
		\newcatcodetable\pgfplots@luabackend@table
		\let\pgfplotsglobalretval=\pgfplots@luabackend@table
	}%
}%
\ifpgfplots@LUA@supported
	\let\pgfplotsutil@directlua=\directlua
	\let\pgfplotsutil@directlua@std=\pgfplotsutil@directlua
	\pgfplotsutil@directlua{
		tex.enableprimitives("pgfplotsutil@",{ "savecatcodetable", "luaescapestring"})}%
	\def\pgfplotsutil@directlua@trace#1{%
		\directlua{%
			texio.write_nl("CALLING\string\n\pgfplotsutil@luaescapestring{#1}");
			#1%
		}%
	}%
	\pgfplots@log3{Initializing with LUA version \pgfplotsutil@directlua{tex.print(_VERSION)}}%
	\pgfplotsutil@directlua{require("pgfplots"); pgfplots.pgfplotsversion = "\pgfplotsversion";}%
	%
	\pgfplots@alloc@catcodetable
	\c@pgf@counta=\pgfplotsglobalretval
	\pgfplotsutil@savecatcodetable\c@pgf@counta
	\pgfplotsutil@directlua{pgfplots.LOAD_TIME_CATCODETABLE = \the\c@pgf@counta}%
\fi

\pgfkeys{
	/pgfplots/lua support/.is choice,
	/pgfplots/lua support/true/.code={\pgfplots@LUA@supportedtrue},%
	/pgfplots/lua support/false/.code={\pgfplots@LUA@supportedfalse},%
	/pgfplots/lua support/auto/.code={%
		\pgfutil@IfUndefined{directlua}{%
			\pgfplots@LUA@supportedfalse
		}{%
			\pgfplots@LUA@supportedtrue
		}%
	},%
	%/pgfplots/lua support/auto,% this is the default anyway
}

\def\pgfplots@error#1{\PackageError{pgfplots}{#1}{}}%
\def\pgfplotstable@error@pkg#1{\PackageError{pgfplotstable}{#1}{}}%
% DO NOT USE THIS METHOD DIRECTLY: use \pgfplotswarning{<warning name>}<args>\pgfeov
\pgfplotsutil@IfUndefined{PackageWarning}{%
	\def\pgfplots@warning#1{\pgfplots@message{! Package pgfplots Warning:
#1}}%
}{%
	\def\pgfplots@warning#1{\PackageWarning{pgfplots}{#1}{}}%
}%

\def\pgfplots@message#1{%
	\immediate\write16{#1}%
}%

\pgfutil@IfUndefined{pgfkeys}{%
	\pgfplots@error{It seems your version of PGF/Tikz is older than 2.00. Unfortunately, pgfplots requires at least version 2.00 ... you may need to update your TeX-Distribution or install PGF manually, sorry}%
}{\relax}

% Throws exception `#1' with arguments `#2'.
%
% #1 : the exception name
% #2: all what comes after the exception name is considered to be argument
%    (or arguments) for the exception '#1'.
%    the \pgfeov is IMPORTANT as it delimits the argument.
%
% Note that all standard pgfplots exceptions provide a feature to
% exchange the error message text: define \pgfplotsexceptionmsg
% set a replacement.
\def\pgfplotsthrow#1#2\pgfeov{%
	\def\pgfkeyscurrentkey{/pgfplots/exception/#1}%
	\pgfkeysvalueof{/pgfplots/exception/#1/.@cmd}#2\pgfeov
}%
\let\pgfplotsthrow@orig=\pgfplotsthrow

% Generates warning `#1' with argument(s) `#2'.
%
% It works in exactly the same way as \pgfplotsthrow.
\def\pgfplotswarning#1#2\pgfeov{%
	\def\pgfkeyscurrentkey{/pgfplots/warning/#1}%
	\pgfkeysvalueof{/pgfplots/warning/#1/.@cmd}#2\pgfeov
}%

\def\pgfplotsthrow@warning#1{%
	\pgfplots@warning{#1}%
	\immediate\write-1{ [warning \pgfkeyscurrentkey]}%
}%

\def\pgfplots@assert@not@zero#1#2{%
	\ifdim#1=0sp %
		\pgfplotsthrow{invalid argument}{\pgfplots@loc@TMPa}{#2}\pgfeov
	\fi
}%

% A primitive try #1 catch #2 end  block.
%
% It tries code #1. If any exception occurs within, it suppresses the
% exception and tries to continue. It then invokes #2 as soon as it
% can.
%
% You can use \pgfplotsrethrow in #2.
%
% ATTENTION: this is a simple attempt to simulate error control. Don't
% rely on it too heavily! In the moment, I am not even sure if it can
% be nested (perhaps you need to introduce extra scopes since
% \pgfplotstry doesn't).
\long\def\pgfplotstry#1\catch#2\endpgfplotstry{%
	\global\let\pgfplotstry@exception\pgfutil@empty
	\def\pgfplotsthrow##1##2\pgfeov{\gdef\pgfplotstry@exception{{##1}{##2}}}%
	#1\relax%
	\let\pgfplotsthrow=\pgfplotsthrow@orig
	\ifx\pgfplotstry@exception\pgfutil@empty
	\else
		#2%
	\fi
	\global\let\pgfplotstry@exception\pgfutil@empty
}%
\def\pgfplotsrethrow{\expandafter\pgfplotsthrow\pgfplotstry@exception\pgfeov}%

\pgfkeys{
	% #1: the argument which should have been assigned.
	% #2: an error message. 
	/pgfplots/exception/invalid argument/.code 2 args={%
		\ifx\pgfplotsexceptionmsg\relax
			\pgfplots@error{#2}%
		\else
			\pgfplots@error{\pgfplotsexceptionmsg}%
		\fi
		\let#1=\pgfutil@empty
	},%
	% #1 : the layer name
	% #2 : additional context information (message)
	/pgfplots/exception/inactive layer/.code 2 args={%
		\ifx\pgfplotsexceptionmsg\relax
			\pgfplots@error{Sorry, the layer '#1' has not been activated but it is referenced in #2. Perhaps you misspelled it? Each referenced layer must be activated (expect for layer 'discard').}%
		\else
			\pgfplots@error{\pgfplotsexceptionmsg}%
		\fi
	},%
	% #1: the argument which should have been assigned.
	% #2: an error message. 
	/pgfplots/exception/no such element/.code 2 args={%
		\ifx\pgfplotsexceptionmsg\relax
			\pgfplots@error{#2}%
		\else
			\pgfplots@error{\pgfplotsexceptionmsg}%
		\fi
		\let#1=\pgfutil@empty
	},%
	/pgfplots/exception/invalid color component/.code 2 args={%
		\ifx\pgfplotsexceptionmsg\relax
			\pgfplots@error{#2}%
		\else
			\pgfplots@error{\pgfplotsexceptionmsg}%
		\fi
		\let#1=\pgfutil@empty
	},%
	% #1: an error message
	/pgfplots/exception/unsupported operation/.code={%
		\ifx\pgfplotsexceptionmsg\relax
			\pgfplots@error{#1}%
		\else
			\pgfplots@error{\pgfplotsexceptionmsg}%
		\fi
	},%
	% #1: the argument which should have been assigned.
	% #2: the file name
	% #3: the error message
	/pgfplots/exception/no such table file/.code args={#1#2#3}{%
		\ifx\pgfplotsexceptionmsg\relax
			\pgfplots@error{#3}%
		\else
			\pgfplots@error{\pgfplotsexceptionmsg}%
		\fi
		\let#1=\relax
	},
	%
	%
	%
	%
	% Generated if you say 'ymin=0' in a log axis.
	% #1: the user argument ('ymin')
	% #2: the offending value.
	/pgfplots/warning/illegal user argument in log axis/.code args={#1#2}{%
		\pgfplotsthrow@warning{Ignoring illegal input argument #1=#2: cannot apply log.}%
	},
	/pgfplots/warning/plot without coordinates/.code={%
		\pgfplotsthrow@warning{the current plot has no coordinates (or all have been filtered away)}%
	},
	/pgfplots/warning/log basis required/.code 2 args={%
		\pgfplotsthrow@warning{Some input argument (#2) explicitly relies on the log basis of #1 (which is e). Please verify that this is intended and set 'log basis #1=10' otherwise. This warning will be shown only once per document.}%
	},
	/pgfplots/warning/no such legend to name/.code={%
		\pgfplotsthrow@warning{There is no `legend to name=#1', sorry. I am skipping it}%
	},
	/pgfplots/warning/no such colorbar to name/.code={%
		\pgfplotsthrow@warning{There is no `colorbar to name=#1', sorry. I am skipping it}%
	},
	/pgfplots/warning/no such addplot label/.code={%
		\pgfplotsthrow@warning{There is no \string\addplot\space label `#1', sorry. I am skipping it}%
	},
	/pgfplots/warning/enlargelimits respects figure size failed/.code={%
		\pgfplotsthrow@warning{enlargelimits respects figure size=true: could not respect the prescribed width/height (\pgfplotsretval). The size and/or aspect ratio of your image will be different than expected. You can try to modify BOTH width and height.}%
	},
	% #1: the axis
	/pgfplots/warning/empty range cleared axis/.code={%
		\pgfplotsthrow@warning{You have an axis with empty range (in direction #1). Replacing it with a default range and clearing all plots.}%
	},
	/pgfplots/warning/approx empty range enlarged/.code args={#1#2#3}{%
		\pgfplotsthrow@warning{Axis range for axis #1 is approximately empty; enlarging it (it is [#2:#3])}%
	},
	/pgfplots/warning/lossy colormap rgb conversion/.code 2 args={%
		\pgfplotsthrow@warning{I am about to convert colormap '#1' from #2\space to rgb (which might be lossy). Perhaps the choice 'colorbar sampled={surf,shader=interp}' yields better results since it supports #2.}%
	},
	/pgfplots/warning/colorbar uninitialised/.code={%
		\pgfplotsthrow@warning{Sorry, the colorbar is uninitialised - your axis has no color data. Consider using 'point meta=f(x)' or something like that. I am using a color range of [0,1] for the colorbar.}%
	},
	/pgfplots/warning/colorbar CMYK unsupported/.code={%
		\pgfplotsthrow@warning{colorbar remark: The colormap '#1' has a CMYK colorspace which is not (yet) supported by PGF. I switched colorbar styles automatically to the visually equivalent 'colorbar sampled={surf,shader=interp}' to generate a proper CMYK colorbar. This has less support in some previewers. Use \string\pgfplotscolorbarCMYKworkaroundfalse\space to disable this improvement. [This message won't be replicated for '#1']}%
	},
	/pgfplots/warning/ticklabel anchor undetermined/.code args={#1#2#3#4}{%
		\pgfplotsthrow@warning{the ticklabel anchor cannot be determined, the normal vector -(#2) and the unit #1 vector (#3) are almost parallel (abs(cos(angle)) = #4)!}%
	},
	/pgfplots/warning/point meta unbounded/.code 2 args={%
		\pgfplotsthrow@warning{The per point meta data `#1' (#2) (and probably others as well) is unbounded - using the minimum value instead.}%
	},
	/pgfplots/warning/gnuplot -V impossible/.code={%
		\pgfplotsthrow@warning{Sorry, I could not determine 'gnuplot -V' to check whether gnuplot and logscale writes results in log() or not. Please set `/pgfplots/gnuplot writes logscale=true|false' manually.}%
	},
	/pgfplots/warning/gnuplot -V format unknown/.code={%
		\pgfplotsthrow@warning{Sorry, I can't reliably check which version of gnuplot is available. I guess it is gnuplot < 4.4. Please set `/pgfplots/gnuplot writes logscale=true|false' manually if anything fails.}%
	},
	/pgfplots/warning/plot3 graphics compatibility mode/.code={%
		\pgfplotsthrow@warning{plot3 graphics is running in backwards compatibility mode. %
					Use \string\pgfplotsset{compat=1.6} or higher to benefit from upgraded scaling capabilites.}%
	},
	/pgfplots/warning/plot3 graphics too few inner anchors/.code={%
		\pgfplotsthrow@warning{plot graphics in 3D axis has just two inner anchors (those with '(x,y,z)=>(imgx,imgy)'). You should provide a third one such that I can check for correct scaling.}%
	},
	/pgfplots/warning/scatter classes no such class/.code={%
		\pgfplotsthrow@warning{scatter/classes: can't find class for '#1'!? Please make sure you have specified 'scatter src=explicit symbolic'. Ignoring class '#1' (this message will not come again).}%
	},
	/pgfplots/warning/lua colormap unavailable/.code={%
		\pgfplotsthrow@warning{there is no LUA colormap representation for #1. Using TeX arithmetics.}%
	},
	/pgfplots/warning/interior colormap inefficient/.code={%
		\pgfplotsthrow@warning{the combination 'shader=interp,interior colormap name' is inefficient (it uses 'shader=faceted interp,faceted color=none') and may slow down the *display* of your document}%
	},
	/pgfplots/warning/matrix delinearize uses sqrt/.code args={#1#2#3#4}{%
		\pgfplotsthrow@warning{%
									mesh processing did not find '[mesh/rows=<N>]' or '[mesh/cols=<N>]'. 
									Assuming [mesh/rows=#1,mesh/cols=#2] 
									as sqrt(num points) = sqrt(#3) [ = #4].}%
	},
	/pgfplots/warning/matrix check/.code args={#1#2#3#4#5}{%
		#1{the arguments of [mesh/rows=#2,mesh/cols=#3] assume  #5\space points, but I got actually N = #4\space points! The data matrix appears to be incomplete or overcomplete!? [Use mesh/check=false to disable this message]}%
	},
	/pgfplots/warning/tick computation failed/.code 2 args={%
		\pgfplotsthrow@warning{Tick computation for direction #1 failed; there are always too few tick labels (try min ticks has already been advanced to #2)}%
	},
	/pgfplots/warning/scale uniformly unsupported/.code={%
		\pgfplotsthrow@warning{Sorry, 'scale uniformly' failed because its actual implementation works only if y_x = 0 and x_y = 0 (for 2d axes) or if z_x = 0 (for 3d axes). The result will not fill the prescribed dimensions. Falling back to 'scale uniformly strategy=units only. (use scale uniformly warning=false to disable this warning)}%
	},
	/pgfplots/warning/axis equal incompatible change/.code={%
		\pgfplotsthrow@warning{the content of your 3d axis has changed compared to previous versions of pgfplots. please review it. ^^J   %
		[continued]   explanation: you have a 3d axis with 'axis equal' and/or 'unit vector ratio' which has (probably) been optimized for an older version of pgfplots. any version older than 1.6 produced wrong output.^^J   %
	[continued]	  to remove this warning, write \string\pgfplotsset{compat=1.6} (or newer) in your preamble (may change all figures in your document) or by adding that to the affected axis.}%
	},
	/pgfplots/warning/unit vector ratio axis undetermined/.code={%
		\pgfplotsthrow@warning{The algorithm to implement 'unit vector ratio' failed! It could not determine the axis which shall be scaled and decided to use 'unit vector ratio axis=#1'.}%
	},
	/pgfplots/warning/linear system singular/.code={%
		\pgfplotsthrow@warning{Linear system is singular up to machine precision! Continuing with small threshold.}%
	},
	/pgfplots/warning/center lines upgrade required/.code={%
		\pgfplotsthrow@warning{Please use '\string\pgfplotsset{compat=1.8}' to enable correct label placement for 3d axes (at least compat/labels=1.8).}%
	},
	/pgfplots/warning/empty column name/.code 2 args={%
		\pgfplotsthrow@warning{Table '#1' has an empty column name; replacing it by '#2'}%
	},
}
\let\pgfplotsexceptionmsg=\relax

\def\pgfplots@EOI{\pgfplots@EOI}%

\newif\ifpgfplots@loc@tmp
\newtoks\t@pgfplots@toka
\newtoks\t@pgfplots@tokb
\newtoks\t@pgfplots@tokc
\newdimen\pgfplots@tmpa
\newcount\c@pgfplots@coordindex
\newcount\c@pgfplots@scanlineindex
\pgfutil@IfUndefined{r@pgf@reada}{%
	\csname newread\endcsname\r@pgfplots@reada
}{%
	\let\r@pgfplots@reada=\r@pgf@reada
}

% use these macros for GLOBAL temporary assignments.
% you can NEVER rely on their values unless you know exactly what you are doing.
\gdef\pgfplots@glob@TMPa{}%
\gdef\pgfplots@glob@TMPb{}%
\gdef\pgfplots@glob@TMPc{}%

% use these macros for LOCAL temporary assigments.
% you can NEVER rely on their values unless you know exactly what you are doing.
\def\pgfplots@loc@TMPa{}%
\def\pgfplots@loc@TMPb{}%
\def\pgfplots@loc@TMPc{}%

% Invokes code #2 if file '#1' exists and #3 if not.
\long\def\pgfplots@iffileexists#1#2#3{%
	\openin\r@pgfplots@reada=#1
	\ifeof\r@pgfplots@reada
		#3\relax
	\else
		\closein\r@pgfplots@reada
		#2\relax
	\fi
}
\let\pgfplotsiffileexists=\pgfplots@iffileexists

\pgfutil@ifundefined{pgfplots@texdist@protect}{%
	\def\pgfplots@texdist@protect{}%
}{}

\input pgfplotssysgeneric.code.tex
\endinput

### 2015-08-19 research_class/pgfplots/tex/generic/pgfplots/sys/pgfplotssysgeneric.code.tex
Add pgfplots_1.12.1.tds.zip

%--------------------------------------------
% Package pgfplots
%
% Provides a user-friendly interface to create function plots (normal
% plots, semi-logplots and double-logplots).
% 
% It is based on Till Tantau's PGF package.
%
% Copyright 2007/2008 by Christian Feuersänger.
%
% This program is free software: you can redistribute it and/or modify
% it under the terms of the GNU General Public License as published by
% the Free Software Foundation, either version 3 of the License, or
% (at your option) any later version.
% 
% This program is distributed in the hope that it will be useful,
% but WITHOUT ANY WARRANTY; without even the implied warranty of
% MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
% GNU General Public License for more details.
% 
% You should have received a copy of the GNU General Public License
% along with this program.  If not, see <http://www.gnu.org/licenses/>.
%
%--------------------------------------------

% NOTE: tex4ht support is lousy. For example, \tikz \node {$10^{-6}$};
% will already break -- because SVG does not support the html tag
% &minus; produced by tex4ht (and TikZ apparently does not know how to
% escape it somehow).
\def\pgfplots@sys@beforenode{%
	\pgfkeysifdefined{/tikz/tex4ht node/escape/.@cmd}{%
		\ifpgfplots@has@cell@picture
			\tikzset{tex4ht node/escape=true}%
			\pgfplots@warning{ATTENTION: you may want to use '\string\pgfplotsset{cell picture=false}' in your preamble. Otherwise, tex4ht will produce TWO SVG files per plot.^^J}%
		\fi
	}{}%
}%

% see compat/show suggested version
\newif\ifpgfplots@show@suggested@version
\pgfplots@show@suggested@versiontrue

\def\pgfplots@show@suggested@compat@value{%
	\ifpgfplots@show@suggested@version
		\begingroup
		\pgfkeysgetvalue{/pgfplots/compat/current}\current
		\pgfkeysgetvalue{/pgfplots/compat/mostrecent}\mostrecent
		%
		\def\b@isnewest{0}%
		\def\b@shouldupgrade{0}%
		\ifx\current\mostrecent
			\def\b@isnewest{1}%
		\else
			\def\pgfplots@loc@TMPa{newest}%
			\ifx\current\pgfplots@loc@TMPa
				\def\b@isnewest{1}%
			\else
				\expandafter\pgfplotsutilifstringequal\expandafter{\current}{pre 1.3}{%
					\def\b@shouldupgrade{1}%
				}{}%
				\expandafter\pgfplotsutilifstringequal\expandafter{\current}{default}{%
					\def\b@shouldupgrade{1}%
				}{}%
			\fi
		\fi
		\if1\b@isnewest
			\pgfplots@message{Package pgfplots notification 'compat/show suggested version=true': document has been generated with the most recent feature set (\string\pgfplotsset{compat=\mostrecent}).^^J}%
		\else
			\if1\b@shouldupgrade
				\pgfplots@warning{running in backwards compatibility mode (unsuitable tick labels; missing features). Consider writing \string\pgfplotsset{compat=\mostrecent} into your preamble.^^J}%
			\else
				\pgfplots@message{Package pgfplots notification 'compat/show suggested version=true': you might benefit from \string\pgfplotsset{compat=\mostrecent} (current compat level: \current).^^J}%
			\fi
		\fi
		\endgroup
	\fi
}

\pgfutil@IfUndefined{AtBeginDocument}{%
}{%
	\AtBeginDocument{%
		\pgfplots@show@suggested@compat@value
	}%
}%

### 2015-08-19 research_class/pgfplots/tex/latex/pgfplots/libs/tikzlibrarypgfplots.clickable.code.tex
Add pgfplots_1.12.1.tds.zip

%--------------------------------------------
%
% Package pgfplots, library for dynamic content in PDF files
% (clickable plots)
%
% Copyright 2007/2008 by Christian Feuersänger.
%
% This program is free software: you can redistribute it and/or modify
% it under the terms of the GNU General Public License as published by
% the Free Software Foundation, either version 3 of the License, or
% (at your option) any later version.
% 
% This program is distributed in the hope that it will be useful,
% but WITHOUT ANY WARRANTY; without even the implied warranty of
% MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
% GNU General Public License for more details.
% 
% You should have received a copy of the GNU General Public License
% along with this program.  If not, see <http://www.gnu.org/licenses/>.
%
%--------------------------------------------

\pgfutil@ifundefined{pgfplotsset}{%
	\PackageError{pgfplots}{Sorry, \string\usetikzlibrary{pgfplots.clickable} needs to hook into PGFPlots routines, disabling it -- please provide the command AFTER \string\usepackage{pgfplots}}{}%
	\endinput
}{%
}%

\edef\pgfplotscatcodeDQ{\the\catcode`\"}%
\catcode`\"=12
\newif\ifpgfplots@clickable
\newif\ifpgfplots@annot@printable

\pgfqkeys{/pgfplots}{
	@backgroundpath@hook/.add code={}{\pgfplots@create@clickable@plotarea@hook},
	clickable/.is if=pgfplots@clickable,
	clickable/.default=true,
	clickable=true,
	/pgfplots/execute at begin axis/.add={}{%
		\pgfplots@clickable@beginaxis
	},%
	clickable coords code/.code={%
		\def\pgfplots@clickable@init@@{%
			\pgfplots@clickable@startcoordcollect
			\pgfplotsset{%
				/pgfplots/execute at end plot@@/.add={}{%
					\pgfplots@clickable@endcoordcollect
				},
				/pgfplots/execute for finished point/.add={}{%
					\pgfplots@clickable@coordcollect{#1}%
				},%
			}%
			\let\pgfplots@clickable@init@@=\relax
		}%
		\pgfkeysalso{%
			/pgfplots/execute at begin plot@@/.add={}{%
				\pgfplots@clickable@init@@
			},
		}%
	},
	clickable coords/.style={
		clickable coords code={\def\pgfplotsretval{#1}},
	},
	clickable coords/.default={(xy)},
	annot/js fillColor/.initial={["RGB",1,1,.855]},
	annot/point format/.initial={(\%.1f, \%.1f)},
	annot/point format 3d/.initial={(\%.1f, \%.1f, \%.1f)},
	annot/slope format/.initial={\%.1f*x \%+.1f},
	annot/printable/.is if=pgfplots@annot@printable,
	annot/printable/.default=true,
	% Available are:
% Times-Roman           font.Times
% Times-Bold            font.TimesB
% Times-Italic          font.TimesI
% Times-BoldItalic      font.TimesBI
% Helvetica             font.Helv
% Helvetica-Bold        font.HelvB
% Helvetica-Oblique     font.HelvI
% Helvetica-BoldOblique font.HelvBI
% Courier               font.Cour
% Courier-Bold          font.CourB
% Courier-Oblique       font.CourI
% Courier-BoldOblique   font.CourBI
% Symbol                font.Symbol
% ZapfDingbats          font.ZapfD
	annot/font/.initial={font.Times},
%	annot/font={"CMR10"},
	annot/textSize/.initial=11,
	annot/richtext/.initial=true,
	annot/richtext/.default=true,
	annot/popup size generic/.initial=auto,
	annot/popup size snap/.initial=auto,
	clickable coords size/.style={
		/pgfplots/annot/popup size snap={#1},
	},
	annot/popup size/.style={
		/pgfplots/annot/popup size generic={#1},
		/pgfplots/annot/popup size snap={#1},
	},
	annot/width/.initial=,
	annot/height/.initial=,
	annot/jsname/.initial=,
	annot/dim/.initial=2,
	annot/xmin/.initial=,
	annot/xmax/.initial=,
	annot/ymin/.initial=,
	annot/ymax/.initial=,
	annot/zmin/.initial=,
	annot/zmax/.initial=,
	annot/xscale/.initial=, % "log" or "linear" or "log:<basis>"
	annot/yscale/.initial=,
	annot/zscale/.initial=,
	annot/minminmin/.initial=,
	annot/xaxis/.initial=,
	annot/yaxis/.initial=,
	annot/zaxis/.initial=,
	% format: 
	% 	[<plot1coords>, <plot2coords>,...,<plotncoords>]
	% with <ploticoords> = [<coord>,<coord>,....,<coord>]
	% and <coord> = [ <x>, <y>, "text" ]
	% for example:
	% 	[[[0.0e0,-4.0849609e-1,": (0,0)"],[1.25e-1,-3.7412109e-1,": (3,1)"],[2.5e-1,-3.0849609e-1,": (2,1)"],[3.75e-1,-2.1162109e-1,": (3,3)"],[5.0e-1,-8.3496094e-2,": (1,1)"],[6.25e-1,7.5878906e-2,": (3,5)"],[7.5e-1,2.6650391e-1,": (2,3)"],[8.75e-1,4.8837891e-1,": (3,7)"],[1.0e0,7.4150391e-1,": (0,1)"]]]
	annot/collected plots/.initial=,
	annot/snap dist/.initial=4,
	every semilogy axis/.append style={/pgfplots/annot/point format={(\%.1f, \%.1e)}},
	every semilogx axis/.append style={/pgfplots/annot/point format={(\%.1e, \%.1f)}},
	every loglog axis/.append style={/pgfplots/annot/point format={(\%.1e, \%.1e)}},
}

\pgfkeysifdefined{/pgfplots/annot/xy pattern}{%
}{%
	\pgfkeyssetvalue{/pgfplots/annot/xy pattern}{(xy)}%
}%
\pgfkeysifdefined{/pgfplots/annot/no such coord}{%
}{%
	\pgfkeyssetvalue{/pgfplots/annot/no such coord}{--}%
}%

% ATTENTION:
% The eforms package creates \begin{Form} and \end{Form} in \AtBeginDocument and \AtEndDocument!
\def\pgfplots@glob@TMPa{pgfsys-pdftex.def}
\ifx\pgfplots@glob@TMPa\pgfsysdriver
	\RequirePackage[pdftex]{insdljs}
	\RequirePackage[pdftex]{eforms}
\else
\def\pgfplots@glob@TMPa{pgfsys-dvipdfm.def}
\ifx\pgfplots@glob@TMPa\pgfsysdriver
	\RequirePackage[dvipdfm]{insdljs}
	\RequirePackage[dvipdfm]{eforms}
\else
\def\pgfplots@glob@TMPa{pgfsys-dvips.def}
\ifx\pgfplots@glob@TMPa\pgfsysdriver
	\RequirePackage[dvips]{insdljs}
	\RequirePackage[dvips]{eforms}
\else
\def\pgfplots@glob@TMPa{pgfsys-textures.def}
\ifx\pgfplots@glob@TMPa\pgfsysdriver
	\RequirePackage[textures]{insdljs}
	\RequirePackage[textures]{eforms}
\else
	\RequirePackage{insdljs}
	\RequirePackage{eforms}
\fi\fi\fi\fi

% Work-around for a bug in \begindljs of acrotex:
% if " is active, \begindljs will be wrong.
%--------------------------------------------------
% HAS BEEN FIXED IN THE MEANTIME
% \def\begindljs
% {%
%     \iwvo{\string\begingroup}
%     {\uccode`c=`\%\uppercase{\iwvo{\string\obeyspaces\string\obeylines\string\global\string\let\string^\string^M=\string\jsR c}}}
%     {\escapechar=-1 \lccode`C=`\%\lowercase{\iwvo{\string\\catcode`\string\\"=12C}}}
% }
%--------------------------------------------------

% Bugfix to fix incompatibility when '<' and '>' are active (as for
% \usepackage[spanish]{babel}:
\def\write@objs@BUGGY
{%
    \iwvo{\begingroup}
    {\lccode`C=`\%\lowercase{\iwvo{\string\ccpdftex C}}}
    {\lccode`C=`\%\lowercase{\iwvo{\string\input{dljscc.def}C\the\dljsobjtoks}}}
    \iwvo{\endgroup}
    \iwvo{\string\endinput}%
}
\def\write@objs@CORRECT
{%
	\iwvo{\begingroup}
	\iwvo{\string\catcode`\string\<=12 }
	\iwvo{\string\catcode`\string\>=12 }
	{\lccode`C=`\%\lowercase{\iwvo{\string\ccpdftex C}}}
	{\lccode`C=`\%\lowercase{\iwvo{\string\input{dljscc.def}C\the\dljsobjtoks}}}
	\iwvo{\endgroup}
	\iwvo{\string\endinput}%
}
\pgfutil@IfUndefined{write@objs}{}{%
	\ifx\write@objs\write@objs@BUGGY
		\let\write@objs=\write@objs@CORRECT
	\fi
}%

\def\pgfplots@clickable@no{0}

% This catcode-hackery is too much for me, I don't get '|' to work in
% conjunction with ltxdoc document style.
% So, I simply use \pgfplotsVERTBAR instead... *sigh*.

% this here is a bugfix to resolve an incompatibility between insdljs
% an \usepackage[spanish]{babel}:
\def\@roman#1{\romannumeral #1}%

% FIXME : write this stuff DIRECTLY into a pdf using \pdfobj! I don't
% need ANY TeX code inside of the methods. Maybe I can even use some
% sort of "include external .js file" command instead of /S /Javascript /JS
%
% FIXME : collect this stuff using \beginpgfplotsverbatim ...
% \endpgfplotsverbatim to avoid the \catcode problems! Who knows
% whether there are more of them...
%
% FIXME : using '\jobname' here produces a bug in insdljs :-(
% The problem: insdljs creates a macro named 'dlsj\jobname' or
% something like that, but if fails to use '\csname' properly. So:
% only normal letters are allowed inside of the argument here.
\begin{insDLJS}[processAnnotatedPlot]{pgfplotsJS}{pgfplots Clickable Plot Code}
/*********************************************************************************
 * function sprintf() - written by Kevin van Zonneveld as part of the php to javascript 
 * conversion project.
 * 
 * More info at: http://kevin.vanzonneveld.net/techblog/article/phpjs_licensing/
 * 
 * This is version: 1.33
 * php.js is copyright 2008 Kevin van Zonneveld.
 * 
 * Portions copyright Michael White (http://crestidg.com), _argos, Jonas
 * Raoni Soares Silva (http://www.jsfromhell.com), Legaev Andrey, Ates Goral
 * (http://magnetiq.com), Philip Peterson, Martijn Wieringa, Webtoolkit.info
 * (http://www.webtoolkit.info/), Carlos R. L. Rodrigues
 * (http://www.jsfromhell.com), Ash Searle (http://hexmen.com/blog/),
 * Erkekjetter, GeekFG (http://geekfg.blogspot.com), Johnny Mast
 * (http://www.phpvrouwen.nl), marrtins, Alfonso Jimenez
 * (http://www.alfonsojimenez.com), Aman Gupta, Arpad Ray
 * (mailto:arpad@php.net), Karol Kowalski, Mirek Slugen, Thunder.m, Tyler
 * Akins (http://rumkin.com), d3x, mdsjack (http://www.mdsjack.bo.it), Alex,
 * Alexander Ermolaev (http://snippets.dzone.com/user/AlexanderErmolaev),
 * Allan Jensen (http://www.winternet.no), Andrea Giammarchi
 * (http://webreflection.blogspot.com), Arno, Bayron Guevara, Ben Bryan,
 * Benjamin Lupton, Brad Touesnard, Brett Zamir, Cagri Ekin, Cord, David,
 * David James, DxGx, FGFEmperor, Felix Geisendoerfer
 * (http://www.debuggable.com/felix), FremyCompany, Gabriel Paderni, Howard
 * Yeend, J A R, Leslie Hoare, Lincoln Ramsay, Luke Godfrey, MeEtc
 * (http://yass.meetcweb.com), Mick@el, Nathan, Nick Callen, Ozh, Pedro Tainha
 * (http://www.pedrotainha.com), Peter-Paul Koch
 * (http://www.quirksmode.org/js/beat.html), Philippe Baumann, Sakimori,
 * Sanjoy Roy, Simon Willison (http://simonwillison.net), Steve Clay, Steve
 * Hilder, Steven Levithan (http://blog.stevenlevithan.com), T0bsn, Thiago
 * Mata (http://thiagomata.blog.com), Tim Wiel, XoraX (http://www.xorax.info),
 * Yannoo, baris ozdil, booeyOH, djmix, dptr1988, duncan, echo is bad, gabriel
 * paderni, ger, gorthaur, jakes, john (http://www.jd-tech.net), kenneth,
 * loonquawl, penutbutterjelly, stensi
 * 
 * Dual licensed under the MIT (MIT-LICENSE.txt)
 * and GPL (GPL-LICENSE.txt) licenses.
 * 
 * Permission is hereby granted, free of charge, to any person obtaining a
 * copy of this software and associated documentation files (the
 * "Software"), to deal in the Software without restriction, including
 * without limitation the rights to use, copy, modify, merge, publish,
 * distribute, sublicense, and/or sell copies of the Software, and to
 * permit persons to whom the Software is furnished to do so, subject to
 * the following conditions:
 * 
 * The above copyright notice and this permission notice shall be included
 * in all copies or substantial portions of the Software.
 * 
 * THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS
 * OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
 * MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT.
 * IN NO EVENT SHALL KEVIN VAN ZONNEVELD BE LIABLE FOR ANY CLAIM, DAMAGES
 * OR OTHER LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE,
 * ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR
 * OTHER DEALINGS IN THE SOFTWARE.
 */ 
// ATTENTION: this method has been masked such that special characters of TeX and javascript 
// don't produce problems.
function sprintf( ) {
    // Return a formatted string
    // 
    // +    discuss at: http://kevin.vanzonneveld.net/techblog/article/javascript_equivalent_for_phps_sprintf/
    // +       version: 804.1712
    // +   original by: Ash Searle (http://hexmen.com/blog/)
    // + namespaced by: Michael White (http://crestidg.com)
    // *     example 1: sprintf("\pgfplotsPERCENT01.2f", 123.1);
    // *     returns 1: 123.10

    var regex = /\pgfplotsPERCENT\pgfplotsPERCENT\pgfplotsVERTBAR\pgfplotsPERCENT(\d+\$)?([-+\pgfplotsHASH0 ]*)(\*\d+\$\pgfplotsVERTBAR\*\pgfplotsVERTBAR\d+)?(\.(\*\d+\$\pgfplotsVERTBAR\*\pgfplotsVERTBAR\d+))?([scboxXuidfegEG])/g;
    var a = arguments, i = 0, format = a[i++];

    // pad()
    var pad = function(str, len, chr, leftJustify) {
        var padding = (str.length >= len) ? '' : Array(1 + len - str.length >>> 0).join(chr);
        return leftJustify ? str + padding : padding + str;
    };

    // justify()
    var justify = function(value, prefix, leftJustify, minWidth, zeroPad) {
        var diff = minWidth - value.length;
        if (diff > 0) {
            if (leftJustify \pgfplotsVERTBAR\pgfplotsVERTBAR !zeroPad) {
            value = pad(value, minWidth, ' ', leftJustify);
            } else {
            value = value.slice(0, prefix.length) + pad('', diff, '0', true) + value.slice(prefix.length);
            }
        }
        return value;
    };

    // formatBaseX()
    var formatBaseX = function(value, base, prefix, leftJustify, minWidth, precision, zeroPad) {
        // Note: casts negative numbers to positive ones
        var number = value >>> 0;
        prefix = prefix && number && {'2': '0b', '8': '0', '16': '0x'}[base] \pgfplotsVERTBAR\pgfplotsVERTBAR '';
        value = prefix + pad(number.toString(base), precision \pgfplotsVERTBAR\pgfplotsVERTBAR 0, '0', false);
        return justify(value, prefix, leftJustify, minWidth, zeroPad);
    };

    // formatString()
    var formatString = function(value, leftJustify, minWidth, precision, zeroPad) {
        if (precision != null) {
            value = value.slice(0, precision);
        }
        return justify(value, '', leftJustify, minWidth, zeroPad);
    };

    // finalFormat()
    var doFormat = function(substring, valueIndex, flags, minWidth, _, precision, type) {
        if (substring == '\pgfplotsPERCENT\pgfplotsPERCENT') return '\pgfplotsPERCENT';

        // parse flags
        var leftJustify = false, positivePrefix = '', zeroPad = false, prefixBaseX = false;
        for (var j = 0; flags && j < flags.length; j++) switch (flags.charAt(j)) {
            case ' ': positivePrefix = ' '; break;
            case '+': positivePrefix = '+'; break;
            case '-': leftJustify = true; break;
            case '0': zeroPad = true; break;
            case '\pgfplotsHASH': prefixBaseX = true; break;
        }

        // parameters may be null, undefined, empty-string or real valued
        // we want to ignore null, undefined and empty-string values
        if (!minWidth) {
            minWidth = 0;
        } else if (minWidth == '*') {
            minWidth = +a[i++];
        } else if (minWidth.charAt(0) == '*') {
            minWidth = +a[minWidth.slice(1, -1)];
        } else {
            minWidth = +minWidth;
        }

        // Note: undocumented perl feature:
        if (minWidth < 0) {
            minWidth = -minWidth;
            leftJustify = true;
        }

        if (!isFinite(minWidth)) {
            throw new Error('sprintf: (minimum-)width must be finite');
        }

        if (!precision) {
            precision = 'fFeE'.indexOf(type) > -1 ? 6 : (type == 'd') ? 0 : void(0);
        } else if (precision == '*') {
            precision = +a[i++];
        } else if (precision.charAt(0) == '*') {
            precision = +a[precision.slice(1, -1)];
        } else {
            precision = +precision;
        }

        // grab value using valueIndex if required?
        var value = valueIndex ? a[valueIndex.slice(0, -1)] : a[i++];

        switch (type) {
            case 's': return formatString(String(value), leftJustify, minWidth, precision, zeroPad);
            case 'c': return formatString(String.fromCharCode(+value), leftJustify, minWidth, precision, zeroPad);
            case 'b': return formatBaseX(value, 2, prefixBaseX, leftJustify, minWidth, precision, zeroPad);
            case 'o': return formatBaseX(value, 8, prefixBaseX, leftJustify, minWidth, precision, zeroPad);
            case 'x': return formatBaseX(value, 16, prefixBaseX, leftJustify, minWidth, precision, zeroPad);
            case 'X': return formatBaseX(value, 16, prefixBaseX, leftJustify, minWidth, precision, zeroPad).toUpperCase();
            case 'u': return formatBaseX(value, 10, prefixBaseX, leftJustify, minWidth, precision, zeroPad);
            case 'i':
            case 'd': {
                        var number = parseInt(+value);
                        var prefix = number < 0 ? '-' : positivePrefix;
                        value = prefix + pad(String(Math.abs(number)), precision, '0', false);
                        return justify(value, prefix, leftJustify, minWidth, zeroPad);
                    }
            case 'e':
            case 'E':
            case 'f':
            case 'F':
            case 'g':
            case 'G':
                        {
                        var number = +value;
                        var prefix = number < 0 ? '-' : positivePrefix;
                        var method = ['toExponential', 'toFixed', 'toPrecision']['efg'.indexOf(type.toLowerCase())];
                        var textTransform = ['toString', 'toUpperCase']['eEfFgG'.indexOf(type) \pgfplotsPERCENT 2];
                        value = prefix + Math.abs(number)[method](precision);
                        return justify(value, prefix, leftJustify, minWidth, zeroPad)[textTransform]();
                    }
            default: return substring;
        }
    };

    return format.replace(regex, doFormat);
}
/*********************************************************************************/

% see https://developer.mozilla.org/en/Core_JavaScript_1.5_Guide/Inheritance

var lastPoint = null;
var posOnMouseDownX = -1;
var posOnMouseDownY = -1;

// preallocation.
var tmpArray1 = new Array(3);
var tmpArray2 = new Array(3);

var clickablePatternForXY="\pgfkeysvalueof{/pgfplots/annot/xy pattern}";
var clickableStringNoSuchCoord = "\pgfkeysvalueof{/pgfplots/annot/no such coord}";

var nan = Number.NaN;
var NAN = Number.NaN;
var inf = Number.POSITIVE_INFINITY;
var INF = Number.POSITIVE_INFINITY;

var pgfplotsAxisRegistry = new Object();
pgfplotsAxisRegistry["rectangle"]	= function( axisAnnotObj ) { return new PGFPlotsAxis( axisAnnotObj ); }
pgfplotsAxisRegistry["ternary"]		= function( axisAnnotObj ) { return new PGFPlotsTernaryAxis( axisAnnotObj ); }

function PGFPlotsClassExtend( child, super )
{
	for (var property in super.prototype) {
		if (typeof child.prototype[property] == "undefined")
			child.prototype[property] = super.prototype[property];
	}
	return child;
}

function PGFPlotsCreateAxisFor( axisAnnotObj, docObject )
{
	var ret = null;
	if( pgfplotsAxisRegistry[axisAnnotObj.axisType] ) {
		ret = pgfplotsAxisRegistry[axisAnnotObj.axisType](axisAnnotObj);
		ret.docObject = docObject;
	}
	return ret;
}

function ClickableCoord(canvasx,canvasy, realx,realy, text)
{
	this.dim=2;
	this.canvasx=canvasx;
	this.canvasy=canvasy;
	this.realx=realx;
	this.realy=realy;
	this.text=text;
}
ClickableCoord.prototype = 
{
	dim : 0,
	canvasx : 0,
	canvasy : 0,
	realx : 0,
	realy : 0,
	realz : 0,
	text : "",
	sourcePlotIdx : -1,
	sourceCoordIdx : -1,

	isSnapToNearestCoord : function() {
		return this.sourcePlotIdx >= 0;
	},

	/**
	 * Takes an already existing TextField, changes its value to point.text and places it at (x,y).
	 * Additional \c displayOpts will be used to format it.
	 */
	draw : function( docObject, canvas, displayOpts )
	{
		var charsX = 1;
		var charsY = 1.5;
		var popupSize = ( this.isSnapToNearestCoord() ? displayOpts.popupSizeSnap : displayOpts.popupSizeGeneric );
		if( popupSize == "auto" ) {
			charsX = Math.max( 5,this.text.length )/2;

		} else {
			var commaOff = popupSize.indexOf( "," );
			if( commaOff >= 0 ) {
				charsX = popupSize.substring( 0, commaOff );
				charsY = popupSize.substring( commaOff+1 );
			} else
				charsX = popupSize;
		}
	//	console.println( this.text );

		if( !displayOpts.richText  )
			canvas.value = ""+this.text;
		else {
			canvas.richText=true;
			var xmlValue = "<?xml version=\"1.0\"?>"+
				"<body> " + % xmlns=\"http://www.w3.org/1999/xhtml\" xmlns:xfa=\"http://www.xfa.org/schema/xfa-data/1.0/\" xfa:APIVersion=\"Acrobat:9.3.2\" xfa:spec=\"2.0.2\">" +
				%"<p>"+
				"<span>"+ % style="font-size:9.0pt;text-align:left;color:#7E0000;font-weight:normal;font-style:normal;font-family:Helvetica,sans-serif;font-stretch:normal"
					this.text + 
				"</span>"+
				%"</p>"+
				"</body>";
			canvas.richValue = util.xmlToSpans(xmlValue);
% debug helper:
% spans= new Array();
% spans[0] = new Object();
% spans[0].text = this.text;
% spans[0].textColor = ["RGB",0.5,0,0];
% canvas.richValue = spans;
% console.println(util.spansToXML(canvas.richValue));
	 	}
		var R = canvas.rect;
		R[0] = this.canvasx;
		R[1] = this.canvasy;
		R[2] = R[0] + charsX*displayOpts.textSize;
		R[3] = R[1] - charsY*displayOpts.textSize;
		canvas.rect = R;
		canvas.textFont = displayOpts.textFont;
		canvas.textSize = displayOpts.textSize;
		canvas.multiline = true;
		canvas.fillColor = displayOpts.fillColor;//["RGB",1,1,.855];
		canvas.doNotSpellCheck = true;
		canvas.readonly = true;
		if( displayOpts.printable )
			canvas.display = display.visible;
		else
			canvas.display = display.noPrint;

		var mark = docObject.getField( canvas.name + "mark");
		if( mark ) {
			R = mark.rect;
			R[0]=this.canvasx-2;
			R[1]=this.canvasy-2;
			R[2]=R[0]+4;
			R[3]=R[1]+4;
			mark.value="";
			mark.rect = R;
			mark.fillColor = ["RGB",0,0,0];
			mark.doNotSpellCheck = true;
			mark.readonly = true;
			if( displayOpts.printable )
				mark.display = display.visible;
			else
				mark.display = display.noPrint;
		}
	}
}

function PGFPlotsAxis(properties) {
	for (var property in properties)
		this[property] = properties[property];
}
PGFPlotsAxis.prototype = {
	/**
	 * @return an instance of ClickableCoord or null.
	 * The returned canvas coordinates are NOT yet initialised.
	 */
	findNearest : function( point, rect, startSearchAt )
	{
		var actx;
		var acty;
		var minSoFar = 1e324;
		var minPtSoFar = null;
		var tmpPt = null;
		var dist =0;
		var startPlot = 0;
		var startCoord = 0;
		var collectedPlots = this.collectedPlots;
		if( point.dim > 2 )
			tmpPt = new ClickableCoord(0,0,0,0,"");

		if( startSearchAt != null && startSearchAt.isSnapToNearestCoord() ) {
			startPlot = startSearchAt.sourcePlotIdx;
			startCoord= startSearchAt.sourceCoordIdx;
		}
		for( var i = startPlot; i<collectedPlots.length; ++i ) {
			for( var j = startCoord; j<collectedPlots[i].length; ++j ) {
				if( point.dim == 2 ) {
					actx = collectedPlots[i][j][0] - point.realx;
					acty = collectedPlots[i][j][1] - point.realy;
				} else {
					// dim > 2 should be compared using CANVAS coordinates, not real coordinates:
					tmpPt.realx = collectedPlots[i][j][0];
					tmpPt.realy = collectedPlots[i][j][1];
					if( collectedPlots[i][j].length-1 >= 3 )
						tmpPt.realz = collectedPlots[i][j][2];
					else
						tmpPt.realz = 0;
					this.computeCanvasFor( tmpPt, rect );
					actx = tmpPt.canvasx - point.canvasx;
					acty = tmpPt.canvasy - point.canvasy;
				}
				dist = actx*actx + acty*acty;
				if( dist < minSoFar ) {
					if( minPtSoFar == null )
						minPtSoFar = new ClickableCoord(-1,-1,0,0,"");
					minPtSoFar.realx = collectedPlots[i][j][0];
					minPtSoFar.realy = collectedPlots[i][j][1];
					if( collectedPlots[i][j].length-1 >= 3 )
						minPtSoFar.realz = collectedPlots[i][j][2];
					else
						minPtSoFar.realz = 0;
					minPtSoFar.text= collectedPlots[i][j][ collectedPlots[i][j].length-1 ];
					minPtSoFar.sourcePlotIdx = i;
					minPtSoFar.sourceCoordIdx = j;
					minSoFar = dist;
				}
			}
		}
		if( minPtSoFar )
			this.computeCanvasFor(minPtSoFar,rect);
		return minPtSoFar;
	},

	/**
	 * Returns either a ClickableCoord describing the point under the mouse cursor or a snap--to--nearest result near the mouse cursor.
	 *
	 * @param this the axis in which we shall search.
	 * @param x,y the input canvas coordinates
	 * @param canvas a pointer to the Drawing object (TextField) whose 'rect' field is the drawing canvas.
	 * @param startSearch either null or an instance of ClickableCoord for which isSnapToNearestCoord() returns true. 
	 *   If it is not null, the next matching point *after* it will be returned (or null if there is no matching snap--to--nearest coord after it).
	 * @return an instance of ClickableCoord or null in case the startSearch!=null and there are no further matches.
	 */
	findClickableCoord : function( x,y, canvas, startSearch )
	{
		// Get and modify bounding box. The mouse movement is only accurate up to one point 
		// (mouseX and mouseY are integers), so the bounding box should be an integer as well.
		var rect = canvas.rect; // rect = [ mincanvasx mincanvasy maxcanvasx maxcanvasy ]; relative to upper left corner
		rect[0] = Math.round(rect[0]);
		rect[1] = Math.round(rect[1]);
		rect[2] = Math.round(rect[2]);
		rect[3] = Math.round(rect[3]);
		canvas.rect= rect;

		var realx=-1;
		var realy=-1;
		if( this.dim == 2 ) {
			// the following code inverts computeCanvasFor():
			var minminminx = rect[0] + this.minminmin[0];
			var minminminy = rect[3] + this.minminmin[1];
			var vecx = x - minminminx;
			var vecy = y - minminminy;
			var A = [ 
				[this.xaxis[0], this.yaxis[0]],
				[this.xaxis[1], this.yaxis[1]] ];
			var b = [ vecx, vecy ];

			var rowpermut = [0, 1];
			if( Math.abs(A[0][0]) < 0.0001 ) {
				rowpermut[0] = 1;
				rowpermut[1] = 0;
			}
			var pivot = -A[rowpermut[1]][0] / A[rowpermut[0]][0];
			var unity = (b[rowpermut[1]] + pivot*b[rowpermut[0]]) / (A[rowpermut[1]][1] + pivot * A[rowpermut[0]][1]);
			var unitx = (b[rowpermut[0]] - A[rowpermut[0]][1] * unity) / A[rowpermut[0]][0];

			realx = this.xmin + unitx * (this.xmax - this.xmin);
			realy = this.ymin + unity * (this.ymax - this.ymin);
			//console.println( "unitx = " + unitx + "; unity " + unity );
		}

		var point = new ClickableCoord( x,y, realx, realy, clickablePatternForXY);
		point.dim = this.dim;

		if( startSearch && !startSearch.isSnapToNearestCoord() ) {
			console.println( "WARNING: startSearch().isSnapToNearestCoord() has been expected!" );
			startSearch = null;
		}

		var nearestClickableCoord = this.findNearest( point, rect, startSearch );
		if( nearestClickableCoord ) {
			if( getDist( point.canvasx,point.canvasy,  nearestClickableCoord.canvasx, nearestClickableCoord.canvasy ) < this.snapDist ) {
				return nearestClickableCoord;
			}
		}
		if( startSearch )
			point = null;
		if( this.dim > 2 ) { // we didn't find a snap--to--nearest point.
			point.text = clickableStringNoSuchCoord;
		}

		return point;
	},

	/**
	 * Takes point's real coordinates and computes its canvas coordinates. The result is written back into \c point.
	 */
	computeCanvasFor : function( point, rect)
	{
		var unitx = (point.realx - this.xmin) / (this.xmax -this.xmin);
		var unity = (point.realy - this.ymin) / (this.ymax -this.ymin);

		point.canvasx = rect[0] + this.minminmin[0] + this.xaxis[0] * unitx + this.yaxis[0] * unity;
		point.canvasy = rect[3] + this.minminmin[1] + this.xaxis[1] * unitx + this.yaxis[1] * unity;
		if( this.dim >= 3 ) {
			var unitz = (point.realz - this.zmin) / (this.zmax -this.zmin);
			point.canvasx += this.zaxis[0] * unitz;
			point.canvasy += this.zaxis[1] * unitz;
		}
	},

	handleDragNDrop : function( formName, displayOpts )
	{
		if( this.dim == 3 )
			return;

		var result = this.docObject.getField( formName + "-result");
		var result2 = this.docObject.getField( formName + "-result2");
		var resultmark = this.docObject.getField( formName + "-resultmark");
		var result2mark = this.docObject.getField( formName + "-result2mark");
		var slope 	= this.docObject.getField( formName + "-slope" );
		if( !result ) {
			console.println( "WARNING: there is no TextField \"" + formName + "-result\" to display results for interactive element \"" + formName + "\"");
			return;
		}

		var a = this.docObject.getField(formName);
		if( ! a ) {
			console.println( "Warning: there is no form named \"" + formName + "\"" );
			return;
		}
		// dragging the mouse results in slope computation:
		// placeClickableCoord shows the endpoint coords and returns the (transformed) coordinates into tmpArray1 and tmpArray2:
		this.placeClickableCoord( 
			this.findClickableCoord( posOnMouseDownX, posOnMouseDownY, a, null ),
			result, displayOpts, tmpArray1 );
		this.placeClickableCoord( 
			this.findClickableCoord( mouseX, mouseY, a, null ),
			result2, displayOpts, tmpArray2 );

		var m =  ( tmpArray2[1] - tmpArray1[1] ) / ( tmpArray2[0] - tmpArray1[0] );
		var n =  tmpArray1[1] - m * tmpArray1[0];

		var slopePoint = new ClickableCoord(
			0.5 * ( mouseX + posOnMouseDownX ),
			0.5 * ( mouseY + posOnMouseDownY ),
			-1,-1,
			sprintf( displayOpts.slopeFormat, m, n ));
		slopePoint.draw(
			this.docObject,
			slope,
			displayOpts );

		// FIXME! these document rights seem to forbid modifications to annotations, although they work for text fields!?
		//var lineobj = this.getAnnot( a.page, formName + '-line' );
		//console.println( 'lineobj = ' + lineobj );
		//lineobj.points = [[mouseX,mouseY],[posOnMouseDownX,posOnMouseDownY]];
		//lineobj.display = display.visible;
	},

	handleClick : function( formName, displayOpts )
	{
		var result = this.docObject.getField( formName + "-result");
		var result2 = this.docObject.getField( formName + "-result2");
		var resultmark = this.docObject.getField( formName + "-resultmark");
		var result2mark = this.docObject.getField( formName + "-result2mark");
		var slope 	= this.docObject.getField( formName + "-slope" );
		if( !result ) {
			console.println( "WARNING: there is no TextField \"" + formName + "-result\" to display results for interactive element \"" + formName + "\"");
			return;
		}
		result2.display = display.hidden;
		slope.display = display.hidden;
		result2mark.display = display.hidden;

		var a = this.docObject.getField(formName);
		if( ! a ) {
			console.println( "Warning: there is no form named \"" + formName + "\"" );
			return;
		}

		var point = null;
		var bSearchPoint = true;
		if( lastPoint ) {
			if( getDist( mouseX,mouseY,  lastPoint.canvasx,lastPoint.canvasy) < this.snapDist ) {
				if( lastPoint.isSnapToNearestCoord() )
					++lastPoint.sourceCoordIdx;
				else
					bSearchPoint = false;

			} else
				lastPoint = null; // no search restriction.
		}
		if( bSearchPoint )
			point = this.findClickableCoord( mouseX, mouseY, a, lastPoint );

		lastPoint = point;

		// clicking twice onto the same point hides it:
		if( point == null ) {
			result.display = display.hidden;
			resultmark.display = display.hidden;
			return;
		}

		this.placeClickableCoord( 
			point,
			result, displayOpts, null );
	},

	/**
	 * Changes all required Field values of \c plotRegionField, inserts the proper
	 * value and displays it at the pdf positions (x,y) .
	 *
	 * @param plotRegionField a reference to a Field object.
	 * @param x the x canvas coordinate where the annotation shall be placed and which is used to determine
	 *  the annotation text.
	 * @param y the corresponding y coord.
	 * @param axis An object containing axis references.
	 * @param displayOpts An object for display flags.
	 * @param[out] retCoords will be filled with the point in axis coordinates (should have length axis.dim).
	 */
	placeClickableCoord : function( point, textField, displayOpts, retCoords )
	{
		var transformedCoordx = point.realx;
		var transformedCoordy = point.realy;

		if( this.xscale.length >= 3 && this.xscale.substr(0,3) == "log" ) {
			if( this.xscale.length > 4 ) // log:<basis>
				point.realx = point.realx * Math.log( this.xscale.substr(4) );
			else {
				// pgfplots handles log plots base e INTERNALLY, but uses base 10 for display.
				// convert to base 10:
				transformedCoordx = point.realx / Math.log(10);
			}
			point.realx = Math.exp(point.realx);
		}
		if( this.yscale.length >= 3 && this.yscale.substr(0,3) == "log" ) {
			if( this.yscale.length > 4 ) // log:<basis>
				point.realy = point.realy * Math.log( this.yscale.substr(4) );
			else {
				// pgfplots handles log plots base e INTERNALLY, but uses base 10 for display.
				// convert to base 10:
				transformedCoordy = point.realy / Math.log(10);
			}
			point.realy = Math.exp(point.realy);
		}
		if( this.dim > 2 ) {
			if( this.zscale.length >= 3 && this.zscale.substr(0,3) == "log" ) {
				if( this.zscale.length > 4 ) // log:<basis>
					point.realz = point.realz * Math.log( this.zscale.substr(4) );
				else {
					// pgfplots handles log plots base e INTERNALLY, but uses base 10 for display.
					// convert to base 10:
					transformedCoordz = point.realz / Math.log(10);
				}
				point.realz = Math.exp(point.realz);
			}
		}

		// replace the text substring "(xy)" with the actual coordinates:
		var coordOff = point.text.indexOf(clickablePatternForXY);
		if( coordOff >= 0 ) {
			point.text = 
				point.text.substring( 0, coordOff ) + 
				sprintf( displayOpts.pointFormat, point.realx,point.realy,point.realz) +
				point.text.substr( coordOff+clickablePatternForXY.length );
		}
		point.draw( this.docObject, textField, displayOpts );

		if( retCoords ) {
			retCoords[0] = transformedCoordx;
			retCoords[1] = transformedCoordy;
			if( this.dim > 2 )
				retCoords[2] = transformedCoordz;
		}

	}

}

function PGFPlotsTernaryAxis(properties) {
	PGFPlotsAxis.call(this,properties);
}
PGFPlotsTernaryAxis.prototype = {
	findClickableCoord : function( x,y, canvas, startSearch )
	{
		// Get and modify bounding box. The mouse movement is only accurate up to one point 
		// (mouseX and mouseY are integers), so the bounding box should be an integer as well.
		var rect = canvas.rect; // rect = [ mincanvasx mincanvasy maxcanvasx maxcanvasy ]; relative to lower left corner
		rect[0] = Math.round(rect[0]);
		rect[1] = Math.round(rect[1]);
		rect[2] = Math.round(rect[2]);
		rect[3] = Math.round(rect[3]);
		canvas.rect= rect;

		var ternaryScale = 1 / ( rect[2] - rect[0] );
		var X = ( x - rect[0] ) * ternaryScale;
		var Y = ( y - rect[3] ) * ternaryScale;

		var realx = 1.15470053837925 * Y; // 2/sqrt(3) 
		var realz = X - 0.5 * realx;
		var realy = 1 - realx - realz;

		if( realx < 0 \pgfplotsVERTBAR\pgfplotsVERTBAR realy < 0 \pgfplotsVERTBAR\pgfplotsVERTBAR realz < 0 )
			return null;

		var point = new ClickableCoord( x,y, realx, realy, clickablePatternForXY);
		point.realz = realz;
		point.dim = this.dim;

		if( startSearch && !startSearch.isSnapToNearestCoord() ) {
			console.println( "WARNING: startSearch().isSnapToNearestCoord() has been expected!" );
			startSearch = null;
		}

		var nearestClickableCoord = this.findNearest( point, rect, startSearch );
		if( nearestClickableCoord ) {
			if( getDist( point.canvasx,point.canvasy,  nearestClickableCoord.canvasx, nearestClickableCoord.canvasy ) < this.snapDist ) {
				return nearestClickableCoord;
			}
		}
		if( startSearch )
			point = null;

		return point;
	},

	/**
	 * Takes point's real coordinates and computes its canvas coordinates. The result is written back into \c point.
	 */
	computeCanvasFor : function( point, rect)
	{
		var unitx = (point.realx - this.xmin) / (this.xmax -this.xmin);
		var unity = (point.realy - this.ymin) / (this.ymax -this.ymin);
		var unitz = (point.realz - this.zmin) / (this.zmax -this.zmin);

		var ternaryScale = ( rect[2] - rect[0] );
		point.canvasx = rect[0] + ternaryScale* 0.5 * (unitx + 2 * unitz);
		point.canvasy = rect[3] + ternaryScale* 0.866025403784* unitx; // sqrt(3)/2
	},
}
PGFPlotsClassExtend( PGFPlotsTernaryAxis, PGFPlotsAxis );

function getVecLen( t1,t2 ) {
	return Math.sqrt( t1*t1 + t2*t2);
}
function getDist( x1,y1, x2,y2 ) {
	var t1 = (x1-x2);
	var t2 = (y1-y2);
	return getVecLen(t1,t2);
}

function axisMouseDown(formName ) 
{
	posOnMouseDownX = mouseX;
	posOnMouseDownY = mouseY;
}

/**
 * @param formName the name of the clickable button. It is expected to be as large as the underlying plot.
 * @param axis an object with the fields
 *   - xmin, xmax
 *   - ymin, ymax
 *   - xscale, yscale
 * @param displayOpts an object with the fields
 *   - pointFormat an sprintf format string to format the final point coordinates.
 *   The default is  "(\pgfplotsPERCENT.2f,\pgfplotsPERCENT.2f)"
 *   - fillColor the fill color for the annotation. Options are
 *    transparent, gray, RGB or CMYK color. Default is
 *       ["RGB",1,1,.855]
 *	 - textFont / textSize
 */
function axisMouseUp(formName, axisAnnotObj, displayOpts)
{
	var axis = PGFPlotsCreateAxisFor(axisAnnotObj,this);
	if( axis == null )
		return;

	if( Math.abs( mouseX - posOnMouseDownX ) > 6 \pgfplotsVERTBAR\pgfplotsVERTBAR
		Math.abs( mouseY - posOnMouseDownY ) > 6 )
	{
		axis.handleDragNDrop( formName, displayOpts );

	} else {
		axis.handleClick( formName, displayOpts );
	}
}

\end{insDLJS}
%--------------------------------------------------
% function hideClickableTextfields() {
% 	for (var i = 0; i < this.numFields; i++) {
% 		var fieldName = this.getNthFieldName(i);
% 		console.println("checking Field " + fieldName + " ... ");
% 		if( fieldName.substr( 0, 13 ) == "clickableplot" && fieldName.indexOf( "-result", 12 ) >= 0 ) {
% 			console.println("hiding Field " + fieldName + " ... ");
% 			this.getField( fieldName ).display = display.hidden;
% 		}
% 	}
% }
% hideClickableTextfields();
%--------------------------------------------------

\pgfkeysdef{/pgfplots/annot/xy pattern}{\pgfplots@clickable@xypat@error}%
\pgfkeysdef{/pgfplots/annot/xy pattern/.initial}{\pgfplots@clickable@xypat@error}%
\def\pgfplots@clickable@xypat@error{\pgfplots@error{Sorry, \string\pgfplotsset{annot/xy pattern/.initial=...} can only be assigned *before* \string\usepgfplotslibrary{clickable}}}%

\pgfkeysdef{/pgfplots/annot/no such coord}{\pgfplots@clickable@nosuchcoord@error}%
\pgfkeysdef{/pgfplots/annot/no such coord/.initial}{\pgfplots@clickable@nosuchcoord@error}%
\def\pgfplots@clickable@nosuchcoord@error{\pgfplots@error{Sorry, \string\pgfplotsset{annot/no such coord/.initial=...} can only be assigned *before* \string\usepgfplotslibrary{clickable}}}%

\def\pgfplots@clickable@beginaxis{%
	\pgfplotsapplistXglobalnewempty\pgfplots@clickable@collectedplots
	\gdef\pgfplots@clickable@collectedplots@isempty{1}%
}%

\def\pgfplots@clickable@startcoordcollect{%
	\pgfplotsapplistXnewempty\pgfplots@clickable@collectedplot
	\def\pgfplots@clickable@collectedplot@isempty{1}%
}%

{\catcode`\|=0 |catcode`|\=12 |gdef|pgfplots@clickable@bslash{\}}%
{
\catcode`\"=13
\catcode`\|=13
\catcode`\#=12
\gdef\pgfplots@clickable@coordcollect@handlechars{%
	\def"{\string\\\string\"}%
	\def\#{#}%
	\expandafter\def\csname \string"\endcsname{\string\\\string\"}%
	\def|{\string"}%
	\def\n{\pgfplots@clickable@bslash\string\n}%
	\def\r{\pgfplots@clickable@bslash\string\r}%
	\def\t{\pgfplots@clickable@bslash\string\t}%
	\def\\{\pgfplots@clickable@bslash\pgfplots@clickable@bslash}%
}%
}
\def\pgfplots@clickable@coordcollect#1{%
	\begingroup
	\ifx\pgfplots@current@point@x\pgfutil@empty
		\def\pgfmathresult{nan}%
	\else
		\pgfplotscoordmath{x}{tostring}\pgfplots@current@point@x
	\fi
	\let\pgfplots@current@point@x=\pgfmathresult
	%
	\ifx\pgfplots@current@point@y\pgfutil@empty
		\def\pgfmathresult{nan}%
	\else
		\pgfplotscoordmath{y}{tostring}\pgfplots@current@point@y
	\fi
	\let\pgfplots@current@point@y=\pgfmathresult
	%
	\ifpgfplots@curplot@threedim
		\ifx\pgfplots@current@point@z\pgfutil@empty
			\def\pgfmathresult{nan}%
		\else
			\pgfplotscoordmath{z}{tostring}\pgfplots@current@point@z
		\fi
		\let\pgfplots@current@point@z=\pgfmathresult
	\fi
	%
	\begingroup
		\pgfplots@clickable@coordcollect@handlechars
		#1%
		\xdef\pgfplots@glob@TMPa{\pgfplotsretval}%
%\message{I have '\pgfplotsretval'}%
	\endgroup
	\xdef\pgfplots@glob@TMPa{\if1\pgfplots@clickable@collectedplot@isempty\else,\fi[\pgfplots@current@point@x,\pgfplots@current@point@y\ifpgfplots@curplot@threedim,\pgfplots@current@point@z\fi,"\pgfplots@glob@TMPa"]}%
	\endgroup
	\expandafter\pgfplotsapplistXpushback\expandafter{\pgfplots@glob@TMPa}\to\pgfplots@clickable@collectedplot
	\def\pgfplots@clickable@collectedplot@isempty{0}%
}%
\def\pgfplots@clickable@endcoordcollect{%
	\pgfplotsapplistXlet\pgfplots@clickable@collectedplot@=\pgfplots@clickable@collectedplot
	\if1\pgfplots@clickable@collectedplots@isempty
		\expandafter\pgfplotsapplistXglobalpushback\expandafter{\expandafter[\pgfplots@clickable@collectedplot@]}\to\pgfplots@clickable@collectedplots
	\else
		\expandafter\pgfplotsapplistXglobalpushback\expandafter{\expandafter,\expandafter[\pgfplots@clickable@collectedplot@]}\to\pgfplots@clickable@collectedplots
	\fi
	\gdef\pgfplots@clickable@collectedplots@isempty{0}%
}%

% Creates an area which is clickable. A click produces a popup which
% contains information about the point under the cursor.
%
% The complete (!) context needs to be provided using key-value-pairs,
% either in '#1' or set before invocation of \pgfplotsclickablecreate.
%
% This command actually creates an AcroForm which employs javascript
% whenever it is clicked. A javascript Object is created which
% represents the context (axis limits and options). This javascript
% object is available at runtime.
%
% @remark This method is public and it is NOT restricted to pgfplots.
% The pgfplots hook simply initialises the required key-value-pairs in
% '#1'.
% @remark This method does not draw anything. It initialises only a
% clickable area and javascript code.
%
% The required key-value-pairs are documented in the pdf-manual or can
% be seen above.
%
% @attention Complete key-value validation is NOT performed here. It
% can happen that invalid options will produce javascript bugs when
% opened with Acrobat Reader. Use the javascript console to find them.
\def\pgfplotsclickablecreate[#1]{%
	\def\pgfplots@loc@TMPa{#1}%
	\ifx\pgfplots@loc@TMPa\pgfutil@empty
	\else
		\pgfqkeys{/pgfplots/annot}{#1}%
	\fi
	\pgfkeysgetvalue{/pgfplots/annot/jsname}\pgfplots@clickable@uniquename
	\ifx\pgfplots@clickable@uniquename\pgfutil@empty
		\edef\pgfplots@clickable@uniquename{clickableplot\pgfplots@clickable@no}%
		\begingroup
			\c@pgf@counta=\pgfplots@clickable@no\relax
			\advance\c@pgf@counta by1
			\xdef\pgfplots@clickable@no{\the\c@pgf@counta}%
		\endgroup
	\fi
	%
	\pgfkeysgetvalue{/pgfplots/annot/collected plots}\pgfplots@clickable@collectedplots@@
	\ifx\pgfplots@clickable@collectedplots@@\pgfutil@empty
		\def\pgfplots@clickable@collectedplots@@{[]}%
	\fi
	%
	\edef\pgfplots@clickable@expanded@args{%
		{\pgfkeysvalueof{/pgfplots/annot/width}}%
		{\pgfkeysvalueof{/pgfplots/annot/height}}%
		{\pgfplots@clickable@uniquename}%
		{{
			axisType:	"\pgfkeysvalueof{/pgfplots/axis type}",
			dim:	\pgfkeysvalueof{/pgfplots/annot/dim},
			xmin:	\pgfkeysvalueof{/pgfplots/annot/xmin}, 
			xmax:	\pgfkeysvalueof{/pgfplots/annot/xmax},
			ymin:	\pgfkeysvalueof{/pgfplots/annot/ymin},
			ymax:	\pgfkeysvalueof{/pgfplots/annot/ymax},
			xscale:	"\pgfkeysvalueof{/pgfplots/annot/xscale}",
			yscale:	"\pgfkeysvalueof{/pgfplots/annot/yscale}",
			\ifnum\pgfkeysvalueof{/pgfplots/annot/dim}=3
			zmin:	\pgfkeysvalueof{/pgfplots/annot/zmin},
			zmax:	\pgfkeysvalueof{/pgfplots/annot/zmax},
			zscale:	"\pgfkeysvalueof{/pgfplots/annot/zscale}",
			zaxis:		\pgfkeysvalueof{/pgfplots/annot/z axis},
			\fi
			collectedPlots: \pgfplots@clickable@collectedplots@@,
			snapDist:	\pgfkeysvalueof{/pgfplots/annot/snap dist},
			minminmin:		\pgfkeysvalueof{/pgfplots/annot/minminmin},
			xaxis:		\pgfkeysvalueof{/pgfplots/annot/x axis},
			yaxis:		\pgfkeysvalueof{/pgfplots/annot/y axis}
		}}%
		{{
			pointFormat: "\ifnum\pgfkeysvalueof{/pgfplots/annot/dim}=3 \pgfkeysvalueof{/pgfplots/annot/point format 3d}\else \pgfkeysvalueof{/pgfplots/annot/point format}\fi",
			slopeFormat: "\pgfkeysvalueof{/pgfplots/annot/slope format}",
			fillColor:\pgfkeysvalueof{/pgfplots/annot/js fillColor},
			textFont:\pgfkeysvalueof{/pgfplots/annot/font},
			textSize:\pgfkeysvalueof{/pgfplots/annot/textSize},
			richText:\pgfkeysvalueof{/pgfplots/annot/richtext},
			popupSizeGeneric:"\pgfkeysvalueof{/pgfplots/annot/popup size generic}",
			popupSizeSnap:"\pgfkeysvalueof{/pgfplots/annot/popup size snap}",
			printable: \ifpgfplots@annot@printable true\else false\fi
		}}%
	}%
	\expandafter\pgfplots@create@clickable@plotarea\pgfplots@clickable@expanded@args
}%

% Defines either \def\pgfplotsretval{1} if the current axis is
% clickable or \def\pgfplotsretval{0} if not.
%
% The initial configuration is `1' for any 2d axis, `1' for 3d axes
% with 'clickable coords' and `0' for 3d axes without `clickable
% coords'.
%
% Overwrite \pgfplotsclickable@check@enable@axistype@<theaxistype>  to
% make axis type specific things.
\def\pgfplotsclickable@check@enable{%
	\def\pgfplotsretval{1}%
	\ifpgfplots@threedim
		\ifx\pgfplots@clickable@collectedplots@@\pgfutil@empty
			\def\pgfplotsretval{0}%
		\fi
	\fi
	\csname pgfplotsclickable@check@enable@axistype@\pgfkeysvalueof{/pgfplots/axis type}\endcsname
}%

% This thing is invoked from within pgfplots. It prepares and invokes
% \pgfplotsclickablecreate.
% 
\def\pgfplots@create@clickable@plotarea@hook{%
	\ifpgfplots@clickable
		\begingroup
		% communicate collected plot coordinates:
		\pgfplotsapplistXgloballet\pgfplots@clickable@collectedplots@@=\pgfplots@clickable@collectedplots
		\pgfplotsapplistXglobalnewempty\pgfplots@clickable@collectedplots
		\t@pgfplots@tokc=\expandafter{\pgfplots@clickable@collectedplots@@}%
		\edef\pgfplots@clickable@collectedplots@@{[\the\t@pgfplots@tokc]}%
		\pgfkeyslet{/pgfplots/annot/collected plots}\pgfplots@clickable@collectedplots@@
		%
		\pgfplotsclickable@check@enable
		\if1\pgfplotsretval
			%
			\pgfplotscoordmath{x}{datascaletrafo inverse}{\pgfplots@xmin}%
			\pgfplotscoordmath{x}{tostring}{\pgfmathresult}%
			\let\pgfplots@clickable@xmin=\pgfmathresult
			\pgfplotscoordmath{x}{datascaletrafo inverse}{\pgfplots@xmax}%
			\pgfplotscoordmath{x}{tostring}{\pgfmathresult}%
			\let\pgfplots@clickable@xmax=\pgfmathresult
			%
			\pgfplotscoordmath{y}{datascaletrafo inverse}{\pgfplots@ymin}%
			\pgfplotscoordmath{y}{tostring}{\pgfmathresult}%
			\let\pgfplots@clickable@ymin=\pgfmathresult
			\pgfplotscoordmath{y}{datascaletrafo inverse}{\pgfplots@ymax}%
			\pgfplotscoordmath{y}{tostring}{\pgfmathresult}%
			\let\pgfplots@clickable@ymax=\pgfmathresult
			%
			\ifpgfplots@threedim
				\pgfplotscoordmath{z}{datascaletrafo inverse}{\pgfplots@zmin}%
				\pgfplotscoordmath{z}{tostring}{\pgfmathresult}%
				\let\pgfplots@clickable@zmin=\pgfmathresult
				\pgfplotscoordmath{z}{datascaletrafo inverse}{\pgfplots@zmax}%
				\pgfplotscoordmath{z}{tostring}{\pgfmathresult}%
				\let\pgfplots@clickable@zmax=\pgfmathresult
			\fi
			%
			% Convert to user interface of clickable lib:
			\pgfplotspointbbdiagonal
			\edef\pgfplots@clickable@wd{\the\pgf@x}%
			\pgfkeyslet{/pgfplots/annot/width}\pgfplots@clickable@wd
			\edef\pgfplots@clickable@ht{\the\pgf@y}%
			\pgfkeyslet{/pgfplots/annot/height}\pgfplots@clickable@ht
			\pgfkeyslet{/pgfplots/annot/xmin}\pgfplots@clickable@xmin
			\pgfkeyslet{/pgfplots/annot/xmax}\pgfplots@clickable@xmax
			\pgfkeyslet{/pgfplots/annot/ymin}\pgfplots@clickable@ymin
			\pgfkeyslet{/pgfplots/annot/ymax}\pgfplots@clickable@ymax
			\ifpgfplots@xislinear
				\pgfkeyssetvalue{/pgfplots/annot/xscale}{linear}%
			\else
				\pgfkeyssetvalue{/pgfplots/annot/xscale}{log:\pgfkeysvalueof{/pgfplots/log basis x}}%
			\fi
			\ifpgfplots@yislinear
				\pgfkeyssetvalue{/pgfplots/annot/yscale}{linear}%
			\else
				\pgfkeyssetvalue{/pgfplots/annot/yscale}{log:\pgfkeysvalueof{/pgfplots/log basis y}}%
			\fi
			\ifpgfplots@threedim
				\pgfkeyslet{/pgfplots/annot/zmin}\pgfplots@clickable@zmin
				\pgfkeyslet{/pgfplots/annot/zmax}\pgfplots@clickable@zmax
				\ifpgfplots@zislinear
					\pgfkeyssetvalue{/pgfplots/annot/zscale}{linear}%
				\else
					\pgfkeyssetvalue{/pgfplots/annot/zscale}{log:\pgfkeysvalueof{/pgfplots/log basis z}}%
				\fi
			\fi
			%
			% note that \pgfplotspointbblowerleft has *not* been
			% transformed after \endtikzpicture in pgfplots.code.tex.
			% That's why I can use it here:
			\pgfpointdiff
				\pgfplotspointbblowerleft
				{\pgfplotspointminminmin}%
			\pgf@sys@bp@correct\pgf@x \pgf@sys@bp@correct\pgf@y
			\edef\pgfplots@loc@TMPa{[\pgf@sys@tonumber\pgf@x, \pgf@sys@tonumber\pgf@y]}%
			\pgfkeyslet{/pgfplots/annot/minminmin}\pgfplots@loc@TMPa
			%
			\pgfplotspointxaxis
			\pgf@sys@bp@correct\pgf@x \pgf@sys@bp@correct\pgf@y
			\edef\pgfplots@loc@TMPa{[\pgf@sys@tonumber\pgf@x, \pgf@sys@tonumber\pgf@y]}%
			\pgfkeyslet{/pgfplots/annot/x axis}\pgfplots@loc@TMPa
			%
			\pgfplotspointyaxis
			\pgf@sys@bp@correct\pgf@x \pgf@sys@bp@correct\pgf@y
			\edef\pgfplots@loc@TMPa{[\pgf@sys@tonumber\pgf@x, \pgf@sys@tonumber\pgf@y]}%
			\pgfkeyslet{/pgfplots/annot/y axis}\pgfplots@loc@TMPa
			%
			\ifpgfplots@threedim
				\pgfplotspointzaxis
				\pgf@sys@bp@correct\pgf@x \pgf@sys@bp@correct\pgf@y
				\edef\pgfplots@loc@TMPa{[\pgf@sys@tonumber\pgf@x, \pgf@sys@tonumber\pgf@y]}%
				\pgfkeyslet{/pgfplots/annot/z axis}\pgfplots@loc@TMPa
			\fi
			%
			\pgfkeyssetvalue{/pgfplots/annot/dim}{\ifpgfplots@threedim 3\else2\fi}%
			%
			%
			\pgfinterruptboundingbox%
			\pgftext[left,bottom,at=\lowerleftinnercorner]{\pgfplotsclickablecreate[]}%
			\endpgfinterruptboundingbox%
		\fi
		\endgroup
	\fi
}

% This is the low-level method which creates Acroforms and Javascript
% code.
% #1: width
% #2: height
% #3: name
% #4: the axisAnnotObj object (see javascript docs above)
% #5: the displayOpts object (see javascript docs above)
\def\pgfplots@create@clickable@plotarea#1#2#3#4#5{%
%\tracingmacros=2\tracingcommands=2
	%\leavevmode
	%--------------------------------------------------
	% hyperref.sty implementation:
	%--------------------------------------------------
	% \PushButton[name=#3,borderwidth=0,bordersep=0,
	%        onclick={processAnnotatedPlot("#3", #4, #5);}]{\vbox to #2{\hsize=#1\vfill\hfill}}%
	% \TextField[name=#3-result,hidden=true]{}%
	%-------------------------------------------------- 
	%
	%-------------------------------------------------- 
	% eforms.sty implementation:
	% it allows more customization.
	%-------------------------------------------------- 
	%
	% Hier kommen Annotation Directories zum Zuge!
	% -> siehe Annotations in pdf reference
	\def\pushButtonDefaults{	
		\W{0}% border width
		\Border{0 0 0}%
	}%
	\def\textFieldDefaults{%
		\W{1}% border width
		\S{B}% border style ( one of SDBIU )
		\F{2}% "display" bitflag. 2^1= hidden (see "Annotation Flags" in pdf reference)
		%ATTENTION: \F 2 (hidden) produces INCOMPATIBILITIES with figure environment!?
		%  ---> ok, that has been fixed by recent versions of hyperref.
		%  (the pdf catalog may be incomplete in older hyperref pdftex drivers)
		\Ff{1}% "Field characteristics" bitflag. 2^0 = "read-only"
	}%
	\pushButton[
		%\A{/S/JavaScript/JS(processAnnotatedPlot("#3", #4, #5);)}%
		\AA{
			% /U = "mouse UP"
			% /D = "mouse DOWN"
			/U << /S/JavaScript/JS(axisMouseUp("#3", #4, #5);) >> 
			/D << /S/JavaScript/JS(axisMouseDown("#3");) >> 
		}%
		]
		{#3}
		{#1}{#2}%
	\textField{#3-result}{0pt}{0pt}%
	\textField{#3-result2}{0pt}{0pt}%
	\textField{#3-resultmark}{0pt}{0pt}%
	\textField{#3-result2mark}{0pt}{0pt}%
	\textField{#3-slope}{0pt}{0pt}%
	% Unfortunately, line annotations can't be created/changed in acrobat
	% javascript due to right problems :-(
	%--------------------------------------------------
	% \hbox to 0pt{\vsize=0pt \pdfannot{
	% 		/Subtype	/Line
	% 		/Open		/false
	% 		/NM			(#3-line)
	% 		/C			
	% 		/CA			
	% 		/Subj		()
	% 		/Contents	()
	% 		/L			[0 0 0 0]
	% 		/LE			[/None /OpenArrow] % PDF 1.4
	% 	%	/Ff			194
	% }}%
	%-------------------------------------------------- 
	%	
	%
	%\setLinkBbox[%
	%	\Border{}%
	%	\A{/S/JavaScript/JS(processme("#3");)}%
	%	\rawPDF{/C[1 0 0] /NM (#3)}%
	%  ]%
	%  {#1}{#2}{}%
}

\catcode`\"=\pgfplotscatcodeDQ
\endinput

### 2015-08-19 research_class/pgfplots/scripts/pgfplots/pgf2pdf.sh

@@ -1,0 +1,287 @@
+#!/bin/sh
+#
+# ATTENTION: this file is more or less deprecated.
+# Please take a look at the 'external' library which has been added to pgf.
+# At the time of this writing, this library is only available for pgf cvs (newer than 2.00).
+
+TEX_FILE=""
+TEX_LOG_FILE=""
+
+TEX_DEFINES=""
+
+OLD_DIR=`pwd`
+
+DRIVER="pdftex"
+
+ALSO_EPS_OUTPUT=0
+WARN_ONLY_IF_TEXFILE_DOESNOT_INCLUDE_TARGET=0
+VERBOSE_LEVEL=0
+
+function dumpHelp() {
+	echo -e \
+		"`basename $0` [OPTIONS] [--texdefs <defsfile> | --mainfile <latexmainfile>.tex ]  [plot1.pgf plot2.pgf .... plotn.pgf]\n"\
+		"converts each plot*.pgf to plot*.pdf.\n"\
+		"This is done by running \n"\
+		"  latex --jobname plot1 latexmainfile\n"\
+		"for each single plot. See the pgfmanual section \"Externalizing graphics\".\n"\
+		"Options:\n"\
+		"--eps\n"\
+		"    will also produce eps output files.\n"\
+		"--driver D\n"\
+		"    will use either \"dvipdfm\", \"dvips\" or \"pdflatex\"\n"\
+		"    please note that only pdflatex works without additional\n"\
+		"    work.\n"\
+		"--mainfile FILE\n"\
+		"    A tex-file which has been configured for externalized graphics.\n"\
+		"    Two conditions must be met to perform the conversion of\n"\
+		"      \"plot.pgf\"  -> \"plot.pdf\":\n"\
+		"    1. FILE needs the command\n"\
+		"         \pgfrealjobname{FILE}\n"\
+		"       (see the pgf manual for details)\n"\
+		"    2. It needs to include \"plot.pgf\" somewhere (using \input{plot.pgf})\n"\
+		"\n"\
+		"--warnonly\n"\
+		"    Use this flag if the argument of --mainfile does not contain\n"\
[…]

### 2015-08-19 research_class/pgfplots/scripts/pgfplots/pgfplots.py

@@ -1,0 +1,95 @@
+"""Module to plot using Pgfplots.
+
+This module provides a means to create and display a graph very quickly.  
+
+In this code, the program used to display the created PDF is 'xpdf'.  Change
+it to your favorite PDF reader, such as Acrobat Reader (called acroread or
+something similar)
+
+The code used to generate the graph is printed in the command line. Edit
+your graph iteratively, and when you are satisfied with the graph, copy and
+paste the relevant part to your TEX file. 
+
+This module requires the numpy module.
+
+For example of usage, see the executable part at the bottom.
+
+
+ATTENTION: this file has been provided by 3rd party users in the hope that it
+may be useful. However, it is not maintained by the pgfplots team as such. Use at your own risk.
+
+See also a related (improved) pgfplots code generation for python on https://github.com/olivierverdier/pygfplots
+
+"""
+import numpy as np
+import subprocess
+import os
+GRAPH_N = 0
+
+class Pgf:
+    def __init__(z, xlabel='', ylabel=''):
+        """Initialize and provide axis labels."""
+        z.buf = []
+        z.options = []
+        z.opt('xlabel={{{0}}}'.format(xlabel))
+        z.opt('ylabel={{{0}}}'.format(ylabel))
+        z.legend = []
+    def opt(z, *args):
+        """Write arguments to the AXIS environment."""
+        for arg in args:
+            z.options.append(arg)
+    def plot(z, x, y, legend=None, *args):
+        """Plot the data contained in the vectors x and y.
+
+        Options to the \addplot command can be provided in *args.
+        """
+        coor = ''.join(['({0}, {1})'. format(u, v) for u, v in zip(x,y)])
+        z.buf.append('\\addplot{0} coordinates {{{1}}};\n'.format(
+                ('[' + ', '.join(args) + ']') if len(args) else '' ,coor))
+        if legend is not None:
+            z.legend.append(legend)
+    def save(z, graph_n=None):
+        """Generate graph.
+        
+        If graph_n is None or a number, the graph in a file beginning with
+        zzz.  This file is meant to be temporary.  If graph_n is a string,
+        that string is used as the file name.
+        """
+        if type(graph_n) is str:
+            file_name = graph_n
+        else:
+            if graph_n is None:
+                global GRAPH_N
+                graph_n = GRAPH_N
[…]

### 2015-08-19 research_class/paper.tex
paper.tex: add testing pgfplots graphs/plots.

@@
 \DeclareGraphicsExtensions{.png,.jpg,.pdf,.eps}
+
+\usepackage{pgfplots}
+\pgfplotsset{compat=1.5}
 
@@
 
-\section{Use Cases}
+\section{Motivation/Use Cases}
 The traditional use of distributed consensus is in the data-center
@@
 
-\section{Outcome}
+\section{Results}
 
-\subsection{Testing}
+\subsection{Manual Testing}
+
+\subsubsection{Setup}
 
@@
 
-\subsection{Results}
+\subsubsection{Outcome}
 
@@
 
+\subsection{Automated Testing}
+
+\subsubsection{Setup}
+
+TBD: Setup for the automated tests.
+
+Description of SlimerJS
+
+Creation of SlimerJS test scripts using JavaScript.
+
+\subsubsection{Outcome}
+
+TBD: Outcome of the automated tests.
+
+\begin{tikzpicture}
+    \begin{axis}[
+            title=Initial Cluster Activation,
+            xlabel={Nodes},
+            ylabel={Time (ms)},
+            grid=major,
+        ]
+        \addplot table {data/data_cluster_up.dat};
+    \end{axis}
+\end{tikzpicture}
+
+\begin{tikzpicture}
+    \begin{axis}[
+            title=Message Propagation,
+            xlabel={Nodes},
+            ylabel={Time (ms)},
+            grid=major,
+        ]
+        \addplot table {data/data_cluster_up.dat};
+    \end{axis}
+\end{tikzpicture}
+
+\begin{tikzpicture}
+    \begin{axis}[
+            title=Node Failure Recovery,
+            xlabel={Nodes},
+            ylabel={Time (ms)},
+            grid=major,
+            legend entries={$Leader Down$,$Half Cluster Down$},
+        ]
+        \addplot table {data/data_kill_leader.dat};
+        \addplot table {data/data_kill_half.dat};
+    \end{axis}
+\end{tikzpicture}
+
 \section{Next Steps}
@@
   becomes completely unresponsive or just very slow.
-\item Implement a full online chat system with linearizable guarantees for
-  the chat message ordering.
 \item Explore models for dynamically adjusting Raft timeout values to
@@
   clients.
-\item Implement and test forwarding client requests rather than default
-  redirecting the client to the leader. In other words, allow client
-  requests to originate from any node of the cluster and be redirected
-  automatically to the leader (rather than the current manual
-  process).
-\item Perform quantitative and real-world testing to better characterize
-  the performance, availability and scalability characteristics of the
-  system.
 \item Implement a user interface for sending commands to the transaction
@@
   system.
-\item Implement a user interface for sending commands to the transaction
-  log and for viewing the current state of the Raft state machine.
-\item Explore models for dealing with the case where the cluster goes from
-  two nodes down to one node. Can usefulness and availability be
-  improved without sacrificing Raft safety constraints.
-\item Test the system in more various network environments. Determine the
-  reliability of establishing direct browser-to-browser communication.
-  Determine if corporate firewalls make this as difficult as is
-  implied in some of the documentation and standards texts.
-\item Explore models for timing out and automatically removing nodes
-  when they become unresponsive without needing to be notified by the
-  signaling server that the nodes have disconnected.
-\item Test larger scale deployments across multiple systems. Determine the
-  factors that affect scalability and characterize scaling limits.
-\item Perform wider cross-browser testing including Internet Explorer and
-  Opera.
-\item Implement Raft log compaction.
-\item Explore models that allow for a combination of strict shared state
-  and high performance bulk data such as media or object blobs. For
-  example, the data with full consensus might contain hashes that to
-  the refer to bulk data.
-\item Implement a Node.js based Raft over WebRTC application. This would
-  allow non-browser contexts to participate in the Raft cluster
-  together with the browser clients. PeerJS issue
-  [103](https://github.com/peers/peerjs/issues/103) must be resolved
-  for this to be feasible.
-\item Explore a design where some nodes of the cluster are non-voting and
-  are not counted as part of transaction commit quorum. However, these
-  nodes would still be included in the log replication algorithm. One
-  model this would enable would be a configuration where a small group
-  of voting nodes serve as the hub for a larger group of non-voting
-  nodes. This may enable higher scaling of the system. It also enables
-  interesting applications such as a logging or backup nodes.
-\item Implement and test forwarding client requests rather than default
-  redirecting the client to the leader. In other words, allow client
-  requests to originate from any node of the cluster and be redirected
-  automatically to the leader (rather than the current manual
-  process).
-\item Perform quantitative and real-world testing to better characterize
-  the performance, availability and scalability characteristics of the
-  system.
-\item Implement a user interface for sending commands to the transaction
-  log and for viewing the current state of the Raft state machine.
-\item Explore models for dealing with the case where the cluster goes from
-  two nodes down to one node. Can usefulness and availability be
-  improved without sacrificing Raft safety constraints.
-\item Test the system in more various network environments. Determine the
-  reliability of establishing direct browser-to-browser communication.
-  Determine if corporate firewalls make this as difficult as is
-  implied in some of the documentation and standards texts.
 \item Explore models for timing out and automatically removing nodes

### 2015-08-19 test/chat.js

@@ -1,193 +1,0 @@
-// Usage:
-// - Start rtc_server.js first:
-//   ./rtc_server --port 8000 --home chat.html
-//
-// - Then use a docker build of slimerjs to run the test:
-//   IP_ADDR=$(hostname -I | awk '{print $1}')
-//   docker run -it -v `pwd`/test/chat.js:/chat.js fentas/slimerjs slimerjs /chat.js http://${IP_ADDR}:8000/
-
-var home = '/chat.html',
-    system = require('system'),
-    webpage = require('webpage'),
-    channel = Math.round(Math.random()*1000000),
-    base_address = system.args[1],
-    server_count = (system.args.length >= 3) ? parseInt(system.args[2]) : 3,
-    page_create_delay = 1,
-    up_timeout = (10 + (server_count*server_count)/4)*1000;
-    pred_timeout = (1 + (server_count*server_count)/6)*1000;
-
-var query = '?channel='+ channel + '&console_logging=true';
-
-var pages = [];
-
-function new_page(page_id, home, firstServer, callback) {
-    console.log("new_page: " + Date.now());
-    var page = webpage.create();
-
-    // Register external handlers
-    page.onConsoleMessage = function (msg, line, origin) {
-        console.log('CONSOLE ' + page_id + ': ' + msg);
-    };
-    page.onCallback = function(msg) {
-        console.log('CALLBACK ' + page_id + ': ' + msg);
-        if (msg === 'QUIT') {
-            console.log('Normal exit');
-            slimer.exit(0);
-        }
-    };
-
-    var full_address = base_address;
-
-    if (!firstServer) {
-        full_address += home;
-    }
-    full_address += query + (firstServer ? '#firstServer' : '');
-
-    console.log('Opening ' + page_id + ': ' + full_address);
-
-    page.open(full_address, function(status) {
-        if (status !== 'success') {
-            console.log('Unable to load the address!');
-            phantom.exit(1);
-        }
-        //var mainTitle = page.evaluate(function () {
-        //    return document.title;
-        //});
-        if (callback) { callback(page); }
-    });
-
-    return page;
-}
-
-// Evaluates in page context so 'node' variable is implicit
-function get_node_info(full) {
-    if (typeof node !== 'undefined' && node) {
-        var data = {id: node._self.id,
-                    state: node._self.state,
-                    serverMapKeys: Object.keys(node._self.serverMap)};
-        if (full) {
-            data.stateMachine = node._self.stateMachine;
[…]

### 2015-08-19 test/common.js

@@ -1,49 +1,31 @@
-// Usage:
-// - Start rtc_server.js first:
-//   ./rtc_server --port 8000
-//
-// - Then use a docker build of slimerjs to run the test:
-//   IP_ADDR=$(hostname -I | awk '{print $1}')
-//   docker run -it -v `pwd`/test/rtc.js:/rtc.js fentas/slimerjs slimerjs /rtc.js http://${IP_ADDR}:8000/
+var webpage = require('webpage');
 
-var home = '/rtc.html',
-    system = require('system'),
-    webpage = require('webpage'),
-    channel = Math.round(Math.random()*1000000),
-    base_address = system.args[1],
-    server_count = (system.args.length >= 3) ? parseInt(system.args[2]) : 3,
-    page_create_delay = 1,
-    up_timeout = (10 + (server_count*server_count)/4)*1000;
-    pred_timeout = (1 + (server_count*server_count)/6)*1000;
+exports.new_page = function (page_idx, firstServer, opts, callback) {
+    var query = '?channel='+ opts.channel + '&console_logging=true';
 
-var query = '?channel='+ channel + '&console_logging=true';
-
-var pages = [];
-
-function new_page(page_id, home, firstServer, callback) {
     console.log("new_page: " + Date.now());
     var page = webpage.create();
 
     // Register external handlers
     page.onConsoleMessage = function (msg, line, origin) {
-        console.log('CONSOLE ' + page_id + ': ' + msg);
+        console.log('CONSOLE ' + page_idx + ': ' + msg);
     };
     page.onCallback = function(msg) {
-        console.log('CALLBACK ' + page_id + ': ' + msg);
+        console.log('CALLBACK ' + page_idx + ': ' + msg);
         if (msg === 'QUIT') {
             console.log('Normal exit');
             slimer.exit(0);
         }
     };
 
-    var full_address = base_address;
+    var full_address = opts.base_address;
 
     if (!firstServer) {
-        full_address += home;
+        full_address += opts.home;
     }
     full_address += query + (firstServer ? '#firstServer' : '');
 
-    console.log('Opening ' + page_id + ': ' + full_address);
+    console.log('Opening ' + page_idx + ': ' + full_address);
 
     page.open(full_address, function(status) {
         if (status !== 'success') {
@@ -59,30 +41,46 @@ function new_page(page_id, home, firstServer, callback) {
     return page;
 }
 
-// Evaluates in page context so 'node' variable is implicit
-function get_node_info(full) {
-    if (typeof node !== 'undefined' && node) {
[…]

### 2015-08-19 test/test_chat_propagate.js

@@ -1,0 +1,64 @@
+// Usage:
+// - Start rtc_server.js first:
+//   ./rtc_server --port 8000 --home chat.html
+//
+// - Then use a docker build of slimerjs to run the test:
+//   IP_ADDR=$(hostname -I | awk '{print $1}')
+//   docker run -it -v `pwd`/test:/test slimerjs-0.9.6 slimerjs /test/test_chat_propagate.js http://${IP_ADDR}:8000/
+
+var common = require('common'),
+    system = require('system'),
+    home = '/chat.html',
+    channel = Math.round(Math.random()*1000000),
+    base_address = system.args[1],
+    server_count = (system.args.length >= 3) ? parseInt(system.args[2]) : 3,
+    up_timeout = (10 + (server_count*server_count)/4)*1000;
+    pred_timeout = (1 + (server_count*server_count)/6)*1000;
+
+var pages = [];
+
+// Start checking the states
+common.wait_cluster_up(pages, server_count, up_timeout, function(status, nodes, elapsed) {
+    if (status) {
+        console.log('Cluster is up after ' + elapsed + 'ms');
+        //common.show_nodes(nodes);
+
+        console.log('Sending chat message via last page/server');
+        pages[pages.length-1].evaluate(function() {
+            $('#talk').val("test line");
+            //sendLine();
+            $('#send').click();
+        });
+        common.wait_cluster_predicate(pages, server_count, pred_timeout, function () {
+            var sm = node._self.stateMachine;
+            //console.log(nodeId + " stateMachine: " + JSON.stringify(sm));
+            if ('history' in sm && 'value' in sm.history) {
+                var lines = sm.history.value,
+                    m = lines[lines.length-1].match(/.*: test line$/);
+                return m ? true : false;
+            } else {
+                return false;
+            }
+        }, function(status, results, elapsed) {
+            if (status) {
+                console.log('Cluster state propagated after ' + elapsed + 'ms');
+                phantom.exit(0);
+            } else {
+                console.log('Cluster state failed to propagate after ' + elapsed + 'ms');
+                phantom.exit(1);
+            }
+        });
+    } else {
+        console.log('Cluster failed to come up after ' + elapsed + 'ms');
+        //common.show_nodes(nodes);
+        phantom.exit(1);
+    }
+});
+
+// Start each page/cluster node
+var opts = {base_address: base_address, home: home, channel: channel};
[…]

### 2015-08-19 test/test_kill_nodes.js

@@ -1,0 +1,72 @@
+// Usage:
+// - Start rtc_server.js first:
+//   ./rtc_server --port 8000 --home chat.html
+//
+// - Then use a docker build of slimerjs to run the test:
+//   IP_ADDR=$(hostname -I | awk '{print $1}')
+//   docker run -it -v `pwd`/test:/test slimerjs-0.9.6 slimerjs /test/test_chat_propagate.js http://${IP_ADDR}:8000/
+
+var common = require('common'),
+    system = require('system'),
+    home = '/chat.html',
+    channel = Math.round(Math.random()*1000000),
+    base_address = system.args[1],
+    server_count = (system.args.length >= 3) ? parseInt(system.args[2]) : 3,
+    up_timeout = (10 + (server_count*server_count)/4)*1000;
+    pred_timeout = (1 + (server_count*server_count)/6)*1000;
+
+var pages = [];
+
+// Kill just less than half (including the leader)
+var kill_count = parseInt((server_count-1)/2, 10);
+
+if (kill_count < 1) {
+    console.log("Must start with at least 3 nodes");
+    phantom.exit(2);
+}
+
+// Start checking the states
+common.wait_cluster_up(pages, server_count, up_timeout, function(status, nodes, elapsed) {
+    if (status) {
+        console.log('Cluster is up after ' + elapsed + 'ms');
+        //common.show_nodes(nodes);
+
+        console.log('Removing ' + kill_count + ' nodes/pages (including the leader)');
+
+        var leader_idx = common.get_leader_idx(common.get_node_info(pages));
+
+        console.log('Removing page index ' + leader_idx + ' (current leader)');
+        pages[leader_idx].close();
+        pages.splice(leader_idx, 1); // NOTE: mutates pages array in place
+
+        for (var j=0; j<kill_count-1; j++) {
+            var kill_idx = parseInt(Math.random()*kill_count, 10);
+            console.log('Removing page index ' + kill_idx);
+            pages[kill_idx].close();
+            pages.splice(kill_idx, 1); // NOTE: mutates pages array in place
+        }
+
+        console.log('Waiting for cluster to stabalize');
+        common.wait_cluster_up(pages, server_count-kill_count, up_timeout, function(status, nodes, elapsed) {
+            if (status) {
+                console.log('Cluster recovered after ' + elapsed + 'ms');
+                phantom.exit(0);
+            } else {
+                console.log('Cluster failed to recover after ' + elapsed + 'ms');
[…]

### 2015-08-19 test/test_up.js

@@ -1,0 +1,40 @@
+// Usage:
+// - Start rtc_server.js first:
+//   ./rtc_server --port 8000
+//
+// - Then use a docker build of slimerjs to run the test:
+//   IP_ADDR=$(hostname -I | awk '{print $1}')
+//   docker run -it -v `pwd`/test:/test slimerjs-0.9.6 slimerjs /test/test_up.js http://${IP_ADDR}:8000/
+
+var common = require('common'),
+    system = require('system'),
+    home = '/rtc.html',
+    channel = Math.round(Math.random()*1000000),
+    base_address = system.args[1],
+    server_count = (system.args.length >= 3) ? parseInt(system.args[2]) : 3,
+    up_timeout = (10 + (server_count*server_count)/4)*1000;
+    pred_timeout = (1 + (server_count*server_count)/6)*1000;
+
+var pages = [];
+
+// Start checking the states
+common.wait_cluster_up(pages, server_count, up_timeout, function(status, nodes, elapsed) {
+    if (status) {
+        console.log('Cluster is up after ' + elapsed + 'ms');
+        //common.show_nodes(nodes);
+
+        phantom.exit(0);
+    } else {
+        console.log('Cluster failed to come up after ' + elapsed + 'ms');
+        //common.show_nodes(nodes);
+        phantom.exit(1);
+    }
+});
+
+// Start each page/cluster node
+var opts = {base_address: base_address, home: home, channel: channel};
+pages.push(common.new_page(0, true, opts, function() {
+    for (var idx = 1; idx < server_count; idx++) {
+        pages.push(common.new_page(idx, false, opts));
+    }
+}));

### 2015-08-20 research_class/paper.tex
paper.tex: abstract, tasks, improved graphs.
- add missed kill_count command line.

@@
 \begin{abstract}
-TODO: write an abstract.
+We describe the creation of a browser-based distributed peer-to-peer
+chat application using WebRTC, PeerJS and Raft.js. WebRTC is
+a protocol that modern browsers have begun to integrate which enables
+browser-to-browser communication. PeerJS is a WebRTC client library
+and extensible signalling server. Raft.js is an implementation in
+JavaScript of the Raft distributed consensus protocol. These
+technologies are combined to create a chat application that guarantees
+a shared, consistent and linearized view of the chat history without
+relying on a single centralized server. We also create a series of
+automated tests using the SlimerJS browser automation tool and use
+them to show the correctness of the chat application and also to
+measure the performance and scalability of the application over
+various cluster sizes.
 \end{abstract}
@@
 
-During the planning phase, the project was broken down into six
-incremental steps:
+During the planning phase, the project was broken down into ten incremental steps:
 
@@
 
+\item Sessionless/Datagram RPC messages: extend Raft RPC requests and response messages so that RPC responses have enough context to be handled separately from RPC requests.
+Specifically this enables Raft to support a datagram (sessionless) network transport.
+
+\item Atomic state machine operations: implement a state machine update operation that succeeds against the specified key only if no other operation has been applied to a key since the last time the client read from that key.
+
+\item Automatic client request forwarding: extend the client request RPC to also be sessionless/datagram based.
+Implement a asynchronous JavaScript function that determines which node is the leader and forwards client request RPCs to the leader.
+This function must also correlate responses to the client request and invoke the callback for the original request.
+
+\item Chat application: add the ability to send a message to the chat log using the client request forwarding functionality with an atomic state machine operation.
+Show the current chat history based on the chat entries in the state machine.
+
 \item Server as Raft.js peer (aspirational): enable server nodes (Raft.js running on Node.js) to fully participate as Raft cluster nodes.
@@
 
-Steps 1-5 were successfully implemented for this paper. Step 6 was an
+Steps 1-9 were successfully implemented for this paper. Step 10 was an
 aspirational goal because this functionality is not yet supported in
@@
 
-TBD: Setup for the automated tests.
-
-Description of SlimerJS
-
-Creation of SlimerJS test scripts using JavaScript.
+TODO: Setup for the automated tests. Description of SlimerJS. Creation of SlimerJS test scripts using JavaScript.
 
@@
 
-TBD: Outcome of the automated tests.
+TODO: Outcome of the automated tests.
+
+The following figure shows the amount of time that the cluster takes to
+elect a leader and then reconfigure the cluster to the the final size
+(adding one node a time).
 
@@
             xlabel={Nodes},
-            ylabel={Time (ms)},
+            ylabel={Time (seconds)},
             grid=major,
@@
 \end{tikzpicture}
+
+The following figure shows the amount of time that a stable cluster takes to
+propagate, commit and apply a single state machine change to the state
+machines on every cluster node.
 
@@
         ]
-        \addplot table {data/data_cluster_up.dat};
+        \addplot table {data/data_propagate.dat};
     \end{axis}
 \end{tikzpicture}
+
+The following figure shows the amount of time that it takes for
+a stable cluster to recover (elect a new leader and remove the failed
+servers from the cluster) when one or more nodes fails. The first line
+shows the recovery time when just the leader fails. The second line
+show the recovery time when one less than half of the nodes fail.
 

### 2015-08-20 test/test_kill_nodes.js

@@ -12,16 +12,14 @@ var common = require('common'),
     channel = Math.round(Math.random()*1000000),
     base_address = system.args[1],
     server_count = (system.args.length >= 3) ? parseInt(system.args[2]) : 3,
+    kill_count = (system.args.length >= 4) ? parseInt(system.args[3]) : parseInt((server_count-1)/2, 10),
     up_timeout = (10 + (server_count*server_count)/4)*1000;
     pred_timeout = (1 + (server_count*server_count)/6)*1000;
 
 var pages = [];
 
-// Kill just less than half (including the leader)
-var kill_count = parseInt((server_count-1)/2, 10);
-
-if (kill_count < 1) {
-    console.log("Must start with at least 3 nodes");
+if (kill_count > parseInt((server_count-1)/2, 10)) {
+    console.log('Kill count must be less than half of server count');
     phantom.exit(2);
 }
 

### 2015-08-23 research_class/paper.tex
paper.tex, data/: drop acmtog, data/plots.

@@
-% Based on: v2-acmtog-sample.tex, dated March 7 2012
-%
-% Compilation using 'acmtog.cls' - version 1.2 (March 2012), Aptara Inc.
-% (c) 2010 Association for Computing Machinery (ACM)
-%
-\documentclass{acmtog} % V1.2
+\documentclass[9pt]{extarticle}
+
+\usepackage{multicol}
+\setlength{\columnsep}{0.8cm}
+
+\usepackage{geometry}
+\geometry{margin=2.0cm}
 
 \usepackage{hyperref}
+\usepackage{caption}
+\usepackage{subcaption}
 
@@
 
-%\acmVolume{VV}
-%\acmNumber{N}
-%\acmYear{YYYY}
-%\acmMonth{Month}
-%\acmArticleNum{XXX}
-%\acmdoi{10.1145/XXXXXXX.YYYYYYY}
-
-\acmVolume{28}
-\acmNumber{4}
-\acmYear{2009}
-\acmMonth{September}
-\acmArticleNum{106}
-\doi{10.1145/1559755.1559763}
-
-%ISSN
-\issn{1234-56789}
-
-% Copyright
-%\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
-%\setcopyright{usgov}
-%\setcopyright{usgovmixed}
-%\setcopyright{cagov}
-%\setcopyright{cagovmixed}
+\title{Raft Distributed Consensus Algorithm over WebRTC (Web Real-Time Communication)} % title
+
+\author{Joel D. Martin and David Levine
+\thanks{University of Texas at Arglington}
+}
 
@@
 
-\markboth{J. D. Martin et al.}{Raft Distributed Consensus Algorithm over WebRTC (Web Real-Time Communication)}
-
-\title{Raft Distributed Consensus Algorithm over WebRTC (Web Real-Time Communication)} % title
-
-\author{Joel D. Martin {\upshape and} David Levine
-\affil{University of Texas at Arglington}
-}
-
-%
-% The code below should be generated by the tool at
-% http://dl.acm.org/ccs.cfm
-% Please copy and paste the code instead of the example below. 
-%
-\begin{CCSXML}
-    <ccs2012>
-    <concept>
-    <concept_id>10003752.10003809.10010172</concept_id>
-    <concept_desc>Theory of computation~Distributed
-    algorithms</concept_desc>
-    <concept_significance>500</concept_significance>
-    </concept>
-    </ccs2012>
-\end{CCSXML}
-
-\ccsdesc[500]{Theory of computation~Distributed algorithms}
-
-%
-% End generated code
-%
-
-\keywords{Distributed systems, Real-time chat, Distributed consensus, WWW, World Wide Web, Browsers}
-
-%\acmformat{Vitor F. Pamplona, Manuel M. Oliveira, Gladimir V. G. Baranoski,
-%and Sean Fogarty. 2009. Photorealistic models for pupil light reflex and iridal pattern deformation.
-%{\em ACM Trans. Graph.} 28, 4, Article 106 (September 2009), 10 pages.\\
-%\doiline}
-
 \maketitle
 
-%\begin{bottomstuff}
-%Manuel M. Oliveira acknowledges a CNPq-Brazil fellowship (305613/2007-3). Gladimir V. G. Baranoski acknowledges a
-%NSERC-Canada grant (238337). Microsoft Brazil provided additional support.
-%Authors' addresses: Sean Fogarty, (Current address) NASA Ames Research Center, Moffett Field, California 94035.
-%\end{bottomstuff}
+\begin{multicols}{2}
 
@@
 \item \emph{Consistent order secure chat}: a chat room system where all users see the same message order but messages never traverse through a central server even though the application code itself is delivered from a central web service.
+This application was implemented for this study.
 \item \emph{Private multiuser document editing/whiteboard system}: Google Docs among other systems has demonstrated a powerful model for collaborative document creation.
@@
 implements most of the Raft algorithm including: leader election, log
-replication, safety, membership changes, client interaction, and
+replication, safety, membership changes, basic client interaction, and
 read-only operation optimizations. Log compaction and full client
 linearizability are not currently implemented (clients and client
-transactions are not assigned unique IDs). These author intends to
-implement these functions in the future, however these do not have
-a significant impact on the ability to test Raft combined with WebRTC.
+transactions are not assigned unique IDs). Log compaction is important
+for long-running production systems in order to limit memory growth
+and allow new nodes to catch up their logs more efficiently. However
+log compaction is not required for verifying overall system
+correctness and does not significantly impact short-running test
+configurations. Full client linearizability is important feature for
+simplifying certain types of distributed applications, however the
+implementation requires some significant additions to the Raft
+implemenation (e.g. assigning client IDs and sequence IDs and tracking
+timed-out clients). For the chat application, simply adding one atomic
+check-and-set state machine command is sufficient to achieve
+a linearized chat history. This does impose a slight additional cost
+on the client because it must retry the atomic operations until they
+succeed. The author intends to implement both log compaction and
+full client linearizability in the future but, however these do not
+have a significant impact on the ability to test Raft combined with
+WebRTC.
 
@@
 committing entries to the log and applying entries to the state
-machine, etc. As part of the work for implementing Raft
-over WebRTC, the Raft.js implementation was updated to reflect the
-most recent Raft paper.
+machine, clearer client interaction, etc. As part of the work for
+implementing Raft over WebRTC, the Raft.js implementation was updated
+to reflect these changes to the Raft protocol definition.
 
@@
 
-\section {Design and Implementation}
-
-\begin{figure*}[Ht]
-\centerline{\includegraphics[width=15cm]{raft_rtc_architecture}}
+\end{multicols}
+
+\begin{figure*}[h]
+\centerline{\includegraphics[width=0.8\textwidth]{raft_rtc_architecture}}
 \caption{Raft over RTC Architecture}
@@
 
+%\begin{multicols}{2}
+%\end{multicols}
+
+\begin{figure*}[h]
+\centerline{\includegraphics[width=0.8\textwidth]{raft_rtc_sequence}}
+\caption{Raft over RTC Sequence Diagram}
+  \label{fig:raft_rtc_sequence}
+\end{figure*}
+
+\begin{multicols}{2}
+
+\section {Design and Implementation}
+
 \subsection{Incremental Steps}
@@
 This step uses a statically configured three node cluster (three browser tabs).
-The PeerJS signaling server is polled to retrieve the list and other clients and the Raft cluster is started when the number of PeerJS clients reaches three.
-
-\item Updated Raft.js: update the Raft.js implementation to match the latest description of the Raft protocol (as described in "Consensus", Ongaro dissertation).
+The PeerJS signaling server is polled to retrieve the list of other clients and the Raft cluster is started when the number of PeerJS clients reaches three.
+
+\item Update Raft.js protocol: update the Raft.js implementation to match the latest description of the Raft protocol (as described in "Consensus", Ongaro dissertation).
 This includes the implementation of the simplified membership change algorithm (sequential changes) and the separation of the log commit and state machine apply concepts.
-[Additionally, the new Raft.js implementation always starts a new Raft cluster as a cluster with a single node and additional nodes are dynamically added to the cluster (as suggested in section 4.4 of the paper).]
-
-\item Dynamic Raft.js over WebRTC: enhance the signaling server with support for sessions/channels to enable multiple separate Raft clusters.
+Additionally, the new Raft.js implementation always starts a new Raft cluster as a cluster with a single node and additional nodes are dynamically added to the cluster (as suggested in section 4.4 of the paper).
+
+\item Dynamic Raft.js over WebRTC: enhance the signaling server with support for multiple sessions/channels to enable separate Raft clusters with the same signalling server.
 Start each cluster with a single node and automatically add and remove Raft nodes as they come and go.
@@
 
-\item Sessionless/Datagram RPC messages: extend Raft RPC requests and response messages so that RPC responses have enough context to be handled separately from RPC requests.
-Specifically this enables Raft to support a datagram (sessionless) network transport.
-
-\item Atomic state machine operations: implement a state machine update operation that succeeds against the specified key only if no other operation has been applied to a key since the last time the client read from that key.
+\item Sessionless/Datagram RPC messages: change Raft RPC requests and response messages so that RPC responses have enough context to be handled separately from RPC requests.
+Specifically this enables Raft to support datagram (sessionless) network transports like WebRTC.
+
+\item Atomic state machine operations: implement a state machine update operation that only succeeds if no other operation has been applied to the same key since the last time this client read from that key.
 
 \item Automatic client request forwarding: extend the client request RPC to also be sessionless/datagram based.
-Implement a asynchronous JavaScript function that determines which node is the leader and forwards client request RPCs to the leader.
+Implement an asynchronous JavaScript function that determines which node is the leader and forwards client request RPCs to the leader.
 This function must also correlate responses to the client request and invoke the callback for the original request.
 
-\item Chat application: add the ability to send a message to the chat log using the client request forwarding functionality with an atomic state machine operation.
+\item Chat application: add the ability to send a message to the chat log using the client request forwarding functionality together with atomic state machine operation.
 Show the current chat history based on the chat entries in the state machine.
@@
 \subsection{Client}
-
-\begin{figure*}[Ht]
-\centerline{\includegraphics[width=15cm]{raft_rtc_sequence}}
-\caption{Raft over RTC Sequence Diagram}
-  \label{fig:raft_rtc_sequence}
-\end{figure*}
 
@@
 
+\subsubsection{Client Request Implementation}
+
+TODO: client request functionality
+
+\subsection{Chat Application}
+
+The chat application extends the basic Raft.js over WebRTC client. The
+chat application provides a large text area containing the chat
+history and a text entry field below it. When the user clicks on the
+\emph{Send} button (or pressed enter), the text line is added to
+a pending send queue and the text entry field is blanked. A function
+to flush sends runs periodically (currently every 100ms) and does the
+following:
+
+\begin{enumerate}
+\item If there is line of text that has been sent but not yet
+    added (applied) to the state machine or there are no lines of text
+    on the pending send queue, then the function does nothing.
+\item A text line is removed from the head of the pending send queue
+    and stored in \emph{curSend} to indicate that a send is in
+    progress. The a client request is made using an atomic
+    \emph{seqAppend} command to add the new text to the state machine.
+    The \emph{seqAppend} operation attempts to append the text line to
+    a list at the \emph{history} key in the state machine.
+\item When the atomic \emph{seqAppend} operation completes, it may
+    have failed (if another text line was appended first), in which
+    case the flush function puts the failed text line back on the
+    pending send queue. Whether the send was successful or failed, the
+    \emph{curSend} variable is set to nil/null to indicate that no
+    send is in progress (the next time the flush function runs it will
+    attempt to send again).
+\end{enumerate}
+
+TODO: screenshot.
+
+TODO: atomic seqAppend operation
+
+TODO: longer delay for propagation in chat application because each
+addition to chat history take several messages to fully propagate.
+
 \section{Results}
@@
 
-\begin{figure*}[Ht]
-\centerline{\includegraphics[width=15cm]{browser_1}}
-\caption{Raft over WebRTC Browser with One Node}
+\begin{center}
+    \includegraphics[width=0.4\textwidth]{browser_1.png}
+    \captionof{figure}{Raft over WebRTC Browser with One Node}
     \label{fig:browser_1}
-\end{figure*}
+\end{center}
 
@@
 
-\begin{figure*}[Ht]
-\centerline{\includegraphics[width=15cm]{browser_2}}
-\caption{Raft over WebRTC Browser with Five Node}
+%\begin{figure*}[Ht]
+%\centerline{\includegraphics[width=15cm]{browser_2}}
+%\caption{Raft over WebRTC Browser with Five Node}
+%  \label{fig:browser_2}
+%\end{figure*}
+
+\begin{center}
+    \includegraphics[width=0.4\textwidth]{browser_2.png}
+    \captionof{figure}{Raft over WebRTC Browser with Five Nodes}
     \label{fig:browser_2}
-\end{figure*}
+\end{center}
+
+In addition, the full chat application was also tested manually. TODO:
+more text.
+
+TODO: image of chat app with one node
+
+TODO: image of chat app five nodes
 
@@
 
-\begin{tikzpicture}
-    \begin{axis}[
-            title=Initial Cluster Activation,
-            xlabel={Nodes},
-            ylabel={Time (seconds)},
-            grid=major,
-        ]
-        \addplot table {data/data_cluster_up.dat};
-    \end{axis}
-\end{tikzpicture}
-
-The following figure shows the amount of time that a stable cluster takes to
-propagate, commit and apply a single state machine change to the state
-machines on every cluster node.
-
-\begin{tikzpicture}
-    \begin{axis}[
-            title=Message Propagation,
-            xlabel={Nodes},
-            ylabel={Time (ms)},
-            grid=major,
-        ]
-        \addplot table {data/data_propagate.dat};
-    \end{axis}
-\end{tikzpicture}
+The following figure shows the amount of time that it takes for
+a stable cluster to recover (elect a new leader and remove the failed
+server from the cluster) when the current leader fails.
 
@@
 
+\end{multicols}
+
+\begin{figure}
+    \begin{subfigure}[b]{0.30\textwidth}
+        \centering
+        \resizebox{\linewidth}{!}{
             \begin{tikzpicture}
-    \begin{axis}[
-            title=Node Failure Recovery,
-            xlabel={Nodes},
-            ylabel={Time (seconds)},
+            \begin{axis}[%
+                height=1.5\textwidth, width=1.4\textwidth,
+                xlabel={Nodes}, ylabel={Time (seconds)},
+                xtick=data,
+                ymin=0, ymax=65,
                 grid=major,
-            legend entries={$Leader Down$,$Half Cluster Down$},
-        ]
-        \addplot table {data/data_kill_leader.dat};
-        \addplot table {data/data_kill_half.dat};
+                scatter/classes={%
+                    et100={mark=x,draw=blue},
+                    et100_fail={mark=x,draw=red},
+                    et100_average={mark=,draw=red}}]
+            \addplot[scatter,only marks,%
+                scatter src=explicit symbolic]%
+            table[meta=label] {data/data_cluster_up_100.dat};
+            \addplot[scatter,%
+                scatter src=explicit symbolic]%
+            table[meta=label] {data/data_cluster_up_100_averages.dat};
             \end{axis}
             \end{tikzpicture}
+        }
+        \caption{100ms election timer}
+        \label{fig:A}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.30\textwidth}
+        \centering
+        \resizebox{\linewidth}{!}{
+            \begin{tikzpicture}
+            \begin{axis}[%
+                height=1.5\textwidth, width=1.4\textwidth,
+                xlabel={Nodes}, ylabel={Time (seconds)},
+                xtick=data,
+                ymin=0, ymax=65,
+                grid=major,
+                scatter/classes={%
+                    et500={mark=x,draw=blue},
+                    et500_average={mark=,draw=red}}]
+            \addplot[scatter,only marks,%
+                scatter src=explicit symbolic]%
+            table[meta=label] {data/data_cluster_up_500.dat};
+            \addplot[scatter,%
+                scatter src=explicit symbolic]%
+            table[meta=label] {data/data_cluster_up_500_averages.dat};
+            \end{axis}
+            \end{tikzpicture}
+        }
+        \caption{500ms election timer}
+        \label{fig:B}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.30\textwidth}
+        \centering
+        \resizebox{\linewidth}{!}{
+            \begin{tikzpicture}
+            \begin{axis}[%
+                height=1.5\textwidth, width=1.4\textwidth,
+                xlabel={Nodes}, ylabel={Time (seconds)},
+                xtick=data,
+                ymin=0, ymax=65,
+                grid=major,
+                scatter/classes={%
+                    et1000={mark=x,draw=blue},
+                    et1000_average={mark=,draw=red}}]
+            \addplot[scatter,only marks,%
+                scatter src=explicit symbolic]%
+            table[meta=label] {data/data_cluster_up_1000.dat};
+            \addplot[scatter,%
+                scatter src=explicit symbolic]%
+            table[meta=label] {data/data_cluster_up_1000_averages.dat};
+            \end{axis}
+            \end{tikzpicture}
+        }
+        \caption{1000ms election timer}
+        \label{fig:C}
+    \end{subfigure}
+    \caption{Initial Cluster Activation}
+    \label{fig:cluster_up}
+
+%%% Next Row
+\rule{\textwidth}{0.3pt}
+\smallskip
+
+    \begin{subfigure}[b]{0.30\textwidth}
+        \centering
+        \resizebox{\linewidth}{!}{
+            \begin{tikzpicture}
+            \begin{axis}[%
+                height=1.5\textwidth, width=1.4\textwidth,
+                xlabel={Nodes}, ylabel={Time (seconds)},
+                xtick=data,
+                ymin=0, ymax=3.5, xmax=23,
+                grid=major,
+                scatter/classes={%
+                    kl={mark=x,draw=blue},
+                    kl_fail={mark=x,draw=red},
+                    kl_average={mark=,draw=red}}]
+            \addplot[scatter,only marks,%
+                scatter src=explicit symbolic]%
+            table[meta=label] {data/data_kill_leader_100.dat};
+            \addplot[scatter,%
+                scatter src=explicit symbolic]%
+            table[meta=label] {data/data_kill_leader_100_averages.dat};
+            \end{axis}
+            \end{tikzpicture}
+        }
+        \caption{100ms election timer}
+        \label{fig:D}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.30\textwidth}
+        \centering
+        \resizebox{\linewidth}{!}{
+            \begin{tikzpicture}
+            \begin{axis}[%
+                height=1.5\textwidth, width=1.4\textwidth,
+                xlabel={Nodes}, ylabel={Time (seconds)},
+                xtick=data,
+                ymin=0, ymax=3.5,
+                grid=major,
+                scatter/classes={%
+                    kl={mark=x,draw=blue},
+                    kl_average={mark=,draw=red}}]
+            \addplot[scatter,only marks,%
+                scatter src=explicit symbolic]%
+            table[meta=label] {data/data_kill_leader_500.dat};
+            \addplot[scatter,%
+                scatter src=explicit symbolic]%
+            table[meta=label] {data/data_kill_leader_500_averages.dat};
+            \end{axis}
+            \end{tikzpicture}
+        }
+        \caption{500ms election timer}
+        \label{fig:E}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.30\textwidth}
+        \centering
+        \resizebox{\linewidth}{!}{
+            \begin{tikzpicture}
+            \begin{axis}[%
+                height=1.5\textwidth, width=1.4\textwidth,
+                xlabel={Nodes}, ylabel={Time (seconds)},
+                xtick=data,
+                ymin=0, ymax=3.5,
+                grid=major,
+                scatter/classes={%
+                    kl={mark=x,draw=blue},
+                    kl_average={mark=,draw=red}}]
+            \addplot[scatter,only marks,%
+                scatter src=explicit symbolic]%
+            table[meta=label] {data/data_kill_leader_1000.dat};
+            \addplot[scatter,%
+                scatter src=explicit symbolic]%
+            table[meta=label] {data/data_kill_leader_1000_averages.dat};
+            \end{axis}
+            \end{tikzpicture}
+        }
+        \caption{1000ms election timer}
+        \label{fig:F}
+    \end{subfigure}
+    \caption{Recovery from Leader Failure}
+    \label{fig:E}
+
+%%% Next Row
+\rule{\textwidth}{0.3pt}
+\smallskip
+
+    \begin{subfigure}[b]{0.30\textwidth}
+        \centering
+        \resizebox{\linewidth}{!}{
+            \begin{tikzpicture}
+            \begin{axis}[%
+                height=1.5\textwidth, width=1.4\textwidth,
+                xlabel={Nodes}, ylabel={Time (seconds)},
+                xtick=data,
+                ymin=0, ymax=3.5, xmax=23,
+                grid=major,
+                scatter/classes={%
+                    kh={mark=x,draw=blue},
+                    kh_average={mark=,draw=red}}]
+            \addplot[scatter,only marks,%
+                scatter src=explicit symbolic]%
+            table[meta=label] {data/data_kill_half_100.dat};
+            \addplot[scatter,%
+                scatter src=explicit symbolic]%
+            table[meta=label] {data/data_kill_half_100_averages.dat};
+            \end{axis}
+            \end{tikzpicture}
+        }
+        \caption{100ms election timer}
+        \label{fig:G}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.30\textwidth}
+        \centering
+        \resizebox{\linewidth}{!}{
+            \begin{tikzpicture}
+            \begin{axis}[%
+                height=1.5\textwidth, width=1.4\textwidth,
+                xlabel={Nodes}, ylabel={Time (seconds)},
+                xtick=data,
+                ymin=0, ymax=3.5,
+                grid=major,
+                scatter/classes={%
+                    kh={mark=x,draw=blue},
+                    kh_average={mark=,draw=red}}]
+            \addplot[scatter,only marks,%
+                scatter src=explicit symbolic]%
+            table[meta=label] {data/data_kill_half_500.dat};
+            \addplot[scatter,%
+                scatter src=explicit symbolic]%
+            table[meta=label] {data/data_kill_half_500_averages.dat};
+            \end{axis}
+            \end{tikzpicture}
+        }
+        \caption{500ms election timer}
+        \label{fig:H}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.30\textwidth}
+        \centering
+        \resizebox{\linewidth}{!}{
+            \begin{tikzpicture}
+            \begin{axis}[%
+                height=1.5\textwidth, width=1.4\textwidth,
+                xlabel={Nodes}, ylabel={Time (seconds)},
+                xtick=data,
+                ymin=0, ymax=3.5,
+                grid=major,
+                scatter/classes={%
+                    kh={mark=x,draw=blue},
+                    kh_average={mark=,draw=red}}]
+            \addplot[scatter,only marks,%
+                scatter src=explicit symbolic]%
+            table[meta=label] {data/data_kill_half_1000.dat};
+            \addplot[scatter,%
+                scatter src=explicit symbolic]%
+            table[meta=label] {data/data_kill_half_1000_averages.dat};
+            \end{axis}
+            \end{tikzpicture}
+        }
+        \caption{1000ms election timer}
+        \label{fig:I}
+    \end{subfigure}
+    \caption{Recovery from Half Cluster Failure}
+    \label{fig:G}
+\end{figure}
+
+\begin{multicols}{2}
+
+The following figure shows the amount of time that a stable cluster takes to
+propagate, commit and apply a single state machine command to the state
+machines on every cluster node.
+
+\begin{center}
+    \begin{tikzpicture}
+    \begin{axis}[%
+        xlabel={Nodes}, ylabel={Time (seconds)},
+        xtick=data,
+        grid=major,
+        scatter/classes={%
+            propagate={mark=x,draw=blue},
+            propagate_average={mark=,draw=red}}]
+    \addplot[scatter,only marks,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/data_propagate.dat};
+    \addplot[scatter,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/data_propagate_averages.dat};
+    \end{axis}
+    \end{tikzpicture}
+\captionof{figure}{Propagate$,$ Commit$,$ and Apply Command}
+\label{fig:propagate}
+\end{center}
 
@@
 
-\begin{acks}
+\section*{Acknowledgements}
 I would like to thank my academic supervisor, David Levine, for his
@@
 weekends I spent on this project. Thanks!
-\end{acks}
-
-% Bibliography
-\bibliographystyle{ACM-Reference-Format-Journals}
+
 % TODO: references
-\bibliography{acmtog-sample-bibfile}
-                                % Sample .bib file with references that match those in
-                                % the 'Specifications Document (V1.5)' as well containing
-                                % 'legacy' bibs and bibs with 'alternate codings'.
-                                % Gerry Murray - March 2012
-
-%\received{September 2008}{March 2009}
-
+
+\end{multicols}
 \end{document}
-% End of v2-acmtog-sample.tex (March 2012) - Gerry Murray, ACM

### 2015-08-23 rtc.js
rtc.js: parameterize addRemoveServer intervals.

@@ -4,7 +4,17 @@
 var verbose = 1,
     debug = true,
     electionTimeout = 1000,
-    connPollDelay = 3000;
+    // addRemoveServersRetry: smaller values increase the number of
+    // PENDING_CONFIG_CHANGE that will happen when there are multiple
+    // changes pending (more than one server added or removed). Larger
+    // values mean that multiple changes may take longer to complete.
+    addRemoveServersRetry = 50,
+    // addRemoveServersPoll: smaller values increase the CPU usage due
+    // to constant scanning of the PeerJS connections and Raft.js
+    // server map to identify nodes that need to be added or dropped.
+    // Larger values means that it takes longer to detect when cluster
+    // changes happen.
+    addRemoveServersPoll = 500;
 
 // Global state
 var messages = document.getElementById('messages'),
@@ -186,9 +196,9 @@ function addRemoveServersAsync() {
 
     if (changes > 1) {
         // If there is still pending changes then cycle around faster
-        setTimeout(addRemoveServersAsync, 50);
+        setTimeout(addRemoveServersAsync, addRemoveServersRetry);
     } else {
-        setTimeout(addRemoveServersAsync, 500);
+        setTimeout(addRemoveServersAsync, addRemoveServersPoll);
     }
 }
 

### 2015-08-23 research_class/paper.tex
paper.tex: add chat images.

@@
 
+\usepackage{authblk}
+
+\usepackage{microtype}
+
 \usepackage{graphicx}
@@
 
-\title{Raft Distributed Consensus Algorithm over WebRTC (Web Real-Time Communication)} % title
-
-\author{Joel D. Martin and David Levine
-\thanks{University of Texas at Arglington}
-}
+\title{Raft Distributed Consensus Algorithm over WebRTC (Web Real-Time Communication)}
+
+\author{Joel D. Martin}
+\author{David Levine}
+\affil{University of Texas at Arglington}
 
@@
 We describe the creation of a browser-based distributed peer-to-peer
-chat application using WebRTC, PeerJS and Raft.js. WebRTC is
-a protocol that modern browsers have begun to integrate which enables
-browser-to-browser communication. PeerJS is a WebRTC client library
-and extensible signalling server. Raft.js is an implementation in
-JavaScript of the Raft distributed consensus protocol. These
-technologies are combined to create a chat application that guarantees
-a shared, consistent and linearized view of the chat history without
-relying on a single centralized server. We also create a series of
-automated tests using the SlimerJS browser automation tool and use
-them to show the correctness of the chat application and also to
-measure the performance and scalability of the application over
-various cluster sizes.
+application using WebRTC, PeerJS and Raft.js. WebRTC is a protocol that modern
+browsers have begun to integrate which enables browser-to-browser
+communication. PeerJS is a WebRTC client library and extensible signalling
+server. Raft.js is an implementation in JavaScript of the Raft distributed
+consensus protocol. These technologies are combined to create a chat
+application that guarantees a shared, consistent and linearized view of the
+chat history without relying on a single centralized server. We also create a
+series of automated tests using the SlimerJS browser automation tool and use
+them to show the correctness of the chat application and also to measure the
+performance and scalability of the application over various cluster sizes.
 \end{abstract}
@@
 
+\section {Design and Implementation}
+
 \end{multicols}
@@
 \begin{figure*}[h]
-\centerline{\includegraphics[width=0.8\textwidth]{raft_rtc_architecture}}
+\centerline{\includegraphics[width=0.8\textwidth]{imgs/raft_rtc_architecture}}
 \caption{Raft over RTC Architecture}
@@
 \begin{figure*}[h]
-\centerline{\includegraphics[width=0.8\textwidth]{raft_rtc_sequence}}
+\centerline{\includegraphics[width=0.8\textwidth]{imgs/raft_rtc_sequence}}
 \caption{Raft over RTC Sequence Diagram}
@@
 \begin{multicols}{2}
-
-\section {Design and Implementation}
 
@@
 \begin{center}
-    \includegraphics[width=0.4\textwidth]{browser_1.png}
-    \captionof{figure}{Raft over WebRTC Browser with One Node}
+    \includegraphics[width=0.45\textwidth]{imgs/browser_1.png}
+    \captionof{figure}{Basic app with one Node/Tab}
     \label{fig:browser_1}
@@
 
-%\begin{figure*}[Ht]
-%\centerline{\includegraphics[width=15cm]{browser_2}}
-%\caption{Raft over WebRTC Browser with Five Node}
-%  \label{fig:browser_2}
-%\end{figure*}
-
 \begin{center}
-    \includegraphics[width=0.4\textwidth]{browser_2.png}
-    \captionof{figure}{Raft over WebRTC Browser with Five Nodes}
-    \label{fig:browser_2}
+    \includegraphics[width=0.45\textwidth]{imgs/browser_5.png}
+    \captionof{figure}{Basic app with Five Nodes/Tabs}
+    \label{fig:browser_5}
 \end{center}
@@
 more text.
+
+\begin{center}
+    \includegraphics[width=0.45\textwidth]{imgs/chat_1a.png}
+    \captionof{figure}{Chat app with one Node/Tab}
+    \label{fig:browser_chat_1}
+\end{center}
+
+\begin{center}
+    \includegraphics[width=0.45\textwidth]{imgs/chat_5a.png}
+    \captionof{figure}{Chat app with Five Nodes/Tabs}
+    \label{fig:browser_chat_5}
+\end{center}
 

### 2015-08-23 research_class/paper.tex
Expand outcome section. Comment out TODOs

@@
 discussed in this paper is based on "Consensus: Bridging Theory and
-Practice" by Diego Ongaro (PhD Dissertation) TODO/CITE.
+Practice" by Diego Ongaro (PhD Dissertation). % TODO/CITE
 
@@
 Raft.js is an implementation of the Raft algorithm in JavaScript that
-was creator by the author of this paper TODO/CITE. Raft.js is designed
-to run in either a browser environment or within node.js (server-side
-JavaScript). Raft.js implements the Raft algorithm as described in
-"Consensus", Ongaro (dissertation TODO/CITE).
+was creator by the author of this paper. %TODO/CITE
+Raft.js is designed to run in either a browser environment or within
+node.js (server-side JavaScript). Raft.js implements the Raft
+algorithm as described in "Consensus", Ongaro (dissertation).
+% TODO/CITE
 
@@
 currently in the process of being standardized by the W3C organization
-(TODO/CITE: http://www.w3.org/TR/webrtc/)
+(http://www.w3.org/TR/webrtc/) % TODO/CITE
 
@@
 PeerServer called ExpressPeerServer that combines a PeerJS signaling
-server with the Node.js based Express web application framework
-(TODO/CITE).
+server with the Node.js based Express web application framework.
+% (TODO/CITE)
 
@@
 the `addServer` and `removeServer` RPC calls. Servers additions are
-prioritized before removals (as recommended in Ongaro 4.4 TODO/CITE)
+prioritized before removals (as recommended in Ongaro 4.4) % TODO/CITE
 in order to maximize availability.
@@
 
-\subsubsection{Client Request Implementation}
-
-TODO: client request functionality
+\subsubsection{Client Request Redirect Implementation}
+
+In the Raft over WebRTC implemenation each local browser is a client
+of the Raft cluster. However in Raft, the leader node is the only node
+that adds entries to the log so clients on follower nodes must have
+a way to communicate with the leader. This is implemented as pair of
+functions named clientRequest and clientRequestResponse. clientRequest
+is called by the client with the arguments of a client request RPC and
+a callback function to be called once the request is completed. The
+clientRequest/clientRequestResponse functions work as follows:
+
+\begin{enumerate}
+\item the clientRequestResponse is registered with the local Raft.js
+    implementation as the callback function anytime
+    clientRequestResponse RPCs are received.
+\item the clientRequest function keeps track of the current leader ID.
+    If the current leader ID is not set or this node is the leader
+    then the client's request is sent directly to the local Raft node.
+    If the current leader ID is set then the request is sent to that
+    node as a normal RPC request over the WebRTC channel.
+\item when the clientRequestResponse function is called and the status
+    is "NOT\_LEADER" then the current leader ID being tracked by the
+    clientRequest function is set to the leader ID returned in the
+    response. Then the original request is re-issued over the normal
+    WebRTC channel to the leader.
+\item when the clientRequestResponse function is called with
+    a successful status, the original callback argument of the
+    clientRequest function is called with the results from the RPC
+    response.
+\end{enumerate}
 
@@
 history and a text entry field below it. When the user clicks on the
-\emph{Send} button (or pressed enter), the text line is added to
+\emph{Send} button (or pressed enter), the current node ID is
+prepended to the beginning of the text line and it is added to
 a pending send queue and the text entry field is blanked. A function
@@
 
-TODO: screenshot.
-
-TODO: atomic seqAppend operation
-
-TODO: longer delay for propagation in chat application because each
-addition to chat history take several messages to fully propagate.
+The chat application also extends the default set of state machine
+operations to add a seqAppend operation. This is an atomic operation
+that is implemented as follows:
+
+\begin{enumerate}
+\item the seqAppend operation operates on a specially structured
+    key/value pair in the state machine map. The value is
+    a associative map that contains a counter and list structure that
+    contains the actual data. seqAppend operations are sent with both
+    a count and then new value to append.
+\item When a seqAppend operation is applied to the state machine and
+    the count value in the command matches the counter in the state
+    machines value, then the new value in the request is added to the
+    list and the counter is incremented. If the count value does not
+    match then counter then the the operation is not applied. In
+    either case, a list is returned that contains two values: whether
+    the seqAppend operation succeeded and what the current value of
+    the counter.
+\item the first time the seqAppend operation is used on a non-existent
+    key, the structure described above is created with the counter set
+    to 0 and the list structure is created with the single value
+    passed in the seqAppend operation.
+\item clients that call the seqAppend operation should check the
+    status value in the result list. If the status is false, then the
+    client should retry the operation using the counter value as the
+    count in the operation.
+\end{enumerate}
 
@@
 
-In addition, the full chat application was also tested manually. TODO:
-more text.
+In addition, the full chat application was also tested manually by
+loading the initial chat application landing page to create a Raft
+over WebRTC cluster with a single leader. Then one or more lines of
+text were entered into the text line each followed by clicking the
+Send button.  Here is a screenshot of the browser after the first
+node/tab of the chat application is created and a line of text was
+sent:
 
@@
 
+Then four more Raft nodes were then added to the cluster by clicking
+on the embedded "create more clients" link in the application.
+Finally, on a non-leader node more text lines were entered and sent by
+clicking the Send button. Here is a screenshot of the browser once
+four new nodes/tabs are added to the cluster (5 total) and another
+text line was sent:
+
 \begin{center}
@@
 \end{center}
-
-TODO: image of chat app with one node
-
-TODO: image of chat app five nodes
 
@@
 
+In the context of the full chat application the slowdown of background
+pages is even more pronounced because each addition to the chat
+history can take several messages to fully propagate the new date.
+
 \subsection{Automated Testing}
@@
 
-TODO: Setup for the automated tests. Description of SlimerJS. Creation of SlimerJS test scripts using JavaScript.
+Automated testing of the Raft over WebRTC system was performed by
+leveraging the SlimerJS scriptable browser environment. SlimerJS
+enables web applications to be scripted by loading pages in a special
+Mozilla rendering environment called XULRunner. The XULRunner instance
+is directed to a X virtual frame buffer (xvfb) which provides
+effectively headless testing.
+
+To simplify the execution of the tests and to make the testing
+infrastructure more portable, a docker container environment was used
+to encapsulate and execute the SlimerJS environment. The container is
+built using \emph{test/Dockerfile} which is in turn derived from
+the "fentas/slimerjs" docker image
+(https://github.com/fentas/docker/blob/master/slimerjs/Dockerfile).
+This custom Dockerfile was created because the upstream version of the
+Dockerfile used an older version of XULRunner that did not support
+more than 7 WebRTC clients communicating together simultaneously.
+% (TODO/CITE)
+
+Writing tests using SlimerJS involves creating JavaScript programs
+that are run within the SlimerJS environment. These scripts can create
+browser instances (basically headless browser tabs), query the status
+of those pages and execute JavaScript within the context of the pages
+themselves.
+
+Three different tests were created to test the Raft over WebRTC
+functionality and the chat application. Each of the tests takes
+a parameter indicating the number of nodes/pages to launch and test
+again. The tests scripts files are:
+
+\begin{enumerate}
+    \item \emph{test/common.js}: a module of common functions used by
+        the other tests scripts.
+    \item \emph{test/test\_up.js}: launch and start an initial leader
+        node/page, then start the rest of the nodes (if any) so that
+        the leader will configure them into the cluster. Then test
+        then waits for every node to be recorded (applied) into the log
+        of every other node of the cluster. The test reports the
+        number of milliseconds for the cluster to reach this state.
+    \item \emph{test/test\_kill\_nodes.js}: similar to
+        \emph{test/test\_up.js} but once all the nodes are configured
+        into the cluster, this script then kills the leader plus
+        a configurable number of nodes. By default the script kills
+        just under half the nodes of the cluster. The test then waits
+        for a new leader to be elected and remove the dead nodes from
+        the cluster. The test reports both the number of milliseconds
+        for the cluster to come up and the recover from the failure.
+    \item \emph{test/test\_chat\_propagate.js}: similar to
+        \emph{test/test\_up.js} but once the cluster is up, the test
+        connects to one of the non-leader nodes and and sends a chat
+        text line. The test then waits for the chat message to
+        propogate to every other node's state machine. The test
+        reports both the number of milliseconds for the cluster to
+        come up and for the messages to propagate to every node.
+\end{enumerate}
+
+The tests were all run on a system with the following configuration:
+
+\begin{enumerate}
+    \item \emph{CPUs}: 8 Intel Core i7 processors running at 3.20GHz
+    \item \emph{Memory}: 12 GB
+    \item \emph{OS}: Ubuntu 14.04.1 LTS
+    \item \emph{Docker}: version 1.3.1
+    \item \emph{Docker image}:
+    \begin{enumerate}
+        \item \emph{OS}: Ubuntu 12.04.5 LTS
+        \item \emph{SlimerJS}: vesion 0.9.6
+        \item \emph{XULRunner}: 38.0.5
+    \end{enumerate}
+\end{enumerate}
+
+Three tests were run at 10 different cluster sizes (odd cluster sizes
+from 3 to 21 nodes): \emph{test/test\_up.js} and
+\emph{test/test\_kill\_nodes.js} in kill one (leader) node and kill
+half nodes. Each of these tests was run with three different Raft election
+timeout values: 100ms, 500ms, 1000ms. The
+\emph{test/test\_chat\_propagate.js} test was run at
+11 different cluster sizes (odd cluster sizes from 1 to 21 nodes).
+Each test mode and cluster size was run 7 times for a total of 707
+test runs (3*10*7*3 + 11*7)
 
@@
 
-TODO: Outcome of the automated tests.
-
-The following figure shows the amount of time that the cluster takes to
-elect a leader and then reconfigure the cluster to the the final size
-(adding one node a time).
-
-The following figure shows the amount of time that it takes for
-a stable cluster to recover (elect a new leader and remove the failed
-server from the cluster) when the current leader fails.
-
-The following figure shows the amount of time that it takes for
-a stable cluster to recover (elect a new leader and remove the failed
-servers from the cluster) when one or more nodes fails. The first line
-shows the recovery time when just the leader fails. The second line
-show the recovery time when one less than half of the nodes fail.
-
-\end{multicols}
-
-\begin{figure}
+% TODO: programmatic figure numbers
+Figure 7 shows the results of running the
+\emph{test/test\_chat\_propagate.js} test to show the amount of time
+that a stable cluster takes to propagate, commit and apply a single
+state machine command to the state machines on every cluster node.
+
+Figure 7 is a scatterplot with each test run as a blue "x" and a black
+line representing the average at each cluster size.
+
+\begin{center}
+    \begin{tikzpicture}
+    \begin{axis}[%
+        xlabel={Nodes}, ylabel={Time (seconds)},
+        xtick=data,
+        grid=major,
+        scatter/classes={%
+            propagate={mark=x,draw=blue},
+            propagate_average={mark=,draw=red}}]
+    \addplot[scatter,only marks,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/data_propagate.dat};
+    \addplot[scatter,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/data_propagate_averages.dat};
+    \end{axis}
+    \end{tikzpicture}
+\captionof{figure}{Propagate$,$ Commit$,$ and Apply Command}
+\label{fig:propagate}
+\end{center}
+
+Figures 8, 9 and 10 show the results of running the
+\emph{test/test\_up.js} and \emph{test/test\_kill\_nodes.js} in both
+modes (kill only leader and kill just under half the nodes). Three
+different scatterplots are shown for each test mode to test different
+values of the Raft election timeout value: 100ms, 500ms, and 1000ms.
+The election timeout value makes very little difference to the average
+time curve line. The reason for this is that the first leader tends to
+stay the leader as new nodes are added so for the most part the
+election timer does not have an impact.
+
+Some specific notes about the figures:
+
+% TODO: link figure numbers
+
+\begin{enumerate}
+    \item \emph{Figure 8 (a)}: The results show that once the cluster
+        size reaches 17, the cluster sometimes does not activate
+        before the test times out (the red "x" marks). Once the
+        cluster reaches 21 nodes, all 7 test runs time out. The reason
+        for this is because at 17 nodes, the leader is not always able
+        to send out appendEntries RPCs to all the other nodes before
+        their election timeout fires. More analysis is needed, but one
+        possible explanation is that the single SlimerJS XULRunner
+        process that is handling all the cluster nodes is not able to
+        give enough CPU time to the leader node to reliable send
+        appendEntries RPCs to all the other nodes within the election
+        timeout value. Increasing the election timeout to 500ms or
+        greater eliminates the failures without significantly changing
+        the average cluster activation time curve.
+    \item \emph{Figure 9 (a)}: The results show that when the cluster
+        size reaches 15, then average amount of time for cluster
+        recovery (after the leader is killed) increases dramatically
+        (not all data points are shown on the plot for size 15). At
+        cluster size 17 the cluster is no longer able to recover
+        within the test timeout period. Increasing the election
+        timeout to 500ms or greater eliminates the failures at the
+        cost of slightly increased recovery time for smaller cluster
+        sizes.
+    \item \emph{Figure 10 (a)}: During this test cluster was unable to
+        activate at cluster size 21 when the election timeout value is
+        set to 100ms (see Figure 8 (a)). This means there was no
+        opportunity for cluster recovery at cluster size 21 so there
+        is no data for that column.
+
+\end{enumerate}
+
+Each figure is a scatterplot with each test run as a blue "x" and
+a black line representing the average at each cluster size.
+
+%\end{multicols}
+
+\begin{figure*}[h]
     \begin{subfigure}[b]{0.30\textwidth}
@@
     \label{fig:G}
-\end{figure}
-
-\begin{multicols}{2}
-
-The following figure shows the amount of time that a stable cluster takes to
-propagate, commit and apply a single state machine command to the state
-machines on every cluster node.
-
-\begin{center}
-    \begin{tikzpicture}
-    \begin{axis}[%
-        xlabel={Nodes}, ylabel={Time (seconds)},
-        xtick=data,
-        grid=major,
-        scatter/classes={%
-            propagate={mark=x,draw=blue},
-            propagate_average={mark=,draw=red}}]
-    \addplot[scatter,only marks,%
-        scatter src=explicit symbolic]%
-    table[meta=label] {data/data_propagate.dat};
-    \addplot[scatter,%
-        scatter src=explicit symbolic]%
-    table[meta=label] {data/data_propagate_averages.dat};
-    \end{axis}
-    \end{tikzpicture}
-\captionof{figure}{Propagate$,$ Commit$,$ and Apply Command}
-\label{fig:propagate}
-\end{center}
+\end{figure*}
+
+%\begin{multicols}{2}
 

### 2015-08-23 research_class/paper.tex
paper.tex: spell check.

@@
 \author{David Levine}
-\affil{University of Texas at Arglington}
+\affil{University of Texas at Arlington}
 
@@
 browsers have begun to integrate which enables browser-to-browser
-communication. PeerJS is a WebRTC client library and extensible signalling
+communication. PeerJS is a WebRTC client library and extensible signaling
 server. Raft.js is an implementation in JavaScript of the Raft distributed
@@
 The Raft algorithm breaks the problem of consensus into three core
-subproblems: leader election, log replication, and safety. A full Raft
+sub-problems: leader election, log replication, and safety. A full Raft
 solution must also address: cluster membership changes, log
@@
 implementation requires some significant additions to the Raft
-implemenation (e.g. assigning client IDs and sequence IDs and tracking
+implementation (e.g. assigning client IDs and sequence IDs and tracking
 timed-out clients). For the chat application, simply adding one atomic
@@
 
-\item Dynamic Raft.js over WebRTC: enhance the signaling server with support for multiple sessions/channels to enable separate Raft clusters with the same signalling server.
+\item Dynamic Raft.js over WebRTC: enhance the signaling server with support for multiple sessions/channels to enable separate Raft clusters with the same signaling server.
 Start each cluster with a single node and automatically add and remove Raft nodes as they come and go.
@@
 
-In the Raft over WebRTC implemenation each local browser is a client
+In the Raft over WebRTC implementation each local browser is a client
 of the Raft cluster. However in Raft, the leader node is the only node
@@
         text line. The test then waits for the chat message to
-        propogate to every other node's state machine. The test
+        propagate to every other node's state machine. The test
         reports both the number of milliseconds for the cluster to
@@
         \item \emph{OS}: Ubuntu 12.04.5 LTS
-        \item \emph{SlimerJS}: vesion 0.9.6
+        \item \emph{SlimerJS}: version 0.9.6
         \item \emph{XULRunner}: 38.0.5
@@
 
-\section*{Acknowledgements}
+\section*{Acknowledgments}
 I would like to thank my academic supervisor, David Levine, for his

### 2015-09-05 research_class/paper.tex
Initial working biblatex bibliography/citations.
Also:

- update pgfplots compat version
- generate bibliography in makefile
- add clean rule to Makefile
- add a couple initial reference files (raft, other)
- start proofread updates.

@@
 \geometry{margin=2.0cm}
+
+%\usepackage[utf8]{inputenc}
+\usepackage[backend=bibtex]{biblatex}
 
@@
 \usepackage{pgfplots}
-\pgfplotsset{compat=1.5}
+\pgfplotsset{compat=1.9}
 
@@
 
+\addbibresource{references/raft.bib}
+\addbibresource{references/other.bib}
+
 \begin{document}
@@
 We describe the creation of a browser-based distributed peer-to-peer
-application using WebRTC, PeerJS and Raft.js. WebRTC is a protocol that modern
-browsers have begun to integrate which enables browser-to-browser
-communication. PeerJS is a WebRTC client library and extensible signaling
-server. Raft.js is an implementation in JavaScript of the Raft distributed
-consensus protocol. These technologies are combined to create a chat
-application that guarantees a shared, consistent and linearized view of the
-chat history without relying on a single centralized server. We also create a
-series of automated tests using the SlimerJS browser automation tool and use
-them to show the correctness of the chat application and also to measure the
-performance and scalability of the application over various cluster sizes.
+application using WebRTC, PeerJS and Raft.js. WebRTC is a protocol
+that modern browsers have begun to integrate which enables
+browser-to-browser communication. PeerJS is a WebRTC client library
+and extensible signaling server. Raft.js is an implementation in
+JavaScript of the Raft distributed consensus protocol. These
+technologies are combined to create a chat application that guarantees
+a shared, consistent and linearized view of the chat history without
+relying on a single centralized server for communication. We also
+create a series of automated tests using the SlimerJS browser
+automation tool and use them to show the correctness of the chat
+application and also to measure the performance and scalability of the
+application for various cluster sizes.
 \end{abstract}
@@
 thank Diego Ongaro for his amazing work on the Raft protocol and paper
-which made implementing a Raft implementation a pleasure. There are
-too many people involved in the creation and standardization of WebRTC
-but I would like to thank them for their tireless work to make
-browser-to-browser communication possible. Last but not least, I would
-like to thank my family for putting up with the long nights and
-weekends I spent on this project. Thanks!
+\cite[p.~215]{lamport78} which made implementing
+a Raft implementation a pleasure \cite{lamport78}.
+There are too many people involved in the creation and standardization
+of WebRTC but I would like to thank them for their tireless work to
+make browser-to-browser communication possible. Last but not least,
+I would like to thank my family for putting up with the long nights
+and weekends I spent on this project. Thanks!
 
@@
 \end{multicols}
+
+\printbibliography
+%\bibliographystyle{plain}
+%\bibliography{references/raft,references/other}
+
 \end{document}

### 2015-09-05 research_class/paper.tex
Proofread and citations through page 3.

@@
 \addbibresource{references/raft.bib}
+\addbibresource{references/w3c.bib}
+\addbibresource{references/ietf.bib}
 \addbibresource{references/other.bib}
@@
 communication including WebSockets for full-duplex browser-to-server
-communication and WebRTC for direct browser-to-browser communication.
-These new capabilities enable software developers to create
-distributed network applications with browser applications as first
-class citizens.
+communication \cite{websockets:rfc6455} and WebRTC for direct
+browser-to-browser communication \cite{webrtc:feb2015}. These new
+capabilities enable software developers to treat web applications as
+first class participants in distributed network applications.
 
@@
 one category of distributed state management in which multiple nodes
-must come to agreement on the state of single value or set of values.
+must come to agreement on the state of single value or set of values. % TODO/CITE
 The grandparent of distributed consensus algorithms is the Paxos
-algorithm described by Leslie Lamport. However, Paxos has been
-notoriously difficult to understand and to implement correctly. Raft
-is a more recent distributed consensus algorithm that aims for
-understandability and practicality of implementation.
-
-In this study we will explore an implementation of the Raft protocol
+algorithm described by Leslie Lamport \cite{paxos:lamport98}. However,
+Paxos has been notoriously difficult to understand and to implement
+correctly \cite{paxos:chandra07}. Raft is a more recent distributed
+consensus algorithm that aims for understandability and practicality
+of implementation \cite{raft_thesis:ongaro14}.
+
+In this paper we will describe an implementation of the Raft protocol
 in a browser context using WebRTC as the communication channel. We
@@
 The traditional use of distributed consensus is in the data-center
-(private or cloud based) especially as the core coordination component
-of a replicated data-storage system. In these examples, the set of
-nodes that compose the cluster tend to be fairly static and changes to
-cluster membership are usually for the purpose of replacing nodes.
-Changes to the long term size of the cluster are uncommon.
-
-Most existing distributed browser applications are extensions of this
-model in which browsers connect to load-balanced web servers which are
-in close network proximity to the distributed consensus cluster. The
-backend web system delivers the web application to the browser and
-also serves as the conduit for all browser-to-browser communication
-and state management.
-
-However, it is not always desirable for application messages and state
-to traverse the server back-end for various reasons including the
-following:
-
-\begin{enumerate}
-\item \emph{security (privacy)}: application state and user messages may need to traverse directly from browser to browser rather than traversing through a server.
-\item \emph{locality}: the nature of the application may allow clusters of applications that are nearby on a network (e.g.
-intranet) to communicate and coordinate directly rather than passing all messages and state through a server back-end which may be considerable more remote.
-\item \emph{bandwidth and scale}: a distributed browser based application may be able to avoid expensive bandwidth into and out of the backend data-center by passing messages and state directly between browsers.
-Depending on the design of the application this may also enable better scalability then a more centralized model.
-\end{enumerate}
-
+(private or cloud based) often as the core coordination component
+of a replicated data-storage system. %TODO/CITE
+In these examples, the set of nodes that compose the cluster tend to
+be fairly static and changes to cluster membership are usually for the
+purpose of replacing nodes. Changes to the long-term size of the
+cluster are uncommon. %TODO/CITE
+
+Most existing distributed browser applications are extensions of the
+data-centric model in which browsers connect to load-balanced web
+servers which are in close network proximity to the distributed
+consensus cluster. %TODO/CITE
+The backend web servers deliver the web application to the browser and
+also serve as the conduit for all browser-to-browser communication and
+state management.
+
+However, requiring application messages and application state to
+traverse the back-end servers is not always desirable for various
+reasons including the following:
+
+\begin{enumerate}
+\item \emph{security (privacy)}: application state and user messages
+    may need to traverse directly from browser to browser rather than
+    traversing through a server.
+\item \emph{locality}: the nature of the application may allow
+    clusters of applications that are nearby on a network (e.g.
+    intranet) to communicate and coordinate directly rather than
+    passing all messages and state through a server back-end which may
+    be considerably more remote.
+\item \emph{bandwidth and scale}: a distributed browser based
+    application may be able to avoid expensive bandwidth into and out
+    of the backend data-center by passing messages and state directly
+    between browsers.  Depending on the design of the application this
+    may also enable better scalability then a more centralized model.
+\end{enumerate}
+
+%TODO: better lead in to this paragraph from above
 Here are some specific use cases in which Raft over WebRTC may be
@@
 \begin{enumerate}
-\item \emph{Consistent order secure chat}: a chat room system where all users see the same message order but messages never traverse through a central server even though the application code itself is delivered from a central web service.
-This application was implemented for this study.
-\item \emph{Private multiuser document editing/whiteboard system}: Google Docs among other systems has demonstrated a powerful model for collaborative document creation.
-However, these systems all user central web services to coordinate shared state and handle browser-to-browser communication.
-Using a Raft over WebRTC could enable tools with similar functionality but without the central coordination and communication.
-\item \emph{Browser based network games}: most network games have certain states that must be agreed upon by all nodes (e.g.
-avatar dead or alive, etc).
-This can be achieved with Raft over WebRTC without requiring any data to traverse a central server.
-In addition to reducing browser to server bandwidth requirements, this may also allow lower latency gaming if games instances are partitioned by latency locality.
+\item \emph{Consistent order secure chat}: a chat room system where
+    all users see the same message history but the messages never
+    traverse through a central server and there is no central arbiter
+    of trusth. The application code itself may still be delivered from
+    a central web service. This is the use case that was implemented
+    for this paper.
+\item \emph{Private multiuser document editing/whiteboard system}:
+    Google Docs, among other examples has demonstrated a powerful
+    model for collaborative document creation. However, these systems
+    all use a centralized servers to coordinate shared state and
+    handle browser-to-browser communication. %TODO/CITE
+    Using a Raft over WebRTC could enable applications with similar
+    functionality but without the central coordination and
+    communication.
+\item \emph{Browser based network games}: most network games have
+    certain states that must be agreed upon by all nodes (e.g. avatar
+    dead or alive, etc). This can be achieved with Raft over WebRTC
+    without requiring any data to traverse a central server. In
+    addition to reducing browser to server bandwidth requirements,
+    this may also allow lower latency gaming if game instances are
+    partitioned into clusters based on latency.
 \end{enumerate}
@@
 understandable and practical without sacrificing safety and
-correctness.
+correctness. The Raft algorithm implemented and described in this
+paper is based on Diego Ongaro's dissertation "Consensus: Bridging
+Theory and Practice" \cite{raft_thesis:ongaro14}.
 
 The Raft algorithm breaks the problem of consensus into three core
-sub-problems: leader election, log replication, and safety. A full Raft
-solution must also address: cluster membership changes, log
-compaction, and client interaction. The Raft algorithm implemented and
-discussed in this paper is based on "Consensus: Bridging Theory and
-Practice" by Diego Ongaro (PhD Dissertation). % TODO/CITE
-
-The Raft protocol is based on the concept of a distributed transaction
-log. Entries in the log represent a sequence of changes to an internal
-state machine. If all members of the Raft cluster have the same log
-entries, then their state machine on each node will have exactly the
-same state. Each node of a Raft cluster can be in one of three states:
-follower, candidate, or leader. The responsibility of the leader is to
-accept new transaction log entries and then replicated these entries
-in the same order to all other members of the cluster. The leader is
-the only member of the cluster that may make changes to the
-transaction log. In order to maintain leadership, the leader sends
-heartbeat RPCs to all the followers in the cluster. Followers that
-do not receive a heartbeat RPC within a certain time period become
-candidates and attempt to be elected by the other nodes as the new
-leader of the cluster.
+sub-problems: leader election, log replication, and safety
+\cite[Section~3]{raft_thesis:ongaro14}. A full Raft system must also
+address: cluster membership changes
+\cite[Section~4]{raft_thesis:ongaro14}, log compaction
+\cite[Section~5]{raft_thesis:ongaro14}, and client
+interaction \cite[Section~6]{raft_thesis:ongaro14}. The Raft protocol
+is based on the concept of a distributed transaction log. Entries in
+the log represent a sequence of commands that will be applied to an
+internal state machine. If all members of the Raft cluster have the
+same log entries, then their state machine on each node will have
+exactly the same state. Each node of a Raft cluster can be in one of
+three states: follower, candidate, or leader. The responsibility of
+the leader is to accept new transaction log entries and then
+replicated these entries in the same order to all other members of the
+cluster. The leader is the only member of the cluster that may make
+changes to the transaction log. In order to maintain leadership, the
+leader sends heartbeat RPCs to all the followers in the cluster.
+Followers that do not receive a heartbeat RPC within a certain time
+period become candidates and attempt to be elected by the other nodes
+as the new leader of the cluster \cite[p.~12]{raft_thesis:ongaro14}.
 
@@
 confirm leadership and reset the election timers on all the other
-nodes.
+nodes \cite[Section~3.4]{raft_thesis:ongaro14}.
 
@@
 apply to the state machine. Each node also keep track of the most
-recent log entry which is known to be committed (exists on more than
-half the nodes in the cluster).
+recent log entry which is known to be committed (exists in the log of
+more than half the nodes in the cluster).
 
@@
 empty appendEntries as heartbeat RPCs to all the nodes until it
-receives a new entry in its transaction log from a client.
+receives a new entry in its transaction log from a client
+\cite[Section~3.5]{raft_thesis:ongaro14}.
 
@@
 Raft leaders may not consider entries from a previous term to be
-committed until it has committed at least one an entry from its own
-term.
+committed until it has committed at least one entry from its own
+term
+\cite[Section~3.6]{raft_thesis:ongaro14}.
 
@@
 before sending or receiving any RPCs. The properties that must be
-persisted are current term, current vote for this term (if any), and
+persisted are: current term, current vote for this term (if any), and
 either the full transaction log or the state machine plus any
-unapplied transaction log entries.
+unapplied transaction log entries
+\cite[Section~3.8]{raft_thesis:ongaro14}.
 
@@
 log entries. As soon as the new entry is added to the log it becomes
-effective without waiting for the entry to be fully committed across
-the cluster. However, a new cluster change entry may not be added to
-the log until the previous one is committed. Once the change entry is
-committed, the initiator of the change is notified, removed servers
-can be shut down and another change entry may be added to the log.
+effective on that node without waiting for the entry to be fully
+committed across the cluster. However, a new cluster change entry may
+not be added to the log until the previous one is committed. Once the
+change entry is committed, the initiator of the change is notified,
+removed servers can be shut down and another change entry may be added
+to the log \cite[Section~4]{raft_thesis:ongaro14}.
 
@@
 the current state machine to other nodes if their transaction log is
-too old (e.g. when a new node is added).
+too old (e.g. when a new node is added)
+\cite[Section~5]{raft_thesis:ongaro14}.
 
@@
 entries to the transaction log. Clients first find the address of any
-node in the cluster either via broadcast or via an external directory
-service. Once the client discovers a node of the cluster, that node
-can either forward messages directly to the leader node (if it is not
-the leader) or it can reject client requests with a response
-indicating the address of the leader for the client to redirect to.
-
-The Raft protocol should also provide strict linearizable semantics for
-client commands/requests (reads and writes). In order to accomplish
-this, each client is given a unique ID, and the client assigns
-a unique serial number to each command. This prevents a single client
-command from being applied twice to the state machine (in case of
-a leader crash, or network duplication).
+node in the cluster either via broadcast, via an external directory
+service, or directly provided by the user. Once the client discovers
+a node of the cluster, that node can either forward messages directly
+to the leader node (if it is not the leader) or it can reject client
+requests with a response indicating the address of the leader where
+client requests should be redirected
+\cite[Section~6.2]{raft_thesis:ongaro14}.
+
+The Raft protocol can also provide strict linearizable semantics for
+client commands/requests (reads and writes) for applications that
+require it. In order to accomplish this, each client is given a unique
+ID, and the client assigns a unique serial number to each command.
+This prevents a single client command from being applied twice (leader
+crash, network duplication) or lost (packet loss)
+\cite[Section~6.3]{raft_thesis:ongaro14}.
 
@@
 already committed new entries).
+\cite[Section~6.4]{raft_thesis:ongaro14}.
 
@@
 Raft.js is an implementation of the Raft algorithm in JavaScript that
-was creator by the author of this paper. %TODO/CITE
+was created by the author of this paper. %TODO/CITE
 Raft.js is designed to run in either a browser environment or within
-node.js (server-side JavaScript). Raft.js implements the Raft
-algorithm as described in "Consensus", Ongaro (dissertation).
-% TODO/CITE
+Node.js (server-side JavaScript). Raft.js implements the Raft
+algorithm as described in \cite[Consensus:~Diego]{raft_thesis:ongaro14}.
 
@@
 The implementation of the Raft algorithm is implemented in the
-RaftServerBase class (in base.js). The base class does not directly
-implement stateful functions such as RPC communication, durable
-storage, scheduling/timeouts, or state machine command/management. In
-order to create a working implementation these functions must be
-provided either as configuration parameters when the class is
-instantiated or by sub-classing the class.
-
-The modular design of the Raft.js implementation allows it to be easily
-used in several different contexts. For example, the RaftServerLocal
-class (local.js) implements RPC calls using plain JavaScript function
+\texttt{RaftServerBase} class (in \texttt{base.js}). The base class
+omits the implementation of functions with side-effect such as RPC
+communication, durable storage, scheduling/timeouts, or state machine
+command/management. In order to create a working implementation these
+functions must be provided either as configuration parameters when the
+class is instantiated or by sub-classing the class.
+
+The modular design of the Raft.js implementation allows it to be
+easily used in different contexts. For example, the \texttt{RaftServerLocal}
+class (\texttt{local.js}) implements RPC calls using plain JavaScript function
 calls between different instances of the class (nodes) in the same
-execution context. The online Raft visualization at
-http://kanaka.github.io/raft.js/ uses the RaftServerLocal class and
-instantiates it using a scheduling functions that fire when the user
-clicks on the "Step" button rather than due to the passage of time.
-
-The RaftServerHttp class (`http.js`) embeds a simple web server into
-each Raft node and then uses normal HTTP requests to send RPC messages
-between Raft nodes.
-
-For this paper, the RaftServerLocal class was extended with the
-capability of sending RPCs over the WebRTC Data Channel.
+execution context. For example, the online Raft visualization at
+\href{http://kanaka.github.io/raft.js/}{http://kanaka.github.io/raft.js/}
+uses the \texttt{RaftServerLocal} class and instantiates it using
+a scheduling functions that fire when the user clicks on the "Step"
+button rather than due to the passage of time.
+
+The \texttt{RaftServerHttp} class (\texttt{http.js}) embeds a simple
+web server into each Raft node and then uses normal HTTP requests to
+send RPC messages between Raft nodes.
+
+For this paper, the \texttt{RaftServerLocal} class was extended with
+the capability of sending RPCs over the WebRTC Data Channel.
 
@@
 
-The current Raft.js implementation is based on "Consensus", Ongaro and
-implements most of the Raft algorithm including: leader election, log
-replication, safety, membership changes, basic client interaction, and
-read-only operation optimizations. Log compaction and full client
-linearizability are not currently implemented (clients and client
-transactions are not assigned unique IDs). Log compaction is important
-for long-running production systems in order to limit memory growth
-and allow new nodes to catch up their logs more efficiently. However
-log compaction is not required for verifying overall system
-correctness and does not significantly impact short-running test
-configurations. Full client linearizability is important feature for
-simplifying certain types of distributed applications, however the
-implementation requires some significant additions to the Raft
-implementation (e.g. assigning client IDs and sequence IDs and tracking
-timed-out clients). For the chat application, simply adding one atomic
-check-and-set state machine command is sufficient to achieve
-a linearized chat history. This does impose a slight additional cost
-on the client because it must retry the atomic operations until they
-succeed. The author intends to implement both log compaction and
-full client linearizability in the future but, however these do not
-have a significant impact on the ability to test Raft combined with
-WebRTC.
-
-The Raft protocol description in "Consensus", Ongaro, has undergone
-some revisions since the original draft paper (including incorporation
-of suggestions by the author of this paper). These changes include
+The current Raft.js implementation most of the Raft algorithm
+including: leader election, log replication, safety, membership
+changes, basic client interaction, and read-only operation
+optimizations. % TODO: read-only operations needs mentioning earlier
+Log compaction and full client linearizability are not currently
+implemented (clients and client transactions are not assigned unique
+IDs). Log compaction is important for long-running production systems
+in order to limit memory growth and allow new nodes to catch up their
+logs more efficiently. However log compaction is not required for
+verifying overall system correctness and does not significantly impact
+short-running test configurations. Full client linearizability is
+important feature for simplifying certain types of distributed
+applications, however the implementation of linearizability requires
+some significant additions to the Raft implementation (e.g. assigning
+client IDs and sequence IDs and tracking timed-out clients). For the
+chat application, simply adding a single atomic check-and-set state
+machine command is sufficient to achieve a linearized chat history.
+This does impose a slight additional cost on the client because it
+must retry the atomic operations until they succeed. The author
+intends to implement both log compaction and full client
+linearizability in the future, however these do not have a significant
+impact on the ability to test Raft combined with WebRTC.
+
+The Raft protocol has undergone some revisions since the original
+draft paper \cite{raft_paper:ongaro14} (including incorporation of
+suggestions by the author of this paper). These changes include
 cluster membership simplifications, separating the concepts of
@@
 
-The are also some other differences between Raft.js and the Ongaro
-paper:
-
-\begin{enumerate}
-\item The paper describes the Raft RPCs in terms of the traditional request/response model where requests and responses are automatically correlated together by a lower level of the software stack.
-For example with HTTP 1.0 a separate network session is established for each request/response pair (HTTP 1.0).
-Other examples including using a unique per session ID to correlate request and response.
-WebRTC DataChannel connections are message based rather than request/response based.
-The original Raft definition of RPC and RPC response messages almost contain enough information to function correctly in an environment without builtin request/response correlation.
-Raft.js includes a small modification to the response RPCs for requestVote and appendEntries to add a 'sourceId' field.
-\item The Raft paper is clear that each membership change must be delayed until all the previous membership change is committed (sequential).
-It is implied that the leader node should track all pending membership change requests and apply them one at a time.
-However, in Raft.js, if a client requests a membership change while another change request is pending (uncommitted), the leader will reject the change request with a status of 'PENDING\_CONFIG\_CHANGE'.
-This keeps the core Raft implementation simpler with the trade-off being that the client implementation of membership change requests is slightly more complex.
+The are also some other differences between Raft.js and Raft as
+described in \cite[Consensus:~Diego]{raft_thesis:ongaro14}:
+
+\begin{enumerate}
+\item The paper describes the Raft RPCs in terms of the traditional
+    request/response model where requests and responses are
+    automatically correlated together by a lower level of the software
+    stack. For example with HTTP 1.0 a separate network session is
+    established for each request/response pair (HTTP 1.0). Other
+    examples including using a unique per session ID to correlate
+    request and response.  WebRTC DataChannel connections are message
+    based rather than request/response based. The original Raft
+    definition of RPC and RPC response messages almost contain enough
+    information to function correctly in an environment without
+    builtin request/response correlation. Raft.js includes a small
+    modification to the response RPCs for requestVote and
+    appendEntries to add a 'sourceId' field.
+\item The Raft paper is clear that each membership change must be
+    delayed until all the previous membership change is committed
+    (sequential).  It is implied that the leader node should track all
+    pending membership change requests and apply them one at a time.
+    However, in Raft.js, if a client requests a membership change
+    while another change request is pending (uncommitted), the leader
+    will reject the change request with a status of
+    'PENDING\_CONFIG\_CHANGE'. This keeps the core Raft implementation
+    simpler with the trade-off being that the client implementation of
+    membership change requests is slightly more complex.
 \end{enumerate}
@@
     \item \href{https://tools.ietf.org/html/rfc2327}{RFC2327: SDP: Session Description Protocol}
-    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-jsep-09}{WebRTC: Javascript Session Establishment Protocol (JSEP)}
+    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-jsep-09}{WebRTC: JavaScript Session Establishment Protocol (JSEP)}
     \item \href{https://tools.ietf.org/html/rfc5245}{RFC5245: Interactive Connectivity Establishment (ICE)}

### 2015-09-06 research_class/paper.tex
Proofread and citations up through section 4.
Move WebRTC protocol standards/drafts links to bibliography sections.

@@
 Raft.js is an implementation of the Raft algorithm in JavaScript that
-was created by the author of this paper. %TODO/CITE
-Raft.js is designed to run in either a browser environment or within
-Node.js (server-side JavaScript). Raft.js implements the Raft
-algorithm as described in \cite[Consensus:~Diego]{raft_thesis:ongaro14}.
+was created by the author of this paper \cite{raft.js}.  Raft.js is
+designed to run in either a browser environment or within Node.js
+(server-side JavaScript). Raft.js implements the Raft algorithm as
+described in \cite[Consensus:~Diego]{raft_thesis:ongaro14}.
 
@@
 \href{http://kanaka.github.io/raft.js/}{http://kanaka.github.io/raft.js/}
-uses the \texttt{RaftServerLocal} class and instantiates it using
-a scheduling functions that fire when the user clicks on the "Step"
-button rather than due to the passage of time.
+\cite{raft.js:visualization} uses the \texttt{RaftServerLocal} class
+and instantiates it using a scheduling function that executes one
+scheduled event each time the user clicks on the "Step" button rather
+than due to the passage of time.
 
@@
 \begin{enumerate}
-\item The paper describes the Raft RPCs in terms of the traditional
-    request/response model where requests and responses are
+\item The paper describes the Raft RPCs in terms of synchronous stream
+    based request/response model where requests and responses are
     automatically correlated together by a lower level of the software
-    stack. For example with HTTP 1.0 a separate network session is
-    established for each request/response pair (HTTP 1.0). Other
-    examples including using a unique per session ID to correlate
-    request and response.  WebRTC DataChannel connections are message
-    based rather than request/response based. The original Raft
-    definition of RPC and RPC response messages almost contain enough
-    information to function correctly in an environment without
-    builtin request/response correlation. Raft.js includes a small
-    modification to the response RPCs for requestVote and
-    appendEntries to add a 'sourceId' field.
+    stack. For example with HTTP 1.0 \cite{http:rfc1945} client
+    requests are followed by a server response on any given TCP
+    stream. Most remote procedure call mechanims also use
+    a synchronous request/response model. %TODO/CITE
+    WebRTC DataChannel connections are message based rather than
+    request/response stream based. %TODO/CITE
+    The original Raft definition of RPC request and response messages
+    almost contain enough information to function correctly in an
+    environment without builtin request/response correlation such as
+    UDP or SCTP when used with reliable transport disabled. %TODO/CITE
+    Raft.js includes a small modification to the response RPCs for
+    requestVote and appendEntries to add a 'sourceId' field that
+    enables the protocol to function with message based transports.
 \item The Raft paper is clear that each membership change must be
     delayed until all the previous membership change is committed
-    (sequential).  It is implied that the leader node should track all
-    pending membership change requests and apply them one at a time.
-    However, in Raft.js, if a client requests a membership change
-    while another change request is pending (uncommitted), the leader
-    will reject the change request with a status of
-    'PENDING\_CONFIG\_CHANGE'. This keeps the core Raft implementation
-    simpler with the trade-off being that the client implementation of
-    membership change requests is slightly more complex.
+    (sequential) \cite[Section~4.1]{raft_thesis:ongaro14}. It is
+    implied that the leader node should track all pending membership
+    change requests and apply them one at a time.  However, in
+    Raft.js, if a client requests a membership change while another
+    change request is pending (uncommitted), the leader will reject
+    the change request with a status of 'PENDING\_CONFIG\_CHANGE'.
+    This keeps the core Raft implementation simpler with the trade-off
+    being that the client implementation of membership change requests
+    is slightly more complex due to retry logic.
 \end{enumerate}
@@
 
-WebRTC is name of a collection of browser APIs and network protocols
-that are currently being standardized the W3C and IETF respectively.
-Collectively, these APIs and protocols work together to enable
-real-time peer-to-peer video, audio and data communication between
-browsers.
+WebRTC is a collection of browser APIs \cite{webrtc:feb2015} and
+network protocols \cite{draft-rtcweb-overview} that are currently
+being standardized the W3C and IETF respectively. Collectively these
+APIs and protocols enable real-time peer-to-peer video, audio and data
+communication between browsers.
 
@@
 and dynamic, inbound connections are denied by default, and outbound
-connections are network address translated (NATed).
+connections are network address translated (NATed)
+\cite[Section~3.4]{draft-rtcweb-transports}.
 
 To establish a direct communication channel between browsers, WebRTC
-uses mechanism that is similar to the Session Initiation Protocol
-(SIP) commonly used for Voice-over-IP (VoIP) communication. This
+uses mechanism called Interactive Connectivity Establishment (ICE)
+that is similar to the Session Initiation Protocol (SIP) commonly used
+for Voice-over-IP (VoIP) communication \cite{RFC5245:ice}. This
 involves a third party server known as a signaling server. Each user
@@
 signaling server is also uses to communicate session control messages
-(open, close, error) and to negotiate media capabilities. The signaling
-server may optionally provide other services such as higher level
-session management, HTTP serving of the web application, address book
-service, client presence/availability, etc.
+(open, close, error) and to negotiate media capabilities. %TODO/CITE
+The signaling server may optionally provide other services such as
+higher level session management, HTTP serving of the web application,
+address book service, client presence/availability, etc. %TODO/CITE
 
@@
 to establish browser-to-browser network connectivity in order to
-stream real-time video, audio and data communication. These APIs are
-currently in the process of being standardized by the W3C organization
-(http://www.w3.org/TR/webrtc/) % TODO/CITE
-
-There are three primary WebRTC API definitions:
-
-\begin{enumerate}
-\item \href{http://www.w3.org/TR/webrtc/\#rtcpeerconnection-interface}{RTCPeerConnection}: The RTCPeerConnection API represents a direct connection between two browsers.
-Each RTCPeerConnection object has an associated Interactive Connectivity Establishment (ICE) agent that is used to establish network connectivity through firewalls and NAT services.
-
-\item \href{http://www.w3.org/TR/webrtc/\#media-stream-api-extensions-for-network-use}{MediaStream (aka getUserMedia}: The MediaStream API represents a stream of audio or video data.
-MediaStreams can easily be sent and received over RTCPeerConnections.
-The MediaStream API definition provides a mechanism that allows JavaScript to enumerate the native MediaStreams that are provided by the browser such as a local web camera or a live video stream of the desktop (for desktop sharing).
-Although support for MediaStreams was the original impetus for the creation of the WebRTC APIs and protocols, the Raft over WebRTC implementation does not make use of the MediaStream API.
-
-\item \href{http://www.w3.org/TR/webrtc/\#rtcdatachannel}{RTCDataChannel}: The RTCDataChannel API provides a mechanism for sending and receiving generic data.
-A single RTCPeerConnection can contain multiple RTCDataChannels.
-The RTCDataChannel API is modeled after the WebSocket API, however, unlike the WebSocket API which is reliable and guaranteed order transport (like TCP), each RTCDataChannel can be configured with different reliability and ordering settings.
-For example, by relaxing the reliability and ordering constraints on a RTCDataChannel, the average latency and jitter per message can be decreased even though some messages may be dropped or arrive out of order.
-The Raft protocol is specifically designed to support dropped or out-of-order messages so the transport options available with RTCDataChannel can be used to gain extra efficiency.
+perform real-time video, audio and data communication. These APIs are
+currently in the process of being standardized by the \emph{Web
+Real-Time Communications Working Group} of the World Wide Web
+Consortium (W3C) \cite{webrtc:wg}.
+
+There are three primary WebRTC interfaces:
+
+\begin{enumerate}
+\item \emph{RTCPeerConnection} \cite{webrtc:rtcpeerconnection:feb2015}:
+    The RTCPeerConnection API represents a direct connection between
+    two browsers. Each RTCPeerConnection object has an associated
+    Interactive Connectivity Establishment (ICE) agent that is used to
+    establish network connectivity through firewalls and NAT services.
+
+\item \emph{MediaStream~(aka~getUserMedia)} \cite{webrtc:media-stream:feb2015}:
+    The MediaStream API represents a stream of audio or video data.
+    MediaStream data can easily be sent and received over
+    RTCPeerConnections. The MediaStream API definition provides
+    a mechanism that allows JavaScript to enumerate the native
+    MediaStreams that are provided by the browser such as a local web
+    camera or a live video stream of the desktop (for desktop
+    sharing).  Although support for MediaStreams was the original
+    impetus for the creation of the WebRTC APIs and protocols, the
+    Raft over WebRTC implementation does not make use of the
+    MediaStream API.
+
+\item
+    \emph{RTCDataChannel} \cite{webrtc:datachannel:feb2015}:
+    The RTCDataChannel API provides a mechanism for sending and
+    receiving generic data between User Agents (browsers). A single
+    RTCPeerConnection can contain multiple RTCDataChannels. The
+    RTCDataChannel API is modeled after the WebSocket API, however,
+    unlike the WebSocket API which is reliable and guaranteed order
+    transport (like TCP), each RTCDataChannel can be configured with
+    different reliability and ordering settings. For example, by
+    relaxing the reliability and ordering constraints on
+    a RTCDataChannel, the average latency and jitter per message can
+    be decreased even though some messages may be dropped or arrive
+    out of order. The Raft protocol is specifically designed to
+    support dropped or out-of-order messages so the transport options
+    available with RTCDataChannel can be used to gain extra
+    efficiency.
+    %TODO/CITE: more of them above
 \end{enumerate}
@@
 protocol is applicable and is referenced or extended as part of the
-WebRTC suite.
-
-Describing all the protocols that make up the WebRTC suite is beyond
-the scope of the this document. However, here is a (incomplete) list
-of WebRTC protocols drafts and related protocols that are
-part of WebRTC:
-
-\begin{enumerate}
-\item General:
-    \begin{enumerate}
-    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-overview-13}{WebRTC: Overview}
-    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-security-arch-11}{WebRTC: Security Architecture}
-    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-security-08}{WebRTC: Security Considerations}
-    \end{enumerate}
-\item Data transport:
-    \begin{enumerate}
-    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-transports-08}{WebRTC: Data transport overview}
-    \item \href{https://tools.ietf.org/html/draft-ietf-tsvwg-rtcweb-qos-03}{WebRTC: DSCP and other packet markings for RTCWeb QoS}
-    \item \href{https://tools.ietf.org/html/draft-ietf-tsvwg-sctp-dtls-encaps-09}{DTLS Encapsulation of SCTP Packets}
-    \item \href{https://tools.ietf.org/html/draft-ietf-mmusic-sctp-sdp-14}{Stream Control Transmission Protocol (SCTP)-Based Media Transport in the Session Description Protocol (SDP)}
-    \item \href{https://tools.ietf.org/html/draft-ietf-tsvwg-sctp-ndata-02}{Stream Schedulers and a New Data Chunk for the Stream Control Transmission Protocol}
-    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-rtp-usage-23}{WebRTC: Media Transport and Use of RTP}
-    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-data-channel-13}{WebRTC: Data Channels}
-    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-data-protocol-09}{WebRTC: Data Channel Establishment Protocol}
-    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-alpn-01}{WebRTC: Application Layer Protocol Negotiation for Web Real-Time Communications (WebRTC)}
-    \item \href{https://tools.ietf.org/html/rfc5245}{RFC5245: Interactive Connectivity Establishment (ICE)}
-    \item \href{https://tools.ietf.org/html/rfc5128}{RFC5128: Peer-to-Peer (P2P) Communication across Network Address Translators (NATs)}
-    \item \href{https://tools.ietf.org/html/rfc5389}{RFC5389: Session Traversal Utilities for NAT (STUN)}
-    \item \href{https://tools.ietf.org/html/rfc3489}{RFC3489: STUN - Simple Traversal of User Datagram Protocol (UDP) Through Network Address Translators (NATs)}
-    \item \href{https://tools.ietf.org/html/rfc5766}{RFC5766: Traversal Using Relays around NAT (TURN)}
-    \item \href{https://tools.ietf.org/html/rfc6156}{RFC6156: Traversal Using Relays around NAT (TURN) Extension for IPv6}
-    \item \href{https://tools.ietf.org/html/rfc6544}{RFC6544: TCP Candidates with Interactive Connectivity  Establishment (ICE)}
-    \item \href{https://tools.ietf.org/html/RFC4571}{RFC4571: Framing Real-time Transport Protocol (RTP) and RTP Control Protocol (RTCP) Packets over Connection-Oriented Transport}
-    \item \href{https://tools.ietf.org/html/rfc5764}{RFC5764: Datagram Transport Layer Security (DTLS) Extension to Establish Keys for the Secure Real-time Transport Protocol (SRTP)}
-    \end{enumerate}
-
-\item Data framing and security
-    \begin{enumerate}
-    \item \href{https://tools.ietf.org/html/rfc3550}{RFC3550: Real-time Transport Protocol (RTP)}
-    \item \href{https://tools.ietf.org/html/rfc3711}{RFC3711: Secure Real-time Transport Protocol (SRTP)}
-    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-rtp-usage-23}{WebRTC: Media Transport and Use of RTP}
-    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-data-channel-13}{WebRTC: Data Channels}
-    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-data-protocol-09}{WebRTC: Data Channel Establishment Protocol}
-    \end{enumerate}
-\item Data formats (audio)
-    \begin{enumerate}
-    \item \href{https://tools.ietf.org/html/draft-cbran-rtcweb-codec-02}{WebRTC: WebRTC Codec and Media Processing Requirements}
-    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-audio-07}{WebRTC: Audio Codec and Processing Requirements}
-    \item \href{https://tools.ietf.org/html/rfc6716}{RFC6716: Opus Audio Codec}
-    \item \href{https://tools.ietf.org/html/draft-ietf-payload-rtp-opus-11}{RTP Payload Format for Opus Speech and Audio Codec}
-    \item \href{https://tools.ietf.org/html/rfc3551}{RFC3551: RTP Profile for Audio and Video Conference}
-    \item \href{https://tools.ietf.org/html/rfc4733}{RFC4733: RTP Payload for DTMF Digits, Telephony Tones, and Telephony Signals}
-    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-audio-codecs-for-interop-01}{WebRTC: Additional WebRTC audio codecs for interoperability with legacy networks}
-    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-video-05}{WebRTC: Video Processing and Codec Requirements}
-    \item \href{https://tools.ietf.org/html/rfc6386}{RFC6386: VP8 Data Format and Decoding Guide}
-    \item \href{http://www.itu.int/rec/T-REC-H.264}{ITU H.264: Advanced video coding for generic audiovisual services}
-    \item \href{https://tools.ietf.org/html/draft-ietf-payload-vp8-15}{RTP Payload Format for VP8 Video}
-    \item \href{https://tools.ietf.org/html/rfc6236}{RFC6236: Negotiation of Generic Image Attributes in the Session Description Protocol (SDP)}
-    \end{enumerate}
-
-\item Connection management
-    \begin{enumerate}
-    \item \href{https://tools.ietf.org/html/rfc3264}{RFC3264: An Offer/Answer Model with the Session Description Protocol (SDP)}
-    \item \href{https://tools.ietf.org/html/rfc2327}{RFC2327: SDP: Session Description Protocol}
-    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-jsep-09}{WebRTC: JavaScript Session Establishment Protocol (JSEP)}
-    \item \href{https://tools.ietf.org/html/rfc5245}{RFC5245: Interactive Connectivity Establishment (ICE)}
-    \item \href{https://tools.ietf.org/html/draft-ietf-mmusic-trickle-ice-02}{RFC5763: Trickle ICE: Incremental Provisioning of Candidates for the Interactive Connectivity Establishment (ICE) Protocol}
-    \end{enumerate}
-\end{enumerate}
-
-Note: the signaling transport itself is not defined as part of WebRTC.
-This is up to the application designer.
+WebRTC suite. %TODO/CITE
+
+The protocols that make up the WebRTC suite are categorized into
+five general functionality groups:
+
+\begin{enumerate}
+    \item data transport
+    \item data framing and security
+    \item connection management
+    \item data formats
+    \item presentation and control
+    \item and local system support functions
+\end{enumerate}
+
+The last three functionality groups focus on interactive audio and video
+applications and are less relevant to Raft over WebRTC. The WebRTC
+section of the bibliography %TODO: reference link
+has a (incomplete) list of WebRTC specific protocols drafts and other
+related protocol standards for the first three functionality groups
+(data transport, data framing and security, and connection
+management).
+
+The signaling transport itself is not defined as part of WebRTC.
+This is up to the application designer. The JavaScript interface used
+for establishing the initial connection (signaling) in the WebRTC 1.0
+standard is somewhat complicated. A competing standard called Object
+Real-time Communication (ORTC) is working to define a simpler and more
+modular JavaScript API for WebRTC signaling \cite{ortc2014}. However,
+at the time of this paper, the ORTC interface was not yet widely
+supported and still in a state of flux. The use of PeerJS provides an
+abstraction that is likely to transparently support ORTC in the future.
 
@@
 also provides a simpler interface for using the WebRTC APIs.
-
-The JavaScript interface use for establishing the initial connection
-(signaling) in the WebRTC 1.0 standard is somewhat complicated.
-A competing standard called Object Real-time Communication (ORTC) is
-working to define a simpler and more modular JavaScript API for WebRTC
-signaling. However, at the time of this work, the ORTC interface was
-not yet widely supported and still in a state of flux.  The use of
-PeerJS provides a suitable abstraction in the meantime (and will
-likely be modified to support ORTC).
 
@@
 PeerServer called ExpressPeerServer that combines a PeerJS signaling
-server with the Node.js based Express web application framework.
+server with the popular Node.js ExpressJS web application framework.
 % (TODO/CITE)
@@
 finalized and so different browsers and browser releases may support
-different versions of the WebRTC draft APIs. The PeerJS client library
-abstracts over these differences and provides a common interface so
-that the application developer does not have to deal directly with
-browser differences.
+different versions of the WebRTC draft APIs. %TODO/CITE
+The PeerJS client library abstracts over these differences and
+provides a common interface so that the application developer does not
+have to deal directly with browser differences.
 
@@
 and protocol is not defined as part of the WebRTC standardization
-effort. The PeerJS organization provides a cloud-hosted version of
-a PeerJS compatible signaling server, however an application developer
-can also run their own PeerJS signaling service using the PeerServer
-server described above.
+effort. %TODO/CITE
+The PeerJS organization provides a cloud-hosted version of a PeerJS
+compatible signaling server, however an application developer can also
+run their own signaling service using PeerJS PeerServer.
 
@@
 
-% TODO: references
-
 \end{multicols}
 
-\printbibliography
-%\bibliographystyle{plain}
-%\bibliography{references/raft,references/other}
+\printbibliography[title={General references},notkeyword=rtc-general,notkeyword=rtc-transport,notkeyword=rtc-framing]
+
+\nocite{draft-rtcweb-overview}
+\nocite{draft-rtcweb-security-arch}
+\nocite{draft-rtcweb-security}
+\printbibliography[title={WebRTC general references},keyword=rtc-general]
+
+\nocite{draft-rtcweb-transports}
+\nocite{draft-rtcweb-qos}
+\nocite{RFC4960:sctp}
+\nocite{RFC3758:sctp-reliability}
+\nocite{draft-sctp-dtls-encaps}
+\nocite{draft-mmusic-sctp-sdp}
+\nocite{draft-sctp-ndata}
+\nocite{draft-rtcweb-rtp-usage}
+\nocite{draft-rtcweb-data-channel}
+\nocite{draft-rtcweb-data-protocol}
+\nocite{draft-rtcweb-alpn}
+\nocite{RFC5245:ice}
+\nocite{RFC5128:p2p-nat}
+\nocite{RFC5389:stun}
+\nocite{RFC3489:stun-udp}
+\nocite{RFC5766:turn}
+\nocite{RFC6156:turn-ipv6}
+\nocite{RFC6544:tcp-ice}
+\nocite{RFC4571:rtp-rtcp}
+\nocite{RFC5764:dtls-srtp}
+\printbibliography[title={WebRTC data transport},keyword=rtc-transport]
+
+\nocite{RFC3550:rtp}
+\nocite{RFC3711:srtp}
+%\nocite{draft-rtcweb-rtp-usage}
+%\nocite{draft-rtcweb-data-channel}
+%\nocite{draft-rtcweb-data-protocol}
+\printbibliography[title={WebRTC framing and security},keyword=rtc-framing]
+
+\nocite{RFC2327:sdp}
+\nocite{RFC3264:sdp-offer-answer}
+\nocite{draft-rtcweb-jsep}
+%\nocite{RFC5245:ice}
+\nocite{draft-mmusic-trickle-ice}
+\nocite{RFC5763:srtp-using-dtls}
+\printbibliography[title={WebRTC connection management},keyword=rtc-connection]
+
+%%%
+
+%\begin{enumerate}
+%\item Data formats (audio)
+%    \begin{enumerate}
+%    \item \href{https://tools.ietf.org/html/draft-cbran-rtcweb-codec-02}{WebRTC: WebRTC Codec and Media Processing Requirements}
+%    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-audio-07}{WebRTC: Audio Codec and Processing Requirements}
+%    \item \href{https://tools.ietf.org/html/rfc6716}{RFC6716: Opus Audio Codec}
+%    \item \href{https://tools.ietf.org/html/draft-ietf-payload-rtp-opus-11}{RTP Payload Format for Opus Speech and Audio Codec}
+%    \item \href{https://tools.ietf.org/html/rfc3551}{RFC3551: RTP Profile for Audio and Video Conference}
+%    \item \href{https://tools.ietf.org/html/rfc4733}{RFC4733: RTP Payload for DTMF Digits, Telephony Tones, and Telephony Signals}
+%    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-audio-codecs-for-interop-01}{WebRTC: Additional WebRTC audio codecs for interoperability with legacy networks}
+%    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-video-05}{WebRTC: Video Processing and Codec Requirements}
+%    \item \href{https://tools.ietf.org/html/rfc6386}{RFC6386: VP8 Data Format and Decoding Guide}
+%    \item \href{http://www.itu.int/rec/T-REC-H.264}{ITU H.264: Advanced video coding for generic audiovisual services}
+%    \item \href{https://tools.ietf.org/html/draft-ietf-payload-vp8-15}{RTP Payload Format for VP8 Video}
+%    \item \href{https://tools.ietf.org/html/rfc6236}{RFC6236: Negotiation of Generic Image Attributes in the Session Description Protocol (SDP)}
+%    \end{enumerate}
+%\end{enumerate}
 

### 2015-09-08 research_class/paper.tex
Fix bib month data. Figure placement.
Proofread and citations. Update seq diagram.

@@
 \cite[Section~6.4]{raft_thesis:ongaro14}.
+
+\end{multicols}
+
+\begin{figure*}[h]
+\centerline{\includegraphics[width=0.8\textwidth]{imgs/raft_rtc_architecture}}
+\caption{Raft over RTC Architecture}
+  \label{fig:raft_rtc_architecture}
+\end{figure*}
+
+\begin{multicols}{2}
 
@@
 
-\section {Design and Implementation}
-
 \end{multicols}
-
-\begin{figure*}[h]
-\centerline{\includegraphics[width=0.8\textwidth]{imgs/raft_rtc_architecture}}
-\caption{Raft over RTC Architecture}
-  \label{fig:raft_rtc_architecture}
-\end{figure*}
-
-%\begin{multicols}{2}
-%\end{multicols}
 
@@
 
+\section {Design and Implementation}
+
 \subsection{Incremental Steps}
@@
 \begin{enumerate}
-\item Simple WebRTC messaging: implement a small JavaScript application that uses the PeerJS client library and an unmodified PeerJS signaling server.
-Demonstrate communication over WebRTC DataChannel using two browser instances (or browser tabs).
-
-\item Raft.js over WebRTC: combine the existing Raft.js implementation with the small JavaScript application above to demonstrate Raft over WebRTC.
-This step uses a statically configured three node cluster (three browser tabs).
-The PeerJS signaling server is polled to retrieve the list of other clients and the Raft cluster is started when the number of PeerJS clients reaches three.
-
-\item Update Raft.js protocol: update the Raft.js implementation to match the latest description of the Raft protocol (as described in "Consensus", Ongaro dissertation).
-This includes the implementation of the simplified membership change algorithm (sequential changes) and the separation of the log commit and state machine apply concepts.
-Additionally, the new Raft.js implementation always starts a new Raft cluster as a cluster with a single node and additional nodes are dynamically added to the cluster (as suggested in section 4.4 of the paper).
-
-\item Dynamic Raft.js over WebRTC: enhance the signaling server with support for multiple sessions/channels to enable separate Raft clusters with the same signaling server.
-Start each cluster with a single node and automatically add and remove Raft nodes as they come and go.
-
-\item Visual representation: display Raft node state and RPC statistics in addition to the internal Raft log messages.
-
-\item Sessionless/Datagram RPC messages: change Raft RPC requests and response messages so that RPC responses have enough context to be handled separately from RPC requests.
-Specifically this enables Raft to support datagram (sessionless) network transports like WebRTC.
-
-\item Atomic state machine operations: implement a state machine update operation that only succeeds if no other operation has been applied to the same key since the last time this client read from that key.
-
-\item Automatic client request forwarding: extend the client request RPC to also be sessionless/datagram based.
-Implement an asynchronous JavaScript function that determines which node is the leader and forwards client request RPCs to the leader.
-This function must also correlate responses to the client request and invoke the callback for the original request.
-
-\item Chat application: add the ability to send a message to the chat log using the client request forwarding functionality together with atomic state machine operation.
-Show the current chat history based on the chat entries in the state machine.
-
-\item Server as Raft.js peer (aspirational): enable server nodes (Raft.js running on Node.js) to fully participate as Raft cluster nodes.
+\item Simple WebRTC messaging: implement a small JavaScript
+    application that uses the PeerJS client library and an unmodified
+    PeerJS signaling server. Demonstrate communication over WebRTC
+    DataChannel using two browser instances (or browser tabs).
+\item Raft.js over WebRTC: combine the existing Raft.js implementation
+    with the small JavaScript application above to demonstrate Raft
+    over WebRTC. This step uses a statically configured three node
+    cluster (three browser tabs). The PeerJS signaling server is
+    polled to retrieve the list of other clients and the Raft cluster
+    is started when the number of PeerJS clients reaches three.
+\item Update Raft.js protocol: update the Raft.js implementation to
+    match the latest description of the Raft protocol
+    \cite{raft_thesis:ongaro14}. This includes the implementation of
+    the simplified membership change algorithm (sequential changes)
+    and the separation of the log commit and state machine apply
+    concepts.  Additionally, the new Raft.js implementation always
+    starts a new Raft cluster as a cluster with a single node and
+    additional nodes are dynamically added to the cluster
+    \cite[Section~4.4]{raft_thesis:ongaro14}.
+\item Dynamic Raft.js over WebRTC: enhance the signaling server with
+    support for multiple sessions/channels to enable separate Raft
+    clusters with the same signaling server. Start each cluster with
+    a single node and automatically add and remove Raft nodes as they
+    come and go.
+\item Visual representation: display Raft node state and RPC
+    statistics in addition to the internal Raft logging messages.
+\item Sessionless/Datagram RPC messages: change Raft RPC requests and
+    response messages so that RPC responses have enough context to be
+    handled separately from RPC requests. Specifically this enables
+    Raft to support datagram (sessionless) network transports like
+    WebRTC.
+\item Atomic state machine operation: implement a state machine
+    update operation that only succeeds if no other operation has been
+    applied to the same key since the last time this client read from
+    that key.
+\item Automatic client request forwarding: extend the client request
+    RPC to also be sessionless/datagram based. Implement an
+    asynchronous JavaScript function that determines which node is the
+    leader and forwards client request RPCs to the leader. This
+    function must also correlate RPC responses to the client request
+    and invoke the callback for the original request.
+\item Chat application: add the ability to send a chat message to the
+    log using the client request forwarding functionality together
+    with atomic state machine operation. Show the current chat history
+    based on the chat entries in the state machine.
+\item Server as Raft.js peer (aspirational): enable server nodes
+    (Raft.js running on Node.js) to fully participate as Raft cluster
+    nodes.
 \end{enumerate}
@@
 aspirational goal because this functionality is not yet supported in
-PeerJS although it is being tracked by issue report:
-https://github.com/peers/peerjs/issues/103
+PeerJS \cite{peerjs:103}.
 
@@
 \begin{enumerate}
-\item Static web service endpoint: this serves the static HTML, CSS and JavaScript that make up the Raft over WebRTC web application.
-
-\item New channel endpoint: when a browser loads this endpoint, a new PeerJS channel is created and the browser is redirected to a URL for loading the main application (via static file service above).
-The redirect URL has a query string appended that contains the channel ID.
-In addition the redirect URL has a fragment identifier (hash/history) that communicates to the web application that this is the first server in the channel.
-
-\item Peer list endpoint: this endpoint takes a channel ID and returns the list of WebRTC peers (browser user agents) that are currently registered in that channel.
-
-Refer to the "4.3 Client" section for a description of how these
+\item Static web service endpoint: this serves the static HTML, CSS
+    and JavaScript that make up the Raft over WebRTC web application.
+\item New channel endpoint: when a browser loads this endpoint, a new
+    PeerJS channel is created and the browser is redirected to a URL
+    for loading the main application (via static file service above).
+    The redirect URL has a query string appended that contains the
+    channel ID of the new channel that was just created. In addition
+    the redirect URL has a fragment identifier (hash/history) that
+    communicates to the web application that this is the first server
+    in the channel.
+\item Peer list endpoint: this endpoint takes a channel ID and returns
+    the list of WebRTC peers (browser user agents) that are currently
+    connected to that channel.
+
+\end{enumerate}
+
+Refer to Section~\ref{section:design-client} and
+Figure~\ref{fig:raft_rtc_sequence} for a description of how these
 endpoints interact.
-\end{enumerate}
 
 \subsection{Client}
+    \label{section:design-client}
 
 The Raft over WebRTC testing application is designed as a single web
-page (`rtc.html`). The application has different starting behavior
-depending on the value of the URL query string (the URL component
-following the "?" and up to the "\#") and the fragment identifier (the
-URL component after the "\#").
-
-When a new Raft cluster is created, the browser (user agent) connects
-to the server new channel endpoint. This endpoint causes the server to
-create a new channel for PeerJS signaling. Once this is complete the
-server responds to the browser with a redirect message that contains
-the new channel identifier in the query parameter and an indication
-that this is the very first Raft cluster node in the fragment
-identifier.
+page (\texttt{rtc.html} or \texttt{chat.html}). The application has
+different starting behavior depending on the value of the URL query
+string (the URL component following the "?" and up to the "\#") and
+the fragment identifier (the URL component after the "\#").
+
+When a new Raft cluster is created, the user agent (browser) connects
+to the server's new channel endpoint
+(Figure~\ref{fig:raft_rtc_sequence}~[1]). This endpoint causes the
+server to create a new channel for PeerJS signaling
+(Figure~\ref{fig:raft_rtc_sequence}~[2]). Once this is complete the
+server responds to the browser with a redirect message
+that contains the new channel identifier in the query parameter and an
+indication that this is the very first Raft cluster node in the
+fragment identifier (Figure~\ref{fig:raft_rtc_sequence}~[3]).
 
 The browser follows the redirect and loads the static web application
-at the static file endpoint provided by the server. This includes the
-main application page (`rtc.html`), the core raft algorithm
-(`base.js` and `local.js`), the PeerJS client library (`peer.js`), the
-actual Raft over WebRTC implementation (`rtc.js`), and a layout
-stylesheet (`web/demo.css`).
+at the static file endpoint provided by the server
+(Figure~\ref{fig:raft_rtc_sequence}~[4]). This includes the
+main application page (\texttt{rtc.html} or \texttt{chat.html}), the
+core raft algorithm (\texttt{base.js} and \texttt{local.js}), the
+PeerJS client library (\texttt{peer.js}), the actual Raft over WebRTC
+implementation (\texttt{rtc.js}), a layout stylesheet
+(\texttt{web/demo.css}), and higher-level application code if that is
+being loaded (\texttt{chat.js})
+(Figure~\ref{fig:raft_rtc_sequence}~[5]).
 
@@
 browser context (separate browser, browser window, or browser tab) and
-loading the same URL (which no longer has the "firstServer" fragment
-identifier). There is a convenience link provided in the application
-page that opens a new window with a new cluster node.
-
-When the page first loads, a PeerJS signaling connection is
-established with the server. Once this connection is established, the
-application will be notified when any PeerJS clients connect to or
-disconnect from the server. After the PeerJS signaling connection is
-established, the client immediately queries the Peer list endpoint to
-discover any clients that were connected prior to when this node
-connected to the server.
+loading the same URL without the "firstServer" fragment identifier
+(Figure~\ref{fig:raft_rtc_sequence}~[9]). There is a convenience link
+provided in the application page that opens a new browser window with
+a new cluster node (see \ref{fig:browser_1}).
+
+When each page first loads, a PeerJS signaling connection is
+established with the server
+(Figure~\ref{fig:raft_rtc_sequence}~[6]~and~[11]). Once this
+connection is established, the application will be notified when any
+PeerJS clients connect to or disconnect from the server.  After the
+PeerJS signaling connection is established, the client immediately
+queries the Peer list endpoint
+(Figure~\ref{fig:raft_rtc_sequence}~[7]~and~[12]~and~[14]]) to discover
+any clients that were connected prior to when this node connected to
+the server (Figure~\ref{fig:raft_rtc_sequence}~[8]~and~[13]~and~[15]]).
 
 The web application also begins a periodic async polling function
-(`addRemoveServersAsync`) when the page loads. When the node is a Raft
+(\texttt{addRemoveServersAsync}) when the page loads. When the node is a Raft
 leader, the function compares the current active list of PeerJS clients to
-the Raft cluster node list. If there are any differences between the
-two lists, then any new clients are added to the Raft cluster and
+the Raft cluster membership list. If there are any differences between
+the two lists, then any new clients are added to the Raft cluster and
 missing clients are removed from the Raft cluster one at a time using
-the `addServer` and `removeServer` RPC calls. Servers additions are
-prioritized before removals (as recommended in Ongaro 4.4) % TODO/CITE
-in order to maximize availability.
+the \texttt{addServer} and \texttt{removeServer} RPC calls. Servers
+additions are prioritized before removals 
+in order to maximize availability
+\cite[Section~4.4]{raft_thesis:ongaro14}.
 
 When an instance of the web application starts that is not
-a "firstServer", the normal Raft timers are not automatically enabled
-until the server receives an RPC. The prevents the server from
-possibly becoming a leader before it is incorporated into the cluster
-by an appendEntries RPC from the "true" leader.
+a "firstServer", the normal Raft timers are not enabled until the
+server receives its first RPC. The prevents the server from
+trying to becoming a leader before it is incorporated into the cluster
+by an appendEntries RPC from the true cluster leader.
 
@@
 \item \emph{Term}: the current Raft leadership term.
-\item \emph{State}: the current Raft node state (follower, candidate, or leader)
-\item \emph{Cluster Size}: a count of the current Raft cluster membership (number of nodes currently participating in the cluster)
+\item \emph{State}: the current Raft node state (follower, candidate,
+    or leader)
+\item \emph{Cluster Size}: a count of the current Raft cluster
+    membership (number of nodes currently participating in the
+    cluster)
 \item \emph{Log Length}: number of entries in the transaction log.
-\item \emph{Request Votes Sent/Received}: the number of `requestVote` RPCs that have been sent from and received by this node.
-\item \emph{Append Entries Sent/Received}: the number of `appendEntries` RPCs that have been sent from and received by this node.
-\end{enumerate}
-
-\subsubsection{Client Request Redirect Implementation}
-
-In the Raft over WebRTC implementation each local browser is a client
+\item \emph{Request Votes Sent/Received}: the number of `requestVote`
+    RPCs that have been sent from and received by this node.
+\item \emph{Append Entries Sent/Received}: the number of
+    `appendEntries` RPCs that have been sent from and received by this
+    node.
+\end{enumerate}
+
+\subsubsection{Client Request Redirects}
+
+% TODO: rename clientRequest* function to not clash with RPC names
+% (maybe raftRequest*?)
+
+In the Raft over WebRqTC implementation each local browser is a client
 of the Raft cluster. However in Raft, the leader node is the only node
 that adds entries to the log so clients on follower nodes must have
-a way to communicate with the leader. This is implemented as pair of
-functions named clientRequest and clientRequestResponse. clientRequest
-is called by the client with the arguments of a client request RPC and
+a way to communicate with the leader. This is implemented as a pair of
+functions named \texttt{clientRequest} and
+\texttt{clientRequestResponse}. The \texttt{clientRequest} function is
+called by the client with the arguments of a client request RPC and
 a callback function to be called once the request is completed. The
-clientRequest/clientRequestResponse functions work as follows:
-
-\begin{enumerate}
-\item the clientRequestResponse is registered with the local Raft.js
-    implementation as the callback function anytime
-    clientRequestResponse RPCs are received.
-\item the clientRequest function keeps track of the current leader ID.
-    If the current leader ID is not set or this node is the leader
-    then the client's request is sent directly to the local Raft node.
-    If the current leader ID is set then the request is sent to that
-    node as a normal RPC request over the WebRTC channel.
-\item when the clientRequestResponse function is called and the status
-    is "NOT\_LEADER" then the current leader ID being tracked by the
-    clientRequest function is set to the leader ID returned in the
-    response. Then the original request is re-issued over the normal
-    WebRTC channel to the leader.
-\item when the clientRequestResponse function is called with
+\texttt{clientRequest}/\texttt{clientRequestResponse} functions work
+as follows:
+
+\begin{enumerate}
+\item the \texttt{clientRequestResponse} is registered with the local
+    Raft.js implementation as the callback function anytime
+    a \texttt{clientRequestResponse} RPC is received.
+\item the \texttt{clientRequest} function keeps track of the current
+    leader ID.  If the current leader ID is not set or this node is
+    the leader then the client request RPC is sent directly to the
+    local Raft node bypassing the network transport layer.  If the
+    current leader ID is set then the request is sent to that node as
+    a normal RPC request over the WebRTC channel.
+\item when the \texttt{clientRequestResponse} function is called and
+    the status is "NOT\_LEADER" then the current leader ID being
+    tracked by the \texttt{clientRequest} function is set to the
+    leader ID specified in the response. Then the original request is
+    re-issued over the normal WebRTC channel to the leader.
+\item when the \texttt{clientRequestResponse} function is called with
     a successful status, the original callback argument of the
-    clientRequest function is called with the results from the RPC
-    response.
+    \texttt{clientRequest} function is called with the results from
+    the RPC response.
 \end{enumerate}
@@
 chat application provides a large text area containing the chat
-history and a text entry field below it. When the user clicks on the
-\emph{Send} button (or pressed enter), the current node ID is
+history and a text entry field below it
+(see Figure~\ref{fig:browser_chat_1}). When the user clicks on the
+\emph{Send} button (or presses enter), the current node ID is
 prepended to the beginning of the text line and it is added to
@@
 following:
+% TODO: 100ms above should be less
 
@@
     and stored in \emph{curSend} to indicate that a send is in
-    progress. The a client request is made using an atomic
-    \emph{seqAppend} command to add the new text to the state machine.
-    The \emph{seqAppend} operation attempts to append the text line to
-    a list at the \emph{history} key in the state machine.
+    progress. Then a client request is made using an atomic
+    \emph{seqAppend} command to add the new text to the state machine
+    (\emph{seqAppend} is describe below).  The \emph{seqAppend}
+    operation attempts to append the text line to a list at the
+    \emph{history} key in the state machine.
 \item When the atomic \emph{seqAppend} operation completes, it may
@@
     \emph{curSend} variable is set to nil/null to indicate that no
-    send is in progress (the next time the flush function runs it will
-    attempt to send again).
-\end{enumerate}
+    send is in progress. The next time the flush function runs it will
+    attempt to send again. % TODO: reword
+\end{enumerate}
+
+% TODO: normalize use of \texttt and \emph
 
 The chat application also extends the default set of state machine
-operations to add a seqAppend operation. This is an atomic operation
-that is implemented as follows:
-
-\begin{enumerate}
-\item the seqAppend operation operates on a specially structured
-    key/value pair in the state machine map. The value is
-    a associative map that contains a counter and list structure that
-    contains the actual data. seqAppend operations are sent with both
-    a count and then new value to append.
-\item When a seqAppend operation is applied to the state machine and
-    the count value in the command matches the counter in the state
-    machines value, then the new value in the request is added to the
-    list and the counter is incremented. If the count value does not
-    match then counter then the the operation is not applied. In
-    either case, a list is returned that contains two values: whether
-    the seqAppend operation succeeded and what the current value of
-    the counter.
-\item the first time the seqAppend operation is used on a non-existent
+operations to add a \emph{seqAppend} operation. This is an atomic
+operation that is implemented as follows:
+
+\begin{enumerate}
+\item the \emph{seqAppend} operation operates on a specially structured
+    key/value pair in the state machine map. The value of the pair is
+    another associative map that contains a counter and list structure
+    where the actual chat lines are accumulated. \emph{seqAppend}
+    operations are sent with both a count and then new value to
+    append.
+\item when a \emph{seqAppend} operation is applied to the state
+    machine and the count value in the command matches the counter in
+    the state machine's value, then the new line in the request is
+    added to the value list and the counter is incremented. If the count
+    value does not match the counter then the the operation is not
+    applied. In either case, a list is returned in the
+    \texttt{clientRequestResponse} that contains two values: a boolean
+    indicating whether the \emph{seqAppend} operation succeeded and
+    a number indicating the current value of the counter.
+\item the first time the \emph{seqAppend} operation is used on a non-existent
     key, the structure described above is created with the counter set
-    to 0 and the list structure is created with the single value
-    passed in the seqAppend operation.
-\item clients that call the seqAppend operation should check the
+    to 0 and the list structure is created with the single value that
+    was passed in the \emph{seqAppend} operation.
+\item clients that call the \emph{seqAppend} operation should check the
     status value in the result list. If the status is false, then the
-    client should retry the operation using the counter value as the
-    count in the operation.
+    client should retry the operation. Regardless of success of
+    failure, then client should store the current value of the counter
+    for the next \emph{seqAppend} call or retry.
 \end{enumerate}
@@
 The initial way that the application was tested was by loading the
-initial landing page to create a Raft over WebRTC cluster with
-a single leader. Here is a screenshot of the browser after the first
-node is created:
+basic application to create a Raft over WebRTC cluster with a single
+node acting as leader. Figure~\ref{fig:browser_1} shows a screenshot
+of the browser after the first node is created:
 
@@
 Then four more Raft nodes were then added to the cluster by clicking
-on the embedded "create a node" link in the application. Here is
-a screenshot of the browser once four new nodes are added to the
-cluster (5 total):
+on the embedded "create more nodes" link in the application.
+Figure~\ref{fig:browser_5} shows a screenshot of the browser once four
+new nodes are added to the cluster (5 total):
 
@@
 loading the initial chat application landing page to create a Raft
-over WebRTC cluster with a single leader. Then one or more lines of
-text were entered into the text line each followed by clicking the
-Send button.  Here is a screenshot of the browser after the first
-node/tab of the chat application is created and a line of text was
-sent:
+over WebRTC cluster with a single leader acting as leader. Then lines
+of text were entered into the text area each followed by clicking the
+Send button. Figure~\ref{fig:browser_chat_1} shows a screenshot of the
+browser after the first node/tab of the chat application is created
+and a line of text was sent:
 
@@
 Finally, on a non-leader node more text lines were entered and sent by
-clicking the Send button. Here is a screenshot of the browser once
-four new nodes/tabs are added to the cluster (5 total) and another
-text line was sent:
+clicking the Send button. Figure~\ref{fig:browser_chat_5} shows
+a screenshot of the browser once four new nodes/tabs are added to the
+cluster (5 total) and another text line was sent from the fifth node:
 
@@
 \item all non-leader nodes are receiving `appendEntries` RPCs from the leader
-\item all nodes have the same perception of the cluster size
+\item all nodes have the same cluster membership state
 \item the transaction log entries are identical on all nodes
@@
 The current leader was then terminated. The other nodes were checked
-for the steady state properties. This was repeated until only two
+again the steady state properties. This was repeated until only two
 nodes remained in the cluster.
 
-After the five node cluster was brought up, state machine commands
-were manually sent to leader to add to the transaction log. This was
-done using the JavaScript console as there is not yet a way to do this
-via the interface. Again the nodes were checked for the steady state
-properties.
+Next the five node cluster was brought up again and state machine
+commands were sent to leader to add to the transaction log. Again the
+nodes were checked for the above steady state properties.
 
 Then the application was tested was by loading the initial cluster
-creation landing page in a browser tab. Then nodes were randomly added
-to the cluster (by clicking on the link embedded in the page) or
-removed from the cluster (by closing a tab). After each change the
-steady state properties were checked.
+creation landing page in a browser tab. Nodes were randomly added to
+the cluster (by clicking on the link embedded in the page) or removed
+from the cluster (by closing a tab). After each change the steady
+state properties were checked.
 
@@
 and the Raft properties were maintained throughout. In addition,
-client interaction was tested although the application does not
-provide a friendly interface for doing so.
-
-One limitation of the system is that the if the cluster membership
-drops from two nodes to one node then the cluster will become
-unavailable without manual intervention. This is because when there
-are two nodes in the cluster, both nodes are required for for
+client interaction was tested to make sure commands were still
+properly added to the state machine and transaction log.
+
+One limitation of the Raft protocol is that the if the cluster
+membership drops from two nodes to one node then the cluster will
+become unavailable without manual intervention. This is because when
+there are two nodes in the cluster, both nodes are required for for
 a majority vote and when one member is forcible shutdown then the
-remaining cluster node will be unable to commit transaction entries
-and the cluster will become unavailable. There are several options to
-get around this problem:
+remaining cluster node will be unable to win elections or even commit
+transaction entries and the cluster will become unavailable. There are
+several options to get around this problem:
+
+% NOTE: optional section
 
@@
   may immediately kill the window/tab without giving the application
-  a change to remove itself from the cluster.
+  a chance to remove itself from the cluster.
 \item The signaling server can serve as a tie-breaker. This would allow
@@
   of a network partition. If a node dies or is killed, this would only
-  allow the cluster to continue to operate if the node that dies
-  happens to have the higher ID.
+  allow the cluster to continue to operate if the node that survives
+  happens to have the lower ID.
 \item The system could allow manual user intervention. However, this is
   only likely to work correctly in the case where the user of the last
-  remaining node can be certain that the other node will never return
-  otherwise they is the risk of data-corruption.
-\end{enumerate}
-
-Running a Raft cluster in a browser context leads to some level of
+  remaining node can be certain that other nodes will never return
+  otherwise there is a risk of data-corruption.
+\end{enumerate}
+
+Running a Raft cluster in a browser context can lead to some level of
 timer instability. The JavaScript execution context is a fully event
@@
 events trigger JavaScript code that is registered to that event. In
-order to implement logic that runs periodically, JavaScript code
-is register as a callback timer using the setTimer, setInterval or
+order to implement logic that runs periodically, JavaScript code is
+registered as a callback timer using the setTimer, setInterval or
 requestAnimationFrame APIs. However, none of these mechanisms is
 a precise timer. When a timer fires, the callback code they refer to
-is added to the JavaScript event queue and events on the queue are
+is added to the JavaScript execution queue and events on the queue are
 processed in the order they arrive. This means callbacks may be
-delayed by other code that is already running or queued first.
+delayed by other code that is already running or queued first
+\cite{resig2013secrets}.
 
@@
 of time. This throttling can cause timers to slow down by an order of
-magnitude or more.
+magnitude or more. %TODO/CITE
 
@@
 foreground are more likely to have an election timeout timer fire and
-they are more likely to become a leader because they are more likely
-to be able to send and receive votes quickly than nodes that are in
-the background. In this sense the problem is self-correcting because
-over the long term because higher performing nodes more are likely to
-start and win elections. This also indicates dynamic adjustment of
-timeout values may be a fruitful area of further study.
+they are more likely to become a leader because they will be able to
+send and receive votes more quickly than nodes that are in the
+background.  In this sense the problem is self-correcting because over
+the long term because higher performing nodes more are likely to start
+and win elections. This also indicates dynamic adjustment of timeout
+values may be a fruitful area of further study.
 
@@
 pages is even more pronounced because each addition to the chat
-history can take several messages to fully propagate the new date.
+history can take several messages to fully propagate the new date. An
+interesting area of further study would be to extend the Raft protocol
+to more proactively adapt to network and node performance conditions
+(delayed timers) and more intentionally transition leadership to more
+capable nodes.
 
@@
 built using \emph{test/Dockerfile} which is in turn derived from
-the "fentas/slimerjs" docker image
-(https://github.com/fentas/docker/blob/master/slimerjs/Dockerfile).
+the "fentas/slimerjs" docker image \cite{fentas:slimerjs}.
 This custom Dockerfile was created because the upstream version of the
 Dockerfile used an older version of XULRunner that did not support
-more than 7 WebRTC clients communicating together simultaneously.
-% (TODO/CITE)
+more than 7 WebRTC clients communicating together simultaneously. This
+issue was discovered during testing.
 
@@
 
-Three different tests were created to test the Raft over WebRTC
-functionality and the chat application. Each of the tests takes
+Three different tests were created to test the basic Raft over WebRTC
+application and the chat application. Each of the tests takes
 a parameter indicating the number of nodes/pages to launch and test
@@
     \caption{Recovery from Leader Failure}
-    \label{fig:E}
+    \label{fig:recovery-leader-failure}
 
@@
     \caption{Recovery from Half Cluster Failure}
-    \label{fig:G}
+    \label{fig:recovery-half-failure}
 \end{figure*}
@@
     protocol is tolerant of out-of-order and dropped messages, it is
-  possible that Raft over WebRTC could be more efficient with these
-  one or both of these constraints relaxed.
-\item Test the survivability of the cluster when the signaling server is
-  becomes completely unresponsive or just very slow.
+    possible that Raft over WebRTC could be more efficient with one or
+    both of these constraints relaxed.
+\item Test the survivability of the cluster when the signaling server
+    becomes completely unresponsive or very slow.
 \item Explore models for dynamically adjusting Raft timeout values to
     account for changing network and browser conditions.
-\item Test the ability of the system to catch up stale nodes by
-  disconnecting one or more nodes, run the cluster for some period
-  time with transactions created periodically, and then reconnect the stale
-  clients.
-\item Implement a user interface for sending commands to the transaction
-  log and for viewing the current state of the Raft state machine.
-\item Explore models for dealing with the case where the cluster goes from
-  two nodes down to one node. Can usefulness and availability be
-  improved without sacrificing Raft safety constraints.
-\item Test the system in more various network environments. Determine the
-  reliability of establishing direct browser-to-browser communication.
-  Determine if corporate firewalls make this as difficult as is
-  implied in some of the documentation and standards texts.
-\item Implement and test forwarding client requests rather than default
-  redirecting the client to the leader. In other words, allow client
-  requests to originate from any node of the cluster and be redirected
-  automatically to the leader (rather than the current manual
-  process).
-\item Perform quantitative and real-world testing to better characterize
-  the performance, availability and scalability characteristics of the
-  system.
+\item Explore models for dealing with the case where the cluster goes
+    from two nodes down to one node. Can usefulness and availability
+    be improved without sacrificing Raft safety constraints.
+\item Test the system in more diverse network environments. Determine
+    the reliability of establishing direct browser-to-browser
+    communication.  Determine if corporate firewalls make this as
+    difficult as is implied in some of the documentation and standards
+    texts. %TODO/CITE
+\item Perform quantitative and real-world testing to better
+    characterize the performance, availability and scalability
+    characteristics of the system.
 \item Explore models for timing out and automatically removing nodes
-  when they become unresponsive without needing to be notified by the
-  signaling server that the nodes have disconnected.
-\item Test larger scale deployments across multiple systems. Determine the
-  factors that affect scalability and characterize scaling limits.
-\item Perform wider cross-browser testing including Internet Explorer and
-  Opera.
-\item Implement Raft log compaction.
-\item Explore models that allow for a combination of strict shared state
-  and high performance bulk data such as media or object blobs. For
-  example, the data with full consensus might contain hashes that to
-  the refer to bulk data.
-\item Implement a Node.js based Raft over WebRTC application. This would
-  allow non-browser contexts to participate in the Raft cluster
-  together with the browser clients. PeerJS issue
-  [103](https://github.com/peers/peerjs/issues/103) must be resolved
-  for this to be feasible.
-\item Explore a design where some nodes of the cluster are non-voting and
-  are not counted as part of transaction commit quorum. However, these
-  nodes would still be included in the log replication algorithm. One
-  model this would enable would be a configuration where a small group
-  of voting nodes serve as the hub for a larger group of non-voting
-  nodes. This may enable higher scaling of the system. It also enables
-  interesting applications such as a logging or backup nodes.
+    when they become unresponsive without needing to be notified by
+    the signaling server that the nodes have disconnected.
+\item Consider extensions to the Raft protocol that enable slow
+    leaders to be proactively detected so that a more capable node
+    (network, CPU, etc) can be elected. This would help address that
+    background page/tag throttling of modern browsers.
+\item Test larger scale deployments spanning larger and more diverse
+    network domains. Determine the factors that affect scalability and
+    characterize scaling limits.
+\item Perform wider cross-browser testing including Internet Explorer
+    and Opera.
+\item Implement Raft log compaction and full client linearizability.
+\item Explore models that allow for a combination of strict shared
+    state and high performance bulk data such as media or object
+    blobs. For example, the data with full consensus might contain
+    hashes that to the refer to bulk data.
+\item Implement a Node.js based Raft over WebRTC application. This
+    would allow non-browser contexts to participate in the Raft
+    cluster together with the browser clients. PeerJS issue \#103 must
+    be resolved for this to be feasible \cite{peerjs:103}.
+\item Explore a design where some nodes of the cluster are non-voting
+    and are not counted as part of transaction commit quorum. However,
+    these nodes would still be included in the log replication
+    algorithm. One model this would enable would be a configuration
+    where a small group of voting nodes serve as the hub for a larger
+    group of non-voting nodes. This may enable higher scaling of the
+    system. It also enables interesting applications such as a logging
+    or backup nodes.
 \end{enumerate}

### 2015-09-11 test/launch.js
Start of twst based test framework.

@@ -1,0 +1,31 @@
+var webpage = require('webpage'),
+    system = require('system'),
+    url = system.args[1],
+    timeout = (system.args.length >= 3) ? parseInt(system.args[2]) : 120;
+
+var page = webpage.create();
+
+page.onConsoleMessage = function (msg, line, origin) {
+    console.log('CONSOLE: ' + msg);
+};
+
+page.onCallback = function(msg) {
+    console.log('CALLBACK: ' + msg);
+    if (msg === 'QUIT') {
+        //console.log('Normal exit');
+        slimer.exit(0);
+    }
+};
+
+console.log('Loading ' + url);
+page.open(url, function(status) {
+    if (status !== 'success') {
+        console.log('Unable to load the address!');
+        phantom.exit(1);
+    }
+});
+
+setTimeout(function() {
+    console.log('Timeout');
+    phantom.exit(1);
+}, timeout*1000);

### 2015-09-11 test/wait_start.js

@@ -1,0 +1,64 @@
+#!/usr/bin/env node
+
+var spawn = require('child_process').spawn,
+    Twst = require('twst').Twst,
+    twst = new Twst(),
+    clientCount = process.argv.length >= 3 ? parseInt(process.argv[2]) : 1,
+    timeout = (clientCount*10)*1000;
+
+var url = 'http://192.168.2.3:9000/test/index.html';
+
+function loadPage(url, opts) {
+    opts = opts || {};
+    var prefix = opts.prefix || "";
+    var docker_args = ['run', '-it', '--rm',
+        '-v', process.cwd() + '/test:/test',
+        'slimerjs-wily',
+        'slimerjs', '/test/launch.js',
+        url,
+        timeout];
+
+    console.log(prefix + 'launching ' + url);
+    //console.log('docker', docker_args.join(" "));
+    var page = spawn('docker', docker_args);
+
+    page.stdout.on('data', function(chunk) {
+        console.log(prefix + chunk.toString().replace(/\n$/,''));
+    });
+
+    page.on('close', function (code) {
+        console.log(prefix + 'docker container exited with code ' + code);
+        process.exit(code);
+    });
+    return page;
+}
+
+var pages = [];
+for (var i=0; i<clientCount; i++) {
+    pages.push(loadPage(url, {prefix: 'p' + i + ': '}));
+}
+
+setInterval(function() {
+    if (Object.keys(twst.clients).length >= clientCount) {
+        process.exit(0);
+    }
+}, 100);
+
+setTimeout(function() {
+    console.log("timeout waiting for clients");
+    process.exit(1);
+}, timeout);
+
+/*
+function test_func() {
+    return location.search + Math.random();
+}
+
+setInterval(function() {
+    var opts = {timeout: 10000,
+                restype: 'return'};
+    twst.collect(test_func, opts, function(result, data) {
+        console.log('collect result:', result, data);
+    });
+}, 6000);
+*/

### 2015-09-12 research_class/paper.tex
Add preferred comma.

@@
 These new browser capabilities also bring with them the traditional
-challenges of building reliable distributed applications such as:
+challenges of building reliable distributed applications, such as:
 concurrency, security, scalability, naming, discovery, replication,

### 2015-09-12 research_class/paper.tex
Add missing indefinite article.

@@
 one category of distributed state management in which multiple nodes
-must come to agreement on the state of single value or set of values. % TODO/CITE
+must come to agreement on the state of a single value or set of values. % TODO/CITE
 The grandparent of distributed consensus algorithms is the Paxos

### 2015-09-12 research_class/paper.tex
Add missing comma.

@@
 will describe the results of the implementation and look at some of
-unique challenges (and opportunities) we encountered with the
+the unique challenges (and opportunities) we encountered with the
 implementation. We will also explore some ideas for further study.
@@
 The traditional use of distributed consensus is in the data-center
-(private or cloud based) often as the core coordination component
+(private or cloud based), often as the core coordination component
 of a replicated data-storage system. %TODO/CITE

### 2015-09-12 research_class/paper.tex
Add preferred comma.

@@
 traverse the back-end servers is not always desirable for various
-reasons including the following:
+reasons, including the following:
 

### 2015-09-12 research_class/paper.tex
Usage: "traverse" doesn't need "through"

@@
     may need to traverse directly from browser to browser rather than
-    traversing through a server.
+    traversing a server.
 \item \emph{locality}: the nature of the application may allow
@@
     all users see the same message history but the messages never
-    traverse through a central server and there is no central arbiter
+    traverse a central server and there is no central arbiter
     of trusth. The application code itself may still be delivered from

### 2015-09-12 research_class/paper.tex
Fix typo: s/trusth/truth/

@@
     traverse a central server and there is no central arbiter
-    of trusth. The application code itself may still be delivered from
+    of truth. The application code itself may still be delivered from
     a central web service. This is the use case that was implemented

### 2015-09-12 research_class/paper.tex
Cut three small words, keep clarity. Save a line!
Add missing comma.
This comma changes the meaning. "Google Docs and other examples of collaborative
document creation" is what I think you intend, in which case you want the comma.

"Google Docs exemplifies collaborative document creation and also other
technical achievements" is what this means without the comma, and I don't think
that's what you intend.

@@
     of truth. The application code itself may still be delivered from
-    a central web service. This is the use case that was implemented
+    a central web service. This use case was implemented
     for this paper.
 \item \emph{Private multiuser document editing/whiteboard system}:
-    Google Docs, among other examples has demonstrated a powerful
+    Google Docs, among other examples, has demonstrated a powerful
     model for collaborative document creation. However, these systems

### 2015-09-12 research_class/paper.tex
"Raft" as a proper noun...
Or you could say "a Raft implementation", but I like "Raft" by itself.
Fix directional LaTeX quoting.

@@
     handle browser-to-browser communication. %TODO/CITE
-    Using a Raft over WebRTC could enable applications with similar
+    Using Raft over WebRTC could enable applications with similar
     functionality but without the central coordination and
@@
 correctness. The Raft algorithm implemented and described in this
-paper is based on Diego Ongaro's dissertation "Consensus: Bridging
+paper is based on Diego Ongaro's dissertation ``Consensus: Bridging
 Theory and Practice" \cite{raft_thesis:ongaro14}.

### 2015-09-12 research_class/paper.tex
Fix typo.

@@
 the leader is to accept new transaction log entries and then
-replicated these entries in the same order to all other members of the
+replicate these entries in the same order to all other members of the
 cluster. The leader is the only member of the cluster that may make

### 2015-09-12 research_class/paper.tex
NOT SURE: is this right?
If this is right, I think it's clearer. If it's wrong, then I'm
misunderstanding/misguessing. You may or may not want to consider clarifying
somehow for the other dunderheads like me, who will read this and possibly be
confused.

@@
 current term value. When a Raft follower node does not receive
-a heartbeat from a leader within a random election timeout period, it
+a heartbeat from a leader within a locally randomized election timeout period, it
 transitions to candidate state, votes for itself, increases its

### 2015-09-12 research_class/paper.tex
Fix typo: subject/verb agreement
Two grammar fixes.

@@
 transaction log each contain a term value in addition to the action to
-apply to the state machine. Each node also keep track of the most
+apply to the state machine. Each node also keeps track of the most
 recent log entry which is known to be committed (exists in the log of
@@
 does not have an entry matching the most recent entry on the leader,
-then it replies with false to indicate that its log and not up to
+then it replies with false to indicate that its log is not up to
 date. The next appendEntries RPC that the leader sends to that
 node will contain the next oldest index and term. This will continue
-until the leader discovers the latest log entry which is agreement
+until the leader discovers the latest log entry which is in agreement
 (the node may not have any entries if it is new). Then the leader will

### 2015-09-12 research_class/paper.tex
Plural -> singular sentence.
This sentence is too complex to be plural; subject/verb agreement would require
a few changes and more complex wording. I think it's clearer and simpler in the
singular.

@@
 provides safety guarantees. In particular, there are some restrictions
-on which Raft nodes are actually eligible to become a leader based on
-the state of their transaction log compared to other node transaction
+on whether a Raft node is actually eligible to become a leader, based on
+the state of its transaction log compared to other node transaction
 logs. A Raft node will only vote for a candidate if the candidate has

### 2015-09-13 rtc.js
rtc.js: fix formatting of start routine.

@@ -205,73 +205,72 @@ function addRemoveServersAsync() {
 function start(opts) {
     opts = opts || {};
 
-var peer = new Peer({host: location.hostname,
-                     port: location.port,
-                     path: '/api/' + channel});
-peer.on('open', function(id) {
-    log('my RTC ID:', id);
-    nodeId = id;
-    // Put ourself into the map
-    nodeMap[nodeId] = null;
+    var peer = new Peer({host: location.hostname,
+                         port: location.port,
+                         path: '/api/' + channel});
+    peer.on('open', function(id) {
+        log('my RTC ID:', id);
+        nodeId = id;
+        // Put ourself into the map
+        nodeMap[nodeId] = null;
 
-    // Load the peer list and connect
-    jQuery.getJSON('/peers/' + channel, function (peers) {
-        log("RTC peer list:", peers);
-        $.each(peers, function (idx, peer_id) {
-            // If it is us, or we are already connected, ignore
-            if (peer_id in nodeMap) return true;
-            if (peer_id === nodeId) {
-                // Ignore
-                return true;
-            }
+        // Load the peer list and connect
+        jQuery.getJSON('/peers/' + channel, function (peers) {
+            log("RTC peer list:", peers);
+            $.each(peers, function (idx, peer_id) {
+                // If it is us, or we are already connected, ignore
+                if (peer_id in nodeMap) return true;
+                if (peer_id === nodeId) {
+                    // Ignore
+                    return true;
+                }
 
-            log("Connecting to RTC peer:", peer_id);
-            var conn = peer.connect(peer_id);
-            nodeMap[peer_id] = conn;
-            conn.on('data', function(data) {
-                //log("received from " + conn.peer + ": " + data);
-                rtcReceive(data);
-            });
-            conn.on('close', function(data) {
-                //log("RTC connection closed:" + conn.peer);
-                delete nodeMap[peer_id];
+                log("Connecting to RTC peer:", peer_id);
+                var conn = peer.connect(peer_id);
+                nodeMap[peer_id] = conn;
+                conn.on('data', function(data) {
+                    //log("received from " + conn.peer + ": " + data);
+                    rtcReceive(data);
+                });
+                conn.on('close', function(data) {
+                    //log("RTC connection closed:" + conn.peer);
+                    delete nodeMap[peer_id];
+                });
             });
         });
+
+        // Create the local raft.js node
[…]

### 2015-09-14 rtc.js
RtcTwst class and wait_start test
- RtcTwst class extends Twst class to add docker RTC page creation
- Fix rtc_server.js so that landing page is the actual page and the
  channel is created whenever it is either not specified or if it is
  specified but doesn't exist yet.

@@ -37,16 +37,21 @@ var messages = document.getElementById('messages'),
         appendEntries: 0,
         appendEntriesResponse: 0};
 
-function getParameterByName(name) {
-    name = name.replace(/[\[]/, "\\[").replace(/[\]]/, "\\]");
-    var regex = new RegExp("[\\?&]" + name + "=([^&#]*)"),
-        results = regex.exec(location.search);
-    return results === null ? "" : decodeURIComponent(results[1].replace(/\+/g, " "));
+// Parse query parameters
+if (!location.queryParams) {
+    location.queryParams = {};
+    location.search.substr(1).split("&").forEach(function (param) {
+        if (param === "") return;
+        var key_value = param.split("=");
+        location.queryParams[key_value[0]] = key_value[1] &&
+            decodeURIComponent(key_value[1].replace(/\+/g, " "));
+    });
 }
 
-// Get the channel
-var channel = getParameterByName('channel');
-var console_logging = getParameterByName('console_logging');
+
+// Get the channel and logging parameters
+var channel = location.queryParams.channel;
+var console_logging = location.queryParams.console_logging;
 
 // Setup the new node link
 node_link.href = location.origin + location.pathname + location.search;

### 2015-09-14 rtc_server.js

@@ -36,8 +36,9 @@ function newPeerServerChannel(channel) {
 }
 
 var channelCnt = 0;
+var channels = {};
 
-app.get('/', function(req, res, next) {
+app.get('/' + opts.home, function(req, res, next) {
     var channel = null;
     // Pass along query parameters
     var query = "";
@@ -47,19 +48,31 @@ app.get('/', function(req, res, next) {
             channel = req.query[k];
         }
     }
-    if (channel === null) {
-        channel = channelCnt++;
+    if (channel === null || (!(channel in channels))) {
+        // Redirect to make sure URL has channel parameter and is also
+        // marked as the firstServer
+        if (channel === null) {
+            // Assign an unused channel number
+            do {
+                channel = channelCnt++;
+            } while (channel in channels);
+            var url = '/' + opts.home + '?channel=' + channel + query + '#firstServer';
+        } else {
+            var url = '/' + opts.home + '?' + query + '#firstServer';
+        }
+        // Create a new channel
+        channels[channel] = newPeerServerChannel(channel);
+        // Do the redirect
+        res.send('<html><head><meta http-equiv="refresh" content="0; url=' + url + '"/></head></html>');
+    } else {
+        // The channel already exists, continue
+        next();
     }
-    newPeerServerChannel(channel);
-
-    // Full redirect URL
-    var url = opts.home + '?channel=' + channel + query + '#firstServer';
-    res.send('<html><head><meta http-equiv="refresh" content="0; url=' + url + '"/></head></html>');
 });
 
 
 // Static file serving
-app.use( express.static('./'));
+app.use(express.static('./'));
 
 // Start Express server
 var server = app.listen(opts.port);

### 2015-09-14 test/rtctwst.js

@@ -1,0 +1,105 @@
+var spawn = require('child_process').spawn,
+    Twst = require('twst').Twst,
+    RtcTwst = null;
+
+// Subclass Twst
+exports.RtcTwst = RtcTwst = function(opts) { Twst.call(this, opts); };
+RtcTwst.prototype = Object.create(Twst.prototype);
+RtcTwst.prototype.constructor = RtcTwst;
+
+RtcTwst.prototype.cleanup_exit = function(exitCode) {
+    console.log("Cleaning up and exiting with code " + exitCode);
+    this.broadcast('window.callPhantom("QUIT")');
+    setTimeout(function() {
+        process.exit(exitCode);
+    }, 1000);
+}
+
+RtcTwst.prototype.dockerPage = function(url, opts) {
+    opts = opts || {};
+    var prefix = opts.prefix || "";
+    var timeout = opts.timeout || 60000;
+    var docker_args = ['run', '-it', '--rm',
+        '-v', process.cwd() + '/test:/test',
+        'slimerjs-wily',
+        'slimerjs', '/test/launch.js',
+        url,
+        timeout];
+
+    console.log(prefix + 'launching ' + url);
+    var page = spawn('docker', docker_args);
+
+    page.stdout.on('data', function(chunk) {
+        var line = chunk.toString('utf8').replace(/\r\n$/,'');
+        if (line === '') { return; }
+        console.log(prefix + line);
+    });
+
+    page.on('close', function (code) {
+        console.log(prefix + 'docker container exited with code ' + code);
+        cleanup_exit(this, code);
+    });
+    return page;
+}
+
+////////////////////////////////////
+// RTC specific
+
+RtcTwst.prototype.get_node_info = function(timeout, callback) {
+    var nodes = [];
+    this.collect(function() {
+        // Evaluates in page context so 'node' variable is implicit
+        if (typeof node !== 'undefined' && node) {
+            var data = {id: node._self.id,
+                        state: node._self.state,
+                        serverMapKeys: Object.keys(node._self.serverMap)};
+            return data;
+        } else {
+            return null;
+        }
+    }, {timeout: timeout}, callback);
+}
+
+RtcTwst.prototype.wait_cluster_up = function(timeout, callback) {
+    var self = this,
+        server_count = Object.keys(self.clients).length,
+        start_time = Date.now();
+    var checkfn = function () {
+        // Gather data from the nodes
+        self.get_node_info(2000, function(status, nodes) {
[…]

### 2015-09-14 test/wait_start.js

@@ -1,64 +1,55 @@
 #!/usr/bin/env node
 
-var spawn = require('child_process').spawn,
-    Twst = require('twst').Twst,
-    twst = new Twst(),
-    clientCount = process.argv.length >= 3 ? parseInt(process.argv[2]) : 1,
-    timeout = (clientCount*10)*1000;
+// To use this, first start an rtc_server.js signalling server (i.e.
+// running at 10.0.0.1:8001). Then start this test:
+//     node test/wait_start.js 10.0.01:8001 3
 
-var url = 'http://192.168.2.3:9000/test/index.html';
+var RtcTwst = require('./rtctwst').RtcTwst,
+    rtwst = new RtcTwst(),
+    rtc_address = process.argv[2],
+    clientCount = process.argv.length >= 4 ? parseInt(process.argv[3]) : 1,
+    timeout = (clientCount*20)*1000;
 
-function loadPage(url, opts) {
-    opts = opts || {};
-    var prefix = opts.prefix || "";
-    var docker_args = ['run', '-it', '--rm',
-        '-v', process.cwd() + '/test:/test',
-        'slimerjs-wily',
-        'slimerjs', '/test/launch.js',
-        url,
-        timeout];
-
-    console.log(prefix + 'launching ' + url);
-    //console.log('docker', docker_args.join(" "));
-    var page = spawn('docker', docker_args);
-
-    page.stdout.on('data', function(chunk) {
-        console.log(prefix + chunk.toString().replace(/\n$/,''));
-    });
-
-    page.on('close', function (code) {
-        console.log(prefix + 'docker container exited with code ' + code);
-        process.exit(code);
-    });
-    return page;
-}
+//var url = 'http://192.168.2.3:9000/test/index.html';
+var channel = Math.round(Math.random()*100000);
+var url = 'http://' + rtc_address +
+          '/rtc.html?channel=' + channel +
+          '&twst_address=' + rtwst.getAddress() + '&paused=1';
 
 var pages = [];
+
 for (var i=0; i<clientCount; i++) {
-    pages.push(loadPage(url, {prefix: 'p' + i + ': '}));
+    pages.push(rtwst.dockerPage(url, {prefix: 'p' + i + ': ',
+                                      timeout: timeout}));
 }
 
-setInterval(function() {
-    if (Object.keys(twst.clients).length >= clientCount) {
-        process.exit(0);
+function poll() {
+    if (Object.keys(rtwst.clients).length >= clientCount) {
[…]

### 2015-09-15 test/common.js
test/wait_chat_propagate and wait_cluster_predicate
- refactor wait_start too.

@@ -10,6 +10,9 @@ exports.new_page = function (page_idx, firstServer, opts, callback) {
     page.onConsoleMessage = function (msg, line, origin) {
         console.log('CONSOLE ' + page_idx + ': ' + msg);
     };
+    page.onError = function (msg) {
+        console.log('ERROR ' + page_idx + ': ' + msg);
+    };
     page.onCallback = function(msg) {
         console.log('CALLBACK ' + page_idx + ': ' + msg);
         if (msg === 'QUIT') {

### 2015-09-15 test/rtctwst.js

@@ -46,7 +46,6 @@ RtcTwst.prototype.dockerPage = function(url, opts) {
 // RTC specific
 
 RtcTwst.prototype.get_node_info = function(timeout, callback) {
-    var nodes = [];
     this.collect(function() {
         // Evaluates in page context so 'node' variable is implicit
         if (typeof node !== 'undefined' && node) {
@@ -103,3 +102,33 @@ RtcTwst.prototype.wait_cluster_up = function(timeout, callback) {
     checkfn();
 }
 
+RtcTwst.prototype.wait_cluster_predicate = function(timeout, predicate, callback) {
+    var self = this,
+        server_count = Object.keys(self.clients).length,
+        start_time = Date.now();
+    var checkfn = function () {
+        self.collect(predicate, {timeout: timeout}, function(status, data) {
+            var elapsed = Date.now() - start_time;
+            if (!status) {
+                callback(false, data, elapsed)
+            }
+            var trueCnt = 0;
+            for (var n in data) {
+                if (data[n].data) { trueCnt += 1; }
+            }
+
+            console.log('Predicate results: ' + JSON.stringify(data) +
+                        ', true count: ' + trueCnt);
+            // Exit if cluster is up or we timeout
+            if (trueCnt >= server_count) {
+                callback(true, data, elapsed);
+            } else if (elapsed > timeout) {
+                callback(false, data, elapsed);
+            } else {
+                setTimeout(checkfn, 25);
+                //setTimeout(checkfn, 2000);
+            }
+        });
+    }
+    checkfn();
+}

### 2015-09-15 test/wait_chat_propagate.js

@@ -17,6 +17,7 @@ var RtcTwst = require('./rtctwst').RtcTwst,
 var channel = Math.round(Math.random()*100000);
 var url = 'http://' + rtc_address +
           '/chat.html?channel=' + channel +
+          '&console_logging=true' +
           '&twst_address=' + rtwst.getAddress() + '&paused=1';
 
 var pages = [];
@@ -36,15 +37,42 @@ function poll() {
 }
 
 function do_start() {
-    rtwst.broadcast('start();');
+    rtwst.broadcast('startChat();');
     rtwst.wait_cluster_up(timeout, function(status, nodes, elapsed) {
         if (status) {
             console.log('Cluster is up after ' + elapsed + 'ms');
             console.log('Delaying for 3 seconds before sending message');
-            rtwst.cleanup_exit(0);
+            setTimeout(do_chat, 3000);
         } else {
             console.log('Cluster failed to come up after ' +
                         elapsed + 'ms');
+            rtwst.cleanup_exit(1);
+        }
+    });
+}
+
+function do_chat() {
+    rtwst.send(function() {
+        $('#talk').val("test line");
+        //sendLine();
+        $('#send').click();
+    }, {index: pages.length-1});
+    rtwst.wait_cluster_predicate(timeout, function() {
+        var sm = node._self.stateMachine;
+        //console.log(nodeId + " stateMachine: " + JSON.stringify(sm));
+        if ('history' in sm && 'value' in sm.history) {
+            var lines = sm.history.value,
+                m = lines[lines.length-1].match(/.*: test line$/);
+            return m ? true : false;
+        } else {
+            return false;
+        }
+    }, function(status, results, elapsed) {
+        if (status) {
+            console.log('Cluster state propagated after ' + elapsed + 'ms');
+            rtwst.cleanup_exit(0);
+        } else {
+            console.log('Cluster state failed to propagate after ' + elapsed + 'ms')
             rtwst.cleanup_exit(1);
         }
     });

### 2015-09-16 research_class/paper.tex
Plural -> singular for subject/verb felicity.
Directionalize all opening double-quotes.

@@
 then the entry with a higher index is more up-to-date. In addition,
-Raft leaders may not consider entries from a previous term to be
+a Raft leader may not consider entries from a previous term to be
 committed until it has committed at least one entry from its own
@@
 and instantiates it using a scheduling function that executes one
-scheduled event each time the user clicks on the "Step" button rather
+scheduled event each time the user clicks on the ``Step" button rather
 than due to the passage of time.
@@
 different starting behavior depending on the value of the URL query
-string (the URL component following the "?" and up to the "\#") and
-the fragment identifier (the URL component after the "\#").
+string (the URL component following the ``?" and up to the ``\#") and
+the fragment identifier (the URL component after the ``\#").
 
@@
 
-If the URL fragment identifier is set to "firstServer", then this Raft
+If the URL fragment identifier is set to ``firstServer", then this Raft
 node is initialized as the leader of a single node cluster and all
@@
 browser context (separate browser, browser window, or browser tab) and
-loading the same URL without the "firstServer" fragment identifier
+loading the same URL without the ``firstServer" fragment identifier
 (Figure~\ref{fig:raft_rtc_sequence}~[9]). There is a convenience link
@@
 When an instance of the web application starts that is not
-a "firstServer", the normal Raft timers are not enabled until the
+a ``firstServer", the normal Raft timers are not enabled until the
 server receives its first RPC. The prevents the server from
@@
 \item when the \texttt{clientRequestResponse} function is called and
-    the status is "NOT\_LEADER" then the current leader ID being
+    the status is ``NOT\_LEADER" then the current leader ID being
     tracked by the \texttt{clientRequest} function is set to the
@@
 Then four more Raft nodes were then added to the cluster by clicking
-on the embedded "create more nodes" link in the application.
+on the embedded ``create more nodes" link in the application.
 Figure~\ref{fig:browser_5} shows a screenshot of the browser once four
@@
 Then four more Raft nodes were then added to the cluster by clicking
-on the embedded "create more clients" link in the application.
+on the embedded ``create more clients" link in the application.
 Finally, on a non-leader node more text lines were entered and sent by
@@
 built using \emph{test/Dockerfile} which is in turn derived from
-the "fentas/slimerjs" docker image \cite{fentas:slimerjs}.
+the ``fentas/slimerjs" docker image \cite{fentas:slimerjs}.
 This custom Dockerfile was created because the upstream version of the
@@
 
-Figure 7 is a scatterplot with each test run as a blue "x" and a black
+Figure 7 is a scatterplot with each test run as a blue ``x" and a black
 line representing the average at each cluster size.
@@
         size reaches 17, the cluster sometimes does not activate
-        before the test times out (the red "x" marks). Once the
+        before the test times out (the red ``x" marks). Once the
         cluster reaches 21 nodes, all 7 test runs time out. The reason
@@
 
-Each figure is a scatterplot with each test run as a blue "x" and
+Each figure is a scatterplot with each test run as a blue ``x" and
 a black line representing the average at each cluster size.

### 2015-09-16 research_class/paper.tex
Directionalize paired single-quotes.

@@
     Raft.js includes a small modification to the response RPCs for
-    requestVote and appendEntries to add a 'sourceId' field that
+    requestVote and appendEntries to add a `sourceId' field that
     enables the protocol to function with message based transports.
@@
     change request is pending (uncommitted), the leader will reject
-    the change request with a status of 'PENDING\_CONFIG\_CHANGE'.
+    the change request with a status of `PENDING\_CONFIG\_CHANGE'.
     This keeps the core Raft implementation simpler with the trade-off
@@
 
-The file `rtc\_server.js` implements the signaling server for the Raft
+The file `rtc\_server.js' implements the signaling server for the Raft
 over WebRTC application. The server extends the PeerJS
-`ExpressPeerServer` object in order to provide WebRTC signaling. In
+`ExpressPeerServer' object in order to provide WebRTC signaling. In
 addition to the signaling function, the server also provides the
@@
 \item \emph{Append Entries Sent/Received}: the number of
-    `appendEntries` RPCs that have been sent from and received by this
+    `appendEntries' RPCs that have been sent from and received by this
     node.
@@
 \item all nodes have the same term and transaction log count
-\item all non-leader nodes are receiving `appendEntries` RPCs from the leader
+\item all non-leader nodes are receiving `appendEntries' RPCs from the leader
 \item all nodes have the same cluster membership state

### 2015-09-16 research_class/paper.tex
Remove redundant "implementation".
Fix typo: make plural noun so.

@@
 
-The implementation of the Raft algorithm is implemented in the
+The implementation of the Raft algorithm is in the
 \texttt{RaftServerBase} class (in \texttt{base.js}). The base class
-omits the implementation of functions with side-effect such as RPC
+omits the implementation of functions with side-effects such as RPC
 communication, durable storage, scheduling/timeouts, or state machine

### 2015-09-16 research_class/paper.tex
Fix grammar, clarify with concision.

@@
 
-The current Raft.js implementation most of the Raft algorithm
+The current version of Raft.js implements most of the Raft algorithm
 including: leader election, log replication, safety, membership

### 2015-09-16 research_class/paper.tex
Fix grammar: add an article.

@@
 short-running test configurations. Full client linearizability is
-important feature for simplifying certain types of distributed
+an important feature for simplifying certain types of distributed
 applications, however the implementation of linearizability requires

### 2015-09-16 research_class/paper.tex
Clarify sentence subject, add commas.
Add missing preposition.

@@
     request/response stream based. %TODO/CITE
-    The original Raft definition of RPC request and response messages
+    In the original Raft definition of RPC, request and response messages
     almost contain enough information to function correctly in an
-    environment without builtin request/response correlation such as
+    environment without builtin request/response correlation, such as
     UDP or SCTP when used with reliable transport disabled. %TODO/CITE
@@
 network protocols \cite{draft-rtcweb-overview} that are currently
-being standardized the W3C and IETF respectively. Collectively these
+being standardized by the W3C and IETF respectively. Collectively these
 APIs and protocols enable real-time peer-to-peer video, audio and data

### 2015-09-16 research_class/paper.tex
Add a missing article.

@@
 To establish a direct communication channel between browsers, WebRTC
-uses mechanism called Interactive Connectivity Establishment (ICE)
+uses a mechanism called Interactive Connectivity Establishment (ICE)
 that is similar to the Session Initiation Protocol (SIP) commonly used

### 2015-09-16 research_class/paper.tex
Fix typo: s/uses/used/
Shorten sentence, subtly broaden meaning.
My first thought on reading this sentence was: "But you could *write* in
ClojureScript, or anything else that targets JS..." And then I realized this
proposed change could save you some words and avoid any implication that one
must author JavaScript to use the WebRTC browser APIs.

@@
 order to establish a direct connection to other user agents. The
-signaling server is also uses to communicate session control messages
+signaling server is also used to communicate session control messages
 (open, close, error) and to negotiate media capabilities. %TODO/CITE
@@
 
-The WebRTC browser APIs enable web applications written in JavaScript
+The WebRTC browser APIs enable JavaScript web applications
 to establish browser-to-browser network connectivity in order to

### 2015-09-16 research_class/paper.tex
s/incomplete/partial/ for article agreement
These both work:
* a partial list
* a list

@@
 section of the bibliography %TODO: reference link
-has a (incomplete) list of WebRTC specific protocols drafts and other
+has a (partial) list of WebRTC specific protocols drafts and other
 related protocol standards for the first three functionality groups

### 2015-09-16 research_class/paper.tex
Fix grammar.

@@
 section of the bibliography %TODO: reference link
-has a (partial) list of WebRTC specific protocols drafts and other
+has a (partial) list of WebRTC specific protocol drafts and other
 related protocol standards for the first three functionality groups

### 2015-09-16 research_class/paper.tex
Remove redundant list of functionality groups?
I think it's clear enough without the repitition.

@@
 has a (partial) list of WebRTC specific protocol drafts and other
-related protocol standards for the first three functionality groups
-(data transport, data framing and security, and connection
-management).
+related protocol standards for the first three functionality groups.
 

### 2015-09-16 research_class/paper.tex
Change "at time of paper" tense to present.
Not sure, but I prefer this.
I would remove this sentence.
It seems redundant in a short paragraph, but it *is* nice to start the paragraph
with a direct comment. Not sure about this one.

@@
 modular JavaScript API for WebRTC signaling \cite{ortc2014}. However,
-at the time of this paper, the ORTC interface was not yet widely
-supported and still in a state of flux. The use of PeerJS provides an
+at the time of this paper, the ORTC interface is not yet widely
+supported and is still in a state of flux. The use of PeerJS provides an
 abstraction that is likely to transparently support ORTC in the future.
@@
 
-The PeerJS client library provides an simple abstraction over the
-native browser WebRTC APIs. The WebRTC API standards are not yet
+The WebRTC API standards are not yet
 finalized and so different browsers and browser releases may support

### 2015-09-16 research_class/paper.tex
Really into opinion territory: s/their/her/
I default to the feminine pronouns in my own writing. I definitely grimace at
"their". But there's no obviously great resolution for this issue.
Fix spelling: s/WebRqTC/WebRTC/
Maybe capitalize full sentences in list?
Not sure at all, but my inclination is to say that full sentences should be
capitalized as such.

@@
 compatible signaling server, however an application developer can also
-run their own signaling service using PeerJS PeerServer.
+run her own signaling service using PeerJS PeerServer.
 
@@
 
-In the Raft over WebRqTC implementation each local browser is a client
+In the Raft over WebRTC implementation each local browser is a client
 of the Raft cluster. However in Raft, the leader node is the only node
@@
 \begin{enumerate}
-\item the \texttt{clientRequestResponse} is registered with the local
+\item The \texttt{clientRequestResponse} is registered with the local
     Raft.js implementation as the callback function anytime
     a \texttt{clientRequestResponse} RPC is received.
-\item the \texttt{clientRequest} function keeps track of the current
+\item The \texttt{clientRequest} function keeps track of the current
     leader ID.  If the current leader ID is not set or this node is
@@
     a normal RPC request over the WebRTC channel.
-\item when the \texttt{clientRequestResponse} function is called and
+\item When the \texttt{clientRequestResponse} function is called and
     the status is ``NOT\_LEADER" then the current leader ID being
@@
     re-issued over the normal WebRTC channel to the leader.
-\item when the \texttt{clientRequestResponse} function is called with
+\item When the \texttt{clientRequestResponse} function is called with
     a successful status, the original callback argument of the

### 2015-09-16 research_class/paper.tex
Fix typo: s/describe/described/
Fix wrong-word typo: s/leader/node/

@@
     \emph{seqAppend} command to add the new text to the state machine
-    (\emph{seqAppend} is describe below).  The \emph{seqAppend}
+    (\emph{seqAppend} is described below).  The \emph{seqAppend}
     operation attempts to append the text line to a list at the
@@
 loading the initial chat application landing page to create a Raft
-over WebRTC cluster with a single leader acting as leader. Then lines
+over WebRTC cluster with a single node acting as leader. Then lines
 of text were entered into the text area each followed by clicking the

### 2015-09-16 research_class/paper.tex
Fix grammar.
Fix grammar: subject/verb agreement

@@
 
-Then the application was tested was by loading the initial cluster
+Then the application was tested by loading the initial cluster
 creation landing page in a browser tab. Nodes were randomly added to
@@
 was able to perform leader elections correctly, the log and state
-machine was replicated correctly to all nodes, and the safety
+machine were replicated correctly to all nodes, and the safety
 properties were maintained. During testing the membership of the

### 2015-09-16 research_class/paper.tex
Fix grammar; remove extra "the".

@@
 
-One limitation of the Raft protocol is that the if the cluster
+One limitation of the Raft protocol is that if the cluster
 membership drops from two nodes to one node then the cluster will

### 2015-09-16 research_class/paper.tex
Fix spelling and reword paragraph.
s/forcible/forcibly/

Some of these sentences ran on a bit; I think the rewording is clearer, and is
almost exactly the same length.

@@
 
-One limitation of the Raft protocol is that if the cluster
-membership drops from two nodes to one node then the cluster will
-become unavailable without manual intervention. This is because when
-there are two nodes in the cluster, both nodes are required for for
-a majority vote and when one member is forcible shutdown then the
-remaining cluster node will be unable to win elections or even commit
-transaction entries and the cluster will become unavailable. There are
-several options to get around this problem:
+One limitation of the Raft protocol is that if the cluster membership drops from
+two nodes to one node, then the cluster will become unavailable without manual
+intervention. This is because both nodes of a two-node cluster are required for
+a majority vote.  When one member of a two-node cluster is forcibly shut down,
+then the remaining cluster node will be unable to win elections or even commit
+transaction entries, and the cluster will become unavailable. There are several
+options to get around this problem:
 

### 2015-09-16 research_class/paper.tex
Split run-on sentence with a semi-colon; it's fun!

@@
   only likely to work correctly in the case where the user of the last
-  remaining node can be certain that other nodes will never return
+  remaining node can be certain that other nodes will never return;
   otherwise there is a risk of data-corruption.

### 2015-09-16 research_class/paper.tex
Fix subject/verb agreement. A timer...it refers.

@@
 requestAnimationFrame APIs. However, none of these mechanisms is
-a precise timer. When a timer fires, the callback code they refer to
+a precise timer. When a timer fires, the callback code it refers to
 is added to the JavaScript execution queue and events on the queue are

### 2015-09-16 research_class/paper.tex
Remove extra "because".
Add demonstrative pronoun for clarity.
I think it's clearer with "that". *shrug*

@@
 send and receive votes more quickly than nodes that are in the
-background.  In this sense the problem is self-correcting because over
+background.  In this sense the problem is self-correcting over
 the long term because higher performing nodes more are likely to start
-and win elections. This also indicates dynamic adjustment of timeout
+and win elections. This also indicates that dynamic adjustment of timeout
 values may be a fruitful area of further study.

### 2015-09-16 research_class/paper.tex
Fix article: "an X"
Not sure: formalize with s/basically/essentially/
I'm not sure this is more formal, but "basically" struck me as being a bit
informal.

@@
 Mozilla rendering environment called XULRunner. The XULRunner instance
-is directed to a X virtual frame buffer (xvfb) which provides
+is directed to an X virtual frame buffer (xvfb) which provides
 effectively headless testing.
@@
 that are run within the SlimerJS environment. These scripts can create
-browser instances (basically headless browser tabs), query the status
+browser instances (essentially headless browser tabs), query the status
 of those pages and execute JavaScript within the context of the pages

### 2015-09-16 research_class/paper.tex
Fix grammar: "The test script files are..."
Maybe I misunderstood, but I took this to be your intent.
Fix grammar: s/reliable/reliably

@@
 a parameter indicating the number of nodes/pages to launch and test
-again. The tests scripts files are:
+again. The test script files are:
 
@@
         process that is handling all the cluster nodes is not able to
-        give enough CPU time to the leader node to reliable send
+        give enough CPU time to the leader node to reliably send
         appendEntries RPCs to all the other nodes within the election

### 2015-09-16 research_class/paper.tex
Fix grammar: add "the"

@@
         sizes.
-    \item \emph{Figure 10 (a)}: During this test cluster was unable to
+    \item \emph{Figure 10 (a)}: During this test the cluster was unable to
         activate at cluster size 21 when the election timeout value is

### 2015-09-16 research_class/paper.tex
Make rhetorical question into a question.
Simplify wording slightly.

@@
     from two nodes down to one node. Can usefulness and availability
-    be improved without sacrificing Raft safety constraints.
+    be improved without sacrificing Raft safety constraints?
 \item Test the system in more diverse network environments. Determine
@@
     communication.  Determine if corporate firewalls make this as
-    difficult as is implied in some of the documentation and standards
-    texts. %TODO/CITE
+    difficult as some of the documentation and standards texts imply. %TODO/CITE
 \item Perform quantitative and real-world testing to better

### 2015-09-16 research_class/paper.tex
Fix grammar.
Clarify the description of testing modes.
These two testing modes are listed in this paragraph, as well as in the "Figures
8, 9, and 10" paragraph of section 5.2.2. I think the wording in the second case
was clearer, so I pulled it back and re-used it in the previous case.

@@
 from 3 to 21 nodes): \emph{test/test\_up.js} and
-\emph{test/test\_kill\_nodes.js} in kill one (leader) node and kill
-half nodes. Each of these tests was run with three different Raft election
+\emph{test/test\_kill\_nodes.js} in both modes (kill only leader and kill just
+under half the nodes).
+Each of these tests was run with three different Raft election
 timeout values: 100ms, 500ms, 1000ms. The
@@
     blobs. For example, the data with full consensus might contain
-    hashes that to the refer to bulk data.
+    hashes that refer to bulk data.
 \item Implement a Node.js based Raft over WebRTC application. This

### 2015-09-17 rtc.js
Lower election and addRemoveServer timeout.

@@ -3,7 +3,7 @@
 // Parmeters
 var verbose = 1,
     debug = true,
-    electionTimeout = 1000,
+    electionTimeout = 500,
     // addRemoveServersRetry: smaller values increase the number of
     // PENDING_CONFIG_CHANGE that will happen when there are multiple
     // changes pending (more than one server added or removed). Larger
@@ -14,7 +14,7 @@ var verbose = 1,
     // server map to identify nodes that need to be added or dropped.
     // Larger values means that it takes longer to detect when cluster
     // changes happen.
-    addRemoveServersPoll = 500;
+    addRemoveServersPoll = 100;
 
 // Global state
 var messages = document.getElementById('messages'),

### 2015-09-17 test/rtctwst.js
Add wait_kill_node.js test.

@@ -26,7 +26,7 @@ RtcTwst.prototype.dockerPage = function(url, opts) {
         url,
         timeout];
 
-    console.log(prefix + 'launching ' + url);
+    console.log(prefix + 'launching slimerjs in docker');
     var page = spawn('docker', docker_args);
 
     page.stdout.on('data', function(chunk) {
@@ -59,6 +59,24 @@ RtcTwst.prototype.get_node_info = function(timeout, callback) {
     }, {timeout: timeout}, callback);
 }
 
+RtcTwst.prototype.get_leader_idx = function(timeout, callback) {
+    var self = this;
+    self.get_node_info(timeout, function(status, nodes) {
+        if (status) {
+            for (var cid in nodes) {
+                var node = nodes[cid].data;
+                if (node.state === 'leader') {
+                    callback(true, cid);
+                    return;
+                }
+            }
+            callback(false, null);
+        } else {
+            callback(false, null);
+        }
+    });
+}
+
 RtcTwst.prototype.wait_cluster_up = function(timeout, callback) {
     var self = this,
         server_count = Object.keys(self.clients).length,
@@ -73,7 +91,7 @@ RtcTwst.prototype.wait_cluster_up = function(timeout, callback) {
             // Pull out some stats
             var states = {leader:[], candidate:[], follower:[]},
                 nodeCnts = [];
-            for (var i=0; i < server_count; i++) {
+            for (var i in nodes) {
                 var node = nodes[i].data;
                 if (node) {
                     //console.log("node:", node, "node.state:", node.state);

### 2015-09-17 test/test_kill_nodes.js

@@ -44,7 +44,7 @@ common.wait_cluster_up(pages, server_count, up_timeout, function(status, nodes, 
             pages.splice(kill_idx, 1); // NOTE: mutates pages array in place
         }
 
-        console.log('Waiting for cluster to stabalize');
+        console.log('Waiting for cluster to stabilize');
         common.wait_cluster_up(pages, server_count-kill_count, up_timeout, function(status, nodes, elapsed) {
             if (status) {
                 console.log('Cluster recovered after ' + elapsed + 'ms');

### 2015-09-17 test/wait_kill_nodes.js

@@ -6,12 +6,13 @@
 //
 // - Now run the test using the rtc_server listen address and the
 //   number of nodes:
-//     node test/wait_chat_propagate.js 10.0.01:8001 3
+//     node test/wait_kill_nodes.js 10.0.01:8001 3
 
 var RtcTwst = require('./rtctwst').RtcTwst,
     rtwst = new RtcTwst(),
     rtc_address = process.argv[2],
     clientCount = process.argv.length >= 4 ? parseInt(process.argv[3]) : 1,
+    killCount = (process.argv.length >= 5) ? parseInt(process.argv[4]) : parseInt((clientCount-1)/2, 10),
     timeout = (clientCount*20)*1000;
 
 var channel = Math.round(Math.random()*100000);
@@ -21,6 +22,11 @@ var url = 'http://' + rtc_address +
           '&twst_address=' + rtwst.getAddress() + '&paused=1';
 
 var pages = [];
+
+if (killCount > parseInt((clientCount-1)/2, 10)) {
+    console.log('Kill count must be less than half of client count');
+    process.exit(2);
+}
 
 for (var i=0; i<clientCount; i++) {
     pages.push(rtwst.dockerPage(url, {prefix: 'p' + i + ': ',
@@ -41,8 +47,8 @@ function do_start() {
     rtwst.wait_cluster_up(timeout, function(status, nodes, elapsed) {
         if (status) {
             console.log('Cluster is up after ' + elapsed + 'ms');
-            console.log('Delaying for 3 seconds before sending message');
-            setTimeout(do_chat, 3000);
+            console.log('Delaying for 3 seconds before killing node(s)');
+            setTimeout(do_kill, 3000);
         } else {
             console.log('Cluster failed to come up after ' +
                         elapsed + 'ms');
@@ -51,30 +57,35 @@ function do_start() {
     });
 }
 
-function do_chat() {
-    rtwst.send(function() {
-        $('#talk').val("test line");
-        //sendLine();
-        $('#send').click();
-    }, {index: pages.length-1});
-    rtwst.wait_cluster_predicate(timeout, function() {
-        var sm = node._self.stateMachine;
-        //console.log(nodeId + " stateMachine: " + JSON.stringify(sm));
-        if ('history' in sm && 'value' in sm.history) {
-            var lines = sm.history.value,
-                m = lines[lines.length-1].match(/.*: test line$/);
-            return m ? true : false;
-        } else {
-            return false;
-        }
-    }, function(status, results, elapsed) {
-        if (status) {
[…]

### 2015-09-19 research_class/paper.tex
Restore neutral "their" but with plural object.
- But even singular "their" is acceptable:
    http://blog.oxforddictionaries.com/2012/06/he-or-she-versus-they/
Fix typo in previous commit.
Try again to fix typos.
IEEEtran style bibliography.

@@
-\documentclass[9pt]{extarticle}
-
-\usepackage{multicol}
-\setlength{\columnsep}{0.8cm}
-
-\usepackage{geometry}
-\geometry{margin=2.0cm}
-
-%\usepackage[utf8]{inputenc}
-\usepackage[backend=bibtex]{biblatex}
-
+\documentclass[conference,compsoc]{./IEEEtran/IEEEtran}
+
+% *** CITATION PACKAGES ***
+% IEEE Computer Society needs nocompress option
+% requires cite.sty v4.0 or later (November 2003)
+\usepackage[nocompress]{cite}
+
+% *** ALIGNMENT PACKAGES ***
+\usepackage{array}
+
+% *** SUBFIGURE PACKAGES ***
+\usepackage[caption=false,font=footnotesize,labelfont=sf,textfont=sf]{subfig}
+% The latest version and documentation of subfig.sty can be obtained at:
+% http://www.ctan.org/pkg/subfig
+
+%%\usepackage{geometry}
+%%\geometry{margin=2.0cm}
+%%
+%%\usepackage[backend=bibtex]{biblatex}
+%%
 \usepackage{hyperref}
-\usepackage{caption}
-\usepackage{subcaption}
-
-\usepackage{authblk}
-
-\usepackage{microtype}
+%%\usepackage{caption}
+%%\usepackage{subcaption}
+%%
+%%\usepackage{authblk}
+%%
+%%\usepackage{microtype}
 
@@
 
-\title{Raft Distributed Consensus Algorithm over WebRTC (Web Real-Time Communication)}
-
-\author{Joel D. Martin}
-\author{David Levine}
-\affil{University of Texas at Arlington}
-
-\addbibresource{references/raft.bib}
-\addbibresource{references/w3c.bib}
-\addbibresource{references/ietf.bib}
-\addbibresource{references/other.bib}
+%%\title{Raft Distributed Consensus Algorithm over WebRTC (Web Real-Time Communication)}
+%%
+%%\author{Joel D. Martin}
+%%\author{David Levine}
+%%\affil{University of Texas at Arlington}
+%%
+%%\addbibresource{references/raft.bib}
+%%\addbibresource{references/w3c.bib}
+%%\addbibresource{references/ietf.bib}
+%%\addbibresource{references/other.bib}
+
+% correct bad hyphenation here
+\hyphenation{op-tical net-works semi-conduc-tor}
 
@@
 
+\title{Raft Distributed Consensus Algorithm over\\ WebRTC (Web Real-Time Communication)}
+
+\author{\IEEEauthorblockN{Joel Martin}
+\IEEEauthorblockA{Department of Computer Science and Engineering\\
+University of Texas at Arlington\\
+Arlington, TX 76010\\
+Email: joel@martintribe.org}
+\and
+\IEEEauthorblockN{David Levine}
+\IEEEauthorblockA{Department of Computer Science and Engineering\\
+University of Texas at Arlington\\
+Arlington, TX 76010\\
+Email: levine@cse.uta.edu}}
+
 \maketitle
-
-\begin{multicols}{2}
 
@@
 
-\end{multicols}
-
-\begin{figure*}[h]
+\begin{figure*}[!t]
 \centerline{\includegraphics[width=0.8\textwidth]{imgs/raft_rtc_architecture}}
@@
 \end{figure*}
-
-\begin{multicols}{2}
 
@@
     \item presentation and control
-    \item and local system support functions
+    \item local system support functions
 \end{enumerate}
@@
 The PeerJS organization provides a cloud-hosted version of a PeerJS
-compatible signaling server, however an application developer can also
-run her own signaling service using PeerJS PeerServer.
-
-\end{multicols}
-
-\begin{figure*}[h]
-\centerline{\includegraphics[width=0.8\textwidth]{imgs/raft_rtc_sequence}}
+compatible signaling server, however application developers can also
+run their own signaling service using PeerJS PeerServer.
+
+%\begin{figure*}[!t]
+\begin{figure}[!t]
+%\centerline{\includegraphics[width=0.8\textwidth]{imgs/raft_rtc_sequence}}
+\centerline{\includegraphics[width=0.45\textwidth]{imgs/raft_rtc_sequence}}
 \caption{Raft over RTC Sequence Diagram}
   \label{fig:raft_rtc_sequence}
-\end{figure*}
-
-\begin{multicols}{2}
+\end{figure}
 
@@
 
-\begin{center}
+\begin{figure}[!t]
+    \centering
     \includegraphics[width=0.45\textwidth]{imgs/browser_1.png}
-    \captionof{figure}{Basic app with one Node/Tab}
+    %\captionof{figure}{Basic app with one Node/Tab}
+    \caption{Basic app with one Node/Tab}
     \label{fig:browser_1}
-\end{center}
+\end{figure}
 
@@
 
-\begin{center}
+\begin{figure}[!t]
+    \centering
     \includegraphics[width=0.45\textwidth]{imgs/browser_5.png}
-    \captionof{figure}{Basic app with Five Nodes/Tabs}
+    %\captionof{figure}{Basic app with Five Nodes/Tabs}
+    \caption{Basic app with Five Nodes/Tabs}
     \label{fig:browser_5}
-\end{center}
+\end{figure}
 
@@
 
-\begin{center}
+\begin{figure}[!t]
+    \centering
     \includegraphics[width=0.45\textwidth]{imgs/chat_1a.png}
-    \captionof{figure}{Chat app with one Node/Tab}
+    %\captionof{figure}{Chat app with one Node/Tab}
+    \caption{Chat app with one Node/Tab}
     \label{fig:browser_chat_1}
-\end{center}
+\end{figure}
 
@@
 
-\begin{center}
+\begin{figure}[!t]
+    \centering
     \includegraphics[width=0.45\textwidth]{imgs/chat_5a.png}
-    \captionof{figure}{Chat app with Five Nodes/Tabs}
+    %\captionof{figure}{Chat app with Five Nodes/Tabs}
+    \caption{Chat app with Five Nodes/Tabs}
     \label{fig:browser_chat_5}
-\end{center}
+\end{figure}
 
@@
 
-\begin{center}
+\begin{figure}[!t]
     \begin{tikzpicture}
@@
     \end{tikzpicture}
-\captionof{figure}{Propagate$,$ Commit$,$ and Apply Command}
+\caption{Propagate$,$ Commit$,$ and Apply Command}
   \label{fig:propagate}
-\end{center}
+\end{figure}
 
@@
 
-%\end{multicols}
-
+%\begin{figure*}[!t]
 \begin{figure*}[h]
-    \begin{subfigure}[b]{0.30\textwidth}
     \centering
-        \resizebox{\linewidth}{!}{
+    %\begin{subfigure}[b]{0.30\textwidth}
+    \subfloat[100ms election timer]{
+        \centering
+        \resizebox{0.3\linewidth}{!}{
             \begin{tikzpicture}
             \begin{axis}[%
-                height=1.5\textwidth, width=1.4\textwidth,
+                %height=1.5\textwidth, width=1.4\textwidth,
                 xlabel={Nodes}, ylabel={Time (seconds)},
@@
         }
-        \caption{100ms election timer}
         \label{fig:A}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.30\textwidth}
+    }
+    %\begin{subfigure}[b]{0.30\textwidth}
+    \subfloat[500ms election timer]{
         \centering
-        \resizebox{\linewidth}{!}{
+        \resizebox{0.3\linewidth}{!}{
             \begin{tikzpicture}
             \begin{axis}[%
-                height=1.5\textwidth, width=1.4\textwidth,
+                %height=1.5\textwidth, width=1.4\textwidth,
                 xlabel={Nodes}, ylabel={Time (seconds)},
@@
         }
-        \caption{500ms election timer}
         \label{fig:B}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.30\textwidth}
+    }
+    %\begin{subfigure}[b]{0.30\textwidth}
+    \subfloat[1000ms election timer]{
         \centering
-        \resizebox{\linewidth}{!}{
+        \resizebox{0.3\linewidth}{!}{
             \begin{tikzpicture}
             \begin{axis}[%
-                height=1.5\textwidth, width=1.4\textwidth,
+                %height=1.5\textwidth, width=1.4\textwidth,
                 xlabel={Nodes}, ylabel={Time (seconds)},
@@
         }
-        \caption{1000ms election timer}
         \label{fig:C}
-    \end{subfigure}
+    }
     \caption{Initial Cluster Activation}
@@
 
-    \begin{subfigure}[b]{0.30\textwidth}
+    %\begin{subfigure}[b]{0.30\textwidth}
+    \subfloat[100ms election timer]{
         \centering
-        \resizebox{\linewidth}{!}{
+        \resizebox{0.3\linewidth}{!}{
             \begin{tikzpicture}
             \begin{axis}[%
-                height=1.5\textwidth, width=1.4\textwidth,
+                %height=1.5\textwidth, width=1.4\textwidth,
                 xlabel={Nodes}, ylabel={Time (seconds)},
@@
         }
-        \caption{100ms election timer}
         \label{fig:D}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.30\textwidth}
+    }
+    %\begin{subfigure}[b]{0.30\textwidth}
+    \subfloat[500ms election timer]{
         \centering
-        \resizebox{\linewidth}{!}{
+        \resizebox{0.3\linewidth}{!}{
             \begin{tikzpicture}
             \begin{axis}[%
-                height=1.5\textwidth, width=1.4\textwidth,
+                %height=1.5\textwidth, width=1.4\textwidth,
                 xlabel={Nodes}, ylabel={Time (seconds)},
@@
         }
-        \caption{500ms election timer}
         \label{fig:E}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.30\textwidth}
+    }
+    %\begin{subfigure}[b]{0.30\textwidth}
+    \subfloat[1000ms election timer]{
         \centering
-        \resizebox{\linewidth}{!}{
+        \resizebox{0.3\linewidth}{!}{
             \begin{tikzpicture}
             \begin{axis}[%
-                height=1.5\textwidth, width=1.4\textwidth,
+                %height=1.5\textwidth, width=1.4\textwidth,
                 xlabel={Nodes}, ylabel={Time (seconds)},
@@
         }
-        \caption{1000ms election timer}
         \label{fig:F}
-    \end{subfigure}
+    }
     \caption{Recovery from Leader Failure}
@@
         }
-        \caption{100ms election timer}
         \label{fig:G}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.30\textwidth}
+    }
+    %\begin{subfigure}[b]{0.30\textwidth}
+    \subfloat[500ms election timer]{
         \centering
-        \resizebox{\linewidth}{!}{
+        \resizebox{0.3\linewidth}{!}{
             \begin{tikzpicture}
             \begin{axis}[%
-                height=1.5\textwidth, width=1.4\textwidth,
+                %height=1.5\textwidth, width=1.4\textwidth,
                 xlabel={Nodes}, ylabel={Time (seconds)},
@@
         }
-        \caption{500ms election timer}
         \label{fig:H}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.30\textwidth}
+    }
+    %\begin{subfigure}[b]{0.30\textwidth}
+    \subfloat[1000ms election timer]{
         \centering
-        \resizebox{\linewidth}{!}{
+        \resizebox{0.3\linewidth}{!}{
             \begin{tikzpicture}
             \begin{axis}[%
-                height=1.5\textwidth, width=1.4\textwidth,
+                %height=1.5\textwidth, width=1.4\textwidth,
                 xlabel={Nodes}, ylabel={Time (seconds)},
@@
         }
-        \caption{1000ms election timer}
         \label{fig:I}
-    \end{subfigure}
+    }
     \caption{Recovery from Half Cluster Failure}
@@
 \end{figure*}
-
-%\begin{multicols}{2}
 
@@
 excellent input and oversight of this project. I would also like to
-thank Diego Ongaro for his amazing work on the Raft protocol and paper
-\cite[p.~215]{lamport78} which made implementing
-a Raft implementation a pleasure \cite{lamport78}.
-There are too many people involved in the creation and standardization
-of WebRTC but I would like to thank them for their tireless work to
-make browser-to-browser communication possible. Last but not least,
-I would like to thank my family for putting up with the long nights
-and weekends I spent on this project. Thanks!
-
-\end{multicols}
-
-\printbibliography[title={General references},notkeyword=rtc-general,notkeyword=rtc-transport,notkeyword=rtc-framing]
-
-\nocite{draft-rtcweb-overview}
-\nocite{draft-rtcweb-security-arch}
-\nocite{draft-rtcweb-security}
-\printbibliography[title={WebRTC general references},keyword=rtc-general]
-
-\nocite{draft-rtcweb-transports}
-\nocite{draft-rtcweb-qos}
-\nocite{RFC4960:sctp}
-\nocite{RFC3758:sctp-reliability}
-\nocite{draft-sctp-dtls-encaps}
-\nocite{draft-mmusic-sctp-sdp}
-\nocite{draft-sctp-ndata}
-\nocite{draft-rtcweb-rtp-usage}
-\nocite{draft-rtcweb-data-channel}
-\nocite{draft-rtcweb-data-protocol}
-\nocite{draft-rtcweb-alpn}
-\nocite{RFC5245:ice}
-\nocite{RFC5128:p2p-nat}
-\nocite{RFC5389:stun}
-\nocite{RFC3489:stun-udp}
-\nocite{RFC5766:turn}
-\nocite{RFC6156:turn-ipv6}
-\nocite{RFC6544:tcp-ice}
-\nocite{RFC4571:rtp-rtcp}
-\nocite{RFC5764:dtls-srtp}
-\printbibliography[title={WebRTC data transport},keyword=rtc-transport]
-
-\nocite{RFC3550:rtp}
-\nocite{RFC3711:srtp}
-%\nocite{draft-rtcweb-rtp-usage}
-%\nocite{draft-rtcweb-data-channel}
-%\nocite{draft-rtcweb-data-protocol}
-\printbibliography[title={WebRTC framing and security},keyword=rtc-framing]
-
-\nocite{RFC2327:sdp}
-\nocite{RFC3264:sdp-offer-answer}
-\nocite{draft-rtcweb-jsep}
-%\nocite{RFC5245:ice}
-\nocite{draft-mmusic-trickle-ice}
-\nocite{RFC5763:srtp-using-dtls}
-\printbibliography[title={WebRTC connection management},keyword=rtc-connection]
-
-%%%
-
-%\begin{enumerate}
-%\item Data formats (audio)
-%    \begin{enumerate}
-%    \item \href{https://tools.ietf.org/html/draft-cbran-rtcweb-codec-02}{WebRTC: WebRTC Codec and Media Processing Requirements}
-%    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-audio-07}{WebRTC: Audio Codec and Processing Requirements}
-%    \item \href{https://tools.ietf.org/html/rfc6716}{RFC6716: Opus Audio Codec}
-%    \item \href{https://tools.ietf.org/html/draft-ietf-payload-rtp-opus-11}{RTP Payload Format for Opus Speech and Audio Codec}
-%    \item \href{https://tools.ietf.org/html/rfc3551}{RFC3551: RTP Profile for Audio and Video Conference}
-%    \item \href{https://tools.ietf.org/html/rfc4733}{RFC4733: RTP Payload for DTMF Digits, Telephony Tones, and Telephony Signals}
-%    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-audio-codecs-for-interop-01}{WebRTC: Additional WebRTC audio codecs for interoperability with legacy networks}
-%    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-video-05}{WebRTC: Video Processing and Codec Requirements}
-%    \item \href{https://tools.ietf.org/html/rfc6386}{RFC6386: VP8 Data Format and Decoding Guide}
-%    \item \href{http://www.itu.int/rec/T-REC-H.264}{ITU H.264: Advanced video coding for generic audiovisual services}
-%    \item \href{https://tools.ietf.org/html/draft-ietf-payload-vp8-15}{RTP Payload Format for VP8 Video}
-%    \item \href{https://tools.ietf.org/html/rfc6236}{RFC6236: Negotiation of Generic Image Attributes in the Session Description Protocol (SDP)}
-%    \end{enumerate}
-%\end{enumerate}
+thank Diego Ongaro for his amazing work on the Raft protocol and paper which
+made implementing a Raft implementation a pleasure.  There are too many people
+involved in the creation and standardization of WebRTC but I would like to
+thank them for their tireless work to make browser-to-browser communication
+possible. Last but not least, I would like to thank my family for putting up
+with the long nights and weekends I spent on this project. Thanks!
+
+\bibliographystyle{./IEEEtranBST2/IEEEtran}
+\bibliography{./references/ietf,./references/w3c,./references/raft,./references/other}
+
+%% % Bibliography with multiplie sectiosn and references not actually
+%% % cited in the paper either
+
+%% \addbibresource{references/raft.bib}
+%% \addbibresource{references/w3c.bib}
+%% \addbibresource{references/ietf.bib}
+%% \addbibresource{references/other.bib}
+%%
+%% \printbibliography[title={General references},notkeyword=rtc-general,notkeyword=rtc-transport,notkeyword=rtc-framing]
+%% 
+%% \nocite{draft-rtcweb-overview}
+%% \nocite{draft-rtcweb-security-arch}
+%% \nocite{draft-rtcweb-security}
+%% \printbibliography[title={WebRTC general references},keyword=rtc-general]
+%% 
+%% \nocite{draft-rtcweb-transports}
+%% \nocite{draft-rtcweb-qos}
+%% \nocite{RFC4960:sctp}
+%% \nocite{RFC3758:sctp-reliability}
+%% \nocite{draft-sctp-dtls-encaps}
+%% \nocite{draft-mmusic-sctp-sdp}
+%% \nocite{draft-sctp-ndata}
+%% \nocite{draft-rtcweb-rtp-usage}
+%% \nocite{draft-rtcweb-data-channel}
+%% \nocite{draft-rtcweb-data-protocol}
+%% \nocite{draft-rtcweb-alpn}
+%% \nocite{RFC5245:ice}
+%% \nocite{RFC5128:p2p-nat}
+%% \nocite{RFC5389:stun}
+%% \nocite{RFC3489:stun-udp}
+%% \nocite{RFC5766:turn}
+%% \nocite{RFC6156:turn-ipv6}
+%% \nocite{RFC6544:tcp-ice}
+%% \nocite{RFC4571:rtp-rtcp}
+%% \nocite{RFC5764:dtls-srtp}
+%% \printbibliography[title={WebRTC data transport},keyword=rtc-transport]
+%% 
+%% \nocite{RFC3550:rtp}
+%% \nocite{RFC3711:srtp}
+%% %\nocite{draft-rtcweb-rtp-usage}
+%% %\nocite{draft-rtcweb-data-channel}
+%% %\nocite{draft-rtcweb-data-protocol}
+%% \printbibliography[title={WebRTC framing and security},keyword=rtc-framing]
+%% 
+%% \nocite{RFC2327:sdp}
+%% \nocite{RFC3264:sdp-offer-answer}
+%% \nocite{draft-rtcweb-jsep}
+%% %\nocite{RFC5245:ice}
+%% \nocite{draft-mmusic-trickle-ice}
+%% \nocite{RFC5763:srtp-using-dtls}
+%% \printbibliography[title={WebRTC connection management},keyword=rtc-connection]
+
+%% % WebRTC standards that aren't relevant
+
+%% %\begin{enumerate}
+%% %\item Data formats (audio)
+%% %    \begin{enumerate}
+%% %    \item \href{https://tools.ietf.org/html/draft-cbran-rtcweb-codec-02}{WebRTC: WebRTC Codec and Media Processing Requirements}
+%% %    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-audio-07}{WebRTC: Audio Codec and Processing Requirements}
+%% %    \item \href{https://tools.ietf.org/html/rfc6716}{RFC6716: Opus Audio Codec}
+%% %    \item \href{https://tools.ietf.org/html/draft-ietf-payload-rtp-opus-11}{RTP Payload Format for Opus Speech and Audio Codec}
+%% %    \item \href{https://tools.ietf.org/html/rfc3551}{RFC3551: RTP Profile for Audio and Video Conference}
+%% %    \item \href{https://tools.ietf.org/html/rfc4733}{RFC4733: RTP Payload for DTMF Digits, Telephony Tones, and Telephony Signals}
+%% %    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-audio-codecs-for-interop-01}{WebRTC: Additional WebRTC audio codecs for interoperability with legacy networks}
+%% %    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-video-05}{WebRTC: Video Processing and Codec Requirements}
+%% %    \item \href{https://tools.ietf.org/html/rfc6386}{RFC6386: VP8 Data Format and Decoding Guide}
+%% %    \item \href{http://www.itu.int/rec/T-REC-H.264}{ITU H.264: Advanced video coding for generic audiovisual services}
+%% %    \item \href{https://tools.ietf.org/html/draft-ietf-payload-vp8-15}{RTP Payload Format for VP8 Video}
+%% %    \item \href{https://tools.ietf.org/html/rfc6236}{RFC6236: Negotiation of Generic Image Attributes in the Session Description Protocol (SDP)}
+%% %    \end{enumerate}
+%% %\end{enumerate}
 

### 2015-09-19 research_class/paper.tex
Build full and minimal versions.
- WebRTC citations in full version.

@@
 \documentclass[conference,compsoc]{./IEEEtran/IEEEtran}
+
+%% Configure which sections are included:
+%\newcommand*{\OPTIONAL}{}
 
@@
 is responsible for specifying and/or standardizing (where necessary)
-the protocols that are part of WebRTC. In many cases an existing
-protocol is applicable and is referenced or extended as part of the
-WebRTC suite. %TODO/CITE
+the protocols that are part of WebRTC \cite{draft-rtcweb-overview}
+\cite{draft-rtcweb-security-arch} \cite{draft-rtcweb-security}. In many
+cases an existing protocol is applicable and is referenced or
+extended as part of the WebRTC suite. %TODO/CITE
 
@@
     \item data transport
+        \ifdefined\OPTIONAL
+        \\
+        \cite{draft-rtcweb-transports}
+        \cite{draft-rtcweb-qos}
+        \cite{RFC4960:sctp}
+        \cite{RFC3758:sctp-reliability}
+        \cite{draft-sctp-dtls-encaps}
+        \cite{draft-mmusic-sctp-sdp}
+        \cite{draft-sctp-ndata}
+        \cite{draft-rtcweb-rtp-usage}
+        \cite{draft-rtcweb-data-channel}
+        \cite{draft-rtcweb-data-protocol}
+        \cite{draft-rtcweb-alpn}
+        \cite{RFC5245:ice}
+        \cite{RFC5128:p2p-nat}
+        \cite{RFC5389:stun}
+        \cite{RFC3489:stun-udp}
+        \cite{RFC5766:turn}
+        \cite{RFC6156:turn-ipv6}
+        \cite{RFC6544:tcp-ice}
+        \cite{RFC4571:rtp-rtcp}
+        \cite{RFC5764:dtls-srtp}
+        \fi
     \item data framing and security
+        \ifdefined\OPTIONAL
+        \\
+        \cite{RFC3550:rtp}
+        \cite{RFC3711:srtp}
+        \cite{draft-rtcweb-rtp-usage}
+        \cite{draft-rtcweb-data-channel}
+        \cite{draft-rtcweb-data-protocol}
+        \fi
     \item connection management
+        \ifdefined\OPTIONAL
+        \\
+        \cite{RFC2327:sdp}
+        \cite{RFC3264:sdp-offer-answer}
+        \cite{draft-rtcweb-jsep}
+        \cite{RFC5245:ice}
+        \cite{draft-mmusic-trickle-ice}
+        \cite{RFC5763:srtp-using-dtls}
+        \fi
     \item data formats
@@
 The last three functionality groups focus on interactive audio and video
-applications and are less relevant to Raft over WebRTC. The WebRTC
-section of the bibliography %TODO: reference link
-has a (partial) list of WebRTC specific protocol drafts and other
-related protocol standards for the first three functionality groups.
+applications and are less relevant to Raft over WebRTC. \ifdefined\OPTIONAL
+Some of the important standards for the first three functionality
+groups are cited above. \fi
 
@@
 
+\ifdefined\OPTIONAL
+
 \subsection{PeerJS}
@@
 
+\fi  % \ifdefined\OPTIONAL
+
 %\begin{figure*}[!t]
@@
 \section {Design and Implementation}
+
+\ifdefined\OPTIONAL
 
@@
 
+\fi  % \ifdefined\OPTIONAL
+
 \subsection{Server}
@@
 
+\ifdefined\OPTIONAL
+
 \subsection{Manual Testing}
@@
 capable nodes.
+
+\fi  % \ifdefined\OPTIONAL
 

### 2015-09-19 chat.js
Fix wait_chat_prop test. Better logging.

@@ -84,7 +84,7 @@ function startChat() {
             sendLine();
         }
     };
-    start({applyCmd: applyCmd});
+    startRaft({applyCmd: applyCmd});
     flushSends();
 
 }

### 2015-09-19 rtc.js

@@ -3,7 +3,7 @@
 // Parmeters
 var verbose = 1,
     debug = true,
-    electionTimeout = 500,
+    electionTimeout = 1000,
     // addRemoveServersRetry: smaller values increase the number of
     // PENDING_CONFIG_CHANGE that will happen when there are multiple
     // changes pending (more than one server added or removed). Larger
@@ -207,7 +207,7 @@ function addRemoveServersAsync() {
     }
 }
 
-function start(opts) {
+function startRtc(opts, nextCallback) {
     opts = opts || {};
 
     var peer = new Peer({host: location.hostname,
@@ -244,23 +244,9 @@ function start(opts) {
             });
         });
 
-        // Create the local raft.js node
-        log("Starting Raft node");
-        var node_opts = {verbose: verbose,
-                         debug: debug,
-                         log: log,
-                         serverData: nodeMap,
-                         firstServer: firstServer,
-                         sendRPC: rtcSend,
-                         electionTimeout: electionTimeout,
-                         clientRequestResponse: clientRequestResponse};
-        for (var k in opts) { node_opts[k] = opts[k] };
-
-        node = new local.RaftServerLocal(nodeId, node_opts);
-
-        // Start scanning for new servers
-        addRemoveServersAsync();
-        updateStats();
+        if (nextCallback) {
+            nextCallback(opts);
+        }
     });
 
     peer.on('error', function(e) {
@@ -278,4 +264,24 @@ function start(opts) {
             delete nodeMap[conn.peer];
         });
     });
+}
+
+function startRaft(opts) {
+    // Create the local raft.js node
+    log("Starting Raft node");
+    var node_opts = {verbose: verbose,
+                     debug: debug,
+                     log: log,
+                     serverData: nodeMap,
+                     firstServer: firstServer,
+                     sendRPC: rtcSend,
+                     electionTimeout: electionTimeout,
+                     clientRequestResponse: clientRequestResponse};
+    for (var k in opts) { node_opts[k] = opts[k] };
+
+    node = new local.RaftServerLocal(nodeId, node_opts);
+
+    // Start scanning for new servers
+    addRemoveServersAsync();
+    updateStats();
 }

### 2015-09-19 test/rtctwst.js

@@ -3,7 +3,26 @@ var spawn = require('child_process').spawn,
     RtcTwst = null;
 
 // Subclass Twst
-exports.RtcTwst = RtcTwst = function(opts) { Twst.call(this, opts); };
+exports.RtcTwst = RtcTwst = function(opts) {
+    Twst.call(this, opts);
+    //rtwst.on('return',        function(idx, msg) { console.log('RETURN:', idx, msg); });
+    //rtwst.on('callback',      function(idx, msg) { console.log('CALLBACK:', idx, msg); });
+    this.on('error', function(idx, msg) {
+        console.error(idx + ' ERROR:', msg.data);
+    });
+    this.on('console.log', function(idx, msg) {
+        console.log(idx + ' CONSOLE.LOG:', msg.data.join(' '));
+    });
+    this.on('console.warn', function(idx, msg) {
+        console.warn(idx + ' CONSOLE.WARN:', msg.data.join(' '));
+    });
+    this.on('console.error', function(idx, msg) {
+        console.error(idx + ' CONSOLE.ERROR:', msg.data.join(' '));
+    });
+    this.on('close', function(idx, msg) {
+        console.log(idx + ' CLOSE:', msg.data);
+    });
+};
 RtcTwst.prototype = Object.create(Twst.prototype);
 RtcTwst.prototype.constructor = RtcTwst;
 
@@ -30,9 +49,11 @@ RtcTwst.prototype.dockerPage = function(url, opts) {
     var page = spawn('docker', docker_args);
 
     page.stdout.on('data', function(chunk) {
-        var line = chunk.toString('utf8').replace(/\r\n$/,'');
-        if (line === '') { return; }
-        console.log(prefix + line);
+        if (opts.verbose) {
+            var line = chunk.toString('utf8').replace(/\r\n$/,'');
+            if (line === '') { return; }
+            console.log(prefix + line);
+        }
     });
 
     page.on('close', function (code) {
@@ -83,9 +104,10 @@ RtcTwst.prototype.wait_cluster_up = function(timeout, callback) {
         start_time = Date.now();
     var checkfn = function () {
         // Gather data from the nodes
-        self.get_node_info(2000, function(status, nodes) {
+        self.get_node_info(2000+timeout/10, function(status, nodes) {
             var elapsed = Date.now() - start_time;
             if (!status) {
+                console.log("get_node_info timed out");
                 callback(false, nodes, elapsed)
             }
             // Pull out some stats

### 2015-09-19 test/wait_chat_propagate.js

@@ -37,7 +37,7 @@ function poll() {
 }
 
 function do_start() {
-    rtwst.broadcast('startChat();');
+    rtwst.broadcast('startChat()');
     rtwst.wait_cluster_up(timeout, function(status, nodes, elapsed) {
         if (status) {
             console.log('Cluster is up after ' + elapsed + 'ms');
@@ -56,7 +56,7 @@ function do_chat() {
         $('#talk').val("test line");
         //sendLine();
         $('#send').click();
-    }, {index: pages.length-1});
+    }, {index: rtwst.clients.length-1});
     rtwst.wait_cluster_predicate(timeout, function() {
         var sm = node._self.stateMachine;
         //console.log(nodeId + " stateMachine: " + JSON.stringify(sm));

### 2015-09-19 test/wait_kill_nodes.js
Fix wait_kill_nodes and wait_start.

@@ -43,7 +43,7 @@ function poll() {
 }
 
 function do_start() {
-    rtwst.broadcast('startChat();');
+    rtwst.broadcast('startChat()');
     rtwst.wait_cluster_up(timeout, function(status, nodes, elapsed) {
         if (status) {
             console.log('Cluster is up after ' + elapsed + 'ms');

### 2015-09-19 test/wait_start.js

@@ -17,6 +17,7 @@ var RtcTwst = require('./rtctwst').RtcTwst,
 var channel = Math.round(Math.random()*100000);
 var url = 'http://' + rtc_address +
           '/chat.html?channel=' + channel +
+          '&console_logging=true' +
           '&twst_address=' + rtwst.getAddress() + '&paused=1';
 
 var pages = [];
@@ -36,7 +37,7 @@ function poll() {
 }
 
 function do_start() {
-    rtwst.broadcast('start();');
+    rtwst.broadcast('startChat();');
     rtwst.wait_cluster_up(timeout, function(status, nodes, elapsed) {
         if (status) {
             console.log('Cluster is up after ' + elapsed + 'ms');

### 2015-09-20 chat.js
wait_chat_propagate: multiple message support.

@@ -11,7 +11,7 @@ function updateHistory() {
 }
 
 function applyCmd(stateMachine, cmd) {
-    log("cmd:", JSON.stringify(cmd));
+    //log("cmd:", JSON.stringify(cmd));
     switch (cmd.op) {
         case 'get':
             stateMachine[cmd.key];

### 2015-09-20 test/rtctwst.js

@@ -152,12 +152,20 @@ RtcTwst.prototype.wait_cluster_predicate = function(timeout, predicate, callback
             if (!status) {
                 callback(false, data, elapsed)
             }
-            var trueCnt = 0;
+            var falseStates = [],
+                trueStates = [],
+                trueCnt = 0;
             for (var n in data) {
-                if (data[n].data) { trueCnt += 1; }
+                if (data[n].data) {
+                    trueCnt += 1;
+                    trueStates.push(n);
+                } else {
+                    falseStates.push(n);
+                }
             }
 
-            console.log('Predicate results: ' + JSON.stringify(data) +
+            console.log('Predicate false: ' + JSON.stringify(falseStates) +
+                        ', true: ' + JSON.stringify(trueStates) +
                         ', true count: ' + trueCnt);
             // Exit if cluster is up or we timeout
             if (trueCnt >= server_count) {
@@ -165,7 +173,7 @@ RtcTwst.prototype.wait_cluster_predicate = function(timeout, predicate, callback
             } else if (elapsed > timeout) {
                 callback(false, data, elapsed);
             } else {
-                setTimeout(checkfn, 25);
+                setTimeout(checkfn, 50);
                 //setTimeout(checkfn, 2000);
             }
         });

### 2015-09-20 test/wait_chat_propagate.js

@@ -12,6 +12,7 @@ var RtcTwst = require('./rtctwst').RtcTwst,
     rtwst = new RtcTwst(),
     rtc_address = process.argv[2],
     clientCount = process.argv.length >= 4 ? parseInt(process.argv[3]) : 1,
+    msgCount = (process.argv.length >= 5) ? parseInt(process.argv[4]) : 1,
     timeout = (clientCount*20)*1000;
 
 var channel = Math.round(Math.random()*100000);
@@ -52,18 +53,32 @@ function do_start() {
 }
 
 function do_chat() {
-    rtwst.send(function() {
-        $('#talk').val("test line");
-        //sendLine();
-        $('#send').click();
-    }, {index: rtwst.clients.length-1});
+    rtwst.broadcast("window._test_msgCount = " + msgCount + ";")
+    for (var idx=0; idx < clientCount; idx++) {
+        var cnt = Math.floor((msgCount + clientCount - idx - 1) / clientCount),
+            msg = 'test msg #' + idx,
+            x = '';
+
+        if (cnt === 0) { continue }
+
+        x += 'for(var j=0; j<' + cnt + '; j++) {pendingSends.push('
+        x += '"msg #" + j + " from node ' + idx + '")}';
+        console.log('Sending to ' + idx + ': ' + x);
+        rtwst.send(x, {id: idx})
+    }
     rtwst.wait_cluster_predicate(timeout, function() {
         var sm = node._self.stateMachine;
         //console.log(nodeId + " stateMachine: " + JSON.stringify(sm));
         if ('history' in sm && 'value' in sm.history) {
-            var lines = sm.history.value,
-                m = lines[lines.length-1].match(/.*: test line$/);
-            return m ? true : false;
+            var lines = sm.history.value;
+            //console.log("lines:", lines);
+            if (lines.length === window._test_msgCount) {
+                var m = lines[lines.length-1].match(/msg .* from node .*$/);
+                //console.log("m:", m);
+                return m ? true : false;
+            } else {
+                return false
+            }
         } else {
             return false;
         }

### 2015-09-20 test/common.js
Remove old all in same process slimerjs tests.

@@ -1,141 +1,0 @@
-var webpage = require('webpage');
-
-exports.new_page = function (page_idx, firstServer, opts, callback) {
-    var query = '?channel='+ opts.channel + '&console_logging=true';
-
-    console.log("new_page: " + Date.now());
-    var page = webpage.create();
-
-    // Register external handlers
-    page.onConsoleMessage = function (msg, line, origin) {
-        console.log('CONSOLE ' + page_idx + ': ' + msg);
-    };
-    page.onError = function (msg) {
-        console.log('ERROR ' + page_idx + ': ' + msg);
-    };
-    page.onCallback = function(msg) {
-        console.log('CALLBACK ' + page_idx + ': ' + msg);
-        if (msg === 'QUIT') {
-            console.log('Normal exit');
-            slimer.exit(0);
-        }
-    };
-
-    var full_address = opts.base_address;
-
-    if (!firstServer) {
-        full_address += opts.home;
-    }
-    full_address += query + (firstServer ? '#firstServer' : '');
-
-    console.log('Opening ' + page_idx + ': ' + full_address);
-
-    page.open(full_address, function(status) {
-        if (status !== 'success') {
-            console.log('Unable to load the address!');
-            phantom.exit(1);
-        }
-        //var mainTitle = page.evaluate(function () {
-        //    return document.title;
-        //});
-        if (callback) { callback(page); }
-    });
-
-    return page;
-}
-
-exports.get_node_info = function (pages, full) {
-    var nodes = [];
-    for (var i=0; i < pages.length; i++) {
-        nodes.push(pages[i].evaluate(function(full) {
-            // Evaluates in page context so 'node' variable is implicit
-            if (typeof node !== 'undefined' && node) {
-                var data = {id: node._self.id,
-                            state: node._self.state,
-                            serverMapKeys: Object.keys(node._self.serverMap)};
-                if (full) {
-                    data.stateMachine = node._self.stateMachine;
-                    data.log = node._self.log;
-                }
-                return data;
-            } else {
-                return null;
-            }
-        }, full));
-    }
-    return nodes;
-}
-
-exports.get_leader_idx = function(nodes) {
-    for (var i=0; i < nodes.length; i++) {
-        if (nodes[i].state === 'leader') { return i; }
-    }
-}
-
-exports.show_nodes = function (nodes) {
-    for (var i=0; i < nodes.length; i++) {
-        console.log('Node ' + i + ': ');
[…]

### 2015-09-20 test/page.js

@@ -1,37 +1,0 @@
-var system = require('system'),
-    page = require('webpage').create(),
-    address = system.args[1],
-    timeout = (system.args.length >= 3) ? parseInt(system.args[2]) : 20;
-
-// Register external handlers
-page.onConsoleMessage = function (msg, line, origin) {
-    console.log("CONSOLE: " + origin + ":" + line + ", " + msg);
-};
-page.onCallback = function(msg) {
-    console.log("CALLBACK: " + msg);
-    if (msg === 'QUIT') {
-        console.log("Normal exit");
-        slimer.exit(0);
-    }
-};
-
-page.open(address, function(status) {
-    if (status !== 'success') {
-        console.log('Unable to load the address!');
-        phantom.exit(1);
-    } else {
-        window.setTimeout(function () {
-            console.log("Timeout after " + timeout + " seconds");
-            slimer.exit(1);
-        }, timeout * 1000);
-    }
-    var mainTitle = page.evaluate(function () {
-        console.log('RTCPeerConnection:', window.RTCPeerConnection);
-        console.log('mozRTCPeerConnection:', window.mozRTCPeerConnection);
-        window.callPhantom("test of callPhantom");
-        //window.callPhantom("QUIT");
-        return document.title;
-    });
-    console.log('Returned from page: ' + mainTitle);
-    //slimer.exit()
-});

### 2015-09-20 test/test_chat_propagate.js

@@ -1,64 +1,0 @@
-// Usage:
-// - Start rtc_server.js first:
-//   ./rtc_server --port 8000 --home chat.html
-//
-// - Then use a docker build of slimerjs to run the test:
-//   IP_ADDR=$(hostname -I | awk '{print $1}')
-//   docker run -it -v `pwd`/test:/test slimerjs-0.9.6 slimerjs /test/test_chat_propagate.js http://${IP_ADDR}:8000/
-
-var common = require('common'),
-    system = require('system'),
-    home = '/chat.html',
-    channel = Math.round(Math.random()*1000000),
-    base_address = system.args[1],
-    server_count = (system.args.length >= 3) ? parseInt(system.args[2]) : 3,
-    up_timeout = (10 + (server_count*server_count)/4)*1000;
-    pred_timeout = (1 + (server_count*server_count)/6)*1000;
-
-var pages = [];
-
-// Start checking the states
-common.wait_cluster_up(pages, server_count, up_timeout, function(status, nodes, elapsed) {
-    if (status) {
-        console.log('Cluster is up after ' + elapsed + 'ms');
-        //common.show_nodes(nodes);
-
-        console.log('Sending chat message via last page/server');
-        pages[pages.length-1].evaluate(function() {
-            $('#talk').val("test line");
-            //sendLine();
-            $('#send').click();
-        });
-        common.wait_cluster_predicate(pages, server_count, pred_timeout, function () {
-            var sm = node._self.stateMachine;
-            //console.log(nodeId + " stateMachine: " + JSON.stringify(sm));
-            if ('history' in sm && 'value' in sm.history) {
-                var lines = sm.history.value,
-                    m = lines[lines.length-1].match(/.*: test line$/);
-                return m ? true : false;
-            } else {
-                return false;
-            }
-        }, function(status, results, elapsed) {
-            if (status) {
-                console.log('Cluster state propagated after ' + elapsed + 'ms');
-                phantom.exit(0);
-            } else {
-                console.log('Cluster state failed to propagate after ' + elapsed + 'ms');
-                phantom.exit(1);
-            }
-        });
-    } else {
-        console.log('Cluster failed to come up after ' + elapsed + 'ms');
-        //common.show_nodes(nodes);
-        phantom.exit(1);
-    }
-});
-
-// Start each page/cluster node
-var opts = {base_address: base_address, home: home, channel: channel};
[…]

### 2015-09-20 test/test_kill_nodes.js

@@ -1,70 +1,0 @@
-// Usage:
-// - Start rtc_server.js first:
-//   ./rtc_server --port 8000 --home chat.html
-//
-// - Then use a docker build of slimerjs to run the test:
-//   IP_ADDR=$(hostname -I | awk '{print $1}')
-//   docker run -it -v `pwd`/test:/test slimerjs-0.9.6 slimerjs /test/test_chat_propagate.js http://${IP_ADDR}:8000/
-
-var common = require('common'),
-    system = require('system'),
-    home = '/chat.html',
-    channel = Math.round(Math.random()*1000000),
-    base_address = system.args[1],
-    server_count = (system.args.length >= 3) ? parseInt(system.args[2]) : 3,
-    kill_count = (system.args.length >= 4) ? parseInt(system.args[3]) : parseInt((server_count-1)/2, 10),
-    up_timeout = (10 + (server_count*server_count)/4)*1000;
-    pred_timeout = (1 + (server_count*server_count)/6)*1000;
-
-var pages = [];
-
-if (kill_count > parseInt((server_count-1)/2, 10)) {
-    console.log('Kill count must be less than half of server count');
-    phantom.exit(2);
-}
-
-// Start checking the states
-common.wait_cluster_up(pages, server_count, up_timeout, function(status, nodes, elapsed) {
-    if (status) {
-        console.log('Cluster is up after ' + elapsed + 'ms');
-        //common.show_nodes(nodes);
-
-        console.log('Removing ' + kill_count + ' nodes/pages (including the leader)');
-
-        var leader_idx = common.get_leader_idx(common.get_node_info(pages));
-
-        console.log('Removing page index ' + leader_idx + ' (current leader)');
-        pages[leader_idx].close();
-        pages.splice(leader_idx, 1); // NOTE: mutates pages array in place
-
-        for (var j=0; j<kill_count-1; j++) {
-            var kill_idx = parseInt(Math.random()*kill_count, 10);
-            console.log('Removing page index ' + kill_idx);
-            pages[kill_idx].close();
-            pages.splice(kill_idx, 1); // NOTE: mutates pages array in place
-        }
-
-        console.log('Waiting for cluster to stabilize');
-        common.wait_cluster_up(pages, server_count-kill_count, up_timeout, function(status, nodes, elapsed) {
-            if (status) {
-                console.log('Cluster recovered after ' + elapsed + 'ms');
-                phantom.exit(0);
-            } else {
[…]

### 2015-09-20 test/test_up.js

@@ -1,40 +1,0 @@
-// Usage:
-// - Start rtc_server.js first:
-//   ./rtc_server --port 8000
-//
-// - Then use a docker build of slimerjs to run the test:
-//   IP_ADDR=$(hostname -I | awk '{print $1}')
-//   docker run -it -v `pwd`/test:/test slimerjs-0.9.6 slimerjs /test/test_up.js http://${IP_ADDR}:8000/
-
-var common = require('common'),
-    system = require('system'),
-    home = '/rtc.html',
-    channel = Math.round(Math.random()*1000000),
-    base_address = system.args[1],
-    server_count = (system.args.length >= 3) ? parseInt(system.args[2]) : 3,
-    up_timeout = (10 + (server_count*server_count)/4)*1000;
-    pred_timeout = (1 + (server_count*server_count)/6)*1000;
-
-var pages = [];
-
-// Start checking the states
-common.wait_cluster_up(pages, server_count, up_timeout, function(status, nodes, elapsed) {
-    if (status) {
-        console.log('Cluster is up after ' + elapsed + 'ms');
-        //common.show_nodes(nodes);
-
-        phantom.exit(0);
-    } else {
-        console.log('Cluster failed to come up after ' + elapsed + 'ms');
-        //common.show_nodes(nodes);
-        phantom.exit(1);
-    }
-});
-
-// Start each page/cluster node
-var opts = {base_address: base_address, home: home, channel: channel};
-pages.push(common.new_page(0, true, opts, function() {
-    for (var idx = 1; idx < server_count; idx++) {
-        pages.push(common.new_page(idx, false, opts));
-    }
-}));

### 2015-10-04 chat.js
Rtc/Chat: major refactor and client request fixes.
- lots of fixes around client requests and seqAppend atomic operation.
    - properly set initial sequence count for seqAppend key/value.
    - seqAppend was sending two requests for every command except for
      the first.
- move docker spawning into rtctwst module.
    - start one slimerjs docker instance at a time. This avoids
      strange problem where initial pages disconnect from twst server
      right away.
- in wait_chat_propagate, send all messages from the leader instead of
  spreading them out. The spread messages have trouble fi

@@ -11,7 +11,7 @@ function updateHistory() {
 }
 
 function applyCmd(stateMachine, cmd) {
-    //log("cmd:", JSON.stringify(cmd));
+    //log("applyCmd:", JSON.stringify(cmd));
     switch (cmd.op) {
         case 'get':
             stateMachine[cmd.key];
@@ -25,35 +25,36 @@ function applyCmd(stateMachine, cmd) {
             // the target key is not a sequence.
             if (!'key' in cmd) { throw new Error("seqAppend missing 'key'") }
             if (!'value' in cmd) { throw new Error("seqAppend missing 'value'") }
-            if (cmd.key in stateMachine) {
-                if (!'cnt' in cmd) { throw new Error("seqAppend missing 'cnt'") }
-                var seq = stateMachine[cmd.key];
-                if (!('cnt' in seq && 'value' in seq)) {
-                    throw new Error("seqAppend on non-sequence");
-                }
-                if (cmd.cnt !== seq.cnt) {
-                    return [false, seq.cnt];
-                } else {
-                    seq.value.push(cmd.value);
-                    seq.cnt += 1;
-                    setTimeout(updateHistory, 1);
-                    return [true, seq.cnt];
-                }
+            if (!'cnt' in cmd) { throw new Error("seqAppend missing 'cnt'") }
+
+            // Add the sequence key if it's not already there
+            if (!(cmd.key in stateMachine)) {
+                stateMachine[cmd.key] = {cnt: 0, value: []};
+            }
+
+            var seq = stateMachine[cmd.key];
+            if (!('cnt' in seq && 'value' in seq)) {
+                throw new Error("seqAppend on non-sequence");
+            }
+            if (cmd.cnt !== seq.cnt) {
+                return [false, seq.cnt];
             } else {
-                stateMachine[cmd.key] = {cnt: 0, value: [cmd.value]};
+                seq.value.push(cmd.value);
+                seq.cnt += 1;
                 setTimeout(updateHistory, 1);
-                return [true, 0];
+                return [true, seq.cnt];
             }
     }
 }
 
 var sendTimeout = 100,
+    sendTimer = null,
     curSeqCnt = 0,
     curSend = null,
     pendingSends = [];
 
 function flushSends() {
-    setTimeout(flushSends, sendTimeout);
+    sendTimer = setTimeout(flushSends, sendTimeout);
     if (curSend || pendingSends.length === 0) { return; }
     var curSend = pendingSends.shift(),
         req = {op: 'seqAppend',
[…]

### 2015-10-04 rtc.js

@@ -103,48 +103,58 @@ function rtcSend(targetId, rpcName, args, callback) {
     }
 }
 function rtcReceive(json) {
-    var resp = JSON.parse(json),
-        rpcName = resp[0],
-        otherNodeId = resp[1],
-        args = resp[2];
-    rpcCounts[rpcName]++;
-    
-    // Call the rpc indicated
-    node[rpcName](args);
+    if (node) {
+        var resp = JSON.parse(json),
+            rpcName = resp[0],
+            otherNodeId = resp[1],
+            args = resp[2];
+        rpcCounts[rpcName]++;
+
+        // Call the rpc indicated
+        node[rpcName](args);
+    } else {
+        console.warn("rtcReceive called before Raft cluster started");
+    }
 }
 
 // Wrap async clientRequest/clientRequestResponse messages into
 // a callback based clientRequest call
 var curLeaderId = null;
-var pendingClientRequest = null;
+var pendingClientRequests = [];
 function clientRequest(args, callback) {
-    //log("clientRequest:", args);
-    if (pendingClientRequest) {
-        // TODO: fix this
-        throw new Error("outstanding clientRequest");
-    }
+    //log("clientRequest:", JSON.stringify(args));
     args['responseId'] = nodeId;
-    pendingClientRequest = {args: args, callback: callback};
-    if (curLeaderId === null || curLeaderId === nodeId) {
-        node.clientRequest(args);
-    } else {
-        rtcSend(curLeaderId, 'clientRequest', args);
-    }
+    pendingClientRequests.push({args: args, callback: callback});
+    clientRequestSend();
 }
-function clientRequestResponse(result) {
-    //log("clientRequestResponse:", result);
-    if (result.status === 'NOT_LEADER') {
-        curLeaderId = result.leaderHint;
-        if (pendingClientRequest) {
-            var args = pendingClientRequest.args;
-            //log("curLeaderId:", curLeaderId, nodeMap[curLeaderId]);
+function clientRequestSend() {
+    if (pendingClientRequests.length > 0) {
+        var args = pendingClientRequests[0].args;
+        if (curLeaderId === null || curLeaderId === nodeId) {
+            //console.log("node.clientRequest:", JSON.stringify(args));
+            node.clientRequest(args);
+        } else {
             rtcSend(curLeaderId, 'clientRequest', args);
         }
-    } else {
-        var callback = pendingClientRequest.callback,
-            args = pendingClientRequest.args;
[…]

### 2015-10-04 test/rtctwst.js

@@ -1,10 +1,12 @@
-var spawn = require('child_process').spawn,
+var child_process = require('child_process'),
     Twst = require('twst').Twst,
     RtcTwst = null;
 
 // Subclass Twst
 exports.RtcTwst = RtcTwst = function(opts) {
     Twst.call(this, opts);
+    this.nextPageIndex = 0;
+
     //rtwst.on('return',        function(idx, msg) { console.log('RETURN:', idx, msg); });
     //rtwst.on('callback',      function(idx, msg) { console.log('CALLBACK:', idx, msg); });
     this.on('error', function(idx, msg) {
@@ -22,6 +24,8 @@ exports.RtcTwst = RtcTwst = function(opts) {
     this.on('close', function(idx, msg) {
         console.log(idx + ' CLOSE:', msg.data);
     });
+
+    if (opts.startPages) { this.startPages(opts); }
 };
 RtcTwst.prototype = Object.create(Twst.prototype);
 RtcTwst.prototype.constructor = RtcTwst;
@@ -36,31 +40,61 @@ RtcTwst.prototype.cleanup_exit = function(exitCode) {
 
 RtcTwst.prototype.dockerPage = function(url, opts) {
     opts = opts || {};
-    var prefix = opts.prefix || "";
-    var timeout = opts.timeout || 60000;
-    var docker_args = ['run', '-it', '--rm',
-        '-v', process.cwd() + '/test:/test',
-        'slimerjs-wily',
-        'slimerjs', '/test/launch.js',
-        url,
-        timeout];
+    var self = this,
+        prefix = opts.prefix || "",
+        timeout = opts.timeout || 60000;
+        docker_args = ['run', '-it', '-d', // '--rm',
+                       '-v', process.cwd() + '/test:/test',
+                       'slimerjs-wily',
+                       'slimerjs', '/test/launch.js',
+                       url,
+                       timeout];
 
     console.log(prefix + 'launching slimerjs in docker');
-    var page = spawn('docker', docker_args);
+    var out = child_process.execFileSync('docker', docker_args,
+                                         {encoding: 'utf8'}),
+        id = out.replace(/[\r\n]*$/,'');
+    //var page = child_process.spawn('docker', docker_args);
+    console.log('launched docker container ' + id);
 
-    page.stdout.on('data', function(chunk) {
-        if (opts.verbose) {
-            var line = chunk.toString('utf8').replace(/\r\n$/,'');
-            if (line === '') { return; }
-            console.log(prefix + line);
[…]

### 2015-10-04 test/wait_chat_propagate.js

@@ -6,35 +6,35 @@
 //
 // - Now run the test using the rtc_server listen address and the
 //   number of nodes:
-//     node test/wait_chat_propagate.js 10.0.01:8001 3
+//     node test/wait_chat_propagate.js 10.0.01:8001 3 1
 
-var RtcTwst = require('./rtctwst').RtcTwst,
-    rtwst = new RtcTwst(),
+var getIP = require('twst').getIP,
+    RtcTwst = require('./rtctwst').RtcTwst,
+    port = 9000,
+    //spread_msgs = true,
+    spread_msgs = false,
     rtc_address = process.argv[2],
     clientCount = process.argv.length >= 4 ? parseInt(process.argv[3]) : 1,
     msgCount = (process.argv.length >= 5) ? parseInt(process.argv[4]) : 1,
-    timeout = (clientCount*20)*1000;
-
-var channel = Math.round(Math.random()*100000);
-var url = 'http://' + rtc_address +
+    timeout = (clientCount*20)*1000,
+    channel = Math.round(Math.random()*100000),
+    url = 'http://' + rtc_address +
           '/chat.html?channel=' + channel +
           '&console_logging=true' +
-          '&twst_address=' + rtwst.getAddress() + '&paused=1';
+          '&twst_address=' + getIP() + ':' + port + '&paused=1',
+    rtwst = null;
 
-var pages = [];
+rtwst = new RtcTwst({port: port,
+                     startPages: true,
+                     url: url,
+                     prefix: 'p',
+                     timeout: timeout,
+                     clientCount: clientCount,
+                     pagesCallback: delay_do_start});
 
-for (var i=0; i<clientCount; i++) {
-    pages.push(rtwst.dockerPage(url, {prefix: 'p' + i + ': ',
-                                      timeout: timeout}));
-}
-
-function poll() {
-    if (Object.keys(rtwst.clients).length >= clientCount) {
-        console.log('Delaying for 5 seconds before starting cluster');
-        setTimeout(do_start, 5000);
-    } else {
-        setTimeout(poll, 100);
-    }
+function delay_do_start() {
+    console.log('All clients started, delaying for 5 seconds before starting cluster');
+    setTimeout(do_start, 5000);
 }
 
 function do_start() {
@@ -54,26 +54,39 @@ function do_start() {
 
 function do_chat() {
     rtwst.broadcast("window._test_msgCount = " + msgCount + ";")
[…]

### 2015-10-04 test/wait_kill_nodes.js

@@ -6,40 +6,38 @@
 //
 // - Now run the test using the rtc_server listen address and the
 //   number of nodes:
-//     node test/wait_kill_nodes.js 10.0.01:8001 3
+//     node test/wait_kill_nodes.js 10.0.01:8001 3 1
 
-var RtcTwst = require('./rtctwst').RtcTwst,
-    rtwst = new RtcTwst(),
+var getIP = require('twst').getIP,
+    RtcTwst = require('./rtctwst').RtcTwst,
+    port = 9000,
     rtc_address = process.argv[2],
     clientCount = process.argv.length >= 4 ? parseInt(process.argv[3]) : 1,
     killCount = (process.argv.length >= 5) ? parseInt(process.argv[4]) : parseInt((clientCount-1)/2, 10),
-    timeout = (clientCount*20)*1000;
-
-var channel = Math.round(Math.random()*100000);
-var url = 'http://' + rtc_address +
+    timeout = (clientCount*20)*1000,
+    channel = Math.round(Math.random()*100000),
+    url = 'http://' + rtc_address +
           '/chat.html?channel=' + channel +
           '&console_logging=true' +
-          '&twst_address=' + rtwst.getAddress() + '&paused=1';
-
-var pages = [];
+          '&twst_address=' + getIP() + ':' + port + '&paused=1',
+    rtwst = null;
 
 if (killCount > parseInt((clientCount-1)/2, 10)) {
     console.log('Kill count must be less than half of client count');
     process.exit(2);
 }
 
-for (var i=0; i<clientCount; i++) {
-    pages.push(rtwst.dockerPage(url, {prefix: 'p' + i + ': ',
-                                      timeout: timeout}));
-}
+rtwst = new RtcTwst({port: port,
+                     startPages: true,
+                     url: url,
+                     prefix: 'p',
+                     timeout: timeout,
+                     clientCount: clientCount,
+                     pagesCallback: delay_do_start});
 
-function poll() {
-    if (Object.keys(rtwst.clients).length >= clientCount) {
-        console.log('Delaying for 5 seconds before starting cluster');
-        setTimeout(do_start, 5000);
-    } else {
-        setTimeout(poll, 100);
-    }
+function delay_do_start() {
+    console.log('All clients started, delaying for 5 seconds before starting cluster');
+    setTimeout(do_start, 5000);
 }
 
 function do_start() {
@@ -88,9 +86,6 @@ function do_kill() {
         });
     });
 }
-
[…]

### 2015-10-04 test/wait_start.js

@@ -8,40 +8,37 @@
 //   number of nodes:
 //     node test/wait_start.js 10.0.01:8001 3
 
-var RtcTwst = require('./rtctwst').RtcTwst,
-    rtwst = new RtcTwst(),
+var getIP = require('twst').getIP,
+    RtcTwst = require('./rtctwst').RtcTwst,
+    port = 9000,
     rtc_address = process.argv[2],
     clientCount = process.argv.length >= 4 ? parseInt(process.argv[3]) : 1,
-    timeout = (clientCount*20)*1000;
-
-var channel = Math.round(Math.random()*100000);
-var url = 'http://' + rtc_address +
+    timeout = (clientCount*20)*1000,
+    channel = Math.round(Math.random()*100000),
+    url = 'http://' + rtc_address +
           '/chat.html?channel=' + channel +
           '&console_logging=true' +
-          '&twst_address=' + rtwst.getAddress() + '&paused=1';
+          '&twst_address=' + getIP() + ':' + port + '&paused=1',
+    rtwst = null;
 
-var pages = [];
+rtwst = new RtcTwst({port: port,
+                     startPages: true,
+                     url: url,
+                     prefix: 'p',
+                     timeout: timeout,
+                     clientCount: clientCount,
+                     pagesCallback: delay_do_start});
 
-for (var i=0; i<clientCount; i++) {
-    pages.push(rtwst.dockerPage(url, {prefix: 'p' + i + ': ',
-                                      timeout: timeout}));
-}
-
-function poll() {
-    if (Object.keys(rtwst.clients).length >= clientCount) {
-        console.log('Delaying for 5 seconds before starting cluster');
-        setTimeout(do_start, 5000);
-    } else {
-        setTimeout(poll, 100);
-    }
+function delay_do_start() {
+    console.log('All clients started, delaying for 5 seconds before starting cluster');
+    setTimeout(do_start, 5000);
 }
 
 function do_start() {
-    rtwst.broadcast('startChat();');
+    rtwst.broadcast('startChat()');
     rtwst.wait_cluster_up(timeout, function(status, nodes, elapsed) {
         if (status) {
             console.log('Cluster is up after ' + elapsed + 'ms');
-            console.log('Delaying for 3 seconds before sending message');
             rtwst.cleanup_exit(0);
         } else {
             console.log('Cluster failed to come up after ' +
@@ -50,8 +47,6 @@ function do_start() {
         }
     });
 }
-
-poll();
 
 
 setTimeout(function() {

### 2015-10-04 test/cleanup.sh
Add process.py for post-processing test output.
- Also, move avg.clj

@@ -1,0 +1,6 @@
+#!/bin/bash
+
+for id in $(docker ps -a | grep slimerjs-wily | awk '{print $1}'); do
+    echo "Removing ${id}"
+    docker rm -f ${id}
+done

### 2015-10-04 test/scripts/process.py

@@ -1,0 +1,36 @@
+#!/usr/bin/env python
+
+from __future__ import print_function
+
+import os, sys
+match = sys.argv[1]
+tag = sys.argv[2]
+files = sys.argv[3:]
+
+grouped = {}
+
+for file in files:
+    lines = open(file).readlines()
+    matches = filter(lambda x:match in x, lines)
+    if not matches:
+        print("no matching line in '%s'" % file, file=sys.stderr)
+        continue
+    line = matches[0].rstrip()
+    ms = int(line.split()[-1][:-2])
+    secs = ms / 1000.0
+    [type, subtype, size, attempt] = file.split('-')
+    size = int(size)
+
+    #print(file, type, subtype, size, attempt, secs, ":", line, file=sys.stderr)
+
+    print("%s %s %s" % (size, secs, tag))
+
+    if not grouped.has_key(size):
+        grouped[size] = []
+
+    grouped[size].append(secs)
+
+for size in sorted(grouped.keys()):
+    times = grouped[size]
+    avg = sum(times) / len(times)
+    print("%s %s %s" % (size, avg, tag+"_average"))

### 2015-10-06 research_class/paper.tex
paper.tex: remove obsolete text.
process.py: split output / average. Update 1000ms data.

@@
 
-%%\usepackage{geometry}
-%%\geometry{margin=2.0cm}
-%%
-%%\usepackage[backend=bibtex]{biblatex}
-%%
+% correct bad hyphenation here
+\hyphenation{op-tical net-works semi-conduc-tor}
+
 \usepackage{hyperref}
-%%\usepackage{caption}
-%%\usepackage{subcaption}
-%%
-%%\usepackage{authblk}
-%%
-%%\usepackage{microtype}
 
@@
 \pgfplotsset{compat=1.9}
-
-%%\title{Raft Distributed Consensus Algorithm over WebRTC (Web Real-Time Communication)}
-%%
-%%\author{Joel D. Martin}
-%%\author{David Levine}
-%%\affil{University of Texas at Arlington}
-%%
-%%\addbibresource{references/raft.bib}
-%%\addbibresource{references/w3c.bib}
-%%\addbibresource{references/ietf.bib}
-%%\addbibresource{references/other.bib}
-
-% correct bad hyphenation here
-\hyphenation{op-tical net-works semi-conduc-tor}
 
@@
 
-% TODO: programmatic figure numbers
-Figure 7 shows the results of running the
-\emph{test/test\_chat\_propagate.js} test to show the amount of time
+Figure~\ref{fig:propagate1} shows the results of running the
+\emph{test/wait\_chat\_propagate.js} test to show the amount of time
 that a stable cluster takes to propagate, commit and apply a single
-state machine command to the state machines on every cluster node.
-
-Figure 7 is a scatterplot with each test run as a blue ``x" and a black
-line representing the average at each cluster size.
+chat message to the state machine on every cluster node using the
+atomic \emph{seqAppend} command.  Figure~\ref{fig:propagate1} is
+a scatterplot with each test run as a blue ``x" and a black line
+representing the average at each cluster size.
 
@@
         xlabel={Nodes}, ylabel={Time (seconds)},
-        xtick=data,
+        %xtick=data,
+        xmin=0, ymin=0,
         grid=major,
         scatter/classes={%
-            propagate={mark=x,draw=blue},
-            propagate_average={mark=,draw=red}}]
+            propagate_1_et1000={mark=x,draw=blue},
+            propagate_1_et1000_average={mark=}}]
     \addplot[scatter,only marks,%
         scatter src=explicit symbolic]%
-    table[meta=label] {data/data_propagate.dat};
+    table[meta=label] {data/propagate_1.dat};
     \addplot[scatter,%
         scatter src=explicit symbolic]%
-    table[meta=label] {data/data_propagate_averages.dat};
+    table[meta=label] {data/propagate_1_average.dat};
     \end{axis}
@@
 \caption{Propagate$,$ Commit$,$ and Apply Command}
-  \label{fig:propagate}
+  \label{fig:propagate1}
 \end{figure}
 
-Figures 8, 9 and 10 show the results of running the
+Figure~\ref{fig:propagate100} is similar to
+Figure~\ref{fig:propagate1} but shows the results of adding 100 chat
+messages instead of just 1. The commands are serialized which means
+that the previous command must be committed before the current command
+can be committed. The test is complete once the last chat message
+command has been both committed and applied to every node of the
+cluster. Figure~\ref{fig:propagate1} is a scatterplot with each test
+run as a red ``x" and a black line representing the average at each
+cluster size.
+
+\begin{figure}[!t]
+    \begin{tikzpicture}
+    \begin{axis}[%
+        xlabel={Nodes}, ylabel={Time (seconds)},
+        %xtick=data,
+        xmin=0, ymin=0,
+        grid=major,
+        scatter/classes={%
+            propagate_100_et1000={mark=x,draw=red},
+            propagate_100_et1000_average={mark=}}]
+    \addplot[scatter,only marks,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/propagate_100.dat};
+    \addplot[scatter,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/propagate_100_average.dat};
+    \end{axis}
+    \end{tikzpicture}
+\caption{Propagate$,$ Commit$,$ and Apply Command}
+  \label{fig:propagate100}
+\end{figure}
+
+Figures \ref{fig:cluster_up}, \ref{fig:recovery-leader-failure} and
+\ref{fig:recovery-half-failure} show the results of running the
 \emph{test/test\_up.js} and \emph{test/test\_kill\_nodes.js} in both
@@
 \begin{enumerate}
-    \item \emph{Figure 8 (a)}: The results show that once the cluster
-        size reaches 17, the cluster sometimes does not activate
-        before the test times out (the red ``x" marks). Once the
-        cluster reaches 21 nodes, all 7 test runs time out. The reason
-        for this is because at 17 nodes, the leader is not always able
-        to send out appendEntries RPCs to all the other nodes before
-        their election timeout fires. More analysis is needed, but one
-        possible explanation is that the single SlimerJS XULRunner
-        process that is handling all the cluster nodes is not able to
-        give enough CPU time to the leader node to reliably send
-        appendEntries RPCs to all the other nodes within the election
-        timeout value. Increasing the election timeout to 500ms or
-        greater eliminates the failures without significantly changing
-        the average cluster activation time curve.
-    \item \emph{Figure 9 (a)}: The results show that when the cluster
-        size reaches 15, then average amount of time for cluster
-        recovery (after the leader is killed) increases dramatically
-        (not all data points are shown on the plot for size 15). At
-        cluster size 17 the cluster is no longer able to recover
-        within the test timeout period. Increasing the election
-        timeout to 500ms or greater eliminates the failures at the
-        cost of slightly increased recovery time for smaller cluster
-        sizes.
-    \item \emph{Figure 10 (a)}: During this test the cluster was unable to
-        activate at cluster size 21 when the election timeout value is
-        set to 100ms (see Figure 8 (a)). This means there was no
+    \item \emph{Figure~\ref{fig:cluster_up}~(a)}: The results show that
+        once the cluster size reaches 17, the cluster sometimes does
+        not activate before the test times out (the red ``x" marks).
+        Once the cluster reaches 21 nodes, all 7 test runs time out.
+        The reason for this is because at 17 nodes, the leader is not
+        always able to send out appendEntries RPCs to all the other
+        nodes before their election timeout fires. More analysis is
+        needed, but one possible explanation is that the single
+        SlimerJS XULRunner process that is handling all the cluster
+        nodes is not able to give enough CPU time to the leader node
+        to reliably send appendEntries RPCs to all the other nodes
+        within the election timeout value. Increasing the election
+        timeout to 500ms or greater eliminates the failures without
+        significantly changing the average cluster activation time
+        curve.
+    \item \emph{Figure~\ref{fig:recovery-leader-failure}~(a)}: The
+        results show that when the cluster size reaches 15, then
+        average amount of time for cluster recovery (after the leader
+        is killed) increases dramatically (not all data points are
+        shown on the plot for size 15). At cluster size 17 the cluster
+        is no longer able to recover within the test timeout period.
+        Increasing the election timeout to 500ms or greater eliminates
+        the failures at the cost of slightly increased recovery time
+        for smaller cluster sizes.
+    \item \emph{Figure~\ref{fig:recovery-half-failure}~(a)}: During
+        this test the cluster was unable to activate at cluster size
+        21 when the election timeout value is set to 100ms (see
+        Figure~\ref{fig:cluster_up}~(a)). This means there was no
         opportunity for cluster recovery at cluster size 21 so there
@@
                 xlabel={Nodes}, ylabel={Time (seconds)},
-                xtick=data,
-                ymin=0, ymax=65,
+                %xtick={1,3,5,7,9,11,13,15,17,19,21,23,25,27,29,31,33},
+                %xticklabels={1,,5,,9,,13,,17,,21,,25,,29,,33},
+                xmin=0, xmax=34,
+                ymin=0, ymax=11,
                 grid=major,
                 scatter/classes={%
-                    et1000={mark=x,draw=blue},
-                    et1000_average={mark=,draw=red}}]
+                    start_et1000={mark=x,draw=blue},
+                    start_et1000_average={mark=,draw=red}}]
             \addplot[scatter,only marks,%
                 scatter src=explicit symbolic]%
-            table[meta=label] {data/data_cluster_up_1000.dat};
+            table[meta=label] {data/start_et1000.dat};
             \addplot[scatter,%
                 scatter src=explicit symbolic]%
-            table[meta=label] {data/data_cluster_up_1000_averages.dat};
+            table[meta=label] {data/start_et1000_average.dat};
             \end{axis}
@@
                 xlabel={Nodes}, ylabel={Time (seconds)},
-                xtick=data,
+                %xtick=data,
+                xmin=0,
                 ymin=0, ymax=3.5,
@@
                 scatter/classes={%
-                    kl={mark=x,draw=blue},
-                    kl_average={mark=,draw=red}}]
+                    kill_1_et1000={mark=x,draw=blue},
+                    kill_1_et1000_average={mark=,draw=red}}]
             \addplot[scatter,only marks,%
                 scatter src=explicit symbolic]%
-            table[meta=label] {data/data_kill_leader_1000.dat};
+            table[meta=label] {data/kill_1_et1000.dat};
             \addplot[scatter,%
                 scatter src=explicit symbolic]%
-            table[meta=label] {data/data_kill_leader_1000_averages.dat};
+            table[meta=label] {data/kill_1_et1000_average.dat};
             \end{axis}
@@
                 xlabel={Nodes}, ylabel={Time (seconds)},
-                xtick=data,
+                xmin=0,
                 ymin=0, ymax=3.5,
@@
                 scatter/classes={%
-                    kh={mark=x,draw=blue},
-                    kh_average={mark=,draw=red}}]
+                    kill_half_et1000={mark=x,draw=blue},
+                    kill_half_et1000_average={mark=,draw=red}}]
             \addplot[scatter,only marks,%
                 scatter src=explicit symbolic]%
-            table[meta=label] {data/data_kill_half_1000.dat};
+            table[meta=label] {data/kill_half_et1000.dat};
             \addplot[scatter,%
                 scatter src=explicit symbolic]%
-            table[meta=label] {data/data_kill_half_1000_averages.dat};
+            table[meta=label] {data/kill_half_et1000_average.dat};
             \end{axis}

### 2015-10-06 test/scripts/all_tests.sh

@@ -1,0 +1,13 @@
+#!/bin/bash
+
+# chat propagate 1 message:
+time for x in $(seq 1 7); do for nodes in $(seq 1 2 31); do time node test/wait_chat_propagate.js 192.168.2.3:8001 ${nodes} 1 | tee test/data_et1000/wcp-1-${nodes}-${x}; sleep 3; test/cleanup.sh; done; done
+
+# chat propagate 100 messages:
+time for x in $(seq 1 7); do for nodes in $(seq 1 2 31); do time node test/wait_chat_propagate.js 192.168.2.3:8001 ${nodes} 100 | tee test/data_et1000/wcp-100-${nodes}-${x}; sleep 3; test/cleanup.sh; done; done
+
+# kill 1 node:
+time for x in $(seq 1 7); do for nodes in $(seq 1 2 31); do time node test/wait_kill_nodes.js 192.168.2.3:8001 ${nodes} 1 | tee test/data_et1000/wkn-1-${nodes}-${x}; sleep 3; test/cleanup.sh; done; done
+
+# kill half nodes (162min):
+time for x in $(seq 1 7); do for nodes in $(seq 1 2 31); do time node test/wait_kill_nodes.js 192.168.2.3:8001 ${nodes} | tee test/data_et1000/wkn-half-${nodes}-${x}; sleep 3; test/cleanup.sh; done; done

### 2015-10-06 test/scripts/process.py

@@ -3,12 +3,15 @@
 from __future__ import print_function
 
 import os, sys
-match = sys.argv[1]
-tag = sys.argv[2]
-files = sys.argv[3:]
+fprefix = sys.argv[1]
+match = sys.argv[2]
+tag = sys.argv[3]
+files = sys.argv[4:]
+
+out = open(fprefix+".dat", "w")
+out.write("nodes time label\n")
 
 grouped = {}
-
 for file in files:
     lines = open(file).readlines()
     matches = filter(lambda x:match in x, lines)
@@ -23,14 +26,17 @@ for file in files:
 
     #print(file, type, subtype, size, attempt, secs, ":", line, file=sys.stderr)
 
-    print("%s %s %s" % (size, secs, tag))
+    out.write("%s %s %s\n" % (size, secs, tag))
 
     if not grouped.has_key(size):
         grouped[size] = []
 
     grouped[size].append(secs)
 
+out_avg = open(fprefix+"_average.dat", "w")
+out_avg.write("nodes time label\n")
+
 for size in sorted(grouped.keys()):
     times = grouped[size]
     avg = sum(times) / len(times)
-    print("%s %s %s" % (size, avg, tag+"_average"))
+    out_avg.write("%s %s %s\n" % (size, avg, tag+"_average"))

### 2015-12-07 research_class/paper-min.tex
Add workshop.tex, paper-min.tex and update data files.

--- 2015-10-06 research_class/paper.tex
@@
 current term value. When a Raft follower node does not receive
-a heartbeat from a leader within a locally randomized election timeout period, it
-transitions to candidate state, votes for itself, increases its
-current term value by one and sends a requestVote RPC (with the
+a heartbeat from a leader within a locally randomized election timeout
+period, it transitions to candidate state, votes for itself, increases
+its current term value by one and sends a requestVote RPC (with the
 new term) to all the other nodes in the cluster. When a node receives
@@
     requests are followed by a server response on any given TCP
-    stream. Most remote procedure call mechanims also use
+    stream. Most remote procedure call mechanisms also use
     a synchronous request/response model. %TODO/CITE
@@
 
-\subsubsection{Setup}
+\subsubsection{Twst Test Framework}
+
+In order to facilitate coordinated testing across a large number of
+browser instances, a new test framework was created called
+\cite{twst}. This \emph{Twst} framework provides a client and server
+library written in JavaScript:
+
+\begin{enumerate}
+    \item \emph{server library}: this library is used to write tests
+        that need to coordinate testing across many browser instances.
+        It provides services such as client tracking/enumeration,
+        broadcast and targeted RPC, state rendezvous, log collection,
+        etc.
+    \item \emph{client library}: this library is included with the web
+        application that is being tested. When the page loads, the
+        library will automatically connect to the \emph{Twst} server
+        instance and begin sending console logs and accepting
+        RPCs/commands from the server.
+\end{enumerate}
+
+\subsubsection{Test Setup}
 
@@
 more than 7 WebRTC clients communicating together simultaneously. This
-issue was discovered during testing.
-
-Writing tests using SlimerJS involves creating JavaScript programs
-that are run within the SlimerJS environment. These scripts can create
-browser instances (essentially headless browser tabs), query the status
-of those pages and execute JavaScript within the context of the pages
-themselves.
-
-Three different tests were created to test the basic Raft over WebRTC
-application and the chat application. Each of the tests takes
-a parameter indicating the number of nodes/pages to launch and test
-again. The test script files are:
+issue was discovered during initial testing.
+
+Writing tests using SlimerJS involves creating JavaScript test programs
+which start a \emph{Twst} server instance and then launch a SlimerJS
+docker container for each node of the cluster. Each container starts
+a XULRunner browser instance of the Raft over WebRTC chat application.
+Also included with the application is the \emph{Twst} client library
+which connects back to the \emph{Twst} service in the test program.
+Once all the chat application nodes have connected to the test
+server/program, then the actual test is started across all the
+application nodes.
+
+Three different tests programs/servers were created to test the basic
+Raft over WebRTC application. Each of the tests takes a parameter
+indicating the number of nodes/pages to launch. The test
+source files are:
 
@@
         the cluster. The test reports both the number of milliseconds
-        for the cluster to come up and the recover from the failure.
+        for the cluster to come up and the number of milliseconds for
+        the cluster to fully recover from the failure and evict the
+        failed nodes.
     \item \emph{test/test\_chat\_propagate.js}: similar to
@@
 
-Three tests were run at 10 different cluster sizes (odd cluster sizes
-from 3 to 21 nodes): \emph{test/test\_up.js} and
-\emph{test/test\_kill\_nodes.js} in both modes (kill only leader and kill just
-under half the nodes).
-Each of these tests was run with three different Raft election
-timeout values: 100ms, 500ms, 1000ms. The
-\emph{test/test\_chat\_propagate.js} test was run at
-11 different cluster sizes (odd cluster sizes from 1 to 21 nodes).
-Each test mode and cluster size was run 7 times for a total of 707
-test runs (3*10*7*3 + 11*7)
+Five different tests (using different options for the three test
+programs) were run at 17 different cluster sizes (odd cluster sizes
+from 3 to 35 nodes). Each of these tests was run with three different
+Raft election timeout values: 250ms, 500ms, 1000ms. Each of the five
+tests and cluster size combinations was run 7 times.
 
 \subsubsection{Outcome}
-
-Figure~\ref{fig:propagate1} shows the results of running the
-\emph{test/wait\_chat\_propagate.js} test to show the amount of time
-that a stable cluster takes to propagate, commit and apply a single
-chat message to the state machine on every cluster node using the
-atomic \emph{seqAppend} command.  Figure~\ref{fig:propagate1} is
-a scatterplot with each test run as a blue ``x" and a black line
-representing the average at each cluster size.
 
@@
     \end{tikzpicture}
-\caption{Propagate$,$ Commit$,$ and Apply Command}
+    \caption{Propagate$,$ Commit$,$ and Apply 1 Command}
     \label{fig:propagate1}
-\end{figure}
-
-Figure~\ref{fig:propagate100} is similar to
-Figure~\ref{fig:propagate1} but shows the results of adding 100 chat
-messages instead of just 1. The commands are serialized which means
-that the previous command must be committed before the current command
-can be committed. The test is complete once the last chat message
-command has been both committed and applied to every node of the
-cluster. Figure~\ref{fig:propagate1} is a scatterplot with each test
-run as a red ``x" and a black line representing the average at each
-cluster size.
-
-\begin{figure}[!t]
+%
+    \vspace{1cm}
+%
     \begin{tikzpicture}
@@
         scatter/classes={%
-            propagate_100_et1000={mark=x,draw=red},
+            propagate_100_et1000={mark=x,draw=blue},
             propagate_100_et1000_average={mark=}}]
@@
     \end{tikzpicture}
-\caption{Propagate$,$ Commit$,$ and Apply Command}
+    \caption{Propagate$,$ Commit$,$ and Apply 100 Commands}
     \label{fig:propagate100}
@@
 
-Figures \ref{fig:cluster_up}, \ref{fig:recovery-leader-failure} and
-\ref{fig:recovery-half-failure} show the results of running the
-\emph{test/test\_up.js} and \emph{test/test\_kill\_nodes.js} in both
-modes (kill only leader and kill just under half the nodes). Three
-different scatterplots are shown for each test mode to test different
-values of the Raft election timeout value: 100ms, 500ms, and 1000ms.
-The election timeout value makes very little difference to the average
-time curve line. The reason for this is that the first leader tends to
-stay the leader as new nodes are added so for the most part the
-election timer does not have an impact.
-
-Some specific notes about the figures:
-
-% TODO: link figure numbers
-
-\begin{enumerate}
-    \item \emph{Figure~\ref{fig:cluster_up}~(a)}: The results show that
-        once the cluster size reaches 17, the cluster sometimes does
-        not activate before the test times out (the red ``x" marks).
-        Once the cluster reaches 21 nodes, all 7 test runs time out.
-        The reason for this is because at 17 nodes, the leader is not
-        always able to send out appendEntries RPCs to all the other
-        nodes before their election timeout fires. More analysis is
-        needed, but one possible explanation is that the single
-        SlimerJS XULRunner process that is handling all the cluster
-        nodes is not able to give enough CPU time to the leader node
-        to reliably send appendEntries RPCs to all the other nodes
-        within the election timeout value. Increasing the election
-        timeout to 500ms or greater eliminates the failures without
-        significantly changing the average cluster activation time
-        curve.
-    \item \emph{Figure~\ref{fig:recovery-leader-failure}~(a)}: The
-        results show that when the cluster size reaches 15, then
-        average amount of time for cluster recovery (after the leader
-        is killed) increases dramatically (not all data points are
-        shown on the plot for size 15). At cluster size 17 the cluster
-        is no longer able to recover within the test timeout period.
-        Increasing the election timeout to 500ms or greater eliminates
-        the failures at the cost of slightly increased recovery time
-        for smaller cluster sizes.
-    \item \emph{Figure~\ref{fig:recovery-half-failure}~(a)}: During
-        this test the cluster was unable to activate at cluster size
-        21 when the election timeout value is set to 100ms (see
-        Figure~\ref{fig:cluster_up}~(a)). This means there was no
-        opportunity for cluster recovery at cluster size 21 so there
-        is no data for that column.
-
-\end{enumerate}
-
-Each figure is a scatterplot with each test run as a blue ``x" and
-a black line representing the average at each cluster size.
-
-%\begin{figure*}[!t]
-\begin{figure*}[h]
-    \centering
-    %\begin{subfigure}[b]{0.30\textwidth}
-    \subfloat[100ms election timer]{
-        \centering
-        \resizebox{0.3\linewidth}{!}{
+\begin{figure}[!t]
     \begin{tikzpicture}
@@
         %height=1.5\textwidth, width=1.4\textwidth,
+        %width=0.45\textwidth,
         xlabel={Nodes}, ylabel={Time (seconds)},
-                xtick=data,
-                ymin=0, ymax=65,
+        xmin=0, xmax=43,
+        ymin=0, ymax=15,
         grid=major,
+        %legend pos=outer north east,
+        legend pos=north west,
         scatter/classes={%
-                    et100={mark=x,draw=blue},
-                    et100_fail={mark=x,draw=red},
-                    et100_average={mark=,draw=red}}]
+            start_et250={mark=x,draw=magenta},
+            start_et250_average={mark=,draw=red},
+            %start_et250_fail={mark=x,/raw=red},
+            start_et500={mark=x,draw=cyan},
+            start_et500_average={mark=,draw=red},
+            %start_et500_fail={mark=x,draw=red},
+            start_et1000={mark=x,draw=blue},
+            start_et1000_average={mark=,draw=red}}]
     \addplot[scatter,only marks,%
         scatter src=explicit symbolic]%
-            table[meta=label] {data/data_cluster_up_100.dat};
-            \addplot[scatter,%
-                scatter src=explicit symbolic]%
-            table[meta=label] {data/data_cluster_up_100_averages.dat};
+    table[meta=label] {data/start_et250.dat};
+    \addplot[magenta,scatter,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/start_et250_average.dat};
+
+    \addplot[scatter,only marks,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/start_et500.dat};
+    \addplot[cyan,scatter,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/start_et500_average.dat};
+    %\addplot[scatter,%
+    %    scatter src=explicit symbolic]%
+    %table[meta=label] {data/start_et500_fail.dat};
+
+    \addplot[scatter,only marks,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/start_et1000.dat};
+    \addplot[blue,scatter,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/start_et1000_average.dat};
+
+    \legend{250ms,,500ms,,1000ms,}
     \end{axis}
     \end{tikzpicture}
-        }
-        \label{fig:A}
-    }
-    %\begin{subfigure}[b]{0.30\textwidth}
-    \subfloat[500ms election timer]{
-        \centering
-        \resizebox{0.3\linewidth}{!}{
+    \caption{Initial Cluster Activation}
+    \label{fig:cluster_up}
+\end{figure}
+
+\begin{figure}[!t]
     \begin{tikzpicture}
@@
         %height=1.5\textwidth, width=1.4\textwidth,
+        %width=0.45\textwidth,
         xlabel={Nodes}, ylabel={Time (seconds)},
-                xtick=data,
-                ymin=0, ymax=65,
+        xmin=0, xmax=37,
+        ymin=0, ymax=4.5,
         grid=major,
+        %legend pos=outer north east,
+        legend pos=north west,
         scatter/classes={%
-                    et500={mark=x,draw=blue},
-                    et500_average={mark=,draw=red}}]
+            kill_1_et250={mark=x,draw=magenta},
+            kill_1_et250_average={mark=,draw=red},
+            %kill_1_et250_fail={mark=x,/raw=red},
+            kill_1_et500={mark=x,draw=cyan},
+            kill_1_et500_average={mark=,draw=red},
+            %kill_1_et500_fail={mark=x,draw=red},
+            kill_1_et1000={mark=x,draw=blue},
+            kill_1_et1000_average={mark=,draw=red}}]
     \addplot[scatter,only marks,%
         scatter src=explicit symbolic]%
-            table[meta=label] {data/data_cluster_up_500.dat};
-            \addplot[scatter,%
-                scatter src=explicit symbolic]%
-            table[meta=label] {data/data_cluster_up_500_averages.dat};
+    table[meta=label] {data/kill_1_et250.dat};
+    \addplot[magenta,scatter,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/kill_1_et250_average.dat};
+
+    \addplot[scatter,only marks,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/kill_1_et500.dat};
+    \addplot[cyan,scatter,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/kill_1_et500_average.dat};
+    %\addplot[scatter,%
+    %    scatter src=explicit symbolic]%
+    %table[meta=label] {data/kill_1_et500_fail.dat};
+
+    \addplot[scatter,only marks,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/kill_1_et1000.dat};
+    \addplot[blue,scatter,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/kill_1_et1000_average.dat};
+
+    \legend{250ms,,500ms,,1000ms,}
+
     \end{axis}
     \end{tikzpicture}
-        }
-        \label{fig:B}
-    }
-    %\begin{subfigure}[b]{0.30\textwidth}
-    \subfloat[1000ms election timer]{
-        \centering
-        \resizebox{0.3\linewidth}{!}{
+    \caption{Recovery from Leader Failure}
+    \label{fig:recovery_leader_failure}
+%
+    \vspace{1cm}
+%
     \begin{tikzpicture}
@@
         %height=1.5\textwidth, width=1.4\textwidth,
+        %width=0.45\textwidth,
         xlabel={Nodes}, ylabel={Time (seconds)},
-                %xtick={1,3,5,7,9,11,13,15,17,19,21,23,25,27,29,31,33},
-                %xticklabels={1,,5,,9,,13,,17,,21,,25,,29,,33},
-                xmin=0, xmax=34,
-                ymin=0, ymax=11,
+        xmin=0, xmax=37,
+        ymin=0, ymax=4.5,
         grid=major,
+        %legend pos=outer north east,
+        legend pos=north west,
         scatter/classes={%
-                    start_et1000={mark=x,draw=blue},
-                    start_et1000_average={mark=,draw=red}}]
-            \addplot[scatter,only marks,%
-                scatter src=explicit symbolic]%
-            table[meta=label] {data/start_et1000.dat};
-            \addplot[scatter,%
-                scatter src=explicit symbolic]%
-            table[meta=label] {data/start_et1000_average.dat};
-            \end{axis}
-            \end{tikzpicture}
-        }
-        \label{fig:C}
-    }
-    \caption{Initial Cluster Activation}
-    \label{fig:cluster_up}
-
-%%% Next Row
-\rule{\textwidth}{0.3pt}
-\smallskip
-
-    %\begin{subfigure}[b]{0.30\textwidth}
-    \subfloat[100ms election timer]{
-        \centering
-        \resizebox{0.3\linewidth}{!}{
-            \begin{tikzpicture}
-            \begin{axis}[%
-                %height=1.5\textwidth, width=1.4\textwidth,
-                xlabel={Nodes}, ylabel={Time (seconds)},
-                xtick=data,
-                ymin=0, ymax=3.5, xmax=23,
-                grid=major,
-                scatter/classes={%
-                    kl={mark=x,draw=blue},
-                    kl_fail={mark=x,draw=red},
-                    kl_average={mark=,draw=red}}]
-            \addplot[scatter,only marks,%
-                scatter src=explicit symbolic]%
-            table[meta=label] {data/data_kill_leader_100.dat};
-            \addplot[scatter,%
-                scatter src=explicit symbolic]%
-            table[meta=label] {data/data_kill_leader_100_averages.dat};
-            \end{axis}
-            \end{tikzpicture}
-        }
-        \label{fig:D}
-    }
-    %\begin{subfigure}[b]{0.30\textwidth}
-    \subfloat[500ms election timer]{
-        \centering
-        \resizebox{0.3\linewidth}{!}{
-            \begin{tikzpicture}
-            \begin{axis}[%
-                %height=1.5\textwidth, width=1.4\textwidth,
-                xlabel={Nodes}, ylabel={Time (seconds)},
-                xtick=data,
-                ymin=0, ymax=3.5,
-                grid=major,
-                scatter/classes={%
-                    kl={mark=x,draw=blue},
-                    kl_average={mark=,draw=red}}]
-            \addplot[scatter,only marks,%
-                scatter src=explicit symbolic]%
-            table[meta=label] {data/data_kill_leader_500.dat};
-            \addplot[scatter,%
-                scatter src=explicit symbolic]%
-            table[meta=label] {data/data_kill_leader_500_averages.dat};
-            \end{axis}
-            \end{tikzpicture}
-        }
-        \label{fig:E}
-    }
-    %\begin{subfigure}[b]{0.30\textwidth}
-    \subfloat[1000ms election timer]{
-        \centering
-        \resizebox{0.3\linewidth}{!}{
-            \begin{tikzpicture}
-            \begin{axis}[%
-                %height=1.5\textwidth, width=1.4\textwidth,
-                xlabel={Nodes}, ylabel={Time (seconds)},
-                %xtick=data,
-                xmin=0,
-                ymin=0, ymax=3.5,
-                grid=major,
-                scatter/classes={%
-                    kill_1_et1000={mark=x,draw=blue},
-                    kill_1_et1000_average={mark=,draw=red}}]
-            \addplot[scatter,only marks,%
-                scatter src=explicit symbolic]%
-            table[meta=label] {data/kill_1_et1000.dat};
-            \addplot[scatter,%
-                scatter src=explicit symbolic]%
-            table[meta=label] {data/kill_1_et1000_average.dat};
-            \end{axis}
-            \end{tikzpicture}
-        }
-        \label{fig:F}
-    }
-    \caption{Recovery from Leader Failure}
-    \label{fig:recovery-leader-failure}
-
-%%% Next Row
-\rule{\textwidth}{0.3pt}
-\smallskip
-
-    %\begin{subfigure}[b]{0.30\textwidth}
-    \subfloat[100ms election timer]{
-        \centering
-        \resizebox{0.3\linewidth}{!}{
-            \begin{tikzpicture}
-            \begin{axis}[%
-                %height=1.5\textwidth, width=1.4\textwidth,
-                xlabel={Nodes}, ylabel={Time (seconds)},
-                xtick=data,
-                ymin=0, ymax=3.5, xmax=23,
-                grid=major,
-                scatter/classes={%
-                    kh={mark=x,draw=blue},
-                    kh_average={mark=,draw=red}}]
-            \addplot[scatter,only marks,%
-                scatter src=explicit symbolic]%
-            table[meta=label] {data/data_kill_half_100.dat};
-            \addplot[scatter,%
-                scatter src=explicit symbolic]%
-            table[meta=label] {data/data_kill_half_100_averages.dat};
-            \end{axis}
-            \end{tikzpicture}
-        }
-        \label{fig:G}
-    }
-    %\begin{subfigure}[b]{0.30\textwidth}
-    \subfloat[500ms election timer]{
-        \centering
-        \resizebox{0.3\linewidth}{!}{
-            \begin{tikzpicture}
-            \begin{axis}[%
-                %height=1.5\textwidth, width=1.4\textwidth,
-                xlabel={Nodes}, ylabel={Time (seconds)},
-                xtick=data,
-                ymin=0, ymax=3.5,
-                grid=major,
-                scatter/classes={%
-                    kh={mark=x,draw=blue},
-                    kh_average={mark=,draw=red}}]
-            \addplot[scatter,only marks,%
-                scatter src=explicit symbolic]%
-            table[meta=label] {data/data_kill_half_500.dat};
-            \addplot[scatter,%
-                scatter src=explicit symbolic]%
-            table[meta=label] {data/data_kill_half_500_averages.dat};
-            \end{axis}
-            \end{tikzpicture}
-        }
-        \label{fig:H}
-    }
-    %\begin{subfigure}[b]{0.30\textwidth}
-    \subfloat[1000ms election timer]{
-        \centering
-        \resizebox{0.3\linewidth}{!}{
-            \begin{tikzpicture}
-            \begin{axis}[%
-                %height=1.5\textwidth, width=1.4\textwidth,
-                xlabel={Nodes}, ylabel={Time (seconds)},
-                xmin=0,
-                ymin=0, ymax=3.5,
-                grid=major,
-                scatter/classes={%
+            kill_half_et250={mark=x,draw=magenta},
+            kill_half_et250_average={mark=,draw=red},
+            %kill_half_et250_fail={mark=x,/raw=red},
+            kill_half_et500={mark=x,draw=cyan},
+            kill_half_et500_average={mark=,draw=red},
+            %kill_half_et500_fail={mark=x,draw=red},
             kill_half_et1000={mark=x,draw=blue},
@@
         scatter src=explicit symbolic]%
+    table[meta=label] {data/kill_half_et250.dat};
+    \addplot[magenta,scatter,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/kill_half_et250_average.dat};
+
+    \addplot[scatter,only marks,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/kill_half_et500.dat};
+    \addplot[cyan,scatter,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/kill_half_et500_average.dat};
+    %\addplot[scatter,%
+    %    scatter src=explicit symbolic]%
+    %table[meta=label] {data/kill_half_et500_fail.dat};
+
+    \addplot[scatter,only marks,%
+        scatter src=explicit symbolic]%
     table[meta=label] {data/kill_half_et1000.dat};
-            \addplot[scatter,%
+    \addplot[blue,scatter,%
         scatter src=explicit symbolic]%
     table[meta=label] {data/kill_half_et1000_average.dat};
+
+    \legend{250ms,,500ms,,1000ms,}
+
     \end{axis}
     \end{tikzpicture}
-        }
-        \label{fig:I}
-    }
     \caption{Recovery from Half Cluster Failure}
-    \label{fig:recovery-half-failure}
-\end{figure*}
+    \label{fig:recovery_half_failure}
+\end{figure}
+
+Figure~\ref{fig:propagate1} and Figure~\ref{fig:propagate100} show the
+result of running the \emph{test/wait\_chat\_propagate.js} test to
+determine the amount of time that a stable cluster takes to propagate,
+commit and apply 1 and 100 chat message commands respectively to the state
+machine on every cluster node using the atomic \emph{seqAppend}
+command. The commands are serialized which means that the previous
+command must be committed before the current command can be committed.
+The test is complete once the last chat message command has been both
+committed and applied to every node of the cluster. Each test execution
+is shown as a blue ``x" and the black line represents the average for
+each cluster size.
+
+In Figure~\ref{fig:propagate1} the time to propagate a single message
+to a single node of the cluster is approximately 0.1 seconds. However,
+this really just represents the tests scanning/checking granularity.
+The time to finish propagating and committing a single message shows
+a slight linear slope upwards from about 0.3 seconds for a three node
+cluster to about 0.5 seconds for a 41 node cluster. The time to
+propagate 100 messages (Figure~\ref{fig:propagate100}) has a much
+higher slope and also appears to be non-linear. TODO: why?
+
+Figure~\ref{fig:cluster_up} shows the results of running the
+\emph{test/test\_up.js} test to determine the amount of time that
+a cluster takes to start. The cluster is fully started once a
+leader has been elected and the leader has successfully added all the
+other nodes to the running configuration. Each election timer value
+is represented as a different color and the line represents the
+average value for each cluster size. The cluster startup times show
+a fairly linear increase as the cluster size increases. The reason
+that the slope varies with election timer is due to the fact that the
+election timer also determines the leader heartbeat value (currently
+set to one fifth of the election timer value). The leader heartbeat
+determines the frequency with which \emph{appendEntries} messages are
+sent from the leader to followers and is also used to propagate log
+entries.  This means that the rate at which log entries (and thus
+cluster configuration updates) can be propagated is proportional to
+the election timer value.
+
+Figures \ref{fig:recovery_leader_failure} and
+\ref{fig:recovery_half_failure} show the results of running the
+\emph{test/test\_kill\_nodes.js} in both modes (kill only leader and
+kill just under half the nodes). Each election timer value is
+represented as a different color and the solid line represents the
+average recovery time for each cluster size. When only the leader is
+killed, the recovery times are constant flat regardless of cluster
+size. This is because after a new leader is elected, there is only
+a single configuration entry to propagate to the rest of the cluster
+(to remove the old failed leader). One interesting thing to note is
+that with a 1000ms timer, the 3 node cluster is actually the most
+inefficient size because the average time until a leader election
+starts is inversely proportional to cluster size so the delay until an
+election happens is a significant contributor to the three node
+cluster with 1000ms timer. With a larger cluster or shorter election
+timer, the contribution of election delay is not noticeable. The
+recovery from almost half the nodes failing is not flat because the
+amount of configuration information that needs to be propagated is
+proportional to the size of the cluster.
+
+At larger cluster sizes (35 and greater) the cluster begins to
+experience failures. For example, when testing with a 250ms election
+timer, once the cluster size reaches 35 nodes, the cluster is
+frequently (more than 50\% of the time) unable to recover from the
+half cluster shutdown test. At a cluster size or 41 and with a 1000ms
+election timer, the cluster begins to experience difficulty
+successfully starting up (about 50\% failure rate). The basic problem
+appears to be due to the average election timeout period being too low
+in relation to communication latency. During leader election, the
+first candidate is unable to gather enough votes to become leader
+before other candidates arise and steal votes which prevents a quorum.
+
+Some (or most) of the communication latency may be a result of CPU or
+memory contention due to all the browser instances running on a single
+host. The use of multiple hosts may show that the system can support
+much larger clusters for a given election timeout value. However, with
+a fixed election timer value, there is always a cluster size for which
+communication latency will begin to prevent the cluster from being
+able to perform leader elections. A dynamically adjusted election
+timer is something that should be explored especially for a browser
+context where the communication latency may be highly variable over
+time.
 

### 2015-12-07 research_class/workshop.tex
Add workshop.tex, paper-min.tex and update data files.

--- 2015-12-07 research_class/paper-min.tex
@@
 
-\title{Raft Distributed Consensus Algorithm over\\ WebRTC (Web Real-Time Communication)}
+\title{Peer-to-peer Browser Middleware for Communication and Consensus in an Unreliable Infrastructure}
 
@@
 \begin{abstract}
-We describe the creation of a browser-based distributed peer-to-peer
-application using WebRTC, PeerJS and Raft.js. WebRTC is a protocol
-that modern browsers have begun to integrate which enables
-browser-to-browser communication. PeerJS is a WebRTC client library
-and extensible signaling server. Raft.js is an implementation in
-JavaScript of the Raft distributed consensus protocol. These
-technologies are combined to create a chat application that guarantees
-a shared, consistent and linearized view of the chat history without
-relying on a single centralized server for communication. We also
-create a series of automated tests using the SlimerJS browser
-automation tool and use them to show the correctness of the chat
-application and also to measure the performance and scalability of the
-application for various cluster sizes.
+We describe a browser-based peer-to-peer computing system/middleware that enables distributed consensus that is tolerant of degraded and/or unreliable centralized infrastructure.
+The system is composed of a JavaScript implementation of the Raft distributed consensus protocol (Raft.js) together with the WebRTC browser-to-browser communication protocol.
+The Raft over WebRTC middleware is used to build a demonstration chat application that guarantees a shared, consistent and linearized view of the chat history without relying on a single centralized server for communication.
+The correctness, robustness, performance, and scalability of the system is verified with a series of automated tests implemented using the SlimerJS browser automation tool.
 \end{abstract}
 
-\section{Introduction}
-In the past few years, browsers have quickly evolved from document
-rendering engines into powerful platforms for building sophisticated
-applications. Modern browsers are integrating new protocols for
-communication including WebSockets for full-duplex browser-to-server
-communication \cite{websockets:rfc6455} and WebRTC for direct
-browser-to-browser communication \cite{webrtc:feb2015}. These new
-capabilities enable software developers to treat web applications as
-first class participants in distributed network applications.
+\section{Use Case/Scenario}
+Consider a scenario where emergency responders and volunteers are responding to a city-scale disaster in which utilities such as electrical power and communication systems are highly degraded or completely unavailable for an extended period.
+Emergency management officials have divided the city into zones each of which is assigned several dozen volunteers who systematically search the neighborhoods for people that are injured or in need and also for damaged infrastructure that needs to be repaired.
 
-These new browser capabilities also bring with them the traditional
-challenges of building reliable distributed applications, such as:
-concurrency, security, scalability, naming, discovery, replication,
-relocation, etc. The solutions to many of these challenges involve
-the proper management of distributed state. Distributed consensus is
-one category of distributed state management in which multiple nodes
-must come to agreement on the state of a single value or set of values. % TODO/CITE
-The grandparent of distributed consensus algorithms is the Paxos
-algorithm described by Leslie Lamport \cite{paxos:lamport98}. However,
-Paxos has been notoriously difficult to understand and to implement
-correctly \cite{paxos:chandra07}. Raft is a more recent distributed
-consensus algorithm that aims for understandability and practicality
-of implementation \cite{raft_thesis:ongaro14}.
+Responders and volunteers have access to a distributed disaster response datastore that is accessed and updated via a browser based application.
+Volunteers use their own mobile device and only require a browser with WebRTC capability to run the application.
+The application communicates over the cellular network (existing or provided by mobile emergency response cell stations) or mesh network (although this may also require additional installed software on mobile devices in order to support it).
+Once loaded and running, the application does not require any centralized servers or database system.
+The application uses the Raft distributed consensus protocol to maintain a distributed but consistent view of the datastore among all participating client nodes.
+Also, the distributed cluster is dynamic and allows nodes to join and leave over time without losing state or synchronization of the datastore.
 
-In this paper we will describe an implementation of the Raft protocol
-in a browser context using WebRTC as the communication channel. We
-will describe the results of the implementation and look at some of
-the unique challenges (and opportunities) we encountered with the
-implementation. We will also explore some ideas for further study.
-
-\section{Motivation/Use Cases}
-The traditional use of distributed consensus is in the data-center
-(private or cloud based), often as the core coordination component
-of a replicated data-storage system. %TODO/CITE
-In these examples, the set of nodes that compose the cluster tend to
-be fairly static and changes to cluster membership are usually for the
-purpose of replacing nodes. Changes to the long-term size of the
-cluster are uncommon. %TODO/CITE
-
-Most existing distributed browser applications are extensions of the
-data-centric model in which browsers connect to load-balanced web
-servers which are in close network proximity to the distributed
-consensus cluster. %TODO/CITE
-The backend web servers deliver the web application to the browser and
-also serve as the conduit for all browser-to-browser communication and
-state management.
-
-However, requiring application messages and application state to
-traverse the back-end servers is not always desirable for various
-reasons, including the following:
-
-\begin{enumerate}
-\item \emph{security (privacy)}: application state and user messages
-    may need to traverse directly from browser to browser rather than
-    traversing a server.
-\item \emph{locality}: the nature of the application may allow
-    clusters of applications that are nearby on a network (e.g.
-    intranet) to communicate and coordinate directly rather than
-    passing all messages and state through a server back-end which may
-    be considerably more remote.
-\item \emph{bandwidth and scale}: a distributed browser based
-    application may be able to avoid expensive bandwidth into and out
-    of the backend data-center by passing messages and state directly
-    between browsers.  Depending on the design of the application this
-    may also enable better scalability then a more centralized model.
-\end{enumerate}
-
-%TODO: better lead in to this paragraph from above
-Here are some specific use cases in which Raft over WebRTC may be
-applicable:
-
-\begin{enumerate}
-\item \emph{Consistent order secure chat}: a chat room system where
-    all users see the same message history but the messages never
-    traverse a central server and there is no central arbiter
-    of truth. The application code itself may still be delivered from
-    a central web service. This use case was implemented
-    for this paper.
-\item \emph{Private multiuser document editing/whiteboard system}:
-    Google Docs, among other examples, has demonstrated a powerful
-    model for collaborative document creation. However, these systems
-    all use a centralized servers to coordinate shared state and
-    handle browser-to-browser communication. %TODO/CITE
-    Using Raft over WebRTC could enable applications with similar
-    functionality but without the central coordination and
-    communication.
-\item \emph{Browser based network games}: most network games have
-    certain states that must be agreed upon by all nodes (e.g. avatar
-    dead or alive, etc). This can be achieved with Raft over WebRTC
-    without requiring any data to traverse a central server. In
-    addition to reducing browser to server bandwidth requirements,
-    this may also allow lower latency gaming if game instances are
-    partitioned into clusters based on latency.
-\end{enumerate}
+The above capabilities are demonstrated using a distributed chat application implemented with the Raft over WebRTC middleware.
+This chat application uses direct browser-to-browser communication over WebRTC and maintains a consistent view of the chat history (exact same messages and ordering is seen by all nodes) and is fault tolerant (almost half the cluster nodes can fail simultaneously and the application will remain available without dataloss for the remaining clients).
 
@@
 
-Raft is a new distributed consensus algorithm designed to be
-understandable and practical without sacrificing safety and
-correctness. The Raft algorithm implemented and described in this
-paper is based on Diego Ongaro's dissertation ``Consensus: Bridging
-Theory and Practice" \cite{raft_thesis:ongaro14}.
+Raft is a new distributed consensus algorithm designed to be understandable and practical without sacrificing safety and correctness.
+The Raft algorithm implemented and described in this paper is based on Diego Ongaro's dissertation ``Consensus: Bridging Theory and Practice" \cite{raft_thesis:ongaro14}.
 
-The Raft algorithm breaks the problem of consensus into three core
-sub-problems: leader election, log replication, and safety
-\cite[Section~3]{raft_thesis:ongaro14}. A full Raft system must also
-address: cluster membership changes
-\cite[Section~4]{raft_thesis:ongaro14}, log compaction
-\cite[Section~5]{raft_thesis:ongaro14}, and client
-interaction \cite[Section~6]{raft_thesis:ongaro14}. The Raft protocol
-is based on the concept of a distributed transaction log. Entries in
-the log represent a sequence of commands that will be applied to an
-internal state machine. If all members of the Raft cluster have the
-same log entries, then their state machine on each node will have
-exactly the same state. Each node of a Raft cluster can be in one of
-three states: follower, candidate, or leader. The responsibility of
-the leader is to accept new transaction log entries and then
-replicate these entries in the same order to all other members of the
-cluster. The leader is the only member of the cluster that may make
-changes to the transaction log. In order to maintain leadership, the
-leader sends heartbeat RPCs to all the followers in the cluster.
-Followers that do not receive a heartbeat RPC within a certain time
-period become candidates and attempt to be elected by the other nodes
-as the new leader of the cluster \cite[p.~12]{raft_thesis:ongaro14}.
+The Raft algorithm breaks the problem of consensus into three core sub-problems: leader election, log replication, and safety \cite[Section~3]{raft_thesis:ongaro14}.
+A full Raft system must also address: cluster membership changes \cite[Section~4]{raft_thesis:ongaro14}, log compaction \cite[Section~5]{raft_thesis:ongaro14}, and client interaction \cite[Section~6]{raft_thesis:ongaro14}.
+The Raft protocol is based on the concept of a distributed transaction log.
+Entries in the log represent a sequence of commands that will be applied to an internal state machine.
+If all members of the Raft cluster have the same log entries, then their state machine on each node will have exactly the same state.
+Each node of a Raft cluster can be in one of three states: follower, candidate, or leader.
+The responsibility of the leader is to accept new transaction log entries and then replicate these entries in the same order to all other members of the cluster.
+The leader is the only member of the cluster that may make changes to the transaction log.
+In order to maintain leadership, the leader sends heartbeat RPCs to all the followers in the cluster.
+Followers that do not receive a heartbeat RPC within a certain time period become candidates and attempt to be elected by the other nodes as the new leader of the cluster \cite[p.~12]{raft_thesis:ongaro14}.
 
-\subsubsection{Leader Election}
-
-Each node of the Raft cluster maintains an ordinally increasing
-current term value. When a Raft follower node does not receive
-a heartbeat from a leader within a locally randomized election timeout
-period, it transitions to candidate state, votes for itself, increases
-its current term value by one and sends a requestVote RPC (with the
-new term) to all the other nodes in the cluster. When a node receives
-a requestVote RPC with a term that is greater than its own it will
-become a follower (if not already), update its term to the new term,
-and send a RPC response back that indicates a vote for that candidate.
-However, a node may only vote for one candidate in a given term. If
-a candidate receives votes from more than half of the cluster, then it
-immediately becomes a leader and sends out a appendEntries RPC to
-confirm leadership and reset the election timers on all the other
-nodes \cite[Section~3.4]{raft_thesis:ongaro14}.
-
-\subsubsection{Log Replication}
-
-Each Raft node maintains a transaction log. The entries in the
-transaction log each contain a term value in addition to the action to
-apply to the state machine. Each node also keeps track of the most
-recent log entry which is known to be committed (exists in the log of
-more than half the nodes in the cluster).
-
-When a new leader is elected, it begins sending appendEntries RPCs
-to other nodes in the Raft cluster. These RPC messages contain
-information about the state of its transaction log: index and term of
-the latest entry, and the most recently committed entry. If the node
-does not have an entry matching the most recent entry on the leader,
-then it replies with false to indicate that its log is not up to
-date. The next appendEntries RPC that the leader sends to that
-node will contain the next oldest index and term. This will continue
-until the leader discovers the latest log entry which is in agreement
-(the node may not have any entries if it is new). Then the leader will
-begin to propagate entries to the node in subsequent appendEntries
-RPCs until the node is caught up. The leader continues sending
-empty appendEntries as heartbeat RPCs to all the nodes until it
-receives a new entry in its transaction log from a client
-\cite[Section~3.5]{raft_thesis:ongaro14}.
-
-\subsubsection{Safety}
-
-In order to ensure that each state machine on every node executes
-exactly the same commands in exactly the same order, the Raft system
-provides safety guarantees. In particular, there are some restrictions
-on whether a Raft node is actually eligible to become a leader, based on
-the state of its transaction log compared to other node transaction
-logs. A Raft node will only vote for a candidate if the candidate has
-a log that is more up-to-date than the voter. A log entry with a later
-term is always more up-to-date. If the log entries have the same term,
-then the entry with a higher index is more up-to-date. In addition,
-a Raft leader may not consider entries from a previous term to be
-committed until it has committed at least one entry from its own
-term
-\cite[Section~3.6]{raft_thesis:ongaro14}.
-
-Raft nodes must also persist certain properties to durable storage
-before sending or receiving any RPCs. The properties that must be
-persisted are: current term, current vote for this term (if any), and
-either the full transaction log or the state machine plus any
-unapplied transaction log entries
-\cite[Section~3.8]{raft_thesis:ongaro14}.
-
-\subsubsection{Membership Changes}
-
-Raft leverages the replicated transaction log to accomplish live
-cluster membership changes. Membership changes are accomplished by
-adding or removing one Raft node at a time using special add/remove
-log entries. As soon as the new entry is added to the log it becomes
-effective on that node without waiting for the entry to be fully
-committed across the cluster. However, a new cluster change entry may
-not be added to the log until the previous one is committed. Once the
-change entry is committed, the initiator of the change is notified,
-removed servers can be shut down and another change entry may be added
-to the log \cite[Section~4]{raft_thesis:ongaro14}.
-
-\subsubsection{Log Compaction}
-
-In order to keep the replicated transaction log from growing
-indefinitely, Raft nodes should periodically compact their logs. There
-are many different ways to accomplish log compaction and the best way
-will depend on application requirements and on the specific nature of
-the distributed state machine. For example, if the state machine is
-simply a single shared counter, then any log entries before the most
-recently committed entry may be safely discarded. The simplest and
-most generic solution is often to provide some way of snapshotting the
-entire state machine (including the term and log index it represents)
-to disk at which point the all previously applied log entries can be
-discarded. Snapshotting the state machine also means that the
-implementation must provide a way for the leader to serialize and send
-the current state machine to other nodes if their transaction log is
-too old (e.g. when a new node is added)
-\cite[Section~5]{raft_thesis:ongaro14}.
-
-\subsubsection{Client Interaction}
-
-Clients of the Raft cluster are able to interact with the distributed
-state machine by sending RPCs to the leader in order to add command
-entries to the transaction log. Clients first find the address of any
-node in the cluster either via broadcast, via an external directory
-service, or directly provided by the user. Once the client discovers
-a node of the cluster, that node can either forward messages directly
-to the leader node (if it is not the leader) or it can reject client
-requests with a response indicating the address of the leader where
-client requests should be redirected
-\cite[Section~6.2]{raft_thesis:ongaro14}.
-
-The Raft protocol can also provide strict linearizable semantics for
-client commands/requests (reads and writes) for applications that
-require it. In order to accomplish this, each client is given a unique
-ID, and the client assigns a unique serial number to each command.
-This prevents a single client command from being applied twice (leader
-crash, network duplication) or lost (packet loss)
-\cite[Section~6.3]{raft_thesis:ongaro14}.
-
-Linearizability is important not just for writes/updates to the state
-machine, but also for reads, which means that read requests must also
-be entered as transactions in the log and committed to more than half
-the Raft nodes before the leader can respond to the client. If reads
-bypass the transaction log then those reads are serializable but not
-necessarily linearizable. For example, a leader may have been deposed
-and not realize it if the cluster is partitioned which could result in
-reads of stale data (the leader of the larger partition may have
-already committed new entries).
-\cite[Section~6.4]{raft_thesis:ongaro14}.
-
-\begin{figure*}[!t]
-\centerline{\includegraphics[width=0.8\textwidth]{imgs/raft_rtc_architecture}}
-\caption{Raft over RTC Architecture}
-  \label{fig:raft_rtc_architecture}
-\end{figure*}
-
-\subsection{Raft.js}
-
-Raft.js is an implementation of the Raft algorithm in JavaScript that
-was created by the author of this paper \cite{raft.js}.  Raft.js is
-designed to run in either a browser environment or within Node.js
-(server-side JavaScript). Raft.js implements the Raft algorithm as
-described in \cite[Consensus:~Diego]{raft_thesis:ongaro14}.
-
-\subsubsection{Modular Design}
-
-The implementation of the Raft algorithm is in the
-\texttt{RaftServerBase} class (in \texttt{base.js}). The base class
-omits the implementation of functions with side-effects such as RPC
-communication, durable storage, scheduling/timeouts, or state machine
-command/management. In order to create a working implementation these
-functions must be provided either as configuration parameters when the
-class is instantiated or by sub-classing the class.
-
-The modular design of the Raft.js implementation allows it to be
-easily used in different contexts. For example, the \texttt{RaftServerLocal}
-class (\texttt{local.js}) implements RPC calls using plain JavaScript function
-calls between different instances of the class (nodes) in the same
-execution context. For example, the online Raft visualization at
-\href{http://kanaka.github.io/raft.js/}{http://kanaka.github.io/raft.js/}
-\cite{raft.js:visualization} uses the \texttt{RaftServerLocal} class
-and instantiates it using a scheduling function that executes one
-scheduled event each time the user clicks on the ``Step" button rather
-than due to the passage of time.
-
-The \texttt{RaftServerHttp} class (\texttt{http.js}) embeds a simple
-web server into each Raft node and then uses normal HTTP requests to
-send RPC messages between Raft nodes.
-
-For this paper, the \texttt{RaftServerLocal} class was extended with
-the capability of sending RPCs over the WebRTC Data Channel.
-
-\subsubsection{Differences between Raft and Raft.js}
-
-The current version of Raft.js implements most of the Raft algorithm
-including: leader election, log replication, safety, membership
-changes, basic client interaction, and read-only operation
-optimizations. % TODO: read-only operations needs mentioning earlier
-Log compaction and full client linearizability are not currently
-implemented (clients and client transactions are not assigned unique
-IDs). Log compaction is important for long-running production systems
-in order to limit memory growth and allow new nodes to catch up their
-logs more efficiently. However log compaction is not required for
-verifying overall system correctness and does not significantly impact
-short-running test configurations. Full client linearizability is
-an important feature for simplifying certain types of distributed
-applications, however the implementation of linearizability requires
-some significant additions to the Raft implementation (e.g. assigning
-client IDs and sequence IDs and tracking timed-out clients). For the
-chat application, simply adding a single atomic check-and-set state
-machine command is sufficient to achieve a linearized chat history.
-This does impose a slight additional cost on the client because it
-must retry the atomic operations until they succeed. The author
-intends to implement both log compaction and full client
-linearizability in the future, however these do not have a significant
-impact on the ability to test Raft combined with WebRTC.
-
-The Raft protocol has undergone some revisions since the original
-draft paper \cite{raft_paper:ongaro14} (including incorporation of
-suggestions by the author of this paper). These changes include
-cluster membership simplifications, separating the concepts of
-committing entries to the log and applying entries to the state
-machine, clearer client interaction, etc. As part of the work for
-implementing Raft over WebRTC, the Raft.js implementation was updated
-to reflect these changes to the Raft protocol definition.
-
-The are also some other differences between Raft.js and Raft as
-described in \cite[Consensus:~Diego]{raft_thesis:ongaro14}:
-
-\begin{enumerate}
-\item The paper describes the Raft RPCs in terms of synchronous stream
-    based request/response model where requests and responses are
-    automatically correlated together by a lower level of the software
-    stack. For example with HTTP 1.0 \cite{http:rfc1945} client
-    requests are followed by a server response on any given TCP
-    stream. Most remote procedure call mechanisms also use
-    a synchronous request/response model. %TODO/CITE
-    WebRTC DataChannel connections are message based rather than
-    request/response stream based. %TODO/CITE
-    In the original Raft definition of RPC, request and response messages
-    almost contain enough information to function correctly in an
-    environment without builtin request/response correlation, such as
-    UDP or SCTP when used with reliable transport disabled. %TODO/CITE
-    Raft.js includes a small modification to the response RPCs for
-    requestVote and appendEntries to add a `sourceId' field that
-    enables the protocol to function with message based transports.
-\item The Raft paper is clear that each membership change must be
-    delayed until all the previous membership change is committed
-    (sequential) \cite[Section~4.1]{raft_thesis:ongaro14}. It is
-    implied that the leader node should track all pending membership
-    change requests and apply them one at a time.  However, in
-    Raft.js, if a client requests a membership change while another
-    change request is pending (uncommitted), the leader will reject
-    the change request with a status of `PENDING\_CONFIG\_CHANGE'.
-    This keeps the core Raft implementation simpler with the trade-off
-    being that the client implementation of membership change requests
-    is slightly more complex due to retry logic.
-\end{enumerate}
+Raft leverages the replicated transaction log to accomplish live cluster membership changes.
+Membership changes are accomplished by adding or removing one Raft node at a time using special add/remove log entries.
+In order to keep the replicated transaction log from growing indefinitely, Raft nodes should periodically compact their logs.
+The simplest and most generic solution is often to provide some way of snapshotting the entire state machine (including the term and log index it represents) to disk at which point the all previously applied log entries can be discarded.
+Clients of the Raft cluster are able to interact with the distributed state machine by sending RPCs to the leader in order to add command entries to the transaction log.
+Clients first find the address of any node in the cluster and are then redirected to communicate directly with the leader.
+The Raft protocol also provides strict linearizable semantics for client commands/requests (reads and writes) for applications that require it.
 
@@
 
-WebRTC is a collection of browser APIs \cite{webrtc:feb2015} and
-network protocols \cite{draft-rtcweb-overview} that are currently
-being standardized by the W3C and IETF respectively. Collectively these
-APIs and protocols enable real-time peer-to-peer video, audio and data
-communication between browsers.
+WebRTC is a collection of browser APIs \cite{webrtc:feb2015} and network protocols \cite{draft-rtcweb-overview} that are currently being standardized by the W3C and IETF respectively.
+Collectively these APIs and protocols enable real-time peer-to-peer video, audio and data communication between browsers.
 
-\subsubsection{WebRTC Discovery and Signaling}
-
-One of the challenges with browser-to-browser communication (and
-direct peer-to-peer communication in general) is that the web browser
-environment or user agent (UA) is usually running behind a firewall in
-a private internet subnet. This complicates direct communication in
-several ways: the public Internet address of the user agent is unknown
-and dynamic, inbound connections are denied by default, and outbound
-connections are network address translated (NATed)
-\cite[Section~3.4]{draft-rtcweb-transports}.
-
-To establish a direct communication channel between browsers, WebRTC
-uses a mechanism called Interactive Connectivity Establishment (ICE)
-that is similar to the Session Initiation Protocol (SIP) commonly used
-for Voice-over-IP (VoIP) communication \cite{RFC5245:ice}. This
-involves a third party server known as a signaling server. Each user
-agent (browser) connects to this server at a well known address in
-order to establish a direct connection to other user agents. The
-signaling server is also used to communicate session control messages
-(open, close, error) and to negotiate media capabilities. %TODO/CITE
-The signaling server may optionally provide other services such as
-higher level session management, HTTP serving of the web application,
-address book service, client presence/availability, etc. %TODO/CITE
+One of the challenges with browser-to-browser communication (and direct peer-to-peer communication in general) is that the browser is usually running behind a firewall in a private internet subnet.
+To establish a direct communication channel between browsers, WebRTC uses a mechanism called Interactive Connectivity Establishment (ICE) that is similar to the Session Initiation Protocol (SIP) commonly used for Voice-over-IP (VoIP) communication \cite{RFC5245:ice}.
+This process requires out-of-band exchange of information usually involving a third party server known as a signaling server.
+The signaling server may optionally provide other services such as higher level session management, HTTP serving of the web application, address book service, client presence/availability, etc.
 
@@
 
-The WebRTC browser APIs enable JavaScript web applications
-to establish browser-to-browser network connectivity in order to
-perform real-time video, audio and data communication. These APIs are
-currently in the process of being standardized by the \emph{Web
-Real-Time Communications Working Group} of the World Wide Web
-Consortium (W3C) \cite{webrtc:wg}.
+The WebRTC browser APIs enable JavaScript web applications to establish browser-to-browser network connectivity in order to perform real-time video, audio and data communication.
+These APIs are currently in the process of being standardized by the \emph{Web Real-Time Communications Working Group} of the World Wide Web Consortium (W3C) \cite{webrtc:wg}.
 
-There are three primary WebRTC interfaces:
+There are three primary WebRTC interfaces: \emph{RTCPeerConnection} \cite{webrtc:rtcpeerconnection:feb2015} for establishing direct connections between browsers, \emph{RTCDataChannel} \cite{webrtc:datachannel:feb2015} for sending and receiving generic data between browsers, and \emph{MediaStream~(aka~getUserMedia)} for managing audio and video streams (not relevant to Raft over WebRTC).
+The RTCDataChannel API supports both multiple modes for reliability and ordering which allows average latency/jitter to be decreased at the cost of some messages being dropped or arriving out of order.
+The Raft protocol is specifically designed to support dropped or out-of-order messages so the transport options available with RTCDataChannel can be used to gain extra efficiency.
 
-\begin{enumerate}
-\item \emph{RTCPeerConnection} \cite{webrtc:rtcpeerconnection:feb2015}:
-    The RTCPeerConnection API represents a direct connection between
-    two browsers. Each RTCPeerConnection object has an associated
-    Interactive Connectivity Establishment (ICE) agent that is used to
-    establish network connectivity through firewalls and NAT services.
+\subsubsection{WebRTC Protocols}
 
-\item \emph{MediaStream~(aka~getUserMedia)} \cite{webrtc:media-stream:feb2015}:
-    The MediaStream API represents a stream of audio or video data.
-    MediaStream data can easily be sent and received over
-    RTCPeerConnections. The MediaStream API definition provides
-    a mechanism that allows JavaScript to enumerate the native
-    MediaStreams that are provided by the browser such as a local web
-    camera or a live video stream of the desktop (for desktop
-    sharing).  Although support for MediaStreams was the original
-    impetus for the creation of the WebRTC APIs and protocols, the
-    Raft over WebRTC implementation does not make use of the
-    MediaStream API.
+There is a large suite of protocols that are required for a working WebRTC system.
+The Internet Engineering Task Force (IETF) organization is responsible for specifying and/or standardizing (where necessary) the protocols that are part of WebRTC \cite{draft-rtcweb-overview} \cite{draft-rtcweb-security-arch} \cite{draft-rtcweb-security}.
+In many cases an existing protocol is applicable and is referenced or extended as part of the WebRTC suite.
+The protocols that make up the WebRTC suite are categorized into five general functionality groups: data transport, data framing and security, connection management, data formats, presentation and control, and local system support functions.
+The last three groups focus on interactive audio and video applications and are not relevant to Raft over WebRTC.
 
-\item
-    \emph{RTCDataChannel} \cite{webrtc:datachannel:feb2015}:
-    The RTCDataChannel API provides a mechanism for sending and
-    receiving generic data between User Agents (browsers). A single
-    RTCPeerConnection can contain multiple RTCDataChannels. The
-    RTCDataChannel API is modeled after the WebSocket API, however,
-    unlike the WebSocket API which is reliable and guaranteed order
-    transport (like TCP), each RTCDataChannel can be configured with
-    different reliability and ordering settings. For example, by
-    relaxing the reliability and ordering constraints on
-    a RTCDataChannel, the average latency and jitter per message can
-    be decreased even though some messages may be dropped or arrive
-    out of order. The Raft protocol is specifically designed to
-    support dropped or out-of-order messages so the transport options
-    available with RTCDataChannel can be used to gain extra
-    efficiency.
-    %TODO/CITE: more of them above
-\end{enumerate}
+\section{Design and Implementation}
 
-\subsubsection{WebRTC Protocol}
+\begin{figure}[!t]
+  \centerline{\includegraphics[width=0.45\textwidth]{imgs/raft_rtc_architecture}}
+  \caption{Raft over RTC Architecture}
+    \label{fig:raft_rtc_architecture}
+\end{figure}
 
-There is a large suite of protocols that are required for a working
-WebRTC system. The Internet Engineering Task Force (IETF) organization
-is responsible for specifying and/or standardizing (where necessary)
-the protocols that are part of WebRTC \cite{draft-rtcweb-overview}
-\cite{draft-rtcweb-security-arch} \cite{draft-rtcweb-security}. In many
-cases an existing protocol is applicable and is referenced or
-extended as part of the WebRTC suite. %TODO/CITE
+\subsection{Raft.js Core Library}
 
-The protocols that make up the WebRTC suite are categorized into
-five general functionality groups:
-
-\begin{enumerate}
-    \item data transport
-        \ifdefined\OPTIONAL
-        \\
-        \cite{draft-rtcweb-transports}
-        \cite{draft-rtcweb-qos}
-        \cite{RFC4960:sctp}
-        \cite{RFC3758:sctp-reliability}
-        \cite{draft-sctp-dtls-encaps}
-        \cite{draft-mmusic-sctp-sdp}
-        \cite{draft-sctp-ndata}
-        \cite{draft-rtcweb-rtp-usage}
-        \cite{draft-rtcweb-data-channel}
-        \cite{draft-rtcweb-data-protocol}
-        \cite{draft-rtcweb-alpn}
-        \cite{RFC5245:ice}
-        \cite{RFC5128:p2p-nat}
-        \cite{RFC5389:stun}
-        \cite{RFC3489:stun-udp}
-        \cite{RFC5766:turn}
-        \cite{RFC6156:turn-ipv6}
-        \cite{RFC6544:tcp-ice}
-        \cite{RFC4571:rtp-rtcp}
-        \cite{RFC5764:dtls-srtp}
-        \fi
-    \item data framing and security
-        \ifdefined\OPTIONAL
-        \\
-        \cite{RFC3550:rtp}
-        \cite{RFC3711:srtp}
-        \cite{draft-rtcweb-rtp-usage}
-        \cite{draft-rtcweb-data-channel}
-        \cite{draft-rtcweb-data-protocol}
-        \fi
-    \item connection management
-        \ifdefined\OPTIONAL
-        \\
-        \cite{RFC2327:sdp}
-        \cite{RFC3264:sdp-offer-answer}
-        \cite{draft-rtcweb-jsep}
-        \cite{RFC5245:ice}
-        \cite{draft-mmusic-trickle-ice}
-        \cite{RFC5763:srtp-using-dtls}
-        \fi
-    \item data formats
-    \item presentation and control
-    \item local system support functions
-\end{enumerate}
-
-The last three functionality groups focus on interactive audio and video
-applications and are less relevant to Raft over WebRTC. \ifdefined\OPTIONAL
-Some of the important standards for the first three functionality
-groups are cited above. \fi
-
-The signaling transport itself is not defined as part of WebRTC.
-This is up to the application designer. The JavaScript interface used
-for establishing the initial connection (signaling) in the WebRTC 1.0
-standard is somewhat complicated. A competing standard called Object
-Real-time Communication (ORTC) is working to define a simpler and more
-modular JavaScript API for WebRTC signaling \cite{ortc2014}. However,
-at the time of this paper, the ORTC interface is not yet widely
-supported and is still in a state of flux. The use of PeerJS provides an
-abstraction that is likely to transparently support ORTC in the future.
-
-\ifdefined\OPTIONAL
+Raft.js is an implementation of the Raft algorithm in JavaScript that was created by the author of this paper \cite{raft.js}.
+Raft.js is designed to run in either a browser environment or within Node.js (server-side JavaScript).
+Raft.js implements the Raft algorithm as described in \cite[Consensus:~Diego]{raft_thesis:ongaro14}.
+The core Raft algorithm is defined in the \texttt{RaftServerBase} abstract class (in \texttt{base.js}) which omits the implementation of methods with side-effects such as RPC communication, durable storage, scheduling/timeouts, or state machine command/management.
+The full instantiation of the class (in \texttt{rtc.js} and \texttt{chat.js}) adds state machine commands and the ability to send RPCs over the WebRTC Data Channel.
 
@@
 
-PeerJS is a project that provides a simplified abstraction for using
-WebRTC. The first component of PeerJS is a node.js server library that
-implements a WebRTC signaling server. The second component is
-a JavaScript library that interacts with that signaling server and
-also provides a simpler interface for using the WebRTC APIs.
+PeerJS is a project that provides a simplified abstraction for using WebRTC.
+The first component of PeerJS is a node.js server library that implements a WebRTC signaling server.
+The second component is a JavaScript library that interacts with that signaling server and also provides a simpler interface for using the WebRTC APIs.
 
-\subsubsection{PeerJS Server}
+\subsection{Signalling Server}
 
-The PeerJS project provides a simple WebRTC signaling server called
-PeerServer. This server uses HTTP and WebSockets transport protocols
-to perform WebRTC signaling on behalf of browser clients. The
-PeerServer server is also extensible so that more advanced services
-can be built with it. The PeerJS project includes an extension to
-PeerServer called ExpressPeerServer that combines a PeerJS signaling
-server with the popular Node.js ExpressJS web application framework.
-% (TODO/CITE)
+The file `rtc\_server.js' implements the signaling server for the Raft over WebRTC application.
+The server extends the PeerJS `ExpressPeerServer' object in order to provide WebRTC signaling.
+In addition to the WebRTC signaling function the server also serves static web resources (HTML, CSS, and JavaScript) and WebRTC channel management endpoints.
 
-\subsubsection{PeerJS Client Library}
+\subsection{Client}
 
-The WebRTC API standards are not yet
-finalized and so different browsers and browser releases may support
-different versions of the WebRTC draft APIs. %TODO/CITE
-The PeerJS client library abstracts over these differences and
-provides a common interface so that the application developer does not
-have to deal directly with browser differences.
+The Raft over WebRTC chat application is a single web page (\texttt{chat.html}).
+When a new Raft cluster is created, the browser connects to the signalling server's new channel endpoint (Figure~\ref{fig:raft_rtc_sequence}~[1]).
+This endpoint causes the server to create a new channel for PeerJS signaling (Figure~\ref{fig:raft_rtc_sequence}~[2]).
+Once this is complete the server responds to the browser with a redirect message that contains the new channel identifier in the query parameter and ``firstServer” in the fragment identifier to indicate that this is the very first Raft cluster node (Figure~\ref{fig:raft_rtc_sequence}~[3]).
 
-The PeerJS client library also provides a default signaling mechanism
-that is designed to operate with a PeerJS signaling server. This is
-particularly helpful because much of the WebRTC signaling transport
-and protocol is not defined as part of the WebRTC standardization
-effort. %TODO/CITE
-The PeerJS organization provides a cloud-hosted version of a PeerJS
-compatible signaling server, however application developers can also
-run their own signaling service using PeerJS PeerServer.
+Subsequent nodes are added to the same Raft cluster by starting a new browser context (separate browser, browser window, or browser tab) and loading the same URL without the ``firstServer" fragment identifier (Figure~\ref{fig:raft_rtc_sequence}~[9]).
+There is a convenience link provided in the application page that opens a new browser window with a new cluster node.
 
-\fi  % \ifdefined\OPTIONAL
+When each page first loads, a PeerJS signaling connection is established with the server (Figure~\ref{fig:raft_rtc_sequence}~[6]~and~[11]).
+The web application also begins a periodic async polling function (\texttt{addRemoveServersAsync}) that compares the current active list of PeerJS clients to the Raft cluster membership list.
+If there are any differences between the two lists, then any new clients are added to the Raft cluster and missing clients are removed from the Raft cluster one at a time using the \texttt{addServer} and \texttt{removeServer} RPC calls.
+Servers additions are prioritized before removals in order to maximize availability \cite[Section~4.4]{raft_thesis:ongaro14}.
 
-%\begin{figure*}[!t]
+When an instance of the web application starts that is not a ``firstServer", the normal Raft timers are not enabled until the server receives its first RPC.
+The prevents the server from trying to becoming a leader before it is incorporated into the cluster by an appendEntries RPC from the true cluster leader.
+
+\subsection{Chat Application}
+
 \begin{figure}[!t]
-%\centerline{\includegraphics[width=0.8\textwidth]{imgs/raft_rtc_sequence}}
   \centerline{\includegraphics[width=0.45\textwidth]{imgs/raft_rtc_sequence}}
@@
 
-\section {Design and Implementation}
-
-\ifdefined\OPTIONAL
-
-\subsection{Incremental Steps}
-
-During the planning phase, the project was broken down into ten incremental steps:
+The chat application extends the basic Raft.js over WebRTC client.
+The chat application provides a large text area containing the chat history and a text entry field below it.
+When the user clicks on the \emph{Send} button (or presses enter), the current node ID is prepended to the beginning of the text line and it is added to a pending send queue and the text entry field is blanked.
+The chat application also extends the default set of state machine operations to add a \emph{seqAppend} operation.
+This is an atomic operation that is implemented as follows:
 
 \begin{enumerate}
-\item Simple WebRTC messaging: implement a small JavaScript
-    application that uses the PeerJS client library and an unmodified
-    PeerJS signaling server. Demonstrate communication over WebRTC
-    DataChannel using two browser instances (or browser tabs).
-\item Raft.js over WebRTC: combine the existing Raft.js implementation
-    with the small JavaScript application above to demonstrate Raft
-    over WebRTC. This step uses a statically configured three node
-    cluster (three browser tabs). The PeerJS signaling server is
-    polled to retrieve the list of other clients and the Raft cluster
-    is started when the number of PeerJS clients reaches three.
-\item Update Raft.js protocol: update the Raft.js implementation to
-    match the latest description of the Raft protocol
-    \cite{raft_thesis:ongaro14}. This includes the implementation of
-    the simplified membership change algorithm (sequential changes)
-    and the separation of the log commit and state machine apply
-    concepts.  Additionally, the new Raft.js implementation always
-    starts a new Raft cluster as a cluster with a single node and
-    additional nodes are dynamically added to the cluster
-    \cite[Section~4.4]{raft_thesis:ongaro14}.
-\item Dynamic Raft.js over WebRTC: enhance the signaling server with
-    support for multiple sessions/channels to enable separate Raft
-    clusters with the same signaling server. Start each cluster with
-    a single node and automatically add and remove Raft nodes as they
-    come and go.
-\item Visual representation: display Raft node state and RPC
-    statistics in addition to the internal Raft logging messages.
-\item Sessionless/Datagram RPC messages: change Raft RPC requests and
-    response messages so that RPC responses have enough context to be
-    handled separately from RPC requests. Specifically this enables
-    Raft to support datagram (sessionless) network transports like
-    WebRTC.
-\item Atomic state machine operation: implement a state machine
-    update operation that only succeeds if no other operation has been
-    applied to the same key since the last time this client read from
-    that key.
-\item Automatic client request forwarding: extend the client request
-    RPC to also be sessionless/datagram based. Implement an
-    asynchronous JavaScript function that determines which node is the
-    leader and forwards client request RPCs to the leader. This
-    function must also correlate RPC responses to the client request
-    and invoke the callback for the original request.
-\item Chat application: add the ability to send a chat message to the
-    log using the client request forwarding functionality together
-    with atomic state machine operation. Show the current chat history
-    based on the chat entries in the state machine.
-\item Server as Raft.js peer (aspirational): enable server nodes
-    (Raft.js running on Node.js) to fully participate as Raft cluster
-    nodes.
+\item The \emph{seqAppend} operation operates on a specially structured key/value pair in the state machine map.
+The value of the pair is another associative map that contains a counter and list structure where the actual chat lines are accumulated.
+\emph{seqAppend} operations are sent with both a count and then new value to append.
+\item When a \emph{seqAppend} operation is applied to the state machine and the count value in the command matches the counter in the state machine's value, then the new line in the request is added to the value list and the counter is incremented.
+If the count value does not match the counter then the the operation is not applied.
+In either case, a list is returned in the \texttt{clientRequestResponse} that contains two values: a boolean indicating whether the \emph{seqAppend} operation succeeded and a number indicating the current value of the counter.
+\item The first time the \emph{seqAppend} operation is used on a non-existent key, the structure described above is created with the counter set to 0 and the list structure is created with the single value that was passed in the \emph{seqAppend} operation.\
+\item Clients that call the \emph{seqAppend} operation should check the status value in the result list.
+If the status is false, then the client should retry the operation.
+Regardless of success of failure, then client should store the current value of the counter for the next \emph{seqAppend} call or retry.
 \end{enumerate}
 
-Steps 1-9 were successfully implemented for this paper. Step 10 was an
-aspirational goal because this functionality is not yet supported in
-PeerJS \cite{peerjs:103}.
+\section{Experimental Results}
 
-\fi  % \ifdefined\OPTIONAL
+\subsection{Test Setup}
 
-\subsection{Server}
+Automated testing of the Raft over WebRTC system was performed by leveraging the SlimerJS scriptable browser environment.
+SlimerJS enables web applications to be scripted by loading pages in a special Mozilla rendering environment called XULRunner.
+The XULRunner instance is directed to an X virtual frame buffer (xvfb) which provides effectively headless testing.
 
-The file `rtc\_server.js' implements the signaling server for the Raft
-over WebRTC application. The server extends the PeerJS
-`ExpressPeerServer' object in order to provide WebRTC signaling. In
-addition to the signaling function, the server also provides the
-following service endpoints:
+To simplify the execution of the tests and to make the testing infrastructure more portable, a docker container environment encapsulates and executes the SlimerJS environment.
+The container is built using \emph{test/Dockerfile} which is in turn derived from the ``fentas/slimerjs" docker image \cite{fentas:slimerjs}.
+This custom Dockerfile was created because the upstream version of the Dockerfile used an older version of XULRunner that did not support more than 7 WebRTC clients communicating together simultaneously.
+This issue was discovered during initial testing.
+
+Writing tests using SlimerJS involves creating JavaScript test programs which start a server instance and then launch a SlimerJS docker container for each node of the cluster.
+Each container starts a XULRunner browser instance of the Raft over WebRTC chat application.
+Once all the chat application nodes have connected to the test server/program, then the actual test is started across all the application nodes.
+
+Three different tests programs/servers were created to test the basic Raft over WebRTC application.
+Each of the tests takes a parameter indicating the number of nodes/pages to launch.
+The test source files are:
 
 \begin{enumerate}
-\item Static web service endpoint: this serves the static HTML, CSS
-    and JavaScript that make up the Raft over WebRTC web application.
-\item New channel endpoint: when a browser loads this endpoint, a new
-    PeerJS channel is created and the browser is redirected to a URL
-    for loading the main application (via static file service above).
-    The redirect URL has a query string appended that contains the
-    channel ID of the new channel that was just created. In addition
-    the redirect URL has a fragment identifier (hash/history) that
-    communicates to the web application that this is the first server
-    in the channel.
-\item Peer list endpoint: this endpoint takes a channel ID and returns
-    the list of WebRTC peers (browser user agents) that are currently
-    connected to that channel.
-
+\item \emph{test/common.js}: a module of common functions used by the other tests scripts.
+\item \emph{test/test\_up.js}: launch and start an initial leader node/page, then start the rest of the nodes (if any) so that the leader will configure them into the cluster.
+Then test then waits for every node to be recorded (applied) into the log of every other node of the cluster.
+The test reports the number of milliseconds for the cluster to reach this state.
+\item \emph{test/test\_kill\_nodes.js}: similar to \emph{test/test\_up.js} but once all the nodes are configured into the cluster, this script then kills the leader plus a configurable number of nodes.
+By default the script kills just under half the nodes of the cluster.
+The test then waits for a new leader to be elected and remove the dead nodes from the cluster.
+The test reports both the number of milliseconds for the cluster to come up and the number of milliseconds for the cluster to fully recover from the failure and evict the failed nodes.
+\item \emph{test/test\_chat\_propagate.js}: similar to \emph{test/test\_up.js} but once the cluster is up, the test connects to one of the non-leader nodes and and sends a chat text line.
+The test then waits for the chat message to propagate to every other node's state machine.
+The test reports both the number of milliseconds for the cluster to come up and for the messages to propagate to every node.
 \end{enumerate}
 
-Refer to Section~\ref{section:design-client} and
-Figure~\ref{fig:raft_rtc_sequence} for a description of how these
-endpoints interact.
+The tests were all run on a system with 8 Intel Core i7 processors running at 3.20GHz and 12 GB of main memory.
+The Docker image used to run the test was built using Ubuntu 12.04.5 LTS with \emph{SlimerJS} version 0.9.6, and \emph{XULRunner} version 38.0.5.
 
-\subsection{Client}
-    \label{section:design-client}
-
-The Raft over WebRTC testing application is designed as a single web
-page (\texttt{rtc.html} or \texttt{chat.html}). The application has
-different starting behavior depending on the value of the URL query
-string (the URL component following the ``?" and up to the ``\#") and
-the fragment identifier (the URL component after the ``\#").
-
-When a new Raft cluster is created, the user agent (browser) connects
-to the server's new channel endpoint
-(Figure~\ref{fig:raft_rtc_sequence}~[1]). This endpoint causes the
-server to create a new channel for PeerJS signaling
-(Figure~\ref{fig:raft_rtc_sequence}~[2]). Once this is complete the
-server responds to the browser with a redirect message
-that contains the new channel identifier in the query parameter and an
-indication that this is the very first Raft cluster node in the
-fragment identifier (Figure~\ref{fig:raft_rtc_sequence}~[3]).
-
-The browser follows the redirect and loads the static web application
-at the static file endpoint provided by the server
-(Figure~\ref{fig:raft_rtc_sequence}~[4]). This includes the
-main application page (\texttt{rtc.html} or \texttt{chat.html}), the
-core raft algorithm (\texttt{base.js} and \texttt{local.js}), the
-PeerJS client library (\texttt{peer.js}), the actual Raft over WebRTC
-implementation (\texttt{rtc.js}), a layout stylesheet
-(\texttt{web/demo.css}), and higher-level application code if that is
-being loaded (\texttt{chat.js})
-(Figure~\ref{fig:raft_rtc_sequence}~[5]).
-
-If the URL fragment identifier is set to ``firstServer", then this Raft
-node is initialized as the leader of a single node cluster and all
-normal timers are started. The fragment identifier is then unset. The
-query parameter contains the ID of this cluster (the PeerJS channel).
-
-Subsequent nodes are added to the same Raft cluster by starting a new
-browser context (separate browser, browser window, or browser tab) and
-loading the same URL without the ``firstServer" fragment identifier
-(Figure~\ref{fig:raft_rtc_sequence}~[9]). There is a convenience link
-provided in the application page that opens a new browser window with
-a new cluster node (see \ref{fig:browser_1}).
-
-When each page first loads, a PeerJS signaling connection is
-established with the server
-(Figure~\ref{fig:raft_rtc_sequence}~[6]~and~[11]). Once this
-connection is established, the application will be notified when any
-PeerJS clients connect to or disconnect from the server.  After the
-PeerJS signaling connection is established, the client immediately
-queries the Peer list endpoint
-(Figure~\ref{fig:raft_rtc_sequence}~[7]~and~[12]~and~[14]]) to discover
-any clients that were connected prior to when this node connected to
-the server (Figure~\ref{fig:raft_rtc_sequence}~[8]~and~[13]~and~[15]]).
-
-The web application also begins a periodic async polling function
-(\texttt{addRemoveServersAsync}) when the page loads. When the node is a Raft
-leader, the function compares the current active list of PeerJS clients to
-the Raft cluster membership list. If there are any differences between
-the two lists, then any new clients are added to the Raft cluster and
-missing clients are removed from the Raft cluster one at a time using
-the \texttt{addServer} and \texttt{removeServer} RPC calls. Servers
-additions are prioritized before removals 
-in order to maximize availability
-\cite[Section~4.4]{raft_thesis:ongaro14}.
-
-When an instance of the web application starts that is not
-a ``firstServer", the normal Raft timers are not enabled until the
-server receives its first RPC. The prevents the server from
-trying to becoming a leader before it is incorporated into the cluster
-by an appendEntries RPC from the true cluster leader.
-
-In addition to the low level Raft message log, the following
-statistics are shown in the application:
-
-\begin{enumerate}
-\item \emph{Term}: the current Raft leadership term.
-\item \emph{State}: the current Raft node state (follower, candidate,
-    or leader)
-\item \emph{Cluster Size}: a count of the current Raft cluster
-    membership (number of nodes currently participating in the
-    cluster)
-\item \emph{Log Length}: number of entries in the transaction log.
-\item \emph{Request Votes Sent/Received}: the number of `requestVote`
-    RPCs that have been sent from and received by this node.
-\item \emph{Append Entries Sent/Received}: the number of
-    `appendEntries' RPCs that have been sent from and received by this
-    node.
-\end{enumerate}
-
-\subsubsection{Client Request Redirects}
-
-% TODO: rename clientRequest* function to not clash with RPC names
-% (maybe raftRequest*?)
-
-In the Raft over WebRTC implementation each local browser is a client
-of the Raft cluster. However in Raft, the leader node is the only node
-that adds entries to the log so clients on follower nodes must have
-a way to communicate with the leader. This is implemented as a pair of
-functions named \texttt{clientRequest} and
-\texttt{clientRequestResponse}. The \texttt{clientRequest} function is
-called by the client with the arguments of a client request RPC and
-a callback function to be called once the request is completed. The
-\texttt{clientRequest}/\texttt{clientRequestResponse} functions work
-as follows:
-
-\begin{enumerate}
-\item The \texttt{clientRequestResponse} is registered with the local
-    Raft.js implementation as the callback function anytime
-    a \texttt{clientRequestResponse} RPC is received.
-\item The \texttt{clientRequest} function keeps track of the current
-    leader ID.  If the current leader ID is not set or this node is
-    the leader then the client request RPC is sent directly to the
-    local Raft node bypassing the network transport layer.  If the
-    current leader ID is set then the request is sent to that node as
-    a normal RPC request over the WebRTC channel.
-\item When the \texttt{clientRequestResponse} function is called and
-    the status is ``NOT\_LEADER" then the current leader ID being
-    tracked by the \texttt{clientRequest} function is set to the
-    leader ID specified in the response. Then the original request is
-    re-issued over the normal WebRTC channel to the leader.
-\item When the \texttt{clientRequestResponse} function is called with
-    a successful status, the original callback argument of the
-    \texttt{clientRequest} function is called with the results from
-    the RPC response.
-\end{enumerate}
-
-\subsection{Chat Application}
-
-The chat application extends the basic Raft.js over WebRTC client. The
-chat application provides a large text area containing the chat
-history and a text entry field below it
-(see Figure~\ref{fig:browser_chat_1}). When the user clicks on the
-\emph{Send} button (or presses enter), the current node ID is
-prepended to the beginning of the text line and it is added to
-a pending send queue and the text entry field is blanked. A function
-to flush sends runs periodically (currently every 100ms) and does the
-following:
-% TODO: 100ms above should be less
-
-\begin{enumerate}
-\item If there is line of text that has been sent but not yet
-    added (applied) to the state machine or there are no lines of text
-    on the pending send queue, then the function does nothing.
-\item A text line is removed from the head of the pending send queue
-    and stored in \emph{curSend} to indicate that a send is in
-    progress. Then a client request is made using an atomic
-    \emph{seqAppend} command to add the new text to the state machine
-    (\emph{seqAppend} is described below).  The \emph{seqAppend}
-    operation attempts to append the text line to a list at the
-    \emph{history} key in the state machine.
-\item When the atomic \emph{seqAppend} operation completes, it may
-    have failed (if another text line was appended first), in which
-    case the flush function puts the failed text line back on the
-    pending send queue. Whether the send was successful or failed, the
-    \emph{curSend} variable is set to nil/null to indicate that no
-    send is in progress. The next time the flush function runs it will
-    attempt to send again. % TODO: reword
-\end{enumerate}
-
-% TODO: normalize use of \texttt and \emph
-
-The chat application also extends the default set of state machine
-operations to add a \emph{seqAppend} operation. This is an atomic
-operation that is implemented as follows:
-
-\begin{enumerate}
-\item the \emph{seqAppend} operation operates on a specially structured
-    key/value pair in the state machine map. The value of the pair is
-    another associative map that contains a counter and list structure
-    where the actual chat lines are accumulated. \emph{seqAppend}
-    operations are sent with both a count and then new value to
-    append.
-\item when a \emph{seqAppend} operation is applied to the state
-    machine and the count value in the command matches the counter in
-    the state machine's value, then the new line in the request is
-    added to the value list and the counter is incremented. If the count
-    value does not match the counter then the the operation is not
-    applied. In either case, a list is returned in the
-    \texttt{clientRequestResponse} that contains two values: a boolean
-    indicating whether the \emph{seqAppend} operation succeeded and
-    a number indicating the current value of the counter.
-\item the first time the \emph{seqAppend} operation is used on a non-existent
-    key, the structure described above is created with the counter set
-    to 0 and the list structure is created with the single value that
-    was passed in the \emph{seqAppend} operation.
-\item clients that call the \emph{seqAppend} operation should check the
-    status value in the result list. If the status is false, then the
-    client should retry the operation. Regardless of success of
-    failure, then client should store the current value of the counter
-    for the next \emph{seqAppend} call or retry.
-\end{enumerate}
-
-\section{Results}
-
-\ifdefined\OPTIONAL
-
-\subsection{Manual Testing}
-
-\subsubsection{Setup}
-
-The initial way that the application was tested was by loading the
-basic application to create a Raft over WebRTC cluster with a single
-node acting as leader. Figure~\ref{fig:browser_1} shows a screenshot
-of the browser after the first node is created:
-
-\begin{figure}[!t]
-    \centering
-    \includegraphics[width=0.45\textwidth]{imgs/browser_1.png}
-    %\captionof{figure}{Basic app with one Node/Tab}
-    \caption{Basic app with one Node/Tab}
-    \label{fig:browser_1}
-\end{figure}
-
-Then four more Raft nodes were then added to the cluster by clicking
-on the embedded ``create more nodes" link in the application.
-Figure~\ref{fig:browser_5} shows a screenshot of the browser once four
-new nodes are added to the cluster (5 total):
-
-\begin{figure}[!t]
-    \centering
-    \includegraphics[width=0.45\textwidth]{imgs/browser_5.png}
-    %\captionof{figure}{Basic app with Five Nodes/Tabs}
-    \caption{Basic app with Five Nodes/Tabs}
-    \label{fig:browser_5}
-\end{figure}
-
-In addition, the full chat application was also tested manually by
-loading the initial chat application landing page to create a Raft
-over WebRTC cluster with a single node acting as leader. Then lines
-of text were entered into the text area each followed by clicking the
-Send button. Figure~\ref{fig:browser_chat_1} shows a screenshot of the
-browser after the first node/tab of the chat application is created
-and a line of text was sent:
-
-\begin{figure}[!t]
-    \centering
-    \includegraphics[width=0.45\textwidth]{imgs/chat_1a.png}
-    %\captionof{figure}{Chat app with one Node/Tab}
-    \caption{Chat app with one Node/Tab}
-    \label{fig:browser_chat_1}
-\end{figure}
-
-Then four more Raft nodes were then added to the cluster by clicking
-on the embedded ``create more clients" link in the application.
-Finally, on a non-leader node more text lines were entered and sent by
-clicking the Send button. Figure~\ref{fig:browser_chat_5} shows
-a screenshot of the browser once four new nodes/tabs are added to the
-cluster (5 total) and another text line was sent from the fifth node:
-
-\begin{figure}[!t]
-    \centering
-    \includegraphics[width=0.45\textwidth]{imgs/chat_5a.png}
-    %\captionof{figure}{Chat app with Five Nodes/Tabs}
-    \caption{Chat app with Five Nodes/Tabs}
-    \label{fig:browser_chat_5}
-\end{figure}
-
-After every change was made to the cluster all the nodes were checked
-to ensure that the following steady state properties were observed:
-\begin{enumerate}
-\item only a single leader
-\item all nodes have the same term and transaction log count
-\item all non-leader nodes are receiving `appendEntries' RPCs from the leader
-\item all nodes have the same cluster membership state
-\item the transaction log entries are identical on all nodes
-\item the state machine is identical on all nodes
-\end{enumerate}
-
-The current leader was then terminated. The other nodes were checked
-again the steady state properties. This was repeated until only two
-nodes remained in the cluster.
-
-Next the five node cluster was brought up again and state machine
-commands were sent to leader to add to the transaction log. Again the
-nodes were checked for the above steady state properties.
-
-Then the application was tested by loading the initial cluster
-creation landing page in a browser tab. Nodes were randomly added to
-the cluster (by clicking on the link embedded in the page) or removed
-from the cluster (by closing a tab). After each change the steady
-state properties were checked.
-
-Similar testing to the above was done with browser instances on two
-separate computers. In addition, cross-browser functionality was
-verified by running nodes from the same cluster running simultaneously
-in Chrome 42 and Firefox 37.
-
-\subsubsection{Outcome}
-
-The results of the manual testing above showed that Raft over WebRTC
-was able to perform leader elections correctly, the log and state
-machine were replicated correctly to all nodes, and the safety
-properties were maintained. During testing the membership of the
-cluster was changed numerous times (both node additions and removals)
-and the Raft properties were maintained throughout. In addition,
-client interaction was tested to make sure commands were still
-properly added to the state machine and transaction log.
-
-One limitation of the Raft protocol is that if the cluster membership drops from
-two nodes to one node, then the cluster will become unavailable without manual
-intervention. This is because both nodes of a two-node cluster are required for
-a majority vote.  When one member of a two-node cluster is forcibly shut down,
-then the remaining cluster node will be unable to win elections or even commit
-transaction entries, and the cluster will become unavailable. There are several
-options to get around this problem:
-
-% NOTE: optional section
-
-\begin{enumerate}
-\item The node can ask the leader to be removed and then wait for the
-  transaction to become effective before it disconnects.  This may not
-  always be an option especially in a browser context where the user
-  may immediately kill the window/tab without giving the application
-  a chance to remove itself from the cluster.
-\item The signaling server can serve as a tie-breaker. This would allow
-  the remaining leader successfully commit the removal of the other
-  node. In the case of a partition, the signaling server must only
-  vote for one of the nodes.
-\item The node with the lower ID could have an implicit tie-breaker vote.
-  This would allow one node of the partition to continue in the case
-  of a network partition. If a node dies or is killed, this would only
-  allow the cluster to continue to operate if the node that survives
-  happens to have the lower ID.
-\item The system could allow manual user intervention. However, this is
-  only likely to work correctly in the case where the user of the last
-  remaining node can be certain that other nodes will never return;
-  otherwise there is a risk of data-corruption.
-\end{enumerate}
-
-Running a Raft cluster in a browser context can lead to some level of
-timer instability. The JavaScript execution context is a fully event
-driven environment with a single thread of execution for a given
-JavaScript context. JavaScript code is run when browser (or Node.js)
-events trigger JavaScript code that is registered to that event. In
-order to implement logic that runs periodically, JavaScript code is
-registered as a callback timer using the setTimer, setInterval or
-requestAnimationFrame APIs. However, none of these mechanisms is
-a precise timer. When a timer fires, the callback code it refers to
-is added to the JavaScript execution queue and events on the queue are
-processed in the order they arrive. This means callbacks may be
-delayed by other code that is already running or queued first
-\cite{resig2013secrets}.
-
-However, there is another browser mechanism that is less well known
-but that causes even larger delays in timer code execution. Modern
-browsers often have dozens or even hundreds of windows and/or tabs
-open simultaneously. In order to reduce CPU load, increase user
-responsiveness and improve battery life, most modern browsers will
-throttle the JavaScript callback timers for all windows and tabs that
-are hidden or have not been directly interacted with for some period
-of time. This throttling can cause timers to slow down by an order of
-magnitude or more. %TODO/CITE
-
-In the context of Raft over WebRTC, the slowdown of background page
-timers means that whenever the leader node is residing on a page that
-is in the background, the propagation of transaction log entries slows
-down substantially. This also means that other nodes which are in the
-foreground are more likely to have an election timeout timer fire and
-they are more likely to become a leader because they will be able to
-send and receive votes more quickly than nodes that are in the
-background.  In this sense the problem is self-correcting over
-the long term because higher performing nodes more are likely to start
-and win elections. This also indicates that dynamic adjustment of timeout
-values may be a fruitful area of further study.
-
-In the context of the full chat application the slowdown of background
-pages is even more pronounced because each addition to the chat
-history can take several messages to fully propagate the new date. An
-interesting area of further study would be to extend the Raft protocol
-to more proactively adapt to network and node performance conditions
-(delayed timers) and more intentionally transition leadership to more
-capable nodes.
-
-\fi  % \ifdefined\OPTIONAL
-
-\subsection{Automated Testing}
-
-\subsubsection{Twst Test Framework}
-
-In order to facilitate coordinated testing across a large number of
-browser instances, a new test framework was created called
-\cite{twst}. This \emph{Twst} framework provides a client and server
-library written in JavaScript:
-
-\begin{enumerate}
-    \item \emph{server library}: this library is used to write tests
-        that need to coordinate testing across many browser instances.
-        It provides services such as client tracking/enumeration,
-        broadcast and targeted RPC, state rendezvous, log collection,
-        etc.
-    \item \emph{client library}: this library is included with the web
-        application that is being tested. When the page loads, the
-        library will automatically connect to the \emph{Twst} server
-        instance and begin sending console logs and accepting
-        RPCs/commands from the server.
-\end{enumerate}
-
-\subsubsection{Test Setup}
-
-Automated testing of the Raft over WebRTC system was performed by
-leveraging the SlimerJS scriptable browser environment. SlimerJS
-enables web applications to be scripted by loading pages in a special
-Mozilla rendering environment called XULRunner. The XULRunner instance
-is directed to an X virtual frame buffer (xvfb) which provides
-effectively headless testing.
-
-To simplify the execution of the tests and to make the testing
-infrastructure more portable, a docker container environment was used
-to encapsulate and execute the SlimerJS environment. The container is
-built using \emph{test/Dockerfile} which is in turn derived from
-the ``fentas/slimerjs" docker image \cite{fentas:slimerjs}.
-This custom Dockerfile was created because the upstream version of the
-Dockerfile used an older version of XULRunner that did not support
-more than 7 WebRTC clients communicating together simultaneously. This
-issue was discovered during initial testing.
-
-Writing tests using SlimerJS involves creating JavaScript test programs
-which start a \emph{Twst} server instance and then launch a SlimerJS
-docker container for each node of the cluster. Each container starts
-a XULRunner browser instance of the Raft over WebRTC chat application.
-Also included with the application is the \emph{Twst} client library
-which connects back to the \emph{Twst} service in the test program.
-Once all the chat application nodes have connected to the test
-server/program, then the actual test is started across all the
-application nodes.
-
-Three different tests programs/servers were created to test the basic
-Raft over WebRTC application. Each of the tests takes a parameter
-indicating the number of nodes/pages to launch. The test
-source files are:
-
-\begin{enumerate}
-    \item \emph{test/common.js}: a module of common functions used by
-        the other tests scripts.
-    \item \emph{test/test\_up.js}: launch and start an initial leader
-        node/page, then start the rest of the nodes (if any) so that
-        the leader will configure them into the cluster. Then test
-        then waits for every node to be recorded (applied) into the log
-        of every other node of the cluster. The test reports the
-        number of milliseconds for the cluster to reach this state.
-    \item \emph{test/test\_kill\_nodes.js}: similar to
-        \emph{test/test\_up.js} but once all the nodes are configured
-        into the cluster, this script then kills the leader plus
-        a configurable number of nodes. By default the script kills
-        just under half the nodes of the cluster. The test then waits
-        for a new leader to be elected and remove the dead nodes from
-        the cluster. The test reports both the number of milliseconds
-        for the cluster to come up and the number of milliseconds for
-        the cluster to fully recover from the failure and evict the
-        failed nodes.
-    \item \emph{test/test\_chat\_propagate.js}: similar to
-        \emph{test/test\_up.js} but once the cluster is up, the test
-        connects to one of the non-leader nodes and and sends a chat
-        text line. The test then waits for the chat message to
-        propagate to every other node's state machine. The test
-        reports both the number of milliseconds for the cluster to
-        come up and for the messages to propagate to every node.
-\end{enumerate}
-
-The tests were all run on a system with the following configuration:
-
-\begin{enumerate}
-    \item \emph{CPUs}: 8 Intel Core i7 processors running at 3.20GHz
-    \item \emph{Memory}: 12 GB
-    \item \emph{OS}: Ubuntu 14.04.1 LTS
-    \item \emph{Docker}: version 1.3.1
-    \item \emph{Docker image}:
-    \begin{enumerate}
-        \item \emph{OS}: Ubuntu 12.04.5 LTS
-        \item \emph{SlimerJS}: version 0.9.6
-        \item \emph{XULRunner}: 38.0.5
-    \end{enumerate}
-\end{enumerate}
-
-Five different tests (using different options for the three test
-programs) were run at 17 different cluster sizes (odd cluster sizes
-from 3 to 35 nodes). Each of these tests was run with three different
-Raft election timeout values: 250ms, 500ms, 1000ms. Each of the five
-tests and cluster size combinations was run 7 times.
-
-\subsubsection{Outcome}
+\subsection{Outcome}
 
@@
         xlabel={Nodes}, ylabel={Time (seconds)},
-        xmin=0, xmax=43,
+        xmin=0, xmax=37,
         ymin=0, ymax=15,
@@
 
-Figure~\ref{fig:propagate1} and Figure~\ref{fig:propagate100} show the
-result of running the \emph{test/wait\_chat\_propagate.js} test to
-determine the amount of time that a stable cluster takes to propagate,
-commit and apply 1 and 100 chat message commands respectively to the state
-machine on every cluster node using the atomic \emph{seqAppend}
-command. The commands are serialized which means that the previous
-command must be committed before the current command can be committed.
-The test is complete once the last chat message command has been both
-committed and applied to every node of the cluster. Each test execution
-is shown as a blue ``x" and the black line represents the average for
-each cluster size.
+Figure~\ref{fig:propagate1} and Figure~\ref{fig:propagate100} show the result of running the \emph{test/wait\_chat\_propagate.js} test to determine the amount of time that a stable cluster takes to propagate, commit and apply 1 and 100 chat message commands respectively to the state machine on every cluster node using the atomic \emph{seqAppend} command.
+The commands are serialized which means that the previous command must be committed before the current command can be committed.
+The test is complete once the last chat message command has been both committed and applied to every node of the cluster.
+Each test execution is shown as a blue ``x" and the black line represents the average for each cluster size.
 
-In Figure~\ref{fig:propagate1} the time to propagate a single message
-to a single node of the cluster is approximately 0.1 seconds. However,
-this really just represents the tests scanning/checking granularity.
-The time to finish propagating and committing a single message shows
-a slight linear slope upwards from about 0.3 seconds for a three node
-cluster to about 0.5 seconds for a 41 node cluster. The time to
-propagate 100 messages (Figure~\ref{fig:propagate100}) has a much
-higher slope and also appears to be non-linear. TODO: why?
+In Figure~\ref{fig:propagate1} the time to propagate a single message to a single node of the cluster is approximately 0.1 seconds.
+However, this really just represents the tests scanning/checking granularity.
+The time to finish propagating and committing a single message shows a slight linear slope upwards from about 0.3 seconds for a three node cluster to about 0.5 seconds for a 41 node cluster.
+The time to propagate 100 messages (Figure~\ref{fig:propagate100}) has a much higher slope and also appears to be non-linear.
 
-Figure~\ref{fig:cluster_up} shows the results of running the
-\emph{test/test\_up.js} test to determine the amount of time that
-a cluster takes to start. The cluster is fully started once a
-leader has been elected and the leader has successfully added all the
-other nodes to the running configuration. Each election timer value
-is represented as a different color and the line represents the
-average value for each cluster size. The cluster startup times show
-a fairly linear increase as the cluster size increases. The reason
-that the slope varies with election timer is due to the fact that the
-election timer also determines the leader heartbeat value (currently
-set to one fifth of the election timer value). The leader heartbeat
-determines the frequency with which \emph{appendEntries} messages are
-sent from the leader to followers and is also used to propagate log
-entries.  This means that the rate at which log entries (and thus
-cluster configuration updates) can be propagated is proportional to
-the election timer value.
+Figure~\ref{fig:cluster_up} shows the results of running the \emph{test/test\_up.js} test to determine the amount of time that a cluster takes to start.
+The cluster is fully started once a leader has been elected and the leader has successfully added all the other nodes to the running configuration.
+Each election timer value is represented as a different color and the line represents the average value for each cluster size.
+The cluster startup times show a fairly linear increase as the cluster size increases.
+The reason that the slope varies with election timer is due to the fact that the election timer also determines the leader heartbeat value (currently set to one fifth of the election timer value).
+The leader heartbeat determines the frequency with which \emph{appendEntries} messages are sent from the leader to followers and is also used to propagate log entries.
+This means that the rate at which log entries (and thus cluster configuration updates) can be propagated is proportional to the election timer value.
 
-Figures \ref{fig:recovery_leader_failure} and
-\ref{fig:recovery_half_failure} show the results of running the
-\emph{test/test\_kill\_nodes.js} in both modes (kill only leader and
-kill just under half the nodes). Each election timer value is
-represented as a different color and the solid line represents the
-average recovery time for each cluster size. When only the leader is
-killed, the recovery times are constant flat regardless of cluster
-size. This is because after a new leader is elected, there is only
-a single configuration entry to propagate to the rest of the cluster
-(to remove the old failed leader). One interesting thing to note is
-that with a 1000ms timer, the 3 node cluster is actually the most
-inefficient size because the average time until a leader election
-starts is inversely proportional to cluster size so the delay until an
-election happens is a significant contributor to the three node
-cluster with 1000ms timer. With a larger cluster or shorter election
-timer, the contribution of election delay is not noticeable. The
-recovery from almost half the nodes failing is not flat because the
-amount of configuration information that needs to be propagated is
-proportional to the size of the cluster.
+Figures \ref{fig:recovery_leader_failure} and \ref{fig:recovery_half_failure} show the results of running the \emph{test/test\_kill\_nodes.js} in both modes (kill only leader and kill just under half the nodes).
+Each election timer value is represented as a different color and the solid line represents the average recovery time for each cluster size.
+When only the leader is killed, the recovery times are constant flat regardless of cluster size.
+This is because after a new leader is elected, there is only a single configuration entry to propagate to the rest of the cluster (to remove the old failed leader).
+One interesting thing to note is that with a 1000ms timer, the 3 node cluster is actually the most inefficient size because the average time until a leader election starts is inversely proportional to cluster size so the delay until an election happens is a significant contributor to the three node cluster with 1000ms timer.
+With a larger cluster or shorter election timer, the contribution of election delay is not noticeable.
+The recovery from almost half the nodes failing is not flat because the amount of configuration information that needs to be propagated is proportional to the size of the cluster.
 
-At larger cluster sizes (35 and greater) the cluster begins to
-experience failures. For example, when testing with a 250ms election
-timer, once the cluster size reaches 35 nodes, the cluster is
-frequently (more than 50\% of the time) unable to recover from the
-half cluster shutdown test. At a cluster size or 41 and with a 1000ms
-election timer, the cluster begins to experience difficulty
-successfully starting up (about 50\% failure rate). The basic problem
-appears to be due to the average election timeout period being too low
-in relation to communication latency. During leader election, the
-first candidate is unable to gather enough votes to become leader
-before other candidates arise and steal votes which prevents a quorum.
+\section{Future Work}
 
-Some (or most) of the communication latency may be a result of CPU or
-memory contention due to all the browser instances running on a single
-host. The use of multiple hosts may show that the system can support
-much larger clusters for a given election timeout value. However, with
-a fixed election timer value, there is always a cluster size for which
-communication latency will begin to prevent the cluster from being
-able to perform leader elections. A dynamically adjusted election
-timer is something that should be explored especially for a browser
-context where the communication latency may be highly variable over
-time.
-
-\section{Next Steps}
-
-One of the useful outcomes of this project was that it brought to
-light many areas where further study of Raft, WebRTC, and Raft over
-WebRTC should be pursued.
-
+One of the useful outcomes of this project was that it brought to light many areas where further study of Raft, WebRTC, and Raft over WebRTC should be pursued.
 Here are some interesting areas for further exploration:
@@
 \begin{enumerate}
-\item Explore the use of alternate WebRTC transport modes for relaxing
-    ordering constraints and delivery guarantees. Since the Raft
-    protocol is tolerant of out-of-order and dropped messages, it is
-    possible that Raft over WebRTC could be more efficient with one or
-    both of these constraints relaxed.
-\item Test the survivability of the cluster when the signaling server
-    becomes completely unresponsive or very slow.
-\item Explore models for dynamically adjusting Raft timeout values to
-    account for changing network and browser conditions.
-\item Explore models for dealing with the case where the cluster goes
-    from two nodes down to one node. Can usefulness and availability
-    be improved without sacrificing Raft safety constraints?
-\item Test the system in more diverse network environments. Determine
-    the reliability of establishing direct browser-to-browser
-    communication.  Determine if corporate firewalls make this as
-    difficult as some of the documentation and standards texts imply. %TODO/CITE
-\item Perform quantitative and real-world testing to better
-    characterize the performance, availability and scalability
-    characteristics of the system.
-\item Explore models for timing out and automatically removing nodes
-    when they become unresponsive without needing to be notified by
-    the signaling server that the nodes have disconnected.
-\item Consider extensions to the Raft protocol that enable slow
-    leaders to be proactively detected so that a more capable node
-    (network, CPU, etc) can be elected. This would help address that
-    background page/tag throttling of modern browsers.
-\item Test larger scale deployments spanning larger and more diverse
-    network domains. Determine the factors that affect scalability and
-    characterize scaling limits.
-\item Perform wider cross-browser testing including Internet Explorer
-    and Opera.
-\item Implement Raft log compaction and full client linearizability.
-\item Explore models that allow for a combination of strict shared
-    state and high performance bulk data such as media or object
-    blobs. For example, the data with full consensus might contain
-    hashes that refer to bulk data.
-\item Implement a Node.js based Raft over WebRTC application. This
-    would allow non-browser contexts to participate in the Raft
-    cluster together with the browser clients. PeerJS issue \#103 must
-    be resolved for this to be feasible \cite{peerjs:103}.
-\item Explore a design where some nodes of the cluster are non-voting
-    and are not counted as part of transaction commit quorum. However,
-    these nodes would still be included in the log replication
-    algorithm. One model this would enable would be a configuration
-    where a small group of voting nodes serve as the hub for a larger
-    group of non-voting nodes. This may enable higher scaling of the
-    system. It also enables interesting applications such as a logging
-    or backup nodes.
+\item Explore the use of alternate WebRTC transport modes for relaxing ordering constraints and delivery guarantees.
+Since the Raft protocol is tolerant of out-of-order and dropped messages, it is possible that Raft over WebRTC could be more efficient with one or both of these constraints relaxed.
+\item Test the survivability of the cluster when the signaling server becomes completely unresponsive or very slow.
+\item Explore models for dynamically adjusting Raft timeout values to account for changing network and browser conditions.
+\item Test the system in more diverse network environments.
+Determine the reliability of establishing direct browser-to-browser communication.
+Determine if corporate firewalls make this as difficult as some of the documentation and standards texts imply.
+\item Explore models for timing out and automatically removing nodes when they become unresponsive without needing to be notified by the signaling server that the nodes have disconnected.
+\item Consider extensions to the Raft protocol that enable slow leaders to be proactively detected so that a more capable node (network, CPU, etc) can be elected.
+This would help address that background page/tag throttling of modern browsers.
+\item Test larger scale deployments spanning larger and more diverse network domains. Determine the factors that affect scalability and characterize scaling limits.
+\item Explore models that allow for a combination of strict shared state and high performance bulk data such as media or object blobs.
+For example, the data with full consensus might contain hashes that refer to bulk data.
 \end{enumerate}
 
-\section*{Acknowledgments}
-I would like to thank my academic supervisor, David Levine, for his
-excellent input and oversight of this project. I would also like to
-thank Diego Ongaro for his amazing work on the Raft protocol and paper which
-made implementing a Raft implementation a pleasure.  There are too many people
-involved in the creation and standardization of WebRTC but I would like to
-thank them for their tireless work to make browser-to-browser communication
-possible. Last but not least, I would like to thank my family for putting up
-with the long nights and weekends I spent on this project. Thanks!
+\section{Conclusion}
+
+The ability to create distributed applications that minimize dependence on centralized infrastructure is important for many use cases.
+The combination of the Raft distributed consensus algorithm with the WebRTC communication protocol enables a powerful middleware system for building powerful distributed browser-to-browser applications.
+The Raft over WebRTC middleware system is able to easily scale to at least 35 node clusters and that it can quickly recover from multiple node failures.
+Further study may show that a dynamically tuned election timer that accounts for communication latency would allow the cluster size to be scaled well beyond 35 nodes.
+
+\section*{Acknowledgements}
+
+I would like to thank my academic supervisor, David Levine, for his excellent input and oversight of this project.
+I would also like to thank Diego Ongaro for his amazing work on the Raft protocol and paper which made implementing a Raft implementation a pleasure.
+There are too many people involved in the creation and standardization of WebRTC but I would like to thank them for their tireless work to make browser-to-browser communication possible.
+Last but not least, I would like to thank my family for putting up with the long nights and weekends I spent on this project.
+Thanks!
 
@@
 
-%% % Bibliography with multiplie sectiosn and references not actually
-%% % cited in the paper either
-
-%% \addbibresource{references/raft.bib}
-%% \addbibresource{references/w3c.bib}
-%% \addbibresource{references/ietf.bib}
-%% \addbibresource{references/other.bib}
-%%
-%% \printbibliography[title={General references},notkeyword=rtc-general,notkeyword=rtc-transport,notkeyword=rtc-framing]
-%% 
-%% \nocite{draft-rtcweb-overview}
-%% \nocite{draft-rtcweb-security-arch}
-%% \nocite{draft-rtcweb-security}
-%% \printbibliography[title={WebRTC general references},keyword=rtc-general]
-%% 
-%% \nocite{draft-rtcweb-transports}
-%% \nocite{draft-rtcweb-qos}
-%% \nocite{RFC4960:sctp}
-%% \nocite{RFC3758:sctp-reliability}
-%% \nocite{draft-sctp-dtls-encaps}
-%% \nocite{draft-mmusic-sctp-sdp}
-%% \nocite{draft-sctp-ndata}
-%% \nocite{draft-rtcweb-rtp-usage}
-%% \nocite{draft-rtcweb-data-channel}
-%% \nocite{draft-rtcweb-data-protocol}
-%% \nocite{draft-rtcweb-alpn}
-%% \nocite{RFC5245:ice}
-%% \nocite{RFC5128:p2p-nat}
-%% \nocite{RFC5389:stun}
-%% \nocite{RFC3489:stun-udp}
-%% \nocite{RFC5766:turn}
-%% \nocite{RFC6156:turn-ipv6}
-%% \nocite{RFC6544:tcp-ice}
-%% \nocite{RFC4571:rtp-rtcp}
-%% \nocite{RFC5764:dtls-srtp}
-%% \printbibliography[title={WebRTC data transport},keyword=rtc-transport]
-%% 
-%% \nocite{RFC3550:rtp}
-%% \nocite{RFC3711:srtp}
-%% %\nocite{draft-rtcweb-rtp-usage}
-%% %\nocite{draft-rtcweb-data-channel}
-%% %\nocite{draft-rtcweb-data-protocol}
-%% \printbibliography[title={WebRTC framing and security},keyword=rtc-framing]
-%% 
-%% \nocite{RFC2327:sdp}
-%% \nocite{RFC3264:sdp-offer-answer}
-%% \nocite{draft-rtcweb-jsep}
-%% %\nocite{RFC5245:ice}
-%% \nocite{draft-mmusic-trickle-ice}
-%% \nocite{RFC5763:srtp-using-dtls}
-%% \printbibliography[title={WebRTC connection management},keyword=rtc-connection]
-
-%% % WebRTC standards that aren't relevant
-
-%% %\begin{enumerate}
-%% %\item Data formats (audio)
-%% %    \begin{enumerate}
-%% %    \item \href{https://tools.ietf.org/html/draft-cbran-rtcweb-codec-02}{WebRTC: WebRTC Codec and Media Processing Requirements}
-%% %    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-audio-07}{WebRTC: Audio Codec and Processing Requirements}
-%% %    \item \href{https://tools.ietf.org/html/rfc6716}{RFC6716: Opus Audio Codec}
-%% %    \item \href{https://tools.ietf.org/html/draft-ietf-payload-rtp-opus-11}{RTP Payload Format for Opus Speech and Audio Codec}
-%% %    \item \href{https://tools.ietf.org/html/rfc3551}{RFC3551: RTP Profile for Audio and Video Conference}
-%% %    \item \href{https://tools.ietf.org/html/rfc4733}{RFC4733: RTP Payload for DTMF Digits, Telephony Tones, and Telephony Signals}
-%% %    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-audio-codecs-for-interop-01}{WebRTC: Additional WebRTC audio codecs for interoperability with legacy networks}
-%% %    \item \href{https://tools.ietf.org/html/draft-ietf-rtcweb-video-05}{WebRTC: Video Processing and Codec Requirements}
-%% %    \item \href{https://tools.ietf.org/html/rfc6386}{RFC6386: VP8 Data Format and Decoding Guide}
-%% %    \item \href{http://www.itu.int/rec/T-REC-H.264}{ITU H.264: Advanced video coding for generic audiovisual services}
-%% %    \item \href{https://tools.ietf.org/html/draft-ietf-payload-vp8-15}{RTP Payload Format for VP8 Video}
-%% %    \item \href{https://tools.ietf.org/html/rfc6236}{RFC6236: Negotiation of Generic Image Attributes in the Session Description Protocol (SDP)}
-%% %    \end{enumerate}
-%% %\end{enumerate}
-
 \end{document}

### 2015-12-07 research_class/workshop.tex
workshop.tex: spell check.

@@
 
-\subsection{Signalling Server}
+\subsection{Signaling Server}
 
@@
 The Raft over WebRTC chat application is a single web page (\texttt{chat.html}).
-When a new Raft cluster is created, the browser connects to the signalling server's new channel endpoint (Figure~\ref{fig:raft_rtc_sequence}~[1]).
+When a new Raft cluster is created, the browser connects to the signaling server's new channel endpoint (Figure~\ref{fig:raft_rtc_sequence}~[1]).
 This endpoint causes the server to create a new channel for PeerJS signaling (Figure~\ref{fig:raft_rtc_sequence}~[2]).
@@
 If the count value does not match the counter then the the operation is not applied.
-In either case, a list is returned in the \texttt{clientRequestResponse} that contains two values: a boolean indicating whether the \emph{seqAppend} operation succeeded and a number indicating the current value of the counter.
+In either case, a list is returned in the \texttt{clientRequestResponse} that contains two values: a Boolean indicating whether the \emph{seqAppend} operation succeeded and a number indicating the current value of the counter.
 \item The first time the \emph{seqAppend} operation is used on a non-existent key, the structure described above is created with the counter set to 0 and the list structure is created with the single value that was passed in the \emph{seqAppend} operation.\
@@
 
-\section*{Acknowledgements}
+\section*{Acknowledgments}
 

### 2015-12-07 research_class/workshop.tex
Format title and cleanup future work items.

@@
 
-\title{Peer-to-peer Browser Middleware for Communication and Consensus in an Unreliable Infrastructure}
+\title{Peer-to-peer Browser Middleware\\
+       for Communication and Consensus in an Unreliable Infrastructure}
 
@@
 \begin{enumerate}
-\item Explore the use of alternate WebRTC transport modes for relaxing ordering constraints and delivery guarantees.
+\item Explore models for dynamically adjusting Raft timeout values to account for changing network and browser conditions.
+\item Explore alternate WebRTC transport modes for relaxing ordering constraints and delivery guarantees.
 Since the Raft protocol is tolerant of out-of-order and dropped messages, it is possible that Raft over WebRTC could be more efficient with one or both of these constraints relaxed.
 \item Test the survivability of the cluster when the signaling server becomes completely unresponsive or very slow.
-\item Explore models for dynamically adjusting Raft timeout values to account for changing network and browser conditions.
-\item Test the system in more diverse network environments.
-Determine the reliability of establishing direct browser-to-browser communication.
-Determine if corporate firewalls make this as difficult as some of the documentation and standards texts imply.
+\item Test larger scale deployments spanning larger and more diverse network domains. Characterize the factors that affect and limit scalability.
 \item Explore models for timing out and automatically removing nodes when they become unresponsive without needing to be notified by the signaling server that the nodes have disconnected.
@@
 This would help address that background page/tag throttling of modern browsers.
-\item Test larger scale deployments spanning larger and more diverse network domains. Determine the factors that affect scalability and characterize scaling limits.
 \item Explore models that allow for a combination of strict shared state and high performance bulk data such as media or object blobs.
-For example, the data with full consensus might contain hashes that refer to bulk data.
 \end{enumerate}

### 2015-12-07 test/scripts/all_tests.sh
Update all_tests.sh and .gitignore
- increase all_tests.sh cluster size to 35 and parameterize outdir
- add *.bak to .gitignore

@@ -1,13 +1,17 @@
 #!/bin/bash
 
+outdir=test/data_et1000
+
+mkdir -p ${outdir}
+
 # chat propagate 1 message:
-time for x in $(seq 1 7); do for nodes in $(seq 1 2 31); do time node test/wait_chat_propagate.js 192.168.2.3:8001 ${nodes} 1 | tee test/data_et1000/wcp-1-${nodes}-${x}; sleep 3; test/cleanup.sh; done; done
+time for x in $(seq 1 7); do for nodes in $(seq 1 2 35); do time node test/wait_chat_propagate.js 192.168.2.3:8001 ${nodes} 1 | tee ${outdir}/wcp-1-${nodes}-${x}; sleep 3; test/cleanup.sh; done; done
 
 # chat propagate 100 messages:
-time for x in $(seq 1 7); do for nodes in $(seq 1 2 31); do time node test/wait_chat_propagate.js 192.168.2.3:8001 ${nodes} 100 | tee test/data_et1000/wcp-100-${nodes}-${x}; sleep 3; test/cleanup.sh; done; done
+time for x in $(seq 1 7); do for nodes in $(seq 1 2 35); do time node test/wait_chat_propagate.js 192.168.2.3:8001 ${nodes} 100 | tee ${outdir}/wcp-100-${nodes}-${x}; sleep 3; test/cleanup.sh; done; done
 
 # kill 1 node:
-time for x in $(seq 1 7); do for nodes in $(seq 1 2 31); do time node test/wait_kill_nodes.js 192.168.2.3:8001 ${nodes} 1 | tee test/data_et1000/wkn-1-${nodes}-${x}; sleep 3; test/cleanup.sh; done; done
+time for x in $(seq 1 7); do for nodes in $(seq 1 2 35); do time node test/wait_kill_nodes.js 192.168.2.3:8001 ${nodes} 1 | tee ${outdir}/wkn-1-${nodes}-${x}; sleep 3; test/cleanup.sh; done; done
 
-# kill half nodes (162min):
-time for x in $(seq 1 7); do for nodes in $(seq 1 2 31); do time node test/wait_kill_nodes.js 192.168.2.3:8001 ${nodes} | tee test/data_et1000/wkn-half-${nodes}-${x}; sleep 3; test/cleanup.sh; done; done
+# kill half nodes:
[…]

### 2015-12-10 research_class/workshop.tex
workshop.tex: version to Percom 2016, CASPer

@@
 
-\usepackage{hyperref}
+%\usepackage{hyperref}
 
@@
 
-The above capabilities are demonstrated using a distributed chat application implemented with the Raft over WebRTC middleware.
+These capabilities are demonstrated using a distributed chat application implemented with the Raft over WebRTC middleware.
 This chat application uses direct browser-to-browser communication over WebRTC and maintains a consistent view of the chat history (exact same messages and ordering is seen by all nodes) and is fault tolerant (almost half the cluster nodes can fail simultaneously and the application will remain available without dataloss for the remaining clients).
@@
 In order to keep the replicated transaction log from growing indefinitely, Raft nodes should periodically compact their logs.
-The simplest and most generic solution is often to provide some way of snapshotting the entire state machine (including the term and log index it represents) to disk at which point the all previously applied log entries can be discarded.
+The simplest and most generic solution is often to provide some way of snapshotting the entire state machine (including the term and log index it represents) to disk at which point all previously applied log entries can be discarded.
 Clients of the Raft cluster are able to interact with the distributed state machine by sending RPCs to the leader in order to add command entries to the transaction log.
@@
 Raft.js is designed to run in either a browser environment or within Node.js (server-side JavaScript).
-Raft.js implements the Raft algorithm as described in \cite[Consensus:~Diego]{raft_thesis:ongaro14}.
+Raft.js implements the Raft algorithm as described in Consensus:~Diego \cite{raft_thesis:ongaro14}.
 The core Raft algorithm is defined in the \texttt{RaftServerBase} abstract class (in \texttt{base.js}) which omits the implementation of methods with side-effects such as RPC communication, durable storage, scheduling/timeouts, or state machine command/management.
@@
 The value of the pair is another associative map that contains a counter and list structure where the actual chat lines are accumulated.
-\emph{seqAppend} operations are sent with both a count and then new value to append.
+\emph{seqAppend} operations are sent with both a count and the new value to append.
 \item When a \emph{seqAppend} operation is applied to the state machine and the count value in the command matches the counter in the state machine's value, then the new line in the request is added to the value list and the counter is incremented.
@@
 In either case, a list is returned in the \texttt{clientRequestResponse} that contains two values: a Boolean indicating whether the \emph{seqAppend} operation succeeded and a number indicating the current value of the counter.
-\item The first time the \emph{seqAppend} operation is used on a non-existent key, the structure described above is created with the counter set to 0 and the list structure is created with the single value that was passed in the \emph{seqAppend} operation.\
+\item The first time the \emph{seqAppend} operation is used on a non-existent key, the structure previously described is created with the counter set to 0 and the list structure is created with the single value that was passed in the \emph{seqAppend} operation.\
 \item Clients that call the \emph{seqAppend} operation should check the status value in the result list.
@@
 The container is built using \emph{test/Dockerfile} which is in turn derived from the ``fentas/slimerjs" docker image \cite{fentas:slimerjs}.
-This custom Dockerfile was created because the upstream version of the Dockerfile used an older version of XULRunner that did not support more than 7 WebRTC clients communicating together simultaneously.
-This issue was discovered during initial testing.
 
@@
         scatter/classes={%
+            start_et1000={mark=x,draw=blue},
+            start_et1000_average={mark=,draw=red},
+            start_et500={mark=x,draw=cyan},
+            %start_et500_fail={mark=x,draw=red},
+            start_et500_average={mark=,draw=red},
             start_et250={mark=x,draw=magenta},
-            start_et250_average={mark=,draw=red},
-            %start_et250_fail={mark=x,/raw=red},
-            start_et500={mark=x,draw=cyan},
-            start_et500_average={mark=,draw=red},
-            %start_et500_fail={mark=x,draw=red},
-            start_et1000={mark=x,draw=blue},
-            start_et1000_average={mark=,draw=red}}]
+            %start_et250_fail={mark=x,/raw=red}
+            start_et250_average={mark=,draw=red}}]
+    \addplot[scatter,only marks,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/start_et1000.dat};
+    \addplot[blue,scatter,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/start_et1000_average.dat};
+
+    \addplot[scatter,only marks,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/start_et500.dat};
+    %\addplot[scatter,%
+    %    scatter src=explicit symbolic]%
+    %table[meta=label] {data/start_et500_fail.dat};
+    \addplot[cyan,scatter,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/start_et500_average.dat};
+
     \addplot[scatter,only marks,%
@@
 
-    \addplot[scatter,only marks,%
-        scatter src=explicit symbolic]%
-    table[meta=label] {data/start_et500.dat};
-    \addplot[cyan,scatter,%
-        scatter src=explicit symbolic]%
-    table[meta=label] {data/start_et500_average.dat};
-    %\addplot[scatter,%
-    %    scatter src=explicit symbolic]%
-    %table[meta=label] {data/start_et500_fail.dat};
-
-    \addplot[scatter,only marks,%
-        scatter src=explicit symbolic]%
-    table[meta=label] {data/start_et1000.dat};
-    \addplot[blue,scatter,%
-        scatter src=explicit symbolic]%
-    table[meta=label] {data/start_et1000_average.dat};
-
-    \legend{250ms,,500ms,,1000ms,}
+    \legend{1000ms,,500ms,,250ms,}
     \end{axis}
@@
         scatter/classes={%
+            kill_1_et1000={mark=x,draw=blue},
+            kill_1_et1000_average={mark=,draw=red},
+            kill_1_et500={mark=x,draw=cyan},
+            %kill_1_et500_fail={mark=x,draw=red},
+            kill_1_et500_average={mark=,draw=red},
             kill_1_et250={mark=x,draw=magenta},
-            kill_1_et250_average={mark=,draw=red},
             %kill_1_et250_fail={mark=x,/raw=red},
-            kill_1_et500={mark=x,draw=cyan},
-            kill_1_et500_average={mark=,draw=red},
-            %kill_1_et500_fail={mark=x,draw=red},
-            kill_1_et1000={mark=x,draw=blue},
-            kill_1_et1000_average={mark=,draw=red}}]
+            kill_1_et250_average={mark=,draw=red}}]
+    \addplot[scatter,only marks,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/kill_1_et1000.dat};
+    \addplot[blue,scatter,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/kill_1_et1000_average.dat};
+
+    \addplot[scatter,only marks,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/kill_1_et500.dat};
+    %\addplot[scatter,%
+    %    scatter src=explicit symbolic]%
+    %table[meta=label] {data/kill_1_et500_fail.dat};
+    \addplot[cyan,scatter,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/kill_1_et500_average.dat};
+
     \addplot[scatter,only marks,%
@@
 
-    \addplot[scatter,only marks,%
-        scatter src=explicit symbolic]%
-    table[meta=label] {data/kill_1_et500.dat};
-    \addplot[cyan,scatter,%
-        scatter src=explicit symbolic]%
-    table[meta=label] {data/kill_1_et500_average.dat};
-    %\addplot[scatter,%
-    %    scatter src=explicit symbolic]%
-    %table[meta=label] {data/kill_1_et500_fail.dat};
-
-    \addplot[scatter,only marks,%
-        scatter src=explicit symbolic]%
-    table[meta=label] {data/kill_1_et1000.dat};
-    \addplot[blue,scatter,%
-        scatter src=explicit symbolic]%
-    table[meta=label] {data/kill_1_et1000_average.dat};
-
-    \legend{250ms,,500ms,,1000ms,}
+    \legend{1000ms,,500ms,,250ms,}
 
@@
         scatter src=explicit symbolic]%
+    table[meta=label] {data/kill_half_et1000.dat};
+    \addplot[blue,scatter,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/kill_half_et1000_average.dat};
+
+    \addplot[scatter,only marks,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/kill_half_et500.dat};
+    %\addplot[scatter,%
+    %    scatter src=explicit symbolic]%
+    %table[meta=label] {data/kill_half_et500_fail.dat};
+    \addplot[cyan,scatter,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/kill_half_et500_average.dat};
+
+    \addplot[scatter,only marks,%
+        scatter src=explicit symbolic]%
     table[meta=label] {data/kill_half_et250.dat};
@@
 
-    \addplot[scatter,only marks,%
-        scatter src=explicit symbolic]%
-    table[meta=label] {data/kill_half_et500.dat};
-    \addplot[cyan,scatter,%
-        scatter src=explicit symbolic]%
-    table[meta=label] {data/kill_half_et500_average.dat};
-    %\addplot[scatter,%
-    %    scatter src=explicit symbolic]%
-    %table[meta=label] {data/kill_half_et500_fail.dat};
-
-    \addplot[scatter,only marks,%
-        scatter src=explicit symbolic]%
-    table[meta=label] {data/kill_half_et1000.dat};
-    \addplot[blue,scatter,%
-        scatter src=explicit symbolic]%
-    table[meta=label] {data/kill_half_et1000_average.dat};
-
-    \legend{250ms,,500ms,,1000ms,}
+    \legend{1000ms,,500ms,,250ms,}
 
@@
 In Figure~\ref{fig:propagate1} the time to propagate a single message to a single node of the cluster is approximately 0.1 seconds.
-However, this really just represents the tests scanning/checking granularity.
+This simply represents the tests scanning/checking granularity.
 The time to finish propagating and committing a single message shows a slight linear slope upwards from about 0.3 seconds for a three node cluster to about 0.5 seconds for a 41 node cluster.
@@
 This is because after a new leader is elected, there is only a single configuration entry to propagate to the rest of the cluster (to remove the old failed leader).
-One interesting thing to note is that with a 1000ms timer, the 3 node cluster is actually the most inefficient size because the average time until a leader election starts is inversely proportional to cluster size so the delay until an election happens is a significant contributor to the three node cluster with 1000ms timer.
-With a larger cluster or shorter election timer, the contribution of election delay is not noticeable.
 The recovery from almost half the nodes failing is not flat because the amount of configuration information that needs to be propagated is proportional to the size of the cluster.
@@
 \item Consider extensions to the Raft protocol that enable slow leaders to be proactively detected so that a more capable node (network, CPU, etc) can be elected.
-This would help address that background page/tag throttling of modern browsers.
+This would help address background page/tag throttling in modern browsers.
 \item Explore models that allow for a combination of strict shared state and high performance bulk data such as media or object blobs.
@@
 
-The ability to create distributed applications that minimize dependence on centralized infrastructure is important for many use cases.
-The combination of the Raft distributed consensus algorithm with the WebRTC communication protocol enables a powerful middleware system for building powerful distributed browser-to-browser applications.
-The Raft over WebRTC middleware system is able to easily scale to at least 35 node clusters and that it can quickly recover from multiple node failures.
-Further study may show that a dynamically tuned election timer that accounts for communication latency would allow the cluster size to be scaled well beyond 35 nodes.
+The wide reach of the browser platform (exceeding most if not all other computing platforms) and the recent advances in browser programming and communication interfaces make the browser a very interesting platform for a wide variety of applications.
+The combination of the Raft distributed consensus algorithm with the WebRTC communication protocol enables a middleware system for building powerful distributed browser-to-browser applications.
+
+An important use case for the Raft over WebRTC middleware system is the ability to create distributed applications that minimize dependence on centralized infrastructure and are able to tolerate unreliable communication and frequent changes in cluster membership (adds, removes, failures).
+The demonstration chat application was able to easily scale to clusters containing several dozen nodes and quickly recover from multiple node failures.
+The current limitation in cluster size is due to communication latency relative to the election timer setting.
+Further study is expected to show that a dynamically adjusted election timer that accounts for communication latency and cluster size would allow the cluster size to be scaled significantly larger (at the cost of slower recovery from a leader failure).
+
+There are other types of applications that can benefit from avoiding central communication and central data storage without sacrificing consistent shared state.
+These include applications where privacy or security requirements prevent messages from traversing central servers or where remote communication latency or bandwidth costs make communication with remote services unattractive or expensive.
+The Raft over WebRTC middleware system enables these types of applications to be developed for the browser environment.
 
@@
 
-I would like to thank my academic supervisor, David Levine, for his excellent input and oversight of this project.
-I would also like to thank Diego Ongaro for his amazing work on the Raft protocol and paper which made implementing a Raft implementation a pleasure.
-There are too many people involved in the creation and standardization of WebRTC but I would like to thank them for their tireless work to make browser-to-browser communication possible.
+I would like to thank Diego Ongaro for his amazing work on the Raft protocol which made implementing a Raft implementation a pleasure.
+I would also like to thank the many people involved in the creation and standardization of WebRTC and for their tireless work to make browser-to-browser communication possible.
 Last but not least, I would like to thank my family for putting up with the long nights and weekends I spent on this project.

### 2015-12-11 research_class/paper.tex
paper.tex: trimed to 10 pages without OPTIONALS.

@@
 \cite[Section~4]{raft_thesis:ongaro14}, log compaction
-\cite[Section~5]{raft_thesis:ongaro14}, and client
-interaction \cite[Section~6]{raft_thesis:ongaro14}. The Raft protocol
-is based on the concept of a distributed transaction log. Entries in
-the log represent a sequence of commands that will be applied to an
-internal state machine. If all members of the Raft cluster have the
-same log entries, then their state machine on each node will have
-exactly the same state. Each node of a Raft cluster can be in one of
-three states: follower, candidate, or leader. The responsibility of
-the leader is to accept new transaction log entries and then
-replicate these entries in the same order to all other members of the
-cluster. The leader is the only member of the cluster that may make
-changes to the transaction log. In order to maintain leadership, the
-leader sends heartbeat RPCs to all the followers in the cluster.
-Followers that do not receive a heartbeat RPC within a certain time
-period become candidates and attempt to be elected by the other nodes
-as the new leader of the cluster \cite[p.~12]{raft_thesis:ongaro14}.
+\cite[Section~5]{raft_thesis:ongaro14}, and client interaction
+\cite[Section~6]{raft_thesis:ongaro14}. The Raft protocol is based on
+the concept of a distributed transaction log. Entries in the log
+represent a sequence of commands that will be applied to an internal
+state machine. If all members of the Raft cluster have the same log
+entries, then their state machine on each node will have exactly the
+same state. Each node of a Raft cluster can be in one of three states:
+follower, candidate, or leader. The responsibility of the leader is to
+accept new transaction log entries and then replicate these entries in
+the same order to all other members of the cluster. The leader is the
+only member of the cluster that may make changes to the transaction
+log. In order to maintain leadership, the leader sends heartbeat RPCs
+to all the followers in the cluster.  Followers that do not receive
+a heartbeat RPC within a certain time period become candidates and
+attempt to be elected by the other nodes as the new leader of the
+cluster \cite[p.~12]{raft_thesis:ongaro14}.
 
@@
 simply a single shared counter, then any log entries before the most
-recently committed entry may be safely discarded. The simplest and
-most generic solution is often to provide some way of snapshotting the
-entire state machine (including the term and log index it represents)
-to disk at which point the all previously applied log entries can be
-discarded. Snapshotting the state machine also means that the
-implementation must provide a way for the leader to serialize and send
-the current state machine to other nodes if their transaction log is
-too old (e.g. when a new node is added)
+recently committed entry may be safely discarded.
+The simplest and most generic solution is often to provide some way of
+snapshotting the entire state machine (including the term and log
+index it represents) to disk at which point all previously applied log
+entries can be discarded. Snapshotting the state machine also means
+that the implementation must provide a way for the leader to serialize
+and send the current state machine to other nodes if their transaction
+log is too old (e.g. when a new node is added)
 \cite[Section~5]{raft_thesis:ongaro14}.
@@
 
-The Raft protocol can also provide strict linearizable semantics for
+The Raft protocol also provides strict linearizable semantics for
 client commands/requests (reads and writes) for applications that
-require it. In order to accomplish this, each client is given a unique
-ID, and the client assigns a unique serial number to each command.
-This prevents a single client command from being applied twice (leader
-crash, network duplication) or lost (packet loss)
+require it.
+In order to accomplish this, each client is given a unique ID, and the
+client assigns a unique serial number to each command.  This prevents
+a single client command from being applied twice (leader crash,
+network duplication) or lost (packet loss)
 \cite[Section~6.3]{raft_thesis:ongaro14}.
@@
 
-\begin{figure*}[!t]
-\centerline{\includegraphics[width=0.8\textwidth]{imgs/raft_rtc_architecture}}
+\subsection{WebRTC}
+
+WebRTC is a collection of browser APIs \cite{webrtc:feb2015} and
+network protocols \cite{draft-rtcweb-overview} that are currently
+being standardized by the W3C and IETF respectively. Collectively
+these APIs and protocols enable real-time peer-to-peer video, audio
+and data communication between browsers.
+
+\subsubsection{WebRTC Discovery and Signaling}
+
+One of the challenges with browser-to-browser communication (and
+direct peer-to-peer communication in general) is that the web browser
+environment or user agent (UA) is usually running behind a firewall in
+a private internet subnet. This complicates direct communication in
+several ways: the public Internet address of the user agent is unknown
+and dynamic, inbound connections are denied by default, and outbound
+connections are network address translated (NATed)
+\cite[Section~3.4]{draft-rtcweb-transports}.
+
+To establish a direct communication channel between browsers, WebRTC
+uses a mechanism called Interactive Connectivity Establishment (ICE)
+that is similar to the Session Initiation Protocol (SIP) commonly used
+for Voice-over-IP (VoIP) communication \cite{RFC5245:ice}. This
+involves a third party server known as a signaling server. Each user
+agent (browser) connects to this server at a well known address in
+order to establish a direct connection to other user agents. The
+signaling server is also used to communicate session control messages
+(open, close, error) and to negotiate media capabilities. %TODO/CITE
+The signaling server may optionally provide other services such as
+higher level session management, HTTP serving of the web application,
+address book service, client presence/availability, etc. %TODO/CITE
+
+\subsubsection{WebRTC APIs}
+
+The WebRTC browser APIs enable JavaScript web applications to
+establish browser-to-browser network connectivity in order to perform
+real-time video, audio and data communication. These APIs are
+currently in the process of being standardized by the \emph{Web
+Real-Time Communications Working Group} of the World Wide Web
+Consortium (W3C) \cite{webrtc:wg}.
+
+There are three primary WebRTC interfaces:
+
+\begin{enumerate}
+\item \emph{RTCPeerConnection} \cite{webrtc:rtcpeerconnection:feb2015}:
+    The RTCPeerConnection API represents a direct connection between
+    two browsers. Each RTCPeerConnection object has an associated
+    Interactive Connectivity Establishment (ICE) agent that is used to
+    establish network connectivity through firewalls and NAT services.
+
+\item \emph{MediaStream~(aka~getUserMedia)} \cite{webrtc:media-stream:feb2015}:
+    The MediaStream API represents a stream of audio or video data.
+    MediaStream data can easily be sent and received over
+    RTCPeerConnections. The MediaStream API definition provides
+    a mechanism that allows JavaScript to enumerate the native
+    MediaStreams that are provided by the browser such as a local web
+    camera or a live video stream of the desktop (for desktop
+    sharing).  Although support for MediaStreams was the original
+    impetus for the creation of the WebRTC APIs and protocols, the
+    Raft over WebRTC implementation does not make use of the
+    MediaStream API.
+
+\item
+    \emph{RTCDataChannel} \cite{webrtc:datachannel:feb2015}:
+    The RTCDataChannel API provides a mechanism for sending and
+    receiving generic data between User Agents (browsers). A single
+    RTCPeerConnection can contain multiple RTCDataChannels. The
+    RTCDataChannel API is modeled after the WebSocket API, however,
+    unlike the WebSocket API which is reliable and guaranteed order
+    transport (like TCP), each RTCDataChannel can be configured with
+    different reliability and ordering settings. For example, by
+    relaxing the reliability and ordering constraints on
+    a RTCDataChannel, the average latency and jitter per message can
+    be decreased even though some messages may be dropped or arrive
+    out of order. The Raft protocol is specifically designed to
+    support dropped or out-of-order messages so the transport options
+    available with RTCDataChannel can be used to gain extra
+    efficiency.
+    %TODO/CITE: more of them above
+\end{enumerate}
+
+\subsubsection{WebRTC Protocols}
+
+There is a large suite of protocols that are required for a working
+WebRTC system. The Internet Engineering Task Force (IETF) organization
+is responsible for specifying and/or standardizing (where necessary)
+the protocols that are part of WebRTC \cite{draft-rtcweb-overview}
+\cite{draft-rtcweb-security-arch} \cite{draft-rtcweb-security}. In
+many cases an existing protocol is applicable and is referenced or
+extended as part of the WebRTC suite. %TODO/CITE
+
+The protocols that make up the WebRTC suite are categorized into five
+general functionality groups:
+\ifdefined\OPTIONAL
+
+\begin{enumerate}
+    \item data transport
+        \\
+        \cite{draft-rtcweb-transports}
+        \cite{draft-rtcweb-qos}
+        \cite{RFC4960:sctp}
+        \cite{RFC3758:sctp-reliability}
+        \cite{draft-sctp-dtls-encaps}
+        \cite{draft-mmusic-sctp-sdp}
+        \cite{draft-sctp-ndata}
+        \cite{draft-rtcweb-rtp-usage}
+        \cite{draft-rtcweb-data-channel}
+        \cite{draft-rtcweb-data-protocol}
+        \cite{draft-rtcweb-alpn}
+        \cite{RFC5245:ice}
+        \cite{RFC5128:p2p-nat}
+        \cite{RFC5389:stun}
+        \cite{RFC3489:stun-udp}
+        \cite{RFC5766:turn}
+        \cite{RFC6156:turn-ipv6}
+        \cite{RFC6544:tcp-ice}
+        \cite{RFC4571:rtp-rtcp}
+        \cite{RFC5764:dtls-srtp}
+    \item data framing and security
+        \\
+        \cite{RFC3550:rtp}
+        \cite{RFC3711:srtp}
+        \cite{draft-rtcweb-rtp-usage}
+        \cite{draft-rtcweb-data-channel}
+        \cite{draft-rtcweb-data-protocol}
+    \item connection management
+        \\
+        \cite{RFC2327:sdp}
+        \cite{RFC3264:sdp-offer-answer}
+        \cite{draft-rtcweb-jsep}
+        \cite{RFC5245:ice}
+        \cite{draft-mmusic-trickle-ice}
+        \cite{RFC5763:srtp-using-dtls}
+    \item data formats
+    \item presentation and control
+    \item local system support functions
+\end{enumerate}
+
+The last three functionality groups focus on interactive audio and video
+applications and are less relevant to Raft over WebRTC.
+Some of the important standards for the first three functionality
+groups are cited above.
+\else
+data transport, data framing and security, connection management, data
+formats, presentation and control, local system support functions.
+The last three functionality groups focus on interactive audio and
+video applications and are less relevant to Raft over WebRTC.
+\fi
+
+The signaling transport itself is not defined as part of WebRTC.
+This is up to the application designer. The JavaScript interface used
+for establishing the initial connection (signaling) in the WebRTC 1.0
+standard is somewhat complicated. A competing standard called Object
+Real-time Communication (ORTC) is working to define a simpler and more
+modular JavaScript API for WebRTC signaling \cite{ortc2014}. However,
+at the time of this paper, the ORTC interface is not yet widely
+supported and is still in a state of flux. The use of PeerJS provides an
+abstraction that is likely to transparently support ORTC in the future.
+
+\ifdefined\OPTIONAL
+
+\subsection{PeerJS}
+
+PeerJS is a project that provides a simplified abstraction for using
+WebRTC. The first component of PeerJS is a node.js server library that
+implements a WebRTC signaling server. The second component is
+a JavaScript library that interacts with that signaling server and
+also provides a simpler interface for using the WebRTC APIs.
+
+\subsubsection{PeerJS Server}
+
+The PeerJS project provides a simple WebRTC signaling server called
+PeerServer. This server uses HTTP and WebSockets transport protocols
+to perform WebRTC signaling on behalf of browser clients. The
+PeerServer server is also extensible so that more advanced services
+can be built with it. The PeerJS project includes an extension to
+PeerServer called ExpressPeerServer that combines a PeerJS signaling
+server with the popular Node.js ExpressJS web application framework.
+% (TODO/CITE)
+
+\subsubsection{PeerJS Client Library}
+
+The WebRTC API standards are not yet
+finalized and so different browsers and browser releases may support
+different versions of the WebRTC draft APIs. %TODO/CITE
+The PeerJS client library abstracts over these differences and
+provides a common interface so that the application developer does not
+have to deal directly with browser differences.
+
+The PeerJS client library also provides a default signaling mechanism
+that is designed to operate with a PeerJS signaling server. This is
+particularly helpful because much of the WebRTC signaling transport
+and protocol is not defined as part of the WebRTC standardization
+effort. %TODO/CITE
+The PeerJS organization provides a cloud-hosted version of a PeerJS
+compatible signaling server, however application developers can also
+run their own signaling service using PeerJS PeerServer.
+
+\fi  % \ifdefined\OPTIONAL
+
+\section{Design and Implementation}
+
+\begin{figure}[!t]
+  \centerline{\includegraphics[width=0.45\textwidth]{imgs/raft_rtc_architecture}}
   \caption{Raft over RTC Architecture}
     \label{fig:raft_rtc_architecture}
-\end{figure*}
-
-\subsection{Raft.js}
+\end{figure}
+
+\subsection{Raft.js Core Library}
 
@@
 execution context. For example, the online Raft visualization at
-\href{http://kanaka.github.io/raft.js/}{http://kanaka.github.io/raft.js/}
+http://kanaka.github.io/raft.js/
+%% \href{http://kanaka.github.io/raft.js/}{http://kanaka.github.io/raft.js/}
 \cite{raft.js:visualization} uses the \texttt{RaftServerLocal} class
@@
 verifying overall system correctness and does not significantly impact
-short-running test configurations. Full client linearizability is
-an important feature for simplifying certain types of distributed
-applications, however the implementation of linearizability requires
-some significant additions to the Raft implementation (e.g. assigning
-client IDs and sequence IDs and tracking timed-out clients). For the
-chat application, simply adding a single atomic check-and-set state
-machine command is sufficient to achieve a linearized chat history.
-This does impose a slight additional cost on the client because it
-must retry the atomic operations until they succeed. The author
-intends to implement both log compaction and full client
-linearizability in the future, however these do not have a significant
-impact on the ability to test Raft combined with WebRTC.
+short-running test configurations.
+
+Full client linearizability is an important feature for simplifying
+certain types of distributed applications, however the implementation
+of linearizability requires some significant additions to the Raft
+implementation (e.g. assigning client IDs and sequence IDs and
+tracking timed-out clients). For the chat application, simply adding
+a single atomic check-and-set state machine command is sufficient to
+achieve a linearized chat history.  This does impose a slight
+additional cost on the client because it must retry the atomic
+operations until they succeed. The author intends to implement both
+log compaction and full client linearizability in the future, however
+these do not have a significant impact on the ability to test Raft
+combined with WebRTC.
 
@@
 \end{enumerate}
-
-\subsection{WebRTC}
-
-WebRTC is a collection of browser APIs \cite{webrtc:feb2015} and
-network protocols \cite{draft-rtcweb-overview} that are currently
-being standardized by the W3C and IETF respectively. Collectively these
-APIs and protocols enable real-time peer-to-peer video, audio and data
-communication between browsers.
-
-\subsubsection{WebRTC Discovery and Signaling}
-
-One of the challenges with browser-to-browser communication (and
-direct peer-to-peer communication in general) is that the web browser
-environment or user agent (UA) is usually running behind a firewall in
-a private internet subnet. This complicates direct communication in
-several ways: the public Internet address of the user agent is unknown
-and dynamic, inbound connections are denied by default, and outbound
-connections are network address translated (NATed)
-\cite[Section~3.4]{draft-rtcweb-transports}.
-
-To establish a direct communication channel between browsers, WebRTC
-uses a mechanism called Interactive Connectivity Establishment (ICE)
-that is similar to the Session Initiation Protocol (SIP) commonly used
-for Voice-over-IP (VoIP) communication \cite{RFC5245:ice}. This
-involves a third party server known as a signaling server. Each user
-agent (browser) connects to this server at a well known address in
-order to establish a direct connection to other user agents. The
-signaling server is also used to communicate session control messages
-(open, close, error) and to negotiate media capabilities. %TODO/CITE
-The signaling server may optionally provide other services such as
-higher level session management, HTTP serving of the web application,
-address book service, client presence/availability, etc. %TODO/CITE
-
-\subsubsection{WebRTC APIs}
-
-The WebRTC browser APIs enable JavaScript web applications
-to establish browser-to-browser network connectivity in order to
-perform real-time video, audio and data communication. These APIs are
-currently in the process of being standardized by the \emph{Web
-Real-Time Communications Working Group} of the World Wide Web
-Consortium (W3C) \cite{webrtc:wg}.
-
-There are three primary WebRTC interfaces:
-
-\begin{enumerate}
-\item \emph{RTCPeerConnection} \cite{webrtc:rtcpeerconnection:feb2015}:
-    The RTCPeerConnection API represents a direct connection between
-    two browsers. Each RTCPeerConnection object has an associated
-    Interactive Connectivity Establishment (ICE) agent that is used to
-    establish network connectivity through firewalls and NAT services.
-
-\item \emph{MediaStream~(aka~getUserMedia)} \cite{webrtc:media-stream:feb2015}:
-    The MediaStream API represents a stream of audio or video data.
-    MediaStream data can easily be sent and received over
-    RTCPeerConnections. The MediaStream API definition provides
-    a mechanism that allows JavaScript to enumerate the native
-    MediaStreams that are provided by the browser such as a local web
-    camera or a live video stream of the desktop (for desktop
-    sharing).  Although support for MediaStreams was the original
-    impetus for the creation of the WebRTC APIs and protocols, the
-    Raft over WebRTC implementation does not make use of the
-    MediaStream API.
-
-\item
-    \emph{RTCDataChannel} \cite{webrtc:datachannel:feb2015}:
-    The RTCDataChannel API provides a mechanism for sending and
-    receiving generic data between User Agents (browsers). A single
-    RTCPeerConnection can contain multiple RTCDataChannels. The
-    RTCDataChannel API is modeled after the WebSocket API, however,
-    unlike the WebSocket API which is reliable and guaranteed order
-    transport (like TCP), each RTCDataChannel can be configured with
-    different reliability and ordering settings. For example, by
-    relaxing the reliability and ordering constraints on
-    a RTCDataChannel, the average latency and jitter per message can
-    be decreased even though some messages may be dropped or arrive
-    out of order. The Raft protocol is specifically designed to
-    support dropped or out-of-order messages so the transport options
-    available with RTCDataChannel can be used to gain extra
-    efficiency.
-    %TODO/CITE: more of them above
-\end{enumerate}
-
-\subsubsection{WebRTC Protocol}
-
-There is a large suite of protocols that are required for a working
-WebRTC system. The Internet Engineering Task Force (IETF) organization
-is responsible for specifying and/or standardizing (where necessary)
-the protocols that are part of WebRTC \cite{draft-rtcweb-overview}
-\cite{draft-rtcweb-security-arch} \cite{draft-rtcweb-security}. In many
-cases an existing protocol is applicable and is referenced or
-extended as part of the WebRTC suite. %TODO/CITE
-
-The protocols that make up the WebRTC suite are categorized into
-five general functionality groups:
-
-\begin{enumerate}
-    \item data transport
-        \ifdefined\OPTIONAL
-        \\
-        \cite{draft-rtcweb-transports}
-        \cite{draft-rtcweb-qos}
-        \cite{RFC4960:sctp}
-        \cite{RFC3758:sctp-reliability}
-        \cite{draft-sctp-dtls-encaps}
-        \cite{draft-mmusic-sctp-sdp}
-        \cite{draft-sctp-ndata}
-        \cite{draft-rtcweb-rtp-usage}
-        \cite{draft-rtcweb-data-channel}
-        \cite{draft-rtcweb-data-protocol}
-        \cite{draft-rtcweb-alpn}
-        \cite{RFC5245:ice}
-        \cite{RFC5128:p2p-nat}
-        \cite{RFC5389:stun}
-        \cite{RFC3489:stun-udp}
-        \cite{RFC5766:turn}
-        \cite{RFC6156:turn-ipv6}
-        \cite{RFC6544:tcp-ice}
-        \cite{RFC4571:rtp-rtcp}
-        \cite{RFC5764:dtls-srtp}
-        \fi
-    \item data framing and security
-        \ifdefined\OPTIONAL
-        \\
-        \cite{RFC3550:rtp}
-        \cite{RFC3711:srtp}
-        \cite{draft-rtcweb-rtp-usage}
-        \cite{draft-rtcweb-data-channel}
-        \cite{draft-rtcweb-data-protocol}
-        \fi
-    \item connection management
-        \ifdefined\OPTIONAL
-        \\
-        \cite{RFC2327:sdp}
-        \cite{RFC3264:sdp-offer-answer}
-        \cite{draft-rtcweb-jsep}
-        \cite{RFC5245:ice}
-        \cite{draft-mmusic-trickle-ice}
-        \cite{RFC5763:srtp-using-dtls}
-        \fi
-    \item data formats
-    \item presentation and control
-    \item local system support functions
-\end{enumerate}
-
-The last three functionality groups focus on interactive audio and video
-applications and are less relevant to Raft over WebRTC. \ifdefined\OPTIONAL
-Some of the important standards for the first three functionality
-groups are cited above. \fi
-
-The signaling transport itself is not defined as part of WebRTC.
-This is up to the application designer. The JavaScript interface used
-for establishing the initial connection (signaling) in the WebRTC 1.0
-standard is somewhat complicated. A competing standard called Object
-Real-time Communication (ORTC) is working to define a simpler and more
-modular JavaScript API for WebRTC signaling \cite{ortc2014}. However,
-at the time of this paper, the ORTC interface is not yet widely
-supported and is still in a state of flux. The use of PeerJS provides an
-abstraction that is likely to transparently support ORTC in the future.
-
-\ifdefined\OPTIONAL
-
-\subsection{PeerJS}
-
-PeerJS is a project that provides a simplified abstraction for using
-WebRTC. The first component of PeerJS is a node.js server library that
-implements a WebRTC signaling server. The second component is
-a JavaScript library that interacts with that signaling server and
-also provides a simpler interface for using the WebRTC APIs.
-
-\subsubsection{PeerJS Server}
-
-The PeerJS project provides a simple WebRTC signaling server called
-PeerServer. This server uses HTTP and WebSockets transport protocols
-to perform WebRTC signaling on behalf of browser clients. The
-PeerServer server is also extensible so that more advanced services
-can be built with it. The PeerJS project includes an extension to
-PeerServer called ExpressPeerServer that combines a PeerJS signaling
-server with the popular Node.js ExpressJS web application framework.
-% (TODO/CITE)
-
-\subsubsection{PeerJS Client Library}
-
-The WebRTC API standards are not yet
-finalized and so different browsers and browser releases may support
-different versions of the WebRTC draft APIs. %TODO/CITE
-The PeerJS client library abstracts over these differences and
-provides a common interface so that the application developer does not
-have to deal directly with browser differences.
-
-The PeerJS client library also provides a default signaling mechanism
-that is designed to operate with a PeerJS signaling server. This is
-particularly helpful because much of the WebRTC signaling transport
-and protocol is not defined as part of the WebRTC standardization
-effort. %TODO/CITE
-The PeerJS organization provides a cloud-hosted version of a PeerJS
-compatible signaling server, however application developers can also
-run their own signaling service using PeerJS PeerServer.
-
-\fi  % \ifdefined\OPTIONAL
-
-%\begin{figure*}[!t]
-\begin{figure}[!t]
-%\centerline{\includegraphics[width=0.8\textwidth]{imgs/raft_rtc_sequence}}
-\centerline{\includegraphics[width=0.45\textwidth]{imgs/raft_rtc_sequence}}
-\caption{Raft over RTC Sequence Diagram}
-  \label{fig:raft_rtc_sequence}
-\end{figure}
-
-\section {Design and Implementation}
 
@@
 `ExpressPeerServer' object in order to provide WebRTC signaling. In
-addition to the signaling function, the server also provides the
+addition to the WebRTC signaling function the server also provides the
 following service endpoints:
@@
 provided in the application page that opens a new browser window with
-a new cluster node (see \ref{fig:browser_1}).
+a new cluster node.  %(see \ref{fig:browser_1}).
 
@@
 
+\subsubsection{Client Statistics}
+
 In addition to the low level Raft message log, the following
 statistics are shown in the application:
+\ifdefined\OPTIONAL
 
@@
 \end{enumerate}
+\else
+the current Raft leadership term, the current Raft node state
+(follower, candidate, or leader), Raft cluster size (number of
+participating nodes), number of entries in the transaction log, and
+number of request vote and append entries RPCs that have been sent and
+received.
+\fi
 
@@
 that adds entries to the log so clients on follower nodes must have
-a way to communicate with the leader. This is implemented as a pair of
+a way to communicate with the leader.
+\ifdefined\OPTIONAL
+This is implemented as a pair of
 functions named \texttt{clientRequest} and
@@
 \end{enumerate}
+\else
+When a \texttt{clientRequest} message is sent to a local Raft node
+that is not the leader, the \texttt{clientRequestResponse} message
+will be returned with a status of ``NOT\_LEADER" and the ID of the
+current leader. The \texttt{clientRequest} message will then be
+re-issued over the WebRTC DataCannel to the leader ID that was
+returned.
+\fi
 
 \subsection{Chat Application}
+
+\begin{figure}[!t]
+  \centerline{\includegraphics[width=0.45\textwidth]{imgs/raft_rtc_sequence}}
+  \caption{Raft over RTC Sequence Diagram}
+    \label{fig:raft_rtc_sequence}
+\end{figure}
 
@@
 chat application provides a large text area containing the chat
-history and a text entry field below it
-(see Figure~\ref{fig:browser_chat_1}). When the user clicks on the
-\emph{Send} button (or presses enter), the current node ID is
-prepended to the beginning of the text line and it is added to
-a pending send queue and the text entry field is blanked. A function
-to flush sends runs periodically (currently every 100ms) and does the
-following:
+history and a text entry field below it. % (see Figure~\ref{fig:browser_chat_1}).
+When the user clicks on the \emph{Send} button (or presses enter), the
+current node ID is prepended to the beginning of the text line and it
+is added to a pending send queue and the text entry field is blanked.
+\ifdefined\OPTIONAL
+A function to flush the send queue runs periodically (currently every
+100ms) and does the following:
 % TODO: 100ms above should be less
@@
 \end{enumerate}
+\else
+A function to flush the send queue runs periodically (currently every
+100ms) to send the chat message at the head of the queue using
+the \emph{seqAppend} operation.
+\fi
 
@@
 
-The chat application also extends the default set of state machine
-operations to add a \emph{seqAppend} operation. This is an atomic
+The chat application extends the default set of state machine
+operations to add the \emph{seqAppend} operation. This is an atomic
 operation that is implemented as follows:
@@
 \begin{enumerate}
-\item the \emph{seqAppend} operation operates on a specially structured
-    key/value pair in the state machine map. The value of the pair is
-    another associative map that contains a counter and list structure
-    where the actual chat lines are accumulated. \emph{seqAppend}
-    operations are sent with both a count and then new value to
-    append.
-\item when a \emph{seqAppend} operation is applied to the state
+\item The \emph{seqAppend} operation operates on a specially
+    structured key/value pair in the state machine map. The value of
+    the pair is another associative map that contains a counter and
+    list structure where the actual chat lines are accumulated.
+    \emph{seqAppend} operations are sent with both a count and the new
+    value to append.
+\item When a \emph{seqAppend} operation is applied to the state
     machine and the count value in the command matches the counter in
     the state machine's value, then the new line in the request is
-    added to the value list and the counter is incremented. If the count
-    value does not match the counter then the the operation is not
-    applied. In either case, a list is returned in the
-    \texttt{clientRequestResponse} that contains two values: a boolean
+    added to the value list and the counter is incremented. If the
+    count value does not match the counter then the the operation is
+    not applied. In either case, a list is returned in the
+    \texttt{clientRequestResponse} that contains two values: a Boolean
     indicating whether the \emph{seqAppend} operation succeeded and
     a number indicating the current value of the counter.
-\item the first time the \emph{seqAppend} operation is used on a non-existent
-    key, the structure described above is created with the counter set
-    to 0 and the list structure is created with the single value that
-    was passed in the \emph{seqAppend} operation.
-\item clients that call the \emph{seqAppend} operation should check the
-    status value in the result list. If the status is false, then the
-    client should retry the operation. Regardless of success of
+\item The first time the \emph{seqAppend} operation is used on
+    a non-existent key, the structure previously described is created
+    with the counter set to 0 and the list structure is created with
+    the single value that was passed in the \emph{seqAppend}
+    operation.
+\item Clients that call the \emph{seqAppend} operation should check
+    the status value in the result list. If the status is false, then
+    the client should retry the operation. Regardless of success of
     failure, then client should store the current value of the counter
@@
 
-\section{Results}
+\section{Experimental Results}
 
@@
 To simplify the execution of the tests and to make the testing
-infrastructure more portable, a docker container environment was used
-to encapsulate and execute the SlimerJS environment. The container is
-built using \emph{test/Dockerfile} which is in turn derived from
-the ``fentas/slimerjs" docker image \cite{fentas:slimerjs}.
+infrastructure more portable, a docker container environment
+encapsulates and executes the SlimerJS environment. The container is
+built using \emph{test/Dockerfile} which is in turn derived from the
+``fentas/slimerjs" docker image \cite{fentas:slimerjs}.
 This custom Dockerfile was created because the upstream version of the
@@
 more than 7 WebRTC clients communicating together simultaneously. This
-issue was discovered during testing.
-
-Writing tests using SlimerJS involves creating JavaScript programs
-that are run within the SlimerJS environment. These scripts can create
-browser instances (essentially headless browser tabs), query the status
-of those pages and execute JavaScript within the context of the pages
-themselves.
-
-Three different tests were created to test the basic Raft over WebRTC
-application and the chat application. Each of the tests takes
-a parameter indicating the number of nodes/pages to launch and test
-again. The test script files are:
+issue was discovered during initial testing.
+
+Writing tests using SlimerJS involves creating JavaScript test
+programs which start a \emph{Twst} server instance and then launch a SlimerJS
+docker container for each node of the cluster. Each container starts
+a XULRunner browser instance of the Raft over WebRTC chat application.
+Also included with the application is the \emph{Twst} client library
+which connects back to the \emph{Twst} service in the test program.
+Once all the chat application nodes have connected to the test
+server/program, then the actual test is started across all the
+application nodes.
+
+Three different tests programs/servers were created to test the basic
+Raft over WebRTC application. Each of the tests takes a parameter
+indicating the number of nodes/pages to launch. The test source files
+are:
 
@@
         the leader will configure them into the cluster. Then test
-        then waits for every node to be recorded (applied) into the log
-        of every other node of the cluster. The test reports the
+        then waits for every node to be recorded (applied) into the
+        log of every other node of the cluster. The test reports the
         number of milliseconds for the cluster to reach this state.
@@
 
+\ifdefined\OPTIONAL
 The tests were all run on a system with the following configuration:
@@
 \end{enumerate}
-
-Three tests were run at 10 different cluster sizes (odd cluster sizes
-from 3 to 21 nodes): \emph{test/test\_up.js} and
-\emph{test/test\_kill\_nodes.js} in both modes (kill only leader and kill just
-under half the nodes).
-Each of these tests was run with three different Raft election
-timeout values: 100ms, 500ms, 1000ms. The
-\emph{test/test\_chat\_propagate.js} test was run at
-11 different cluster sizes (odd cluster sizes from 1 to 21 nodes).
-Each test mode and cluster size was run 7 times for a total of 707
-test runs (3*10*7*3 + 11*7)
+\else
+The tests were all run on a system with 8 Intel Core i7 processors
+running at 3.20GHz and 12 GB of main memory. The Docker image used to
+run the test was built using Ubuntu 12.04.5 LTS with \emph{SlimerJS}
+version 0.9.6, and \emph{XULRunner} version 38.0.5.
+\fi
+
+Five different tests (using different options for the three test
+programs) were run at 17 different cluster sizes (odd cluster sizes
+from 3 to 35 nodes). Each of these tests was run with three different
+Raft election timeout values: 250ms, 500ms, 1000ms. Each of the five
+tests and cluster size combinations was run 7 times.
 
 \subsubsection{Outcome}
-
-Figure~\ref{fig:propagate1} shows the results of running the
-\emph{test/wait\_chat\_propagate.js} test to show the amount of time
-that a stable cluster takes to propagate, commit and apply a single
-chat message to the state machine on every cluster node using the
-atomic \emph{seqAppend} command.  Figure~\ref{fig:propagate1} is
-a scatterplot with each test run as a blue ``x" and a black line
-representing the average at each cluster size.
 
@@
         %height=1.5\textwidth, width=1.4\textwidth,
+        %width=0.45\textwidth,
         xlabel={Nodes}, ylabel={Time (seconds)},
-                xtick=data,
-                ymin=0, ymax=65,
+        xmin=0, xmax=37,
+        ymin=0, ymax=15,
         grid=major,
+        %legend pos=outer north east,
+        legend pos=north west,
         scatter/classes={%
-                    et100={mark=x,draw=blue},
-                    et100_fail={mark=x,draw=red},
-                    et100_average={mark=,draw=red}}]
+            start_et1000={mark=x,draw=blue},
+            start_et1000_average={mark=,draw=red},
+            start_et500={mark=x,draw=cyan},
+            %start_et500_fail={mark=x,draw=red},
+            start_et500_average={mark=,draw=red},
+            start_et250={mark=x,draw=magenta},
+            %start_et250_fail={mark=x,/raw=red}
+            start_et250_average={mark=,draw=red}}]
     \addplot[scatter,only marks,%
         scatter src=explicit symbolic]%
-            table[meta=label] {data/data_cluster_up_100.dat};
-            \addplot[scatter,%
-                scatter src=explicit symbolic]%
-            table[meta=label] {data/data_cluster_up_100_averages.dat};
+    table[meta=label] {data/start_et1000.dat};
+    \addplot[blue,scatter,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/start_et1000_average.dat};
+
+    \addplot[scatter,only marks,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/start_et500.dat};
+    %\addplot[scatter,%
+    %    scatter src=explicit symbolic]%
+    %table[meta=label] {data/start_et500_fail.dat};
+    \addplot[cyan,scatter,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/start_et500_average.dat};
+
+    \addplot[scatter,only marks,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/start_et250.dat};
+    \addplot[magenta,scatter,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/start_et250_average.dat};
+
+    \legend{1000ms,,500ms,,250ms,}
     \end{axis}
     \end{tikzpicture}
-        }
-        \label{fig:A}
-    }
-    %\begin{subfigure}[b]{0.30\textwidth}
-    \subfloat[500ms election timer]{
-        \centering
-        \resizebox{0.3\linewidth}{!}{
+    \caption{Initial Cluster Activation}
+    \label{fig:cluster_up}
+\end{figure}
+
+\begin{figure}[!t]
     \begin{tikzpicture}
@@
         %height=1.5\textwidth, width=1.4\textwidth,
+        %width=0.45\textwidth,
         xlabel={Nodes}, ylabel={Time (seconds)},
-                xtick=data,
-                ymin=0, ymax=65,
+        xmin=0, xmax=37,
+        ymin=0, ymax=4.5,
         grid=major,
+        %legend pos=outer north east,
+        legend pos=north west,
         scatter/classes={%
-                    et500={mark=x,draw=blue},
-                    et500_average={mark=,draw=red}}]
+            kill_1_et1000={mark=x,draw=blue},
+            kill_1_et1000_average={mark=,draw=red},
+            kill_1_et500={mark=x,draw=cyan},
+            %kill_1_et500_fail={mark=x,draw=red},
+            kill_1_et500_average={mark=,draw=red},
+            kill_1_et250={mark=x,draw=magenta},
+            %kill_1_et250_fail={mark=x,/raw=red},
+            kill_1_et250_average={mark=,draw=red}}]
     \addplot[scatter,only marks,%
         scatter src=explicit symbolic]%
-            table[meta=label] {data/data_cluster_up_500.dat};
-            \addplot[scatter,%
-                scatter src=explicit symbolic]%
-            table[meta=label] {data/data_cluster_up_500_averages.dat};
+    table[meta=label] {data/kill_1_et1000.dat};
+    \addplot[blue,scatter,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/kill_1_et1000_average.dat};
+
+    \addplot[scatter,only marks,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/kill_1_et500.dat};
+    %\addplot[scatter,%
+    %    scatter src=explicit symbolic]%
+    %table[meta=label] {data/kill_1_et500_fail.dat};
+    \addplot[cyan,scatter,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/kill_1_et500_average.dat};
+
+    \addplot[scatter,only marks,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/kill_1_et250.dat};
+    \addplot[magenta,scatter,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/kill_1_et250_average.dat};
+
+    \legend{1000ms,,500ms,,250ms,}
+
     \end{axis}
     \end{tikzpicture}
-        }
-        \label{fig:B}
-    }
-    %\begin{subfigure}[b]{0.30\textwidth}
-    \subfloat[1000ms election timer]{
-        \centering
-        \resizebox{0.3\linewidth}{!}{
+    \caption{Recovery from Leader Failure}
+    \label{fig:recovery_leader_failure}
+%
+    \vspace{1cm}
+%
     \begin{tikzpicture}
@@
     table[meta=label] {data/kill_half_et1000.dat};
-            \addplot[scatter,%
+    \addplot[blue,scatter,%
         scatter src=explicit symbolic]%
     table[meta=label] {data/kill_half_et1000_average.dat};
+
+    \addplot[scatter,only marks,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/kill_half_et500.dat};
+    %\addplot[scatter,%
+    %    scatter src=explicit symbolic]%
+    %table[meta=label] {data/kill_half_et500_fail.dat};
+    \addplot[cyan,scatter,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/kill_half_et500_average.dat};
+
+    \addplot[scatter,only marks,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/kill_half_et250.dat};
+    \addplot[magenta,scatter,%
+        scatter src=explicit symbolic]%
+    table[meta=label] {data/kill_half_et250_average.dat};
+
+    \legend{1000ms,,500ms,,250ms,}
+
     \end{axis}
     \end{tikzpicture}
-        }
-        \label{fig:I}
-    }
     \caption{Recovery from Half Cluster Failure}
-    \label{fig:recovery-half-failure}
-\end{figure*}
-
-\section{Next Steps}
+    \label{fig:recovery_half_failure}
+\end{figure}
+
+Figure~\ref{fig:propagate1} and Figure~\ref{fig:propagate100} show the
+result of running the \emph{test/wait\_chat\_propagate.js} test to
+determine the amount of time that a stable cluster takes to propagate,
+commit and apply 1 and 100 chat message commands respectively to the state
+machine on every cluster node using the atomic \emph{seqAppend}
+command. The commands are serialized which means that the previous
+command must be committed before the current command can be committed.
+The test is complete once the last chat message command has been both
+committed and applied to every node of the cluster. Each test execution
+is shown as a blue ``x" and the black line represents the average for
+each cluster size.
+
+In Figure~\ref{fig:propagate1} the time to propagate a single message
+to a single node of the cluster is approximately 0.1 seconds. This
+simply represents the tests scanning/checking granularity.  The time
+to finish propagating and committing a single message shows a slight
+linear slope upwards from about 0.3 seconds for a three node cluster
+to about 0.5 seconds for a 41 node cluster. The time to propagate 100
+messages (Figure~\ref{fig:propagate100}) has a much higher slope and
+also appears to be non-linear. % TODO: why?
+
+Figure~\ref{fig:cluster_up} shows the results of running the
+\emph{test/test\_up.js} test to determine the amount of time that
+a cluster takes to start. The cluster is fully started once a leader
+has been elected and the leader has successfully added all the other
+nodes to the running configuration. Each election timer value is
+represented as a different color and the line represents the average
+value for each cluster size. The cluster startup times show a fairly
+linear increase as the cluster size increases. The reason that the
+slope varies with election timer is due to the fact that the election
+timer also determines the leader heartbeat value (currently set to one
+fifth of the election timer value). The leader heartbeat determines
+the frequency with which \emph{appendEntries} messages are sent from
+the leader to followers and is also used to propagate log entries.
+This means that the rate at which log entries (and thus cluster
+configuration updates) can be propagated is proportional to the
+election timer value.
+
+Figures \ref{fig:recovery_leader_failure} and
+\ref{fig:recovery_half_failure} show the results of running the
+\emph{test/test\_kill\_nodes.js} in both modes (kill only leader and
+kill just under half the nodes). Each election timer value is
+represented as a different color and the solid line represents the
+average recovery time for each cluster size. When only the leader is
+killed, the recovery times are constant flat regardless of cluster
+size. This is because after a new leader is elected, there is only
+a single configuration entry to propagate to the rest of the cluster
+(to remove the old failed leader). One interesting thing to note is
+that with a 1000ms timer, the 3 node cluster is actually the most
+inefficient size because the average time until a leader election
+starts is inversely proportional to cluster size so the delay until an
+election happens is a significant contributor to the three node
+cluster with 1000ms timer. With a larger cluster or shorter election
+timer, the contribution of election delay is not noticeable. The
+recovery from almost half the nodes failing is not flat because the
+amount of configuration information that needs to be propagated is
+proportional to the size of the cluster.
+
+At larger cluster sizes (35 and greater) the cluster begins to
+experience failures. For example, when testing with a 250ms election
+timer, once the cluster size reaches 35 nodes, the cluster is
+frequently (more than 50\% of the time) unable to recover from the
+half cluster shutdown test. At a cluster size or 41 and with a 1000ms
+election timer, the cluster begins to experience difficulty
+successfully starting up (about 50\% failure rate). The basic problem
+appears to be due to the average election timeout period being too low
+in relation to communication latency. During leader election, the
+first candidate is unable to gather enough votes to become leader
+before other candidates arise and steal votes which prevents a quorum.
+
+Some (or most) of the communication latency may be a result of CPU or
+memory contention due to all the browser instances running on a single
+host. The use of multiple hosts may show that the system can support
+much larger clusters for a given election timeout value. However, with
+a fixed election timer value, there is always a cluster size for which
+communication latency will begin to prevent the cluster from being
+able to perform leader elections. A dynamically adjusted election
+timer is something that should be explored especially for a browser
+context where the communication latency may be highly variable over
+time.
+
+\section{Future Work}
 
@@
 light many areas where further study of Raft, WebRTC, and Raft over
-WebRTC should be pursued.
-
-Here are some interesting areas for further exploration:
-
-\begin{enumerate}
-\item Explore the use of alternate WebRTC transport modes for relaxing
-    ordering constraints and delivery guarantees. Since the Raft
-    protocol is tolerant of out-of-order and dropped messages, it is
-    possible that Raft over WebRTC could be more efficient with one or
-    both of these constraints relaxed.
+WebRTC should be pursued. Here are some interesting areas for further
+exploration:
+
+\begin{enumerate}
+\item Explore models for dynamically adjusting Raft timeout values to
+    account for changing network and browser conditions.
+\item Explore alternate WebRTC transport modes for relaxing ordering
+    constraints and delivery guarantees. Since the Raft protocol is
+    tolerant of out-of-order and dropped messages, it is possible that
+    Raft over WebRTC could be more efficient with one or both of these
+    constraints relaxed.
 \item Test the survivability of the cluster when the signaling server
     becomes completely unresponsive or very slow.
-\item Explore models for dynamically adjusting Raft timeout values to
-    account for changing network and browser conditions.
+\item Test larger scale deployments spanning larger and more diverse
+    network domains. Characterize the factors that affect and limit
+    scalability.
 \item Explore models for dealing with the case where the cluster goes
@@
     be improved without sacrificing Raft safety constraints?
-\item Test the system in more diverse network environments. Determine
-    the reliability of establishing direct browser-to-browser
-    communication.  Determine if corporate firewalls make this as
-    difficult as some of the documentation and standards texts imply. %TODO/CITE
 \item Perform quantitative and real-world testing to better
@@
     leaders to be proactively detected so that a more capable node
-    (network, CPU, etc) can be elected. This would help address that
-    background page/tag throttling of modern browsers.
-\item Test larger scale deployments spanning larger and more diverse
-    network domains. Determine the factors that affect scalability and
-    characterize scaling limits.
-\item Perform wider cross-browser testing including Internet Explorer
-    and Opera.
+    (network, CPU, etc) can be elected. This would help address
+    background page/tag throttling in modern browsers.
 \item Implement Raft log compaction and full client linearizability.
-\item Explore models that allow for a combination of strict shared
-    state and high performance bulk data such as media or object
-    blobs. For example, the data with full consensus might contain
-    hashes that refer to bulk data.
 \item Implement a Node.js based Raft over WebRTC application. This
@@
     be resolved for this to be feasible \cite{peerjs:103}.
+\ifdefined\OPTIONAL
 \item Explore a design where some nodes of the cluster are non-voting
@@
     or backup nodes.
-\end{enumerate}
+\item Perform wider cross-browser testing including Internet Explorer
+    and Opera.
+\item Explore models that allow for a combination of strict shared
+    state and high performance bulk data such as media or object
+    blobs. For example, the data with full consensus might contain
+    hashes that refer to bulk data.
+\fi
+\end{enumerate}
+
+\section{Conclusion}
+
+The wide reach of the browser platform (exceeding most if not all
+other computing platforms) and the recent advances in browser
+programming and communication interfaces make the browser a very
+interesting platform for a wide variety of applications. The
+combination of the Raft distributed consensus algorithm with the
+WebRTC communication protocol enables a middleware system for building
+powerful distributed browser-to-browser applications.
+
+An important use case for the Raft over WebRTC middleware system is
+the ability to create distributed applications that minimize
+dependence on centralized infrastructure and are able to tolerate
+unreliable communication and frequent changes in cluster membership
+(adds, removes, failures). The demonstration chat application was able
+to easily scale to clusters containing several dozen nodes and quickly
+recover from multiple node failures. Further study is expected to show
+that a dynamically adjusted election timer that accounts for
+communication latency and cluster size would allow the cluster size to
+be scaled significantly larger (at the cost of slower recovery from
+a leader failure).
 
 \section*{Acknowledgments}
-I would like to thank my academic supervisor, David Levine, for his
-excellent input and oversight of this project. I would also like to
-thank Diego Ongaro for his amazing work on the Raft protocol and paper which
-made implementing a Raft implementation a pleasure.  There are too many people
-involved in the creation and standardization of WebRTC but I would like to
-thank them for their tireless work to make browser-to-browser communication
-possible. Last but not least, I would like to thank my family for putting up
-with the long nights and weekends I spent on this project. Thanks!
+
+I would like to thank Diego Ongaro for his amazing work on the Raft
+protocol which made implementing a Raft implementation a pleasure.
+I would also like to thank the many people involved in the creation
+and standardization of WebRTC and for their tireless work to make
+browser-to-browser communication possible. Last but not least, I would
+like to thank my family for putting up with the long nights and
+weekends I spent on this project. Thanks!
 

### 2015-12-13 research_class/paper.tex
paper.tex: update abstract.
- Also, remove paper-min.tex intermediate file.

@@
 \begin{abstract}
-We describe the creation of a browser-based distributed peer-to-peer
-application using WebRTC, PeerJS and Raft.js. WebRTC is a protocol
-that modern browsers have begun to integrate which enables
-browser-to-browser communication. PeerJS is a WebRTC client library
-and extensible signaling server. Raft.js is an implementation in
-JavaScript of the Raft distributed consensus protocol. These
-technologies are combined to create a chat application that guarantees
-a shared, consistent and linearized view of the chat history without
-relying on a single centralized server for communication. We also
-create a series of automated tests using the SlimerJS browser
-automation tool and use them to show the correctness of the chat
-application and also to measure the performance and scalability of the
-application for various cluster sizes.
+We describe a browser-based middleware system for
+creating distributed peer-to-peer applications. The system integrates
+Raft.js, an implementation of the Raft distributed consensus
+algorithm, with the WebRTC communication protocol which enables direct
+browser-to-browser communication. The middleware system is used to
+create a chat application that guarantees a shared, consistent and
+linearized view of the chat history without relying on a single
+centralized server for communication. The Raft over WebRTC middleware
+and the chat application are validated with a series of automated
+tests using the SlimerJS browser automation tool. The results
+demonstrate the correctness and robustness of the system and also
+characterize the system in terms of performance and scalability.
 \end{abstract}
